### 2016-09-23 submission/paper.tex
Initial structure for paper

%\documentclass[wrr]{agutex}
\documentclass[wrr, draft]{agutex}
% Author names in capital letters:
\authorrunninghead{1st AUTHOR ET AL.}

% Shorter version of title entered in capital letters:
\titlerunninghead{ADAPTING TO UNCERTAIN SEA LEVEL RISE}

%Corresponding author mailing address and e-mail address:
\authoraddr{Corresponding author: Name, Address. (email)}

\usepackage{bbm, amsmath}
%\usepackage[dvips]{graphicx}
\usepackage{graphicx}
\setkeys{Gin}{draft=false}
\usepackage{paralist, booktabs}
\usepackage{url}
\usepackage{color}

\usepackage{lineno}
 \linenumbers*[1]
%  To add line numbers to lines with equations:

\begin{document}

%% ------------------------------------------------------------------------ %%
%  TITLE
%% ------------------------------------------------------------------------ %%

\title{I don't know, are you sure you want to do this?}

%% ------------------------------------------------------------------------ %%
%  AUTHORS AND AFFILIATIONS
%% ------------------------------------------------------------------------ %%

% \altaffilmark will produce footnote;
% matching \altaffiltext will appear at bottom of page.

\authors{N.N.\altaffilmark{1}, N.N.\altaffilmark{2,3},...}

\altaffiltext{1}{Affiliation one}

\altaffiltext{2}{Affiliation two}

\altaffiltext{3}{Affiliation three}

%% ------------------------------------------------------------------------ %%
%  ABSTRACT
%% ------------------------------------------------------------------------ %%

\begin{abstract}
...
\end{abstract}

%% ------------------------------------------------------------------------ %%
%  BEGIN ARTICLE
%% ------------------------------------------------------------------------ %%

\begin{article}

\section{Introduction}\label{sec:intro}

\section{Methods}

\subsection{Sea level projections {\color{blue} (PG)}}

\subsubsection{Limitations of the sea level projections}

\subsection{Decision tools {\color{blue} (KdB, MD, TT)}}

\subsubsection{Limitations of the decision tools}

\section{Case studies}

\subsection{Data {\color{blue} (PG)}}

\subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}

A case study focusing on and comparing different cities in Norway.

\subsection{Selection of adaptation measures(?) {\color{blue} (MD)}}

A case study focusing on Denmark.

\section{Conclusions}

%  ACKNOWLEDGMENTS
\begin{acknowledgments}
This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL). The source code for the analysis is implemented in the statistical programming language {\tt R} (\url{http://www.R-project.org}) and is available on GitHub at \url{http://github.com/eSACP/...}.
\end{acknowledgments}

%%  REFERENCE LIST AND TEXT CITATIONS
% 5\bibliographystyle{../BibTeX/agufull08}
\bibliographystyle{agufull08}
% \bibliography{ref.bib}
% Please use ONLY \citet and \citep for reference citations.

\begin{thebibliography}{37}
%%   Before submitting: copy all the contents into the .bbl LaTeX file here
%%   and run latex again
\providecommand{\natexlab}[1]{#1}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

\bibitem[{\textit{Robert and Casella}(2004)}]{RobertCasella2004}
Robert, C.~P., and G.~Casella (2004), \textit{Monte Carlo Statistical Methods},
  2nd ed., Springer, New York.

\bibitem[{\textit{Rue et~al.}(2009)\textit{Rue, Martino, and
  Chopin}}]{Rueetal2009}
Rue, H., S.~Martino, and N.~Chopin (2009), {Approximate {B}ayesian Inference
  for Latent {G}aussian Models Using Integrated Nested {L}aplace Approximations
  (with Discussion)}, \textit{Journal of the Royal Statistical Society, Series
  B}, \textit{71}, 319--392.

\end{thebibliography}

%% ------------------------------------------------------------------------ %%
%  END ARTICLE
%% ------------------------------------------------------------------------ %%
\end{article}

%% Enter Figures and Tables here:

\end{document}

### 2016-10-03 submission/paper.tex
Section headings
Minor additions

@@
 
-\authors{N.N.\altaffilmark{1}, N.N.\altaffilmark{2,3},...}
+\authors{T. Thorarinsdottir.\altaffilmark{1}, P. Guttorp.\altaffilmark{1},...}
 
-\altaffiltext{1}{Affiliation one}
+\altaffiltext{1}{Norwegian Computing Centere}
 
-\altaffiltext{2}{Affiliation two}
+\altaffiltext{2}{Danish Technological University}
 
@@
 
-\section{Methods}
+\section{Sea level projections {\color{blue} (PG)}}
 
-\subsection{Sea level projections {\color{blue} (PG)}}
+\subsection{Global sea level}
 
-\subsubsection{Limitations of the sea level projections}
+\subsection{Local sea level}
 
-\subsection{Decision tools {\color{blue} (KdB, MD, TT)}}
+\subsection{Uncertainty assessment}
 
-\subsubsection{Limitations of the decision tools}
+\subsection{Limitations of the sea level projections}
+
+\section{Decision tools {\color{blue} (KdB, MD, TT)}}
+
+\subsection{Timing of adaptation measures}
+
+We consider adaptation decision making related to the timing of proactive adaptation measures.
+That is, the goal is to adapt to sea level rise before major damages occur.
+In a cost-benefit framework, an investment should be delayed as long as the benefits of delay (avoided investment costs) are greater than the associated costs (higher climate change damages) \citep{Fankhauser&1999}.
+
+\cite{Fankhauser&1999} describe a deterministic framework where an adaptation investment of $C^0$ now (at time $n=0$) leads to unmitigated damage of $d_0^0$ in period $0$, and a stream of partially mitigated damages $d_t^0$ in periods $t=1,2,\ldots$.
+If $r$ is the discount rate, the net present value damage, $D^0$, associated with this investment is
+\begin{linenomath*}
+\begin{equation}\label{eq:deterministic damage}
+D^0 = C^0 + d_0^0 + \frac{d_1^0}{1+r} + \frac{d_2^0}{(1+r)^2} + \cdots  
+\end{equation}
+\end{linenomath*}
+In comparison, postponing the adaptation investment to time period $n=1$ would lead to unmitigated damages in periods $0$ and $1$, and partially mitigated damages, $d_t^1$, thereafter.
+The delay would be preferable if
+\begin{linenomath*}
+\[
+C^0 - \frac{C^1}{(1+r)} > (d_0^1 - d_0^0) + \frac{d_1^1 - d_1^0}{1+r} + \frac{d_2^1 - d_2^0}{(1+r)^2} + \cdots
+\]
+\end{linenomath*}
+Here, the expression on the left describes the benefits of the delay while the expression on the right describes the cost of the delay.
+In the simplest case, there is no change in investment costs ($C^0 = C^1 = C$) and the delay has no lasting effects beyond period 1 ($d_t^1 = d_t^0$ for $t > 1$).
+In this case, the comparison is between the expected return $r$ earned on the captial while implementation is delayed and one addtional time period of unmitigated damage,
+\begin{linenomath*}
+  \[
+  r C > d_1^1 - d_1^0.
+  \]
+  \end{linenomath*}
+
+\subsection{Limitations of the decision framework}
 
@@
 
+\bibitem[{\textit{Fankhauser et~al.}(1999)}]{Fankhauser&1999}
+  Fankhauser, S., J.~B.~Smith, and R.~S.~J.~Tol (1999), {Weathering climate
+    change: some simple rules to guide adaptation decisions},
+  \textit{Ecological Economics}, \textit{30}, 67--78.
+
 \bibitem[{\textit{Robert and Casella}(2004)}]{RobertCasella2004}

### 2016-10-05 submission/paper.tex
Updated description of sea level projections

@@
 
-\authors{T. Thorarinsdottir.\altaffilmark{1}, P. Guttorp.\altaffilmark{1},...}
+\authors{T. Thorarinsdottir\altaffilmark{1}, P. Guttorp\altaffilmark{1}, M. Drews\altaffilmark{2}}
 
@@
 \subsection{Global sea level}
+Most climate models do not explicitly provide sea level as an output of the calculations.
+
+We will use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
+
+%Figure showing historical relationship
+
+We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level. 
+%Figure with sea level projections & IPCC projections
 
 \subsection{Local sea level}
+In order to get from global sea level projections to local ones, it is important to note that sea level rise not is uniform over the globe.
+Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
+%Gravitaional effects
+Another major effect in Fennoskandia is the land rise due to isostatic effects of the glaciers from the last ice age.
+Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
+%Figure showing corrected and uncorrected Bergen data and regression model.
+
+The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one.
+Each climate model temperature projection yields a different local sea level projection.
 
 \subsection{Uncertainty assessment}
+Following the approach of \citet{Guttorp2014} we assess the uncertainty in the local sea level projections taking into account the variability between the climate projections used, the uncertainties in the regressuions of global mean temperature on global mean sealevel and of global on local sea level.
+We express the sea level projection uncertainty in terms of a confidence band that is simultaneously of the intended  level  for all projection years.
+This allows us, for example, to get a confidence band for the years when a given sea level rise is obtained.
+
+%Figure showing Bergen and Copenhagen slr projections
 
 \subsection{Limitations of the sea level projections}
+The main assumption is using historical relationships in statistical projections of the type used in this paper is that there is no major change in how temperature relates to sea level, globally and locally.
+Among the factors that may invalidate this approach are changes in water storage on land, changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
+For example, the rate of ice melt on Greenland may suddenly increase substantially due to intense warming of both air and sea water.
+Our current climate models are not able to resolve the ice processes sufficiently to include such so called tipping points into the projections.
 
@@
 
+\bibitem[{Bolin et~al.(2014)}]{Bolin2014a}
+Bolin, D., P. Guttorp, A. Januzzi, M. Novak, H. Podschwit, L. Richardson, C. Sowder, A. S\"{a}rkk\"{a} and A. Zimmerman (2014), Statistical prediction of global sea level from
+  global temperature. \textit{Statistica Sinica},~{\em25}, 351--367.
+
+\bibitem[{Church and White (2011)}]{csiro}
+Church, J. and N.~White (2011),
+Sea-level rise from the late 19th to the early 21st century.
+{\em Surveys in Geophysics\/}~{\em 32}, 585--602.
+
 \bibitem[{\textit{Fankhauser et~al.}(1999)}]{Fankhauser&1999}
@@
 
-\bibitem[{\textit{Robert and Casella}(2004)}]{RobertCasella2004}
+ \bibitem[{Guttorp et~al. (2014)}]{Guttorp2014}
+Guttorp, P., D. Bolin, A. Januzzi, D. Jones, M. Novak, H. Podschwit, L. Richardson, A.
+S\"{a}rkk\"{a}, C. Sowder and A Zimmerman (2014), Assessing the uncertainty in projecting local
+mean sea level from global temperature. \textit{Journal of Applied Meteorology and Climatology}.
+\textit{53}, 2163--2170.
+
+  \bibitem[{Hansen et~al.} (2001)]{giss}
+Hansen, J., R.~Rued, M.~Sato, M.~Imhoff, W.~Lawrence, D.~Easterling,
+  T.~Peterson, and T.~Karl (2001),
+A closer look at {U}nited {S}tates and global surface temperature
+  change. {\em J. Geophys. Res.\/}~{\em 106}, 23947--23963.
+
+ \bibitem[{Rahmstorf}{2007}]{Rahmstorf07}
+Rahmstorf, S. (2007).
+\newblock {A} semi-empirical approach to projecting future sea-level rise.
+\newblock {\em Science\/}~{\em 315}, 368--370.
+
+\bibitem[{Rahmstorf, Perrette, and Vermeer}{Rahmstorf
+  et~al.}{2011}]{Rahmstorf11}
+Rahmstorf, S., M.~Perrette, and M.~Vermeer (2011),
+Testing the robustness of semi-empirical sea level projections.
+  climate dynamics.
+{\em Climate Dynamics}~{\em 39}, 861--875.
+
+\bibitem[{Robert and Casella}(2004)]{RobertCasella2004}
 Robert, C.~P., and G.~Casella (2004), \textit{Monte Carlo Statistical Methods},
@@
 
-\bibitem[{\textit{Rue et~al.}(2009)\textit{Rue, Martino, and
-  Chopin}}]{Rueetal2009}
+\bibitem[{Rue et~al.}(2009)]{Rueetal2009}
 Rue, H., S.~Martino, and N.~Chopin (2009), {Approximate {B}ayesian Inference
@@
   B}, \textit{71}, 319--392.
+
+\bibitem[{Taylor et~al.(2012)Taylor, Stouffer, and Meehl}]{cmip5}
+Taylor, K., R.~Stouffer, and G.~Meehl, 2012: An overview of {CMIP}5 and the
+  experiment design. \textit{Bull. Amer. Meteor. Soc.}, \textbf{93}, 485--498.
 

### 2016-10-07 submission/paper.tex
Bib file for references

@@
 %Gravitaional effects
-Another major effect in Fennoskandia is the land rise due to isostatic effects of the glaciers from the last ice age.
+Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
 Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
@@
 \bibliographystyle{agufull08}
-% \bibliography{ref.bib}
+\bibliography{ref.bib}
 % Please use ONLY \citet and \citep for reference citations.
 
-\begin{thebibliography}{37}
+%\begin{thebibliography}{37}
 %%   Before submitting: copy all the contents into the .bbl LaTeX file here
 %%   and run latex again
-\providecommand{\natexlab}[1]{#1}
-\expandafter\ifx\csname urlstyle\endcsname\relax
-  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
-  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
-  \urlstyle{rm}\Url}\fi
+%\providecommand{\natexlab}[1]{#1}
+%\expandafter\ifx\csname urlstyle\endcsname\relax
+%  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+%  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+%  \urlstyle{rm}\Url}\fi
 
-\bibitem[{Bolin et~al.(2014)}]{Bolin2014a}
-Bolin, D., P. Guttorp, A. Januzzi, M. Novak, H. Podschwit, L. Richardson, C. Sowder, A. S\"{a}rkk\"{a} and A. Zimmerman (2014), Statistical prediction of global sea level from
-  global temperature. \textit{Statistica Sinica},~{\em25}, 351--367.
-
-\bibitem[{Church and White (2011)}]{csiro}
-Church, J. and N.~White (2011),
-Sea-level rise from the late 19th to the early 21st century.
-{\em Surveys in Geophysics\/}~{\em 32}, 585--602.
-
-\bibitem[{\textit{Fankhauser et~al.}(1999)}]{Fankhauser&1999}
-  Fankhauser, S., J.~B.~Smith, and R.~S.~J.~Tol (1999), {Weathering climate
-    change: some simple rules to guide adaptation decisions},
-  \textit{Ecological Economics}, \textit{30}, 67--78.
-
- \bibitem[{Guttorp et~al. (2014)}]{Guttorp2014}
-Guttorp, P., D. Bolin, A. Januzzi, D. Jones, M. Novak, H. Podschwit, L. Richardson, A.
-S\"{a}rkk\"{a}, C. Sowder and A Zimmerman (2014), Assessing the uncertainty in projecting local
-mean sea level from global temperature. \textit{Journal of Applied Meteorology and Climatology}.
-\textit{53}, 2163--2170.
-
-  \bibitem[{Hansen et~al.} (2001)]{giss}
-Hansen, J., R.~Rued, M.~Sato, M.~Imhoff, W.~Lawrence, D.~Easterling,
-  T.~Peterson, and T.~Karl (2001),
-A closer look at {U}nited {S}tates and global surface temperature
-  change. {\em J. Geophys. Res.\/}~{\em 106}, 23947--23963.
-
- \bibitem[{Rahmstorf}{2007}]{Rahmstorf07}
-Rahmstorf, S. (2007).
-\newblock {A} semi-empirical approach to projecting future sea-level rise.
-\newblock {\em Science\/}~{\em 315}, 368--370.
-
-\bibitem[{Rahmstorf, Perrette, and Vermeer}{Rahmstorf
-  et~al.}{2011}]{Rahmstorf11}
-Rahmstorf, S., M.~Perrette, and M.~Vermeer (2011),
-Testing the robustness of semi-empirical sea level projections.
-  climate dynamics.
-{\em Climate Dynamics}~{\em 39}, 861--875.
-
-\bibitem[{Robert and Casella}(2004)]{RobertCasella2004}
-Robert, C.~P., and G.~Casella (2004), \textit{Monte Carlo Statistical Methods},
-  2nd ed., Springer, New York.
-
-\bibitem[{Rue et~al.}(2009)]{Rueetal2009}
-Rue, H., S.~Martino, and N.~Chopin (2009), {Approximate {B}ayesian Inference
-  for Latent {G}aussian Models Using Integrated Nested {L}aplace Approximations
-  (with Discussion)}, \textit{Journal of the Royal Statistical Society, Series
-  B}, \textit{71}, 319--392.
-
-\bibitem[{Taylor et~al.(2012)Taylor, Stouffer, and Meehl}]{cmip5}
-Taylor, K., R.~Stouffer, and G.~Meehl, 2012: An overview of {CMIP}5 and the
-  experiment design. \textit{Bull. Amer. Meteor. Soc.}, \textbf{93}, 485--498.
-
-\end{thebibliography}
+%\end{thebibliography}
 

### 2016-10-08 submission/paper.tex
Additional text and references in section 2

@@
 Most climate models do not explicitly provide sea level as an output of the calculations.
+Rather, the IPCC AR5 report \citep[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age.
+Judging from the supplementary material to \citet[ch.~13]{ipcc}, the uncertainty assessment is only based on the spread of the ensemble of temperature projections, not on the additional uncertainty in the ice models used.
 
-We will use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
+We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
 
-%Figure showing historical relationship
+%Figure showing historical relationship?
 
-We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level. 
-%Figure with sea level projections & IPCC projections
+We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}).
+%Figure with sea level projections & IPCC projections?
 
@@
 \subsection{Uncertainty assessment}
-Following the approach of \citet{Guttorp2014} we assess the uncertainty in the local sea level projections taking into account the variability between the climate projections used, the uncertainties in the regressuions of global mean temperature on global mean sealevel and of global on local sea level.
+\label{unc_ass}
+Following the approach of \citet{Guttorp2014} we assess the uncertainty in the local sea level projections taking into account the variability between the climate projections used, the uncertainties in the regressions of global mean temperature on global mean sea level and of global on local sea level.
 We express the sea level projection uncertainty in terms of a confidence band that is simultaneously of the intended  level  for all projection years.
@@
 The main assumption is using historical relationships in statistical projections of the type used in this paper is that there is no major change in how temperature relates to sea level, globally and locally.
-Among the factors that may invalidate this approach are changes in water storage on land, changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
+Among the factors that may invalidate this approach are changes in water storage on land (removing water from the oceans), excessive siphoning of groundwater (resulting in land subsidence), changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
 For example, the rate of ice melt on Greenland may suddenly increase substantially due to intense warming of both air and sea water.
 Our current climate models are not able to resolve the ice processes sufficiently to include such so called tipping points into the projections.
+Also, the IPCC scenarios do not include changes in water usage \citep{change}.
 

### 2016-10-10 submission/paper.tex
Code for Bergen simulations. Inclusion of one picture in paper.

@@
 Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model.
-%Figure showing corrected and uncorrected Bergen data and regression model.
+Figure \ref{fig:obs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
+The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}.
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{bergenfit.png}
+\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
+The straight line is the time series regression line.}
+\label{fig:obs}
+\end{center}
+\end{figure}
 
@@
 
-%Figure showing Bergen and Copenhagen slr projections
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{bergen_ci.png}
+\caption{ }
+\label{fig:ci}
+\end{center}
+\end{figure}
 

### 2016-10-10 submission/paper.tex
Added confidence interval figure for Bergen

@@
 \includegraphics[width=39pc]{bergen_ci.png}
-\caption{ }
+\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
+The sea level data are shown in blue and end in 2015.
+The thin red lines are the projections without uncertainty based on each of the climate models.
+The dashed purple lines connect pointwise confidence intervals for each year.
+}
 \label{fig:ci}

### 2016-10-11 submission/paper.tex
solving conflict

--- 2016-10-08 submission/paper.tex
@@
-%\documentclass[wrr]{agutex}
-\documentclass[wrr, draft]{agutex}
-% Author names in capital letters:
-\authorrunninghead{1st AUTHOR ET AL.}
+<!DOCTYPE html>
+<html lang="en" class=" is-copy-enabled is-u2f-enabled">
+  <head prefix="og: http://ogp.me/ns# fb: http://ogp.me/ns/fb# object: http://ogp.me/ns/object# article: http://ogp.me/ns/article# profile: http://ogp.me/ns/profile#">
+    <meta charset='utf-8'>
+    <meta content="origin-when-cross-origin" name="referrer" />
 
-% Shorter version of title entered in capital letters:
-\titlerunninghead{ADAPTING TO UNCERTAIN SEA LEVEL RISE}
+    <link crossorigin="anonymous" href="https://assets-cdn.github.com/assets/frameworks-cb4ede7df6d8670c4051172e4d6bc6916b3c765fa15b4ee9b348f157fdb85114.css" integrity="sha256-y07effbYZwxAURcuTWvGkWs8dl+hW07ps0jxV/24URQ=" media="all" rel="stylesheet" />
+    <link crossorigin="anonymous" href="https://assets-cdn.github.com/assets/github-f32b0e7dbc73ab49677d48d959274568a459569efc1cf486e83fe9aed5890998.css" integrity="sha256-8ysOfbxzq0lnfUjZWSdFaKRZVp78HPSG6D/prtWJCZg=" media="all" rel="stylesheet" />
 
-%Corresponding author mailing address and e-mail address:
-\authoraddr{Corresponding author: Name, Address. (email)}
+    <meta http-equiv="X-UA-Compatible" content="IE=edge">
+    <meta http-equiv="Content-Language" content="en">
+    <meta name="viewport" content="width=device-width">
 
-\usepackage{bbm, amsmath}
-%\usepackage[dvips]{graphicx}
-\usepackage{graphicx}
-\setkeys{Gin}{draft=false}
-\usepackage{paralist, booktabs}
-\usepackage{url}
-\usepackage{color}
+    <title>SeaLevelDecisions/paper.tex at master · eSACP/SeaLevelDecisions</title>
+    <link rel="search" type="application/opensearchdescription+xml" href="/opensearch.xml" title="GitHub">
+    <link rel="fluid-icon" href="https://github.com/fluidicon.png" title="GitHub">
+    <link rel="apple-touch-icon" href="/apple-touch-icon.png">
+    <link rel="apple-touch-icon" sizes="57x57" href="/apple-touch-icon-57x57.png">
+    <link rel="apple-touch-icon" sizes="60x60" href="/apple-touch-icon-60x60.png">
+    <link rel="apple-touch-icon" sizes="72x72" href="/apple-touch-icon-72x72.png">
+    <link rel="apple-touch-icon" sizes="76x76" href="/apple-touch-icon-76x76.png">
+    <link rel="apple-touch-icon" sizes="114x114" href="/apple-touch-icon-114x114.png">
+    <link rel="apple-touch-icon" sizes="120x120" href="/apple-touch-icon-120x120.png">
+    <link rel="apple-touch-icon" sizes="144x144" href="/apple-touch-icon-144x144.png">
+    <link rel="apple-touch-icon" sizes="152x152" href="/apple-touch-icon-152x152.png">
+    <link rel="apple-touch-icon" sizes="180x180" href="/apple-touch-icon-180x180.png">
+    <meta property="fb:app_id" content="1401488693436528">
 
-\usepackage{lineno}
- \linenumbers*[1]
-%  To add line numbers to lines with equations:
+      <meta content="https://avatars1.githubusercontent.com/u/14945390?v=3&amp;s=400" name="twitter:image:src" /><meta content="@github" name="twitter:site" /><meta content="summary" name="twitter:card" /><meta content="eSACP/SeaLevelDecisions" name="twitter:title" /><meta content="SeaLevelDecisions - Paper on sea level change adaptation decision making under uncertainty " name="twitter:description" />
+      <meta content="https://avatars1.githubusercontent.com/u/14945390?v=3&amp;s=400" property="og:image" /><meta content="GitHub" property="og:site_name" /><meta content="object" property="og:type" /><meta content="eSACP/SeaLevelDecisions" property="og:title" /><meta content="https://github.com/eSACP/SeaLevelDecisions" property="og:url" /><meta content="SeaLevelDecisions - Paper on sea level change adaptation decision making under uncertainty " property="og:description" />
+      <meta name="browser-stats-url" content="https://api.github.com/_private/browser/stats">
+    <meta name="browser-errors-url" content="https://api.github.com/_private/browser/errors">
+    <link rel="assets" href="https://assets-cdn.github.com/">
+    <link rel="web-socket" href="wss://live.github.com/_sockets/MTQ5NDUyMTM6MDUyNDYzZDIwMmVhNjNkYzMwNWI5ZWU1M2Q3NmQ3MWQ6NmY5MTBmMGJhNDA4NDZkNzQ4MGRkNDBjODdkYjBmYTI4ZjBkODAzMWJiY2IzODI3YWE5MDI3MGRjZTc0MjBkMQ==--89d25bdd8d4ae8902ca35bc19bf81c83badb6ba1">
+    <meta name="pjax-timeout" content="1000">
+    <link rel="sudo-modal" href="/sessions/sudo_modal">
+    <meta name="request-id" content="9C740835:0C52:28EFDD9:57FCCA5A" data-pjax-transient>
 
-\begin{document}
+    <meta name="msapplication-TileImage" content="/windows-tile.png">
+    <meta name="msapplication-TileColor" content="#ffffff">
+    <meta name="selected-link" value="repo_source" data-pjax-transient>
 
-%% ------------------------------------------------------------------------ %%
-%  TITLE
-%% ------------------------------------------------------------------------ %%
+    <meta name="google-site-verification" content="KT5gs8h0wvaagLKAVWq8bbeNwnZZK1r1XQysX3xurLU">
+<meta name="google-site-verification" content="ZzhVyEFwb7w3e0-uOTltm8Jsck2F5StVihD0exw2fsA">
+    <meta name="google-analytics" content="UA-3769691-2">
 
-\title{I don't know, are you sure you want to do this?}
+<meta content="collector.githubapp.com" name="octolytics-host" /><meta content="github" name="octolytics-app-id" /><meta content="9C740835:0C52:28EFDD9:57FCCA5A" name="octolytics-dimension-request_id" /><meta content="14945213" name="octolytics-actor-id" /><meta content="thordislinda" name="octolytics-actor-login" /><meta content="25d4553f8e8b0f7d5f53a85da67d6ac37df8675bf0532ac0ed5d5f1cf648984f" name="octolytics-actor-hash" />
+<meta content="/&lt;user-name&gt;/&lt;repo-name&gt;/blob/show" data-pjax-transient="true" name="analytics-location" />
 
-%% ------------------------------------------------------------------------ %%
-%  AUTHORS AND AFFILIATIONS
-%% ------------------------------------------------------------------------ %%
+  <meta class="js-ga-set" name="dimension1" content="Logged In">
 
-% \altaffilmark will produce footnote;
-% matching \altaffiltext will appear at bottom of page.
+        <meta name="hostname" content="github.com">
+    <meta name="user-login" content="thordislinda">
 
-\authors{T. Thorarinsdottir\altaffilmark{1}, P. Guttorp\altaffilmark{1}, M. Drews\altaffilmark{2}}
+        <meta name="expected-hostname" content="github.com">
+      <meta name="js-proxy-site-detection-payload" content="Y2ExZjFiMDljMmQxNzBiYjUzNWQzYjgxYmI2OGM1NzY0Nzc1NGRmNWJhNGVmNDNjOTJkYjc0NmJlOTA5ZGU1MXx7InJlbW90ZV9hZGRyZXNzIjoiMTU2LjExNi44LjUzIiwicmVxdWVzdF9pZCI6IjlDNzQwODM1OjBDNTI6MjhFRkREOTo1N0ZDQ0E1QSIsInRpbWVzdGFtcCI6MTQ3NjE4NDY3MX0=">
 
-\altaffiltext{1}{Norwegian Computing Centere}
+      <link rel="mask-icon" href="https://assets-cdn.github.com/pinned-octocat.svg" color="#4078c0">
+      <link rel="icon" type="image/x-icon" href="https://assets-cdn.github.com/favicon.ico">
 
-\altaffiltext{2}{Danish Technological University}
+    <meta name="html-safe-nonce" content="45264389604d630702d1147e97655f012ab39c1a">
+    <meta content="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" name="form-nonce" />
 
-\altaffiltext{3}{Affiliation three}
+    <meta http-equiv="x-pjax-version" content="5773d60f3601ed21fb6ed8e0c6eff2f6">
 
-%% ------------------------------------------------------------------------ %%
-%  ABSTRACT
-%% ------------------------------------------------------------------------ %%
+  <meta name="description" content="SeaLevelDecisions - Paper on sea level change adaptation decision making under uncertainty ">
+  <meta name="go-import" content="github.com/eSACP/SeaLevelDecisions git https://github.com/eSACP/SeaLevelDecisions.git">
 
-\begin{abstract}
-...
-\end{abstract}
+  <meta content="14945390" name="octolytics-dimension-user_id" /><meta content="eSACP" name="octolytics-dimension-user_login" /><meta content="69018538" name="octolytics-dimension-repository_id" /><meta content="eSACP/SeaLevelDecisions" name="octolytics-dimension-repository_nwo" /><meta content="false" name="octolytics-dimension-repository_public" /><meta content="false" name="octolytics-dimension-repository_is_fork" /><meta content="69018538" name="octolytics-dimension-repository_network_root_id" /><meta content="eSACP/SeaLevelDecisions" name="octolytics-dimension-repository_network_root_nwo" />
+  <link href="https://github.com/eSACP/SeaLevelDecisions/commits/master.atom?token=AOQLvX3BJqaD2T7TXulS-nsUHBg8-hiIks62CLlvwA%3D%3D" rel="alternate" title="Recent Commits to SeaLevelDecisions:master" type="application/atom+xml">
 
-%% ------------------------------------------------------------------------ %%
-%  BEGIN ARTICLE
-%% ------------------------------------------------------------------------ %%
+      <link rel="canonical" href="https://github.com/eSACP/SeaLevelDecisions/blob/master/submission/paper.tex" data-pjax-transient>
+  </head>
 
-\begin{article}
+  <body class="logged-in  env-production linux vis-private page-blob">
+    <div id="js-pjax-loader-bar" class="pjax-loader-bar"><div class="progress"></div></div>
+    <a href="#start-of-content" tabindex="1" class="accessibility-aid js-skip-to-content">Skip to content</a>
 
-\section{Introduction}\label{sec:intro}
+        <div class="header header-logged-in true" role="banner">
+  <div class="container clearfix">
 
-\section{Sea level projections {\color{blue} (PG)}}
+    <a class="header-logo-invertocat" href="https://github.com/" data-hotkey="g d" aria-label="Homepage" data-ga-click="Header, go to dashboard, icon:logo">
+  <svg aria-hidden="true" class="octicon octicon-mark-github" height="28" version="1.1" viewBox="0 0 16 16" width="28"><path d="M8 0C3.58 0 0 3.58 0 8c0 3.54 2.29 6.53 5.47 7.59.4.07.55-.17.55-.38 0-.19-.01-.82-.01-1.49-2.01.37-2.53-.49-2.69-.94-.09-.23-.48-.94-.82-1.13-.28-.15-.68-.52-.01-.53.63-.01 1.08.58 1.23.82.72 1.21 1.87.87 2.33.66.07-.52.28-.87.51-1.07-1.78-.2-3.64-.89-3.64-3.95 0-.87.31-1.59.82-2.15-.08-.2-.36-1.02.08-2.12 0 0 .67-.21 2.2.82.64-.18 1.32-.27 2-.27.68 0 1.36.09 2 .27 1.53-1.04 2.2-.82 2.2-.82.44 1.1.16 1.92.08 2.12.51.56.82 1.27.82 2.15 0 3.07-1.87 3.75-3.65 3.95.29.25.54.73.54 1.48 0 1.07-.01 1.93-.01 2.2 0 .21.15.46.55.38A8.013 8.013 0 0 0 16 8c0-4.42-3.58-8-8-8z"></path></svg>
+</a>
 
-\subsection{Global sea level}
-Most climate models do not explicitly provide sea level as an output of the calculations.
-Rather, the IPCC AR5 report \citep[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age.
-Judging from the supplementary material to \citet[ch.~13]{ipcc}, the uncertainty assessment is only based on the spread of the ensemble of temperature projections, not on the additional uncertainty in the ice models used.
+        <div class="header-search scoped-search site-scoped-search js-site-search" role="search">
+  <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/search" class="js-site-search-form" data-scoped-search-url="/eSACP/SeaLevelDecisions/search" data-unscoped-search-url="/search" method="get"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /></div>
+    <label class="form-control header-search-wrapper js-chromeless-input-container">
+      <div class="header-search-scope">This repository</div>
+      <input type="text"
+        class="form-control header-search-input js-site-search-focus js-site-search-field is-clearable"
+        data-hotkey="s"
+        name="q"
+        placeholder="Search"
+        aria-label="Search this repository"
+        data-unscoped-placeholder="Search GitHub"
+        data-scoped-placeholder="Search"
+        autocapitalize="off">
+    </label>
+</form></div>
 
-We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
+      <ul class="header-nav float-left" role="navigation">
+        <li class="header-nav-item">
+          <a href="/pulls" aria-label="Pull requests you created" class="js-selected-navigation-item header-nav-link" data-ga-click="Header, click, Nav menu - item:pulls context:user" data-hotkey="g p" data-selected-links="/pulls /pulls/assigned /pulls/mentioned /pulls">
+            Pull requests
+</a>        </li>
+        <li class="header-nav-item">
+          <a href="/issues" aria-label="Issues you created" class="js-selected-navigation-item header-nav-link" data-ga-click="Header, click, Nav menu - item:issues context:user" data-hotkey="g i" data-selected-links="/issues /issues/assigned /issues/mentioned /issues">
+            Issues
+</a>        </li>
+          <li class="header-nav-item">
+            <a class="header-nav-link" href="https://gist.github.com/" data-ga-click="Header, go to gist, text:gist">Gist</a>
+          </li>
+      </ul>
 
-%Figure showing historical relationship?
+<ul class="header-nav user-nav float-right" id="user-links">
+  <li class="header-nav-item">
 
-We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}).
-%Figure with sea level projections & IPCC projections?
+    <a href="/notifications" aria-label="You have no unread notifications" class="header-nav-link notification-indicator tooltipped tooltipped-s js-socket-channel js-notification-indicator" data-channel="tenant:1:notification-changed:14945213" data-ga-click="Header, go to notifications, icon:read" data-hotkey="g n">
+        <span class="mail-status "></span>
+        <svg aria-hidden="true" class="octicon octicon-bell" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M14 12v1H0v-1l.73-.58c.77-.77.81-2.55 1.19-4.42C2.69 3.23 6 2 6 2c0-.55.45-1 1-1s1 .45 1 1c0 0 3.39 1.23 4.16 5 .38 1.88.42 3.66 1.19 4.42l.66.58H14zm-7 4c1.11 0 2-.89 2-2H5c0 1.11.89 2 2 2z"></path></svg>
+</a>
+  </li>
 
-\subsection{Local sea level}
-In order to get from global sea level projections to local ones, it is important to note that sea level rise not is uniform over the globe.
-Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
-%Gravitaional effects
-Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
-Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
-%Figure showing corrected and uncorrected Bergen data and regression model.
+  <li class="header-nav-item dropdown js-menu-container">
+    <a class="header-nav-link tooltipped tooltipped-s js-menu-target" href="/new"
+       aria-label="Create new…"
+       data-ga-click="Header, create new, icon:add">
+      <svg aria-hidden="true" class="octicon octicon-plus float-left" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M12 9H7v5H5V9H0V7h5V2h2v5h5z"></path></svg>
+      <span class="dropdown-caret"></span>
+    </a>
 
-The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one.
-Each climate model temperature projection yields a different local sea level projection.
+    <div class="dropdown-menu-content js-menu-content">
+      <ul class="dropdown-menu dropdown-menu-sw">
 
-\subsection{Uncertainty assessment}
-\label{unc_ass}
-Following the approach of \citet{Guttorp2014} we assess the uncertainty in the local sea level projections taking into account the variability between the climate projections used, the uncertainties in the regressions of global mean temperature on global mean sea level and of global on local sea level.
+<a class="dropdown-item" href="/new" data-ga-click="Header, create new repository">
+  New repository
+</a>
+
+  <a class="dropdown-item" href="/new/import" data-ga-click="Header, import a repository">
+    Import repository
+  </a>
+
+  <a class="dropdown-item" href="/organizations/new" data-ga-click="Header, create new organization">
+    New organization
+  </a>
+
+  <div class="dropdown-divider"></div>
+  <div class="dropdown-header">
+    <span title="eSACP">This organization</span>
+  </div>
+  <a class="dropdown-item" href="/orgs/eSACP/people#invite-member" data-ga-click="Header, invite someone">
+    Invite someone
+  </a>
+  <a class="dropdown-item" href="/orgs/eSACP/new-team" data-ga-click="Header, create new team">
+    New team
+  </a>
+  <a class="dropdown-item" href="/organizations/eSACP/repositories/new" data-ga-click="Header, create new organization repository, icon:repo">
+    New repository
+  </a>
+
+  <div class="dropdown-divider"></div>
+  <div class="dropdown-header">
+    <span title="eSACP/SeaLevelDecisions">This repository</span>
+  </div>
+    <a class="dropdown-item" href="/eSACP/SeaLevelDecisions/issues/new" data-ga-click="Header, create new issue">
+      New issue
+    </a>
+    <a class="dropdown-item" href="/eSACP/SeaLevelDecisions/settings/collaboration" data-ga-click="Header, create new collaborator">
+      New collaborator
+    </a>
+
+      </ul>
+    </div>
+  </li>
+
+  <li class="header-nav-item dropdown js-menu-container">
+    <a class="header-nav-link name tooltipped tooltipped-sw js-menu-target" href="/thordislinda"
+       aria-label="View profile and more"
+       data-ga-click="Header, show menu, icon:avatar">
+      <img alt="@thordislinda" class="avatar" height="20" src="https://avatars1.githubusercontent.com/u/14945213?v=3&amp;s=40" width="20" />
+      <span class="dropdown-caret"></span>
+    </a>
+
+    <div class="dropdown-menu-content js-menu-content">
+      <div class="dropdown-menu dropdown-menu-sw">
+        <div class="dropdown-header header-nav-current-user css-truncate">
+          Signed in as <strong class="css-truncate-target">thordislinda</strong>
+        </div>
+
+        <div class="dropdown-divider"></div>
+
+        <a class="dropdown-item" href="/thordislinda" data-ga-click="Header, go to profile, text:your profile">
+          Your profile
+        </a>
+        <a class="dropdown-item" href="/thordislinda?tab=stars" data-ga-click="Header, go to starred repos, text:your stars">
+          Your stars
+        </a>
+        <a class="dropdown-item" href="/explore" data-ga-click="Header, go to explore, text:explore">
+          Explore
+        </a>
+          <a class="dropdown-item" href="/integrations" data-ga-click="Header, go to integrations, text:integrations">
+            Integrations
+          </a>
+        <a class="dropdown-item" href="https://help.github.com" data-ga-click="Header, go to help, text:help">
+          Help
+        </a>
+
+        <div class="dropdown-divider"></div>
+
+        <a class="dropdown-item" href="/settings/profile" data-ga-click="Header, go to settings, icon:settings">
+          Settings
+        </a>
+
+        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/logout" class="logout-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="3yXD4WZhwnWJytG55fL3EwgkMSCMjYlEpuhXXi1XrsllPkR2n1RGivgoBFTGcsqjYxx3Jj0qwtsKsr9Lfl+eRQ==" /></div>
+          <button type="submit" class="dropdown-item dropdown-signout" data-ga-click="Header, sign out, icon:logout">
+            Sign out
+          </button>
+</form>      </div>
+    </div>
+  </li>
+</ul>
+
+  </div>
+</div>
+
+    <div id="start-of-content" class="accessibility-aid"></div>
+
+      <div id="js-flash-container">
+</div>
+
+    <div role="main">
+        <div itemscope itemtype="http://schema.org/SoftwareSourceCode">
+    <div id="js-repo-pjax-container" data-pjax-container>
+
+<div class="pagehead repohead instapaper_ignore readability-menu experiment-repo-nav">
+  <div class="container repohead-details-container">
+
+<ul class="pagehead-actions">
+
+  <li>
+        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/notifications/subscribe" class="js-social-container" data-autosubmit="true" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="ow3eH8530MPPaY+NBuQQ2cKNOZgKM52lzUCopLpBByIcEQvPDDXmUb+PxjlwNE8m1o6yftRt4YK14tIqIWRMUQ==" /></div>      <input class="form-control" id="repository_id" name="repository_id" type="hidden" value="69018538" />
+
+        <div class="select-menu js-menu-container js-select-menu">
+          <a href="/eSACP/SeaLevelDecisions/subscription"
+            class="btn btn-sm btn-with-count select-menu-button js-menu-target" role="button" tabindex="0" aria-haspopup="true"
+            data-ga-click="Repository, click Watch settings, action:blob#show">
+            <span class="js-select-button">
+              <svg aria-hidden="true" class="octicon octicon-eye" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M8.06 2C3 2 0 8 0 8s3 6 8.06 6C13 14 16 8 16 8s-3-6-7.94-6zM8 12c-2.2 0-4-1.78-4-4 0-2.2 1.8-4 4-4 2.22 0 4 1.8 4 4 0 2.22-1.78 4-4 4zm2-4c0 1.11-.89 2-2 2-1.11 0-2-.89-2-2 0-1.11.89-2 2-2 1.11 0 2 .89 2 2z"></path></svg>
+              Unwatch
+            </span>
+          </a>
+          <a class="social-count js-social-count"
+            href="/eSACP/SeaLevelDecisions/watchers"
+            aria-label="10 users are watching this repository">
+            10
+          </a>
+
+        <div class="select-menu-modal-holder">
+          <div class="select-menu-modal subscription-menu-modal js-menu-content" aria-hidden="true">
+            <div class="select-menu-header js-navigation-enable" tabindex="-1">
+              <svg aria-label="Close" class="octicon octicon-x js-menu-close" height="16" role="img" version="1.1" viewBox="0 0 12 16" width="12"><path d="M7.48 8l3.75 3.75-1.48 1.48L6 9.48l-3.75 3.75-1.48-1.48L4.52 8 .77 4.25l1.48-1.48L6 6.52l3.75-3.75 1.48 1.48z"></path></svg>
+              <span class="select-menu-title">Notifications</span>
+            </div>
+
+              <div class="select-menu-list js-navigation-container" role="menu">
+
+                <div class="select-menu-item js-navigation-item " role="menuitem" tabindex="0">
+                  <svg aria-hidden="true" class="octicon octicon-check select-menu-item-icon" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M12 5l-8 8-4-4 1.5-1.5L4 10l6.5-6.5z"></path></svg>
+                  <div class="select-menu-item-text">
+                    <input id="do_included" name="do" type="radio" value="included" />
+                    <span class="select-menu-item-heading">Not watching</span>
+                    <span class="description">Be notified when participating or @mentioned.</span>
+                    <span class="js-select-button-text hidden-select-button-text">
+                      <svg aria-hidden="true" class="octicon octicon-eye" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M8.06 2C3 2 0 8 0 8s3 6 8.06 6C13 14 16 8 16 8s-3-6-7.94-6zM8 12c-2.2 0-4-1.78-4-4 0-2.2 1.8-4 4-4 2.22 0 4 1.8 4 4 0 2.22-1.78 4-4 4zm2-4c0 1.11-.89 2-2 2-1.11 0-2-.89-2-2 0-1.11.89-2 2-2 1.11 0 2 .89 2 2z"></path></svg>
+                      Watch
+                    </span>
+                  </div>
+                </div>
+
+                <div class="select-menu-item js-navigation-item selected" role="menuitem" tabindex="0">
+                  <svg aria-hidden="true" class="octicon octicon-check select-menu-item-icon" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M12 5l-8 8-4-4 1.5-1.5L4 10l6.5-6.5z"></path></svg>
+                  <div class="select-menu-item-text">
+                    <input checked="checked" id="do_subscribed" name="do" type="radio" value="subscribed" />
+                    <span class="select-menu-item-heading">Watching</span>
+                    <span class="description">Be notified of all conversations.</span>
+                    <span class="js-select-button-text hidden-select-button-text">
+                      <svg aria-hidden="true" class="octicon octicon-eye" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M8.06 2C3 2 0 8 0 8s3 6 8.06 6C13 14 16 8 16 8s-3-6-7.94-6zM8 12c-2.2 0-4-1.78-4-4 0-2.2 1.8-4 4-4 2.22 0 4 1.8 4 4 0 2.22-1.78 4-4 4zm2-4c0 1.11-.89 2-2 2-1.11 0-2-.89-2-2 0-1.11.89-2 2-2 1.11 0 2 .89 2 2z"></path></svg>
+                      Unwatch
+                    </span>
+                  </div>
+                </div>
+
+                <div class="select-menu-item js-navigation-item " role="menuitem" tabindex="0">
+                  <svg aria-hidden="true" class="octicon octicon-check select-menu-item-icon" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M12 5l-8 8-4-4 1.5-1.5L4 10l6.5-6.5z"></path></svg>
+                  <div class="select-menu-item-text">
+                    <input id="do_ignore" name="do" type="radio" value="ignore" />
+                    <span class="select-menu-item-heading">Ignoring</span>
+                    <span class="description">Never be notified.</span>
+                    <span class="js-select-button-text hidden-select-button-text">
+                      <svg aria-hidden="true" class="octicon octicon-mute" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M8 2.81v10.38c0 .67-.81 1-1.28.53L3 10H1c-.55 0-1-.45-1-1V7c0-.55.45-1 1-1h2l3.72-3.72C7.19 1.81 8 2.14 8 2.81zm7.53 3.22l-1.06-1.06-1.97 1.97-1.97-1.97-1.06 1.06L11.44 8 9.47 9.97l1.06 1.06 1.97-1.97 1.97 1.97 1.06-1.06L13.56 8l1.97-1.97z"></path></svg>
+                      Stop ignoring
+                    </span>
+                  </div>
+                </div>
+
+              </div>
+
+            </div>
+          </div>
+        </div>
+</form>
+  </li>
+
+  <li>
+
+  <div class="js-toggler-container js-social-container starring-container ">
+
+    <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/unstar" class="starred" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="8jsT4FwhRy7KoSwMyDFFxWSrBfQf9q6Lff0fzQ8MdEJMZInhum5hVH/n3iXzf6+xNASq1AZqI4hBSUEtSLrMxA==" /></div>
+      <button
+        type="submit"
+        class="btn btn-sm btn-with-count js-toggler-target"
+        aria-label="Unstar this repository" title="Unstar eSACP/SeaLevelDecisions"
+        data-ga-click="Repository, click unstar button, action:blob#show; text:Unstar">
+        <svg aria-hidden="true" class="octicon octicon-star" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M14 6l-4.9-.64L7 1 4.9 5.36 0 6l3.6 3.26L2.67 14 7 11.67 11.33 14l-.93-4.74z"></path></svg>
+        Unstar
+      </button>
+        <a class="social-count js-social-count" href="/eSACP/SeaLevelDecisions/stargazers"
+           aria-label="0 users starred this repository">
+          0
+        </a>
+</form>
+    <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/star" class="unstarred" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="ks2hkl+V8bZA/NAEAqFMMvvxD9L863yDs4o1AGfGlBRqFH/bqDT71HtOaVIWzDHnNEB4JjA7PZjbQqVS5B4k9w==" /></div>
+      <button
+        type="submit"
+        class="btn btn-sm btn-with-count js-toggler-target"
+        aria-label="Star this repository" title="Star eSACP/SeaLevelDecisions"
+        data-ga-click="Repository, click star button, action:blob#show; text:Star">
+        <svg aria-hidden="true" class="octicon octicon-star" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M14 6l-4.9-.64L7 1 4.9 5.36 0 6l3.6 3.26L2.67 14 7 11.67 11.33 14l-.93-4.74z"></path></svg>
+        Star
+      </button>
+        <a class="social-count js-social-count" href="/eSACP/SeaLevelDecisions/stargazers"
+           aria-label="0 users starred this repository">
+          0
+        </a>
+</form>  </div>
+
+  </li>
+
+  <li>
+          <a href="#fork-destination-box" class="btn btn-sm btn-with-count"
+              title="Fork your own copy of eSACP/SeaLevelDecisions to your account"
+              aria-label="Fork your own copy of eSACP/SeaLevelDecisions to your account"
+              rel="facebox"
+              data-ga-click="Repository, show fork modal, action:blob#show; text:Fork">
+              <svg aria-hidden="true" class="octicon octicon-repo-forked" height="16" version="1.1" viewBox="0 0 10 16" width="10"><path d="M8 1a1.993 1.993 0 0 0-1 3.72V6L5 8 3 6V4.72A1.993 1.993 0 0 0 2 1a1.993 1.993 0 0 0-1 3.72V6.5l3 3v1.78A1.993 1.993 0 0 0 5 15a1.993 1.993 0 0 0 1-3.72V9.5l3-3V4.72A1.993 1.993 0 0 0 8 1zM2 4.2C1.34 4.2.8 3.65.8 3c0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2zm3 10c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2zm3-10c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2z"></path></svg>
+            Fork
+          </a>
+
+          <div id="fork-destination-box" style="display: none;">
+            <h2 class="facebox-header" data-facebox-id="facebox-header">Where should we fork this repository?</h2>
+            <include-fragment src=""
+                class="js-fork-select-fragment fork-select-fragment"
+                data-url="/eSACP/SeaLevelDecisions/fork?fragment=1">
+              <img alt="Loading" height="64" src="https://assets-cdn.github.com/images/spinners/octocat-spinner-128.gif" width="64" />
+            </include-fragment>
+          </div>
+
+    <a href="/eSACP/SeaLevelDecisions/network" class="social-count"
+       aria-label="0 users are forked this repository">
+      0
+    </a>
+  </li>
+</ul>
+
+    <h1 class="private ">
+  <svg aria-hidden="true" class="octicon octicon-lock" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M4 13H3v-1h1v1zm8-6v7c0 .55-.45 1-1 1H1c-.55 0-1-.45-1-1V7c0-.55.45-1 1-1h1V4c0-2.2 1.8-4 4-4s4 1.8 4 4v2h1c.55 0 1 .45 1 1zM3.8 6h4.41V4c0-1.22-.98-2.2-2.2-2.2-1.22 0-2.2.98-2.2 2.2v2H3.8zM11 7H2v7h9V7zM4 8H3v1h1V8zm0 2H3v1h1v-1z"></path></svg>
+  <span class="author" itemprop="author"><a href="/eSACP" class="url fn" rel="author">eSACP</a></span><!--
+--><span class="path-divider">/</span><!--
+--><strong itemprop="name"><a href="/eSACP/SeaLevelDecisions" data-pjax="#js-repo-pjax-container">SeaLevelDecisions</a></strong>
+    <span class="label label-private v-align-middle">Private</span>
+
+</h1>
+
+  </div>
+  <div class="container">
+
+<nav class="reponav js-repo-nav js-sidenav-container-pjax"
+     itemscope
+     itemtype="http://schema.org/BreadcrumbList"
+     role="navigation"
+     data-pjax="#js-repo-pjax-container">
+
+  <span itemscope itemtype="http://schema.org/ListItem" itemprop="itemListElement">
+    <a href="/eSACP/SeaLevelDecisions" aria-selected="true" class="js-selected-navigation-item selected reponav-item" data-hotkey="g c" data-selected-links="repo_source repo_downloads repo_commits repo_releases repo_tags repo_branches /eSACP/SeaLevelDecisions" itemprop="url">
+      <svg aria-hidden="true" class="octicon octicon-code" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M9.5 3L8 4.5 11.5 8 8 11.5 9.5 13 14 8 9.5 3zm-5 0L0 8l4.5 5L6 11.5 2.5 8 6 4.5 4.5 3z"></path></svg>
+      <span itemprop="name">Code</span>
+      <meta itemprop="position" content="1">
+</a>  </span>
+
+    <span itemscope itemtype="http://schema.org/ListItem" itemprop="itemListElement">
+      <a href="/eSACP/SeaLevelDecisions/issues" class="js-selected-navigation-item reponav-item" data-hotkey="g i" data-selected-links="repo_issues repo_labels repo_milestones /eSACP/SeaLevelDecisions/issues" itemprop="url">
+        <svg aria-hidden="true" class="octicon octicon-issue-opened" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M7 2.3c3.14 0 5.7 2.56 5.7 5.7s-2.56 5.7-5.7 5.7A5.71 5.71 0 0 1 1.3 8c0-3.14 2.56-5.7 5.7-5.7zM7 1C3.14 1 0 4.14 0 8s3.14 7 7 7 7-3.14 7-7-3.14-7-7-7zm1 3H6v5h2V4zm0 6H6v2h2v-2z"></path></svg>
+        <span itemprop="name">Issues</span>
+        <span class="counter">0</span>
+        <meta itemprop="position" content="2">
+</a>    </span>
+
+  <span itemscope itemtype="http://schema.org/ListItem" itemprop="itemListElement">
+    <a href="/eSACP/SeaLevelDecisions/pulls" class="js-selected-navigation-item reponav-item" data-hotkey="g p" data-selected-links="repo_pulls /eSACP/SeaLevelDecisions/pulls" itemprop="url">
+      <svg aria-hidden="true" class="octicon octicon-git-pull-request" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M11 11.28V5c-.03-.78-.34-1.47-.94-2.06C9.46 2.35 8.78 2.03 8 2H7V0L4 3l3 3V4h1c.27.02.48.11.69.31.21.2.3.42.31.69v6.28A1.993 1.993 0 0 0 10 15a1.993 1.993 0 0 0 1-3.72zm-1 2.92c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2zM4 3c0-1.11-.89-2-2-2a1.993 1.993 0 0 0-1 3.72v6.56A1.993 1.993 0 0 0 2 15a1.993 1.993 0 0 0 1-3.72V4.72c.59-.34 1-.98 1-1.72zm-.8 10c0 .66-.55 1.2-1.2 1.2-.65 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2zM2 4.2C1.34 4.2.8 3.65.8 3c0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2z"></path></svg>
+      <span itemprop="name">Pull requests</span>
+      <span class="counter">0</span>
+      <meta itemprop="position" content="3">
+</a>  </span>
+
+  <a href="/eSACP/SeaLevelDecisions/projects" class="js-selected-navigation-item reponav-item" data-selected-links="repo_projects new_repo_project repo_project /eSACP/SeaLevelDecisions/projects">
+    <svg class="octicon" aria-hidden="true" version="1.1" width="15" height="16" viewBox="0 0 15 16">
+      <path d="M1 15h13V1H1v14zM15 1v14a1 1 0 0 1-1 1H1a1 1 0 0 1-1-1V1a1 1 0 0 1 1-1h13a1 1 0 0 1 1 1zm-4.41 11h1.82c.59 0 .59-.41.59-1V3c0-.59 0-1-.59-1h-1.82C10 2 10 2.41 10 3v8c0 .59 0 1 .59 1zm-4-2h1.82C9 10 9 9.59 9 9V3c0-.59 0-1-.59-1H6.59C6 2 6 2.41 6 3v6c0 .59 0 1 .59 1zM2 13V3c0-.59 0-1 .59-1h1.82C5 2 5 2.41 5 3v10c0 .59 0 1-.59 1H2.59C2 14 2 13.59 2 13z"></path>
+    </svg>
+    Projects
+    <span class="counter">0</span>
+</a>
+    <a href="/eSACP/SeaLevelDecisions/wiki" class="js-selected-navigation-item reponav-item" data-hotkey="g w" data-selected-links="repo_wiki /eSACP/SeaLevelDecisions/wiki">
+      <svg aria-hidden="true" class="octicon octicon-book" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M3 5h4v1H3V5zm0 3h4V7H3v1zm0 2h4V9H3v1zm11-5h-4v1h4V5zm0 2h-4v1h4V7zm0 2h-4v1h4V9zm2-6v9c0 .55-.45 1-1 1H9.5l-1 1-1-1H2c-.55 0-1-.45-1-1V3c0-.55.45-1 1-1h5.5l1 1 1-1H15c.55 0 1 .45 1 1zm-8 .5L7.5 3H2v9h6V3.5zm7-.5H9.5l-.5.5V12h6V3z"></path></svg>
+      Wiki
+</a>
+
+  <a href="/eSACP/SeaLevelDecisions/pulse" class="js-selected-navigation-item reponav-item" data-selected-links="pulse /eSACP/SeaLevelDecisions/pulse">
+    <svg aria-hidden="true" class="octicon octicon-pulse" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M11.5 8L8.8 5.4 6.6 8.5 5.5 1.6 2.38 8H0v2h3.6l.9-1.8.9 5.4L9 8.5l1.6 1.5H14V8z"></path></svg>
+    Pulse
+</a>
+  <a href="/eSACP/SeaLevelDecisions/graphs" class="js-selected-navigation-item reponav-item" data-selected-links="repo_graphs repo_contributors /eSACP/SeaLevelDecisions/graphs">
+    <svg aria-hidden="true" class="octicon octicon-graph" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M16 14v1H0V0h1v14h15zM5 13H3V8h2v5zm4 0H7V3h2v10zm4 0h-2V6h2v7z"></path></svg>
+    Graphs
+</a>
+    <a href="/eSACP/SeaLevelDecisions/settings" class="js-selected-navigation-item reponav-item" data-selected-links="repo_settings repo_branch_settings hooks integration_installations /eSACP/SeaLevelDecisions/settings">
+      <svg aria-hidden="true" class="octicon octicon-gear" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M14 8.77v-1.6l-1.94-.64-.45-1.09.88-1.84-1.13-1.13-1.81.91-1.09-.45-.69-1.92h-1.6l-.63 1.94-1.11.45-1.84-.88-1.13 1.13.91 1.81-.45 1.09L0 7.23v1.59l1.94.64.45 1.09-.88 1.84 1.13 1.13 1.81-.91 1.09.45.69 1.92h1.59l.63-1.94 1.11-.45 1.84.88 1.13-1.13-.92-1.81.47-1.09L14 8.75v.02zM7 11c-1.66 0-3-1.34-3-3s1.34-3 3-3 3 1.34 3 3-1.34 3-3 3z"></path></svg>
+      Settings
+</a>
+</nav>
+
+  </div>
+</div>
+
+<div class="container new-discussion-timeline experiment-repo-nav">
+  <div class="repository-content">
+
+<a href="/eSACP/SeaLevelDecisions/blob/7bf571a93d136b78617ea889ed7d591e0e09c657/submission/paper.tex" class="d-none js-permalink-shortcut" data-hotkey="y">Permalink</a>
+
+<!-- blob contrib key: blob_contributors:v21:af5a697fb1872010c04d610dfe2db4fd -->
+
+<div class="file-navigation js-zeroclipboard-container">
+
+<div class="select-menu branch-select-menu js-menu-container js-select-menu float-left">
+  <button class="btn btn-sm select-menu-button js-menu-target css-truncate" data-hotkey="w"
+
+    type="button" aria-label="Switch branches or tags" tabindex="0" aria-haspopup="true">
+    <i>Branch:</i>
+    <span class="js-select-button css-truncate-target">master</span>
+  </button>
+
+  <div class="select-menu-modal-holder js-menu-content js-navigation-container" data-pjax aria-hidden="true">
+
+    <div class="select-menu-modal">
+      <div class="select-menu-header">
+        <svg aria-label="Close" class="octicon octicon-x js-menu-close" height="16" role="img" version="1.1" viewBox="0 0 12 16" width="12"><path d="M7.48 8l3.75 3.75-1.48 1.48L6 9.48l-3.75 3.75-1.48-1.48L4.52 8 .77 4.25l1.48-1.48L6 6.52l3.75-3.75 1.48 1.48z"></path></svg>
+        <span class="select-menu-title">Switch branches/tags</span>
+      </div>
+
+      <div class="select-menu-filters">
+        <div class="select-menu-text-filter">
+          <input type="text" aria-label="Find or create a branch…" id="context-commitish-filter-field" class="form-control js-filterable-field js-navigation-enable" placeholder="Find or create a branch…">
+        </div>
+        <div class="select-menu-tabs">
+          <ul>
+            <li class="select-menu-tab">
+              <a href="#" data-tab-filter="branches" data-filter-placeholder="Find or create a branch…" class="js-select-menu-tab" role="tab">Branches</a>
+            </li>
+            <li class="select-menu-tab">
+              <a href="#" data-tab-filter="tags" data-filter-placeholder="Find a tag…" class="js-select-menu-tab" role="tab">Tags</a>
+            </li>
+          </ul>
+        </div>
+      </div>
+
+      <div class="select-menu-list select-menu-tab-bucket js-select-menu-tab-bucket" data-tab-filter="branches" role="menu">
+
+        <div data-filterable-for="context-commitish-filter-field" data-filterable-type="substring">
+
+            <a class="select-menu-item js-navigation-item js-navigation-open selected"
+               href="/eSACP/SeaLevelDecisions/blob/master/submission/paper.tex"
+               data-name="master"
+               data-skip-pjax="true"
+               rel="nofollow">
+              <svg aria-hidden="true" class="octicon octicon-check select-menu-item-icon" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M12 5l-8 8-4-4 1.5-1.5L4 10l6.5-6.5z"></path></svg>
+              <span class="select-menu-item-text css-truncate-target js-select-menu-filter-text">
+                master
+              </span>
+            </a>
+        </div>
+
+          <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/branches" class="js-create-branch select-menu-item select-menu-new-item-form js-navigation-item js-new-item-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="DcMyzVBLpaCbk6kx2+mnhjRNC9PhE1BLNhUynTbqPElsNsRr3UzZrSuzCAJO6wgj6TKTs9w6R/B+N0T7ENYGjQ==" /></div>
+          <svg aria-hidden="true" class="octicon octicon-git-branch select-menu-item-icon" height="16" version="1.1" viewBox="0 0 10 16" width="10"><path d="M10 5c0-1.11-.89-2-2-2a1.993 1.993 0 0 0-1 3.72v.3c-.02.52-.23.98-.63 1.38-.4.4-.86.61-1.38.63-.83.02-1.48.16-2 .45V4.72a1.993 1.993 0 0 0-1-3.72C.88 1 0 1.89 0 3a2 2 0 0 0 1 1.72v6.56c-.59.35-1 .99-1 1.72 0 1.11.89 2 2 2 1.11 0 2-.89 2-2 0-.53-.2-1-.53-1.36.09-.06.48-.41.59-.47.25-.11.56-.17.94-.17 1.05-.05 1.95-.45 2.75-1.25S8.95 7.77 9 6.73h-.02C9.59 6.37 10 5.73 10 5zM2 1.8c.66 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2C1.35 4.2.8 3.65.8 3c0-.65.55-1.2 1.2-1.2zm0 12.41c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2zm6-8c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2z"></path></svg>
+            <div class="select-menu-item-text">
+              <span class="select-menu-item-heading">Create branch: <span class="js-new-item-name"></span></span>
+              <span class="description">from ‘master’</span>
+            </div>
+            <input type="hidden" name="name" id="name" class="js-new-item-value">
+            <input type="hidden" name="branch" id="branch" value="master">
+            <input type="hidden" name="path" id="path" value="submission/paper.tex">
+</form>
+      </div>
+
+      <div class="select-menu-list select-menu-tab-bucket js-select-menu-tab-bucket" data-tab-filter="tags">
+        <div data-filterable-for="context-commitish-filter-field" data-filterable-type="substring">
+
+        </div>
+
+        <div class="select-menu-no-results">Nothing to show</div>
+      </div>
+
+    </div>
+  </div>
+</div>
+
+  <div class="BtnGroup float-right">
+    <a href="/eSACP/SeaLevelDecisions/find/master"
+          class="js-pjax-capture-input btn btn-sm BtnGroup-item"
+          data-pjax
+          data-hotkey="t">
+      Find file
+    </a>
+    <button aria-label="Copy file path to clipboard" class="js-zeroclipboard btn btn-sm BtnGroup-item tooltipped tooltipped-s" data-copied-hint="Copied!" type="button">Copy path</button>
+  </div>
+  <div class="breadcrumb js-zeroclipboard-target">
+    <span class="repo-root js-repo-root"><span class="js-path-segment"><a href="/eSACP/SeaLevelDecisions"><span>SeaLevelDecisions</span></a></span></span><span class="separator">/</span><span class="js-path-segment"><a href="/eSACP/SeaLevelDecisions/tree/master/submission"><span>submission</span></a></span><span class="separator">/</span><strong class="final-path">paper.tex</strong>
+  </div>
+</div>
+
+<include-fragment class="commit-tease" src="/eSACP/SeaLevelDecisions/contributors/master/submission/paper.tex">
+  <div>
+    Fetching contributors&hellip;
+  </div>
+
+  <div class="commit-tease-contributors">
+    <img alt="" class="loader-loading float-left" height="16" src="https://assets-cdn.github.com/images/spinners/octocat-spinner-32-EAF2F5.gif" width="16" />
+    <span class="loader-error">Cannot retrieve contributors at this time</span>
+  </div>
+</include-fragment>
+<div class="file">
+  <div class="file-header">
+  <div class="file-actions">
+
+    <div class="BtnGroup">
+      <a href="/eSACP/SeaLevelDecisions/raw/master/submission/paper.tex" class="btn btn-sm BtnGroup-item" id="raw-url">Raw</a>
+        <a href="/eSACP/SeaLevelDecisions/blame/master/submission/paper.tex" class="btn btn-sm js-update-url-with-hash BtnGroup-item">Blame</a>
+      <a href="/eSACP/SeaLevelDecisions/commits/master/submission/paper.tex" class="btn btn-sm BtnGroup-item" rel="nofollow">History</a>
+    </div>
+
+        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/edit/master/submission/paper.tex" class="inline-form js-update-url-with-hash" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="JHAATQHJKo7GYe0P+sOvc4ri8m7QuFAOQBjGHnIR/IVCje5LFQxPY/7jS0hKVjDbuBVmyoHhmKRJQS4L7b7bMA==" /></div>
+          <button class="btn-octicon tooltipped tooltipped-nw" type="submit"
+            aria-label="Edit this file" data-hotkey="e" data-disable-with>
+            <svg aria-hidden="true" class="octicon octicon-pencil" height="16" version="1.1" viewBox="0 0 14 16" width="14"><path d="M0 12v3h3l8-8-3-3-8 8zm3 2H1v-2h1v1h1v1zm10.3-9.3L12 6 9 3l1.3-1.3a.996.996 0 0 1 1.41 0l1.59 1.59c.39.39.39 1.02 0 1.41z"></path></svg>
+          </button>
+</form>        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/delete/master/submission/paper.tex" class="inline-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="GKa51xlrlUKP7iYS4nEBuU41UNrSrGU/IsCaU7p1GOZljm23SsoIOnuLz6TcbAiZmWeku1jesTDXHdkYwL0JMg==" /></div>
+          <button class="btn-octicon btn-octicon-danger tooltipped tooltipped-nw" type="submit"
+            aria-label="Delete this file" data-disable-with>
+            <svg aria-hidden="true" class="octicon octicon-trashcan" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M11 2H9c0-.55-.45-1-1-1H5c-.55 0-1 .45-1 1H2c-.55 0-1 .45-1 1v1c0 .55.45 1 1 1v9c0 .55.45 1 1 1h7c.55 0 1-.45 1-1V5c.55 0 1-.45 1-1V3c0-.55-.45-1-1-1zm-1 12H3V5h1v8h1V5h1v8h1V5h1v8h1V5h1v9zm1-10H2V3h9v1z"></path></svg>
+          </button>
+</form>  </div>
+
+  <div class="file-info">
+      186 lines (128 sloc)
+      <span class="file-info-divider"></span>
+    10.1 KB
+  </div>
+</div>
+
+  <div itemprop="text" class="blob-wrapper data type-tex">
+      <table class="highlight tab-size js-file-line-container" data-tab-size="8">
+      <tr>
+        <td id="L1" class="blob-num js-line-number" data-line-number="1"></td>
+        <td id="LC1" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%\documentclass[wrr]{agutex}</span></td>
+      </tr>
+      <tr>
+        <td id="L2" class="blob-num js-line-number" data-line-number="2"></td>
+        <td id="LC2" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\documentclass</span>[wrr, draft]{agutex}</td>
+      </tr>
+      <tr>
+        <td id="L3" class="blob-num js-line-number" data-line-number="3"></td>
+        <td id="LC3" class="blob-code blob-code-inner js-file-line"><span class="pl-c">% Author names in capital letters:</span></td>
+      </tr>
+      <tr>
+        <td id="L4" class="blob-num js-line-number" data-line-number="4"></td>
+        <td id="LC4" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\authorrunninghead</span>{1st AUTHOR ET AL.}</td>
+      </tr>
+      <tr>
+        <td id="L5" class="blob-num js-line-number" data-line-number="5"></td>
+        <td id="LC5" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L6" class="blob-num js-line-number" data-line-number="6"></td>
+        <td id="LC6" class="blob-code blob-code-inner js-file-line"><span class="pl-c">% Shorter version of title entered in capital letters:</span></td>
+      </tr>
+      <tr>
+        <td id="L7" class="blob-num js-line-number" data-line-number="7"></td>
+        <td id="LC7" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\titlerunninghead</span>{ADAPTING TO UNCERTAIN SEA LEVEL RISE}</td>
+      </tr>
+      <tr>
+        <td id="L8" class="blob-num js-line-number" data-line-number="8"></td>
+        <td id="LC8" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L9" class="blob-num js-line-number" data-line-number="9"></td>
+        <td id="LC9" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%Corresponding author mailing address and e-mail address:</span></td>
+      </tr>
+      <tr>
+        <td id="L10" class="blob-num js-line-number" data-line-number="10"></td>
+        <td id="LC10" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\authoraddr</span>{Corresponding author: Name, Address. (email)}</td>
+      </tr>
+      <tr>
+        <td id="L11" class="blob-num js-line-number" data-line-number="11"></td>
+        <td id="LC11" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L12" class="blob-num js-line-number" data-line-number="12"></td>
+        <td id="LC12" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\usepackage</span>{bbm, amsmath}</td>
+      </tr>
+      <tr>
+        <td id="L13" class="blob-num js-line-number" data-line-number="13"></td>
+        <td id="LC13" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%\usepackage[dvips]{graphicx}</span></td>
+      </tr>
+      <tr>
+        <td id="L14" class="blob-num js-line-number" data-line-number="14"></td>
+        <td id="LC14" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\usepackage</span>{graphicx}</td>
+      </tr>
+      <tr>
+        <td id="L15" class="blob-num js-line-number" data-line-number="15"></td>
+        <td id="LC15" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\setkeys</span>{Gin}{draft=false}</td>
+      </tr>
+      <tr>
+        <td id="L16" class="blob-num js-line-number" data-line-number="16"></td>
+        <td id="LC16" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\usepackage</span>{paralist, booktabs}</td>
+      </tr>
+      <tr>
+        <td id="L17" class="blob-num js-line-number" data-line-number="17"></td>
+        <td id="LC17" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\usepackage</span>{url}</td>
+      </tr>
+      <tr>
+        <td id="L18" class="blob-num js-line-number" data-line-number="18"></td>
+        <td id="LC18" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\usepackage</span>{color}</td>
+      </tr>
+      <tr>
+        <td id="L19" class="blob-num js-line-number" data-line-number="19"></td>
+        <td id="LC19" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L20" class="blob-num js-line-number" data-line-number="20"></td>
+        <td id="LC20" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L21" class="blob-num js-line-number" data-line-number="21"></td>
+        <td id="LC21" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\usepackage</span>{lineno}</td>
+      </tr>
+      <tr>
+        <td id="L22" class="blob-num js-line-number" data-line-number="22"></td>
+        <td id="LC22" class="blob-code blob-code-inner js-file-line"> <span class="pl-c1">\linenumbers</span>*[1]</td>
+      </tr>
+      <tr>
+        <td id="L23" class="blob-num js-line-number" data-line-number="23"></td>
+        <td id="LC23" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  To add line numbers to lines with equations:</span></td>
+      </tr>
+      <tr>
+        <td id="L24" class="blob-num js-line-number" data-line-number="24"></td>
+        <td id="LC24" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L25" class="blob-num js-line-number" data-line-number="25"></td>
+        <td id="LC25" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{document}</td>
+      </tr>
+      <tr>
+        <td id="L26" class="blob-num js-line-number" data-line-number="26"></td>
+        <td id="LC26" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L27" class="blob-num js-line-number" data-line-number="27"></td>
+        <td id="LC27" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L28" class="blob-num js-line-number" data-line-number="28"></td>
+        <td id="LC28" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  TITLE</span></td>
+      </tr>
+      <tr>
+        <td id="L29" class="blob-num js-line-number" data-line-number="29"></td>
+        <td id="LC29" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L30" class="blob-num js-line-number" data-line-number="30"></td>
+        <td id="LC30" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L31" class="blob-num js-line-number" data-line-number="31"></td>
+        <td id="LC31" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\title</span>{I don&#39;t know, are you sure you want to do this?}</td>
+      </tr>
+      <tr>
+        <td id="L32" class="blob-num js-line-number" data-line-number="32"></td>
+        <td id="LC32" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L33" class="blob-num js-line-number" data-line-number="33"></td>
+        <td id="LC33" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L34" class="blob-num js-line-number" data-line-number="34"></td>
+        <td id="LC34" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  AUTHORS AND AFFILIATIONS</span></td>
+      </tr>
+      <tr>
+        <td id="L35" class="blob-num js-line-number" data-line-number="35"></td>
+        <td id="LC35" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L36" class="blob-num js-line-number" data-line-number="36"></td>
+        <td id="LC36" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L37" class="blob-num js-line-number" data-line-number="37"></td>
+        <td id="LC37" class="blob-code blob-code-inner js-file-line"><span class="pl-c">% \altaffilmark will produce footnote;</span></td>
+      </tr>
+      <tr>
+        <td id="L38" class="blob-num js-line-number" data-line-number="38"></td>
+        <td id="LC38" class="blob-code blob-code-inner js-file-line"><span class="pl-c">% matching \altaffiltext will appear at bottom of page.</span></td>
+      </tr>
+      <tr>
+        <td id="L39" class="blob-num js-line-number" data-line-number="39"></td>
+        <td id="LC39" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L40" class="blob-num js-line-number" data-line-number="40"></td>
+        <td id="LC40" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\authors</span>{T.
+Thorarinsdottir<span class="pl-c1">\altaffilmark</span>{1}, P.
+Guttorp<span class="pl-c1">\altaffilmark</span>{1}, M.
+Drews<span class="pl-c1">\altaffilmark</span>{2}}</td>
+      </tr>
+      <tr>
+        <td id="L41" class="blob-num js-line-number" data-line-number="41"></td>
+        <td id="LC41" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L42" class="blob-num js-line-number" data-line-number="42"></td>
+        <td id="LC42" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\altaffiltext</span>{1}{Norwegian Computing Centere}</td>
+      </tr>
+      <tr>
+        <td id="L43" class="blob-num js-line-number" data-line-number="43"></td>
+        <td id="LC43" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L44" class="blob-num js-line-number" data-line-number="44"></td>
+        <td id="LC44" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\altaffiltext</span>{2}{Danish Technological University}</td>
+      </tr>
+      <tr>
+        <td id="L45" class="blob-num js-line-number" data-line-number="45"></td>
+        <td id="LC45" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L46" class="blob-num js-line-number" data-line-number="46"></td>
+        <td id="LC46" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\altaffiltext</span>{3}{Affiliation three}</td>
+      </tr>
+      <tr>
+        <td id="L47" class="blob-num js-line-number" data-line-number="47"></td>
+        <td id="LC47" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L48" class="blob-num js-line-number" data-line-number="48"></td>
+        <td id="LC48" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L49" class="blob-num js-line-number" data-line-number="49"></td>
+        <td id="LC49" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  ABSTRACT</span></td>
+      </tr>
+      <tr>
+        <td id="L50" class="blob-num js-line-number" data-line-number="50"></td>
+        <td id="LC50" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L51" class="blob-num js-line-number" data-line-number="51"></td>
+        <td id="LC51" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L52" class="blob-num js-line-number" data-line-number="52"></td>
+        <td id="LC52" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{abstract}</td>
+      </tr>
+      <tr>
+        <td id="L53" class="blob-num js-line-number" data-line-number="53"></td>
+        <td id="LC53" class="blob-code blob-code-inner js-file-line">...</td>
+      </tr>
+      <tr>
+        <td id="L54" class="blob-num js-line-number" data-line-number="54"></td>
+        <td id="LC54" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{abstract}</td>
+      </tr>
+      <tr>
+        <td id="L55" class="blob-num js-line-number" data-line-number="55"></td>
+        <td id="LC55" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L56" class="blob-num js-line-number" data-line-number="56"></td>
+        <td id="LC56" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L57" class="blob-num js-line-number" data-line-number="57"></td>
+        <td id="LC57" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L58" class="blob-num js-line-number" data-line-number="58"></td>
+        <td id="LC58" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  BEGIN ARTICLE</span></td>
+      </tr>
+      <tr>
+        <td id="L59" class="blob-num js-line-number" data-line-number="59"></td>
+        <td id="LC59" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L60" class="blob-num js-line-number" data-line-number="60"></td>
+        <td id="LC60" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L61" class="blob-num js-line-number" data-line-number="61"></td>
+        <td id="LC61" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{article}</td>
+      </tr>
+      <tr>
+        <td id="L62" class="blob-num js-line-number" data-line-number="62"></td>
+        <td id="LC62" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L63" class="blob-num js-line-number" data-line-number="63"></td>
+        <td id="LC63" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L64" class="blob-num js-line-number" data-line-number="64"></td>
+        <td id="LC64" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L65" class="blob-num js-line-number" data-line-number="65"></td>
+        <td id="LC65" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\section</span>{Introduction}<span class="pl-c1">\label</span>{sec:intro}</td>
+      </tr>
+      <tr>
+        <td id="L66" class="blob-num js-line-number" data-line-number="66"></td>
+        <td id="LC66" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L67" class="blob-num js-line-number" data-line-number="67"></td>
+        <td id="LC67" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\section</span>{Sea level projections {<span class="pl-c1">\color</span>{blue} (PG)}}</td>
+      </tr>
+      <tr>
+        <td id="L68" class="blob-num js-line-number" data-line-number="68"></td>
+        <td id="LC68" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L69" class="blob-num js-line-number" data-line-number="69"></td>
+        <td id="LC69" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L70" class="blob-num js-line-number" data-line-number="70"></td>
+        <td id="LC70" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L71" class="blob-num js-line-number" data-line-number="71"></td>
+        <td id="LC71" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Global sea level}</td>
+      </tr>
+      <tr>
+        <td id="L72" class="blob-num js-line-number" data-line-number="72"></td>
+        <td id="LC72" class="blob-code blob-code-inner js-file-line">Most climate models do not explicitly provide sea level as an output of the calculations.
+Rather, the IPCC AR5 report <span class="pl-c1">\citep</span>[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age.
+Judging from the supplementary material to <span class="pl-c1">\citet</span>[ch.~13]{ipcc}, the uncertainty assessment is only based on the spread of the ensemble of temperature projections, not on the additional uncertainty in the ice models used.</td>
+      </tr>
+      <tr>
+        <td id="L73" class="blob-num js-line-number" data-line-number="73"></td>
+        <td id="LC73" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L74" class="blob-num js-line-number" data-line-number="74"></td>
+        <td id="LC74" class="blob-code blob-code-inner js-file-line">We will instead use the empirical approach of Rahmstorf and collaborators <span class="pl-c1">\citep</span>{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of <span class="pl-c1">\citet</span>{Bolin2014a} to relate global annual mean temperature anomalies <span class="pl-c1">\citep</span>{giss} to global mean sea level anomalies <span class="pl-c1">\citep</span>{csiro}.
+</td>
+      </tr>
+      <tr>
+        <td id="L75" class="blob-num js-line-number" data-line-number="75"></td>
+        <td id="LC75" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L76" class="blob-num js-line-number" data-line-number="76"></td>
+        <td id="LC76" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%Figure showing historical relationship?</span></td>
+      </tr>
+      <tr>
+        <td id="L77" class="blob-num js-line-number" data-line-number="77"></td>
+        <td id="LC77" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L78" class="blob-num js-line-number" data-line-number="78"></td>
+        <td id="LC78" class="blob-code blob-code-inner js-file-line">We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment <span class="pl-c1">\citep</span>{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection <span class="pl-c1">\ref</span>{unc_ass}).
+</td>
+      </tr>
+      <tr>
+        <td id="L79" class="blob-num js-line-number" data-line-number="79"></td>
+        <td id="LC79" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%Figure with sea level projections &amp; IPCC projections?</span></td>
+      </tr>
+      <tr>
+        <td id="L80" class="blob-num js-line-number" data-line-number="80"></td>
+        <td id="LC80" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L81" class="blob-num js-line-number" data-line-number="81"></td>
+        <td id="LC81" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L82" class="blob-num js-line-number" data-line-number="82"></td>
+        <td id="LC82" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Local sea level}</td>
+      </tr>
+      <tr>
+        <td id="L83" class="blob-num js-line-number" data-line-number="83"></td>
+        <td id="LC83" class="blob-code blob-code-inner js-file-line">In order to get from global sea level projections to local ones, it is important to note that sea level rise not is uniform over the globe.
+Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.</td>
+      </tr>
+      <tr>
+        <td id="L84" class="blob-num js-line-number" data-line-number="84"></td>
+        <td id="LC84" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%Gravitaional effects</span></td>
+      </tr>
+      <tr>
+        <td id="L85" class="blob-num js-line-number" data-line-number="85"></td>
+        <td id="LC85" class="blob-code blob-code-inner js-file-line">Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. </td>
+      </tr>
+      <tr>
+        <td id="L86" class="blob-num js-line-number" data-line-number="86"></td>
+        <td id="LC86" class="blob-code blob-code-inner js-file-line">Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model.
+Figure <span class="pl-c1">\ref</span>{fig:obs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
+The time series regression uses an ARMA(1,1)-model <span class="pl-c1">\citep</span>{boxjenkins}.</td>
+      </tr>
+      <tr>
+        <td id="L87" class="blob-num js-line-number" data-line-number="87"></td>
+        <td id="LC87" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{figure}</td>
+      </tr>
+      <tr>
+        <td id="L88" class="blob-num js-line-number" data-line-number="88"></td>
+        <td id="LC88" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{center}</td>
+      </tr>
+      <tr>
+        <td id="L89" class="blob-num js-line-number" data-line-number="89"></td>
+        <td id="LC89" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\includegraphics</span>[width=39pc]{bergenfit.png}</td>
+      </tr>
+      <tr>
+        <td id="L90" class="blob-num js-line-number" data-line-number="90"></td>
+        <td id="LC90" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\caption</span>{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of <span class="pl-c1">\citet</span>{csiro}.
+The straight line is the time series regression line.}</td>
+      </tr>
+      <tr>
+        <td id="L91" class="blob-num js-line-number" data-line-number="91"></td>
+        <td id="LC91" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\label</span>{fig:obs}</td>
+      </tr>
+      <tr>
+        <td id="L92" class="blob-num js-line-number" data-line-number="92"></td>
+        <td id="LC92" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{center}</td>
+      </tr>
+      <tr>
+        <td id="L93" class="blob-num js-line-number" data-line-number="93"></td>
+        <td id="LC93" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{figure}</td>
+      </tr>
+      <tr>
+        <td id="L94" class="blob-num js-line-number" data-line-number="94"></td>
+        <td id="LC94" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L95" class="blob-num js-line-number" data-line-number="95"></td>
+        <td id="LC95" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L96" class="blob-num js-line-number" data-line-number="96"></td>
+        <td id="LC96" class="blob-code blob-code-inner js-file-line">The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one.
+Each climate model temperature projection yields a different local sea level projection.</td>
+      </tr>
+      <tr>
+        <td id="L97" class="blob-num js-line-number" data-line-number="97"></td>
+        <td id="LC97" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L98" class="blob-num js-line-number" data-line-number="98"></td>
+        <td id="LC98" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Uncertainty assessment}</td>
+      </tr>
+      <tr>
+        <td id="L99" class="blob-num js-line-number" data-line-number="99"></td>
+        <td id="LC99" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\label</span>{unc_ass}</td>
+      </tr>
+      <tr>
+        <td id="L100" class="blob-num js-line-number" data-line-number="100"></td>
+        <td id="LC100" class="blob-code blob-code-inner js-file-line">Following the approach of <span class="pl-c1">\citet</span>{Guttorp2014} we assess the uncertainty in the local sea level projections taking into account the variability between the climate projections used, the uncertainties in the regressions of global mean temperature on global mean sea level and of global on local sea level.
 We express the sea level projection uncertainty in terms of a confidence band that is simultaneously of the intended  level  for all projection years.
-This allows us, for example, to get a confidence band for the years when a given sea level rise is obtained.
-
-%Figure showing Bergen and Copenhagen slr projections
-
-\subsection{Limitations of the sea level projections}
-The main assumption is using historical relationships in statistical projections of the type used in this paper is that there is no major change in how temperature relates to sea level, globally and locally.
-Among the factors that may invalidate this approach are changes in water storage on land (removing water from the oceans), excessive siphoning of groundwater (resulting in land subsidence), changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
+This allows us, for example, to get a confidence band for the years when a given sea level rise is obtained.</td>
+      </tr>
+      <tr>
+        <td id="L101" class="blob-num js-line-number" data-line-number="101"></td>
+        <td id="LC101" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L102" class="blob-num js-line-number" data-line-number="102"></td>
+        <td id="LC102" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{figure}</td>
+      </tr>
+      <tr>
+        <td id="L103" class="blob-num js-line-number" data-line-number="103"></td>
+        <td id="LC103" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{center}</td>
+      </tr>
+      <tr>
+        <td id="L104" class="blob-num js-line-number" data-line-number="104"></td>
+        <td id="LC104" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\includegraphics</span>[width=39pc]{bergen_ci.png}</td>
+      </tr>
+      <tr>
+        <td id="L105" class="blob-num js-line-number" data-line-number="105"></td>
+        <td id="LC105" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\caption</span>{Simultaneous 90<span class="pl-cce">\%</span> confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
+The sea level data are shown in blue and end in 2015.
+The thin red lines are the projections without uncertainty based on each of the climate models.
+The dashed purple lines connect pointwise confidence intervals for each year.
+}</td>
+      </tr>
+      <tr>
+        <td id="L106" class="blob-num js-line-number" data-line-number="106"></td>
+        <td id="LC106" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\label</span>{fig:ci}</td>
+      </tr>
+      <tr>
+        <td id="L107" class="blob-num js-line-number" data-line-number="107"></td>
+        <td id="LC107" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{center}</td>
+      </tr>
+      <tr>
+        <td id="L108" class="blob-num js-line-number" data-line-number="108"></td>
+        <td id="LC108" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{figure}</td>
+      </tr>
+      <tr>
+        <td id="L109" class="blob-num js-line-number" data-line-number="109"></td>
+        <td id="LC109" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L110" class="blob-num js-line-number" data-line-number="110"></td>
+        <td id="LC110" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L111" class="blob-num js-line-number" data-line-number="111"></td>
+        <td id="LC111" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Limitations of the sea level projections}</td>
+      </tr>
+      <tr>
+        <td id="L112" class="blob-num js-line-number" data-line-number="112"></td>
+        <td id="LC112" class="blob-code blob-code-inner js-file-line">The main assumption is using historical relationships in statistical projections of the type used in this paper is that there is no major change in how temperature relates to sea level, globally and locally.
+Among the factors that may invalidate this approach are changes in water storage on land (removing water from the oceans), excessive siphoning of groundwater (resulting in land subsidence), changes in the rates of glacial and land ice melt, and changes in Earth&#39;s gravitational field due to transfer of mass from land ice to ocean water.
 For example, the rate of ice melt on Greenland may suddenly increase substantially due to intense warming of both air and sea water.
 Our current climate models are not able to resolve the ice processes sufficiently to include such so called tipping points into the projections.
-Also, the IPCC scenarios do not include changes in water usage \citep{change}.
+Also, the IPCC scenarios do not include changes in water usage <span class="pl-c1">\citep</span>{change}.</td>
+      </tr>
+      <tr>
+        <td id="L113" class="blob-num js-line-number" data-line-number="113"></td>
+        <td id="LC113" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L114" class="blob-num js-line-number" data-line-number="114"></td>
+        <td id="LC114" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\section</span>{Decision tools {<span class="pl-c1">\color</span>{blue} (KdB, MD, TT)}}</td>
+      </tr>
+      <tr>
+        <td id="L115" class="blob-num js-line-number" data-line-number="115"></td>
+        <td id="LC115" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L116" class="blob-num js-line-number" data-line-number="116"></td>
+        <td id="LC116" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Timing of adaptation measures}</td>
+      </tr>
+      <tr>
+        <td id="L117" class="blob-num js-line-number" data-line-number="117"></td>
+        <td id="LC117" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L118" class="blob-num js-line-number" data-line-number="118"></td>
+        <td id="LC118" class="blob-code blob-code-inner js-file-line">We consider adaptation decision making related to the timing of proactive adaptation measures.
+That is, the goal is to adapt to sea level rise before major damages occur.
+In a cost-benefit framework, an investment should be delayed as long as the benefits of delay (avoided investment costs) are greater than the associated costs (higher climate change damages) <span class="pl-c1">\citep</span>{Fankhauser&amp;1999}.</td>
+      </tr>
+      <tr>
+        <td id="L119" class="blob-num js-line-number" data-line-number="119"></td>
+        <td id="LC119" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L120" class="blob-num js-line-number" data-line-number="120"></td>
+        <td id="LC120" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\cite</span>{Fankhauser&amp;1999} describe a deterministic framework where an adaptation investment of <span class="pl-s"><span class="pl-pds">$</span>C^<span class="pl-c1">0</span><span class="pl-pds">$</span></span> now (at time <span class="pl-s"><span class="pl-pds">$</span>n=<span class="pl-c1">0</span><span class="pl-pds">$</span></span>) leads to unmitigated damage of <span class="pl-s"><span class="pl-pds">$</span>d_<span class="pl-c1">0</span>^<span class="pl-c1">0</span><span class="pl-pds">$</span></span> in period <span class="pl-s"><span class="pl-pds">$</span><span class="pl-c1">0</span><span class="pl-pds">$</span></span>, and a stream of partially mitigated damages <span class="pl-s"><span class="pl-pds">$</span>d_t^<span class="pl-c1">0</span><span class="pl-pds">$</span></span> in periods <span class="pl-s"><span class="pl-pds">$</span>t=<span class="pl-c1">1</span>,<span class="pl-c1">2</span>,<span class="pl-c1">\ldots</span><span class="pl-pds">$</span></span>.
+If <span class="pl-s"><span class="pl-pds">$</span>r<span class="pl-pds">$</span></span> is the discount rate, the net present value damage, <span class="pl-s"><span class="pl-pds">$</span>D^<span class="pl-c1">0</span><span class="pl-pds">$</span></span>, associated with this investment is</td>
+      </tr>
+      <tr>
+        <td id="L121" class="blob-num js-line-number" data-line-number="121"></td>
+        <td id="LC121" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{linenomath*}</td>
+      </tr>
+      <tr>
+        <td id="L122" class="blob-num js-line-number" data-line-number="122"></td>
+        <td id="LC122" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{equation}<span class="pl-c1">\label</span>{eq:deterministic damage}</td>
+      </tr>
+      <tr>
+        <td id="L123" class="blob-num js-line-number" data-line-number="123"></td>
+        <td id="LC123" class="blob-code blob-code-inner js-file-line">D^0 = C^0 + d_0^0 + <span class="pl-c1">\frac</span>{d_1^0}{1+r} + <span class="pl-c1">\frac</span>{d_2^0}{(1+r)^2} + <span class="pl-c1">\cdots</span>  </td>
+      </tr>
+      <tr>
+        <td id="L124" class="blob-num js-line-number" data-line-number="124"></td>
+        <td id="LC124" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{equation}</td>
+      </tr>
+      <tr>
+        <td id="L125" class="blob-num js-line-number" data-line-number="125"></td>
+        <td id="LC125" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{linenomath*}</td>
+      </tr>
+      <tr>
+        <td id="L126" class="blob-num js-line-number" data-line-number="126"></td>
+        <td id="LC126" class="blob-code blob-code-inner js-file-line">In comparison, postponing the adaptation investment to time period <span class="pl-s"><span class="pl-pds">$</span>n=<span class="pl-c1">1</span><span class="pl-pds">$</span></span> would lead to unmitigated damages in periods <span class="pl-s"><span class="pl-pds">$</span><span class="pl-c1">0</span><span class="pl-pds">$</span></span> and <span class="pl-s"><span class="pl-pds">$</span><span class="pl-c1">1</span><span class="pl-pds">$</span></span>, and partially mitigated damages, <span class="pl-s"><span class="pl-pds">$</span>d_t^<span class="pl-c1">1</span><span class="pl-pds">$</span></span>, thereafter.
+The delay would be preferable if</td>
+      </tr>
+      <tr>
+        <td id="L127" class="blob-num js-line-number" data-line-number="127"></td>
+        <td id="LC127" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{linenomath*}</td>
+      </tr>
+      <tr>
+        <td id="L128" class="blob-num js-line-number" data-line-number="128"></td>
+        <td id="LC128" class="blob-code blob-code-inner js-file-line"><span class="pl-cce">\[</span></td>
+      </tr>
+      <tr>
+        <td id="L129" class="blob-num js-line-number" data-line-number="129"></td>
+        <td id="LC129" class="blob-code blob-code-inner js-file-line">C^0 - <span class="pl-c1">\frac</span>{C^1}{(1+r)} &gt; (d_0^1 - d_0^0) + <span class="pl-c1">\frac</span>{d_1^1 - d_1^0}{1+r} + <span class="pl-c1">\frac</span>{d_2^1 - d_2^0}{(1+r)^2} + <span class="pl-c1">\cdots</span></td>
+      </tr>
+      <tr>
+        <td id="L130" class="blob-num js-line-number" data-line-number="130"></td>
+        <td id="LC130" class="blob-code blob-code-inner js-file-line"><span class="pl-cce">\]</span></td>
+      </tr>
+      <tr>
+        <td id="L131" class="blob-num js-line-number" data-line-number="131"></td>
+        <td id="LC131" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{linenomath*}</td>
+      </tr>
+      <tr>
+        <td id="L132" class="blob-num js-line-number" data-line-number="132"></td>
+        <td id="LC132" class="blob-code blob-code-inner js-file-line">Here, the expression on the left describes the benefits of the delay while the expression on the right describes the cost of the delay.
+In the simplest case, there is no change in investment costs (<span class="pl-s"><span class="pl-pds">$</span>C^<span class="pl-c1">0</span> = C^<span class="pl-c1">1</span> = C<span class="pl-pds">$</span></span>) and the delay has no lasting effects beyond period 1 (<span class="pl-s"><span class="pl-pds">$</span>d_t^<span class="pl-c1">1</span> = d_t^<span class="pl-c1">0</span><span class="pl-pds">$</span></span> for <span class="pl-s"><span class="pl-pds">$</span>t &gt; <span class="pl-c1">1</span><span class="pl-pds">$</span></span>).
+In this case, the comparison is between the expected return <span class="pl-s"><span class="pl-pds">$</span>r<span class="pl-pds">$</span></span> earned on the captial while implementation is delayed and one addtional time period of unmitigated damage,</td>
+      </tr>
+      <tr>
+        <td id="L133" class="blob-num js-line-number" data-line-number="133"></td>
+        <td id="LC133" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{linenomath*}</td>
+      </tr>
+      <tr>
+        <td id="L134" class="blob-num js-line-number" data-line-number="134"></td>
+        <td id="LC134" class="blob-code blob-code-inner js-file-line">  <span class="pl-cce">\[</span></td>
+      </tr>
+      <tr>
+        <td id="L135" class="blob-num js-line-number" data-line-number="135"></td>
+        <td id="LC135" class="blob-code blob-code-inner js-file-line">  r C &gt; d_1^1 - d_1^0.</td>
+      </tr>
+      <tr>
+        <td id="L136" class="blob-num js-line-number" data-line-number="136"></td>
+        <td id="LC136" class="blob-code blob-code-inner js-file-line">  <span class="pl-cce">\]</span></td>
+      </tr>
+      <tr>
+        <td id="L137" class="blob-num js-line-number" data-line-number="137"></td>
+        <td id="LC137" class="blob-code blob-code-inner js-file-line">  <span class="pl-c1">\end</span>{linenomath*}</td>
+      </tr>
+      <tr>
+        <td id="L138" class="blob-num js-line-number" data-line-number="138"></td>
+        <td id="LC138" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L139" class="blob-num js-line-number" data-line-number="139"></td>
+        <td id="LC139" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Limitations of the decision framework}</td>
+      </tr>
+      <tr>
+        <td id="L140" class="blob-num js-line-number" data-line-number="140"></td>
+        <td id="LC140" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L141" class="blob-num js-line-number" data-line-number="141"></td>
+        <td id="LC141" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\section</span>{Case studies}</td>
+      </tr>
+      <tr>
+        <td id="L142" class="blob-num js-line-number" data-line-number="142"></td>
+        <td id="LC142" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L143" class="blob-num js-line-number" data-line-number="143"></td>
+        <td id="LC143" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Data {<span class="pl-c1">\color</span>{blue} (PG)}}</td>
+      </tr>
+      <tr>
+        <td id="L144" class="blob-num js-line-number" data-line-number="144"></td>
+        <td id="LC144" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L145" class="blob-num js-line-number" data-line-number="145"></td>
+        <td id="LC145" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Timing of adaptation measures {<span class="pl-c1">\color</span>{blue} (KdB, TT)}}</td>
+      </tr>
+      <tr>
+        <td id="L146" class="blob-num js-line-number" data-line-number="146"></td>
+        <td id="LC146" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L147" class="blob-num js-line-number" data-line-number="147"></td>
+        <td id="LC147" class="blob-code blob-code-inner js-file-line">A case study focusing on and comparing different cities in Norway.</td>
+      </tr>
+      <tr>
+        <td id="L148" class="blob-num js-line-number" data-line-number="148"></td>
+        <td id="LC148" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L149" class="blob-num js-line-number" data-line-number="149"></td>
+        <td id="LC149" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\subsection</span>{Selection of adaptation measures(?) {<span class="pl-c1">\color</span>{blue} (MD)}}</td>
+      </tr>
+      <tr>
+        <td id="L150" class="blob-num js-line-number" data-line-number="150"></td>
+        <td id="LC150" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L151" class="blob-num js-line-number" data-line-number="151"></td>
+        <td id="LC151" class="blob-code blob-code-inner js-file-line">A case study focusing on Denmark. </td>
+      </tr>
+      <tr>
+        <td id="L152" class="blob-num js-line-number" data-line-number="152"></td>
+        <td id="LC152" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L153" class="blob-num js-line-number" data-line-number="153"></td>
+        <td id="LC153" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\section</span>{Conclusions}</td>
+      </tr>
+      <tr>
+        <td id="L154" class="blob-num js-line-number" data-line-number="154"></td>
+        <td id="LC154" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L155" class="blob-num js-line-number" data-line-number="155"></td>
+        <td id="LC155" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  ACKNOWLEDGMENTS</span></td>
+      </tr>
+      <tr>
+        <td id="L156" class="blob-num js-line-number" data-line-number="156"></td>
+        <td id="LC156" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\begin</span>{acknowledgments}</td>
+      </tr>
+      <tr>
+        <td id="L157" class="blob-num js-line-number" data-line-number="157"></td>
+        <td id="LC157" class="blob-code blob-code-inner js-file-line">This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections&#39;&#39; (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets&#39;&#39; (ClimateXL).
+The source code for the analysis is implemented in the statistical programming language {<span class="pl-c1">\tt</span> R} (<span class="pl-c1">\url</span>{http://www.R-project.org}) and is available on GitHub at <span class="pl-c1">\url</span>{http://github.com/eSACP/...}.</td>
+      </tr>
+      <tr>
+        <td id="L158" class="blob-num js-line-number" data-line-number="158"></td>
+        <td id="LC158" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{acknowledgments}</td>
+      </tr>
+      <tr>
+        <td id="L159" class="blob-num js-line-number" data-line-number="159"></td>
+        <td id="LC159" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L160" class="blob-num js-line-number" data-line-number="160"></td>
+        <td id="LC160" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%%  REFERENCE LIST AND TEXT CITATIONS</span></td>
+      </tr>
+      <tr>
+        <td id="L161" class="blob-num js-line-number" data-line-number="161"></td>
+        <td id="LC161" class="blob-code blob-code-inner js-file-line"><span class="pl-c">% 5\bibliographystyle{../BibTeX/agufull08}</span></td>
+      </tr>
+      <tr>
+        <td id="L162" class="blob-num js-line-number" data-line-number="162"></td>
+        <td id="LC162" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\bibliographystyle</span>{agufull08}</td>
+      </tr>
+      <tr>
+        <td id="L163" class="blob-num js-line-number" data-line-number="163"></td>
+        <td id="LC163" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\bibliography</span>{ref.bib}</td>
+      </tr>
+      <tr>
+        <td id="L164" class="blob-num js-line-number" data-line-number="164"></td>
+        <td id="LC164" class="blob-code blob-code-inner js-file-line"><span class="pl-c">% Please use ONLY \citet and \citep for reference citations.</span></td>
+      </tr>
+      <tr>
+        <td id="L165" class="blob-num js-line-number" data-line-number="165"></td>
+        <td id="LC165" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L166" class="blob-num js-line-number" data-line-number="166"></td>
+        <td id="LC166" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%\begin{thebibliography}{37}</span></td>
+      </tr>
+      <tr>
+        <td id="L167" class="blob-num js-line-number" data-line-number="167"></td>
+        <td id="LC167" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%%   Before submitting: copy all the contents into the .bbl LaTeX file here</span></td>
+      </tr>
+      <tr>
+        <td id="L168" class="blob-num js-line-number" data-line-number="168"></td>
+        <td id="LC168" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%%   and run latex again</span></td>
+      </tr>
+      <tr>
+        <td id="L169" class="blob-num js-line-number" data-line-number="169"></td>
+        <td id="LC169" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%\providecommand{\natexlab}[1]{#1}</span></td>
+      </tr>
+      <tr>
+        <td id="L170" class="blob-num js-line-number" data-line-number="170"></td>
+        <td id="LC170" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%\expandafter\ifx\csname urlstyle\endcsname\relax</span></td>
+      </tr>
+      <tr>
+        <td id="L171" class="blob-num js-line-number" data-line-number="171"></td>
+        <td id="LC171" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else</span></td>
+      </tr>
+      <tr>
+        <td id="L172" class="blob-num js-line-number" data-line-number="172"></td>
+        <td id="LC172" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup</span></td>
+      </tr>
+      <tr>
+        <td id="L173" class="blob-num js-line-number" data-line-number="173"></td>
+        <td id="LC173" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  \urlstyle{rm}\Url}\fi</span></td>
+      </tr>
+      <tr>
+        <td id="L174" class="blob-num js-line-number" data-line-number="174"></td>
+        <td id="LC174" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L175" class="blob-num js-line-number" data-line-number="175"></td>
+        <td id="LC175" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L176" class="blob-num js-line-number" data-line-number="176"></td>
+        <td id="LC176" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%\end{thebibliography}</span></td>
+      </tr>
+      <tr>
+        <td id="L177" class="blob-num js-line-number" data-line-number="177"></td>
+        <td id="LC177" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L178" class="blob-num js-line-number" data-line-number="178"></td>
+        <td id="LC178" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L179" class="blob-num js-line-number" data-line-number="179"></td>
+        <td id="LC179" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%  END ARTICLE</span></td>
+      </tr>
+      <tr>
+        <td id="L180" class="blob-num js-line-number" data-line-number="180"></td>
+        <td id="LC180" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% ------------------------------------------------------------------------ %%</span></td>
+      </tr>
+      <tr>
+        <td id="L181" class="blob-num js-line-number" data-line-number="181"></td>
+        <td id="LC181" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{article}</td>
+      </tr>
+      <tr>
+        <td id="L182" class="blob-num js-line-number" data-line-number="182"></td>
+        <td id="LC182" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L183" class="blob-num js-line-number" data-line-number="183"></td>
+        <td id="LC183" class="blob-code blob-code-inner js-file-line"><span class="pl-c">%% Enter Figures and Tables here:</span></td>
+      </tr>
+      <tr>
+        <td id="L184" class="blob-num js-line-number" data-line-number="184"></td>
+        <td id="LC184" class="blob-code blob-code-inner js-file-line">
+</td>
+      </tr>
+      <tr>
+        <td id="L185" class="blob-num js-line-number" data-line-number="185"></td>
+        <td id="LC185" class="blob-code blob-code-inner js-file-line"><span class="pl-c1">\end</span>{document}</td>
+      </tr>
+</table>
 
-\section{Decision tools {\color{blue} (KdB, MD, TT)}}
+  </div>
 
-\subsection{Timing of adaptation measures}
+</div>
 
-We consider adaptation decision making related to the timing of proactive adaptation measures.
-That is, the goal is to adapt to sea level rise before major damages occur.
-In a cost-benefit framework, an investment should be delayed as long as the benefits of delay (avoided investment costs) are greater than the associated costs (higher climate change damages) \citep{Fankhauser&1999}.
+<button type="button" data-facebox="#jump-to-line" data-facebox-class="linejump" data-hotkey="l" class="d-none">Jump to Line</button>
+<div id="jump-to-line" style="display:none">
+  <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="" class="js-jump-to-line-form" method="get"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /></div>
+    <input class="form-control linejump-input js-jump-to-line-field" type="text" placeholder="Jump to line&hellip;" aria-label="Jump to line" autofocus>
+    <button type="submit" class="btn">Go</button>
+</form></div>
 
-\cite{Fankhauser&1999} describe a deterministic framework where an adaptation investment of $C^0$ now (at time $n=0$) leads to unmitigated damage of $d_0^0$ in period $0$, and a stream of partially mitigated damages $d_t^0$ in periods $t=1,2,\ldots$.
-If $r$ is the discount rate, the net present value damage, $D^0$, associated with this investment is
-\begin{linenomath*}
-\begin{equation}\label{eq:deterministic damage}
-D^0 = C^0 + d_0^0 + \frac{d_1^0}{1+r} + \frac{d_2^0}{(1+r)^2} + \cdots  
-\end{equation}
-\end{linenomath*}
-In comparison, postponing the adaptation investment to time period $n=1$ would lead to unmitigated damages in periods $0$ and $1$, and partially mitigated damages, $d_t^1$, thereafter.
-The delay would be preferable if
-\begin{linenomath*}
-\[
-C^0 - \frac{C^1}{(1+r)} > (d_0^1 - d_0^0) + \frac{d_1^1 - d_1^0}{1+r} + \frac{d_2^1 - d_2^0}{(1+r)^2} + \cdots
-\]
-\end{linenomath*}
-Here, the expression on the left describes the benefits of the delay while the expression on the right describes the cost of the delay.
-In the simplest case, there is no change in investment costs ($C^0 = C^1 = C$) and the delay has no lasting effects beyond period 1 ($d_t^1 = d_t^0$ for $t > 1$).
-In this case, the comparison is between the expected return $r$ earned on the captial while implementation is delayed and one addtional time period of unmitigated damage,
-\begin{linenomath*}
-  \[
-  r C > d_1^1 - d_1^0.
-  \]
-  \end{linenomath*}
+  </div>
+  <div class="modal-backdrop js-touch-events"></div>
+</div>
 
-\subsection{Limitations of the decision framework}
+    </div>
+  </div>
 
-\section{Case studies}
+    </div>
 
-\subsection{Data {\color{blue} (PG)}}
+        <div class="container site-footer-container">
+  <div class="site-footer" role="contentinfo">
+    <ul class="site-footer-links float-right">
+        <li><a href="https://github.com/contact" data-ga-click="Footer, go to contact, text:contact">Contact GitHub</a></li>
+      <li><a href="https://developer.github.com" data-ga-click="Footer, go to api, text:api">API</a></li>
+      <li><a href="https://training.github.com" data-ga-click="Footer, go to training, text:training">Training</a></li>
+      <li><a href="https://shop.github.com" data-ga-click="Footer, go to shop, text:shop">Shop</a></li>
+        <li><a href="https://github.com/blog" data-ga-click="Footer, go to blog, text:blog">Blog</a></li>
+        <li><a href="https://github.com/about" data-ga-click="Footer, go to about, text:about">About</a></li>
 
-\subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
+    </ul>
 
-A case study focusing on and comparing different cities in Norway.
+    <a href="https://github.com" aria-label="Homepage" class="site-footer-mark" title="GitHub">
+      <svg aria-hidden="true" class="octicon octicon-mark-github" height="24" version="1.1" viewBox="0 0 16 16" width="24"><path d="M8 0C3.58 0 0 3.58 0 8c0 3.54 2.29 6.53 5.47 7.59.4.07.55-.17.55-.38 0-.19-.01-.82-.01-1.49-2.01.37-2.53-.49-2.69-.94-.09-.23-.48-.94-.82-1.13-.28-.15-.68-.52-.01-.53.63-.01 1.08.58 1.23.82.72 1.21 1.87.87 2.33.66.07-.52.28-.87.51-1.07-1.78-.2-3.64-.89-3.64-3.95 0-.87.31-1.59.82-2.15-.08-.2-.36-1.02.08-2.12 0 0 .67-.21 2.2.82.64-.18 1.32-.27 2-.27.68 0 1.36.09 2 .27 1.53-1.04 2.2-.82 2.2-.82.44 1.1.16 1.92.08 2.12.51.56.82 1.27.82 2.15 0 3.07-1.87 3.75-3.65 3.95.29.25.54.73.54 1.48 0 1.07-.01 1.93-.01 2.2 0 .21.15.46.55.38A8.013 8.013 0 0 0 16 8c0-4.42-3.58-8-8-8z"></path></svg>
+</a>
+    <ul class="site-footer-links">
+      <li>&copy; 2016 <span title="0.17424s from github-fe164-cp1-prd.iad.github.net">GitHub</span>, Inc.</li>
+        <li><a href="https://github.com/site/terms" data-ga-click="Footer, go to terms, text:terms">Terms</a></li>
+        <li><a href="https://github.com/site/privacy" data-ga-click="Footer, go to privacy, text:privacy">Privacy</a></li>
+        <li><a href="https://github.com/security" data-ga-click="Footer, go to security, text:security">Security</a></li>
+        <li><a href="https://status.github.com/" data-ga-click="Footer, go to status, text:status">Status</a></li>
+        <li><a href="https://help.github.com" data-ga-click="Footer, go to help, text:help">Help</a></li>
+    </ul>
+  </div>
+</div>
 
-\subsection{Selection of adaptation measures(?) {\color{blue} (MD)}}
+    <div id="ajax-error-message" class="ajax-error-message flash flash-error">
+      <svg aria-hidden="true" class="octicon octicon-alert" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M8.865 1.52c-.18-.31-.51-.5-.87-.5s-.69.19-.87.5L.275 13.5c-.18.31-.18.69 0 1 .19.31.52.5.87.5h13.7c.36 0 .69-.19.86-.5.17-.31.18-.69.01-1L8.865 1.52zM8.995 13h-2v-2h2v2zm0-3h-2V6h2v4z"></path></svg>
+      <button type="button" class="flash-close js-flash-close js-ajax-error-dismiss" aria-label="Dismiss error">
+        <svg aria-hidden="true" class="octicon octicon-x" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M7.48 8l3.75 3.75-1.48 1.48L6 9.48l-3.75 3.75-1.48-1.48L4.52 8 .77 4.25l1.48-1.48L6 6.52l3.75-3.75 1.48 1.48z"></path></svg>
+      </button>
+      You can't perform that action at this time.
+    </div>
 
-A case study focusing on Denmark.
+      <script crossorigin="anonymous" integrity="sha256-u8ByW3Q46gf3lKMTt8lghhPBANPkJ7NpsnoMo37E37w=" src="https://assets-cdn.github.com/assets/frameworks-bbc0725b7438ea07f794a313b7c9608613c100d3e427b369b27a0ca37ec4dfbc.js"></script>
+      <script async="async" crossorigin="anonymous" integrity="sha256-PXu924XPoLit0LXTJLIDib8M+UnB8jQ2Yrvb0FeYMyA=" src="https://assets-cdn.github.com/assets/github-3d7bbddb85cfa0b8add0b5d324b20389bf0cf949c1f2343662bbdbd057983320.js"></script>
 
-\section{Conclusions}
+    <div class="js-stale-session-flash stale-session-flash flash flash-warn flash-banner d-none">
+      <svg aria-hidden="true" class="octicon octicon-alert" height="16" version="1.1" viewBox="0 0 16 16" width="16"><path d="M8.865 1.52c-.18-.31-.51-.5-.87-.5s-.69.19-.87.5L.275 13.5c-.18.31-.18.69 0 1 .19.31.52.5.87.5h13.7c.36 0 .69-.19.86-.5.17-.31.18-.69.01-1L8.865 1.52zM8.995 13h-2v-2h2v2zm0-3h-2V6h2v4z"></path></svg>
+      <span class="signed-in-tab-flash">You signed in with another tab or window. <a href="">Reload</a> to refresh your session.</span>
+      <span class="signed-out-tab-flash">You signed out in another tab or window. <a href="">Reload</a> to refresh your session.</span>
+    </div>
+    <div class="facebox" id="facebox" style="display:none;">
+  <div class="facebox-popup">
+    <div class="facebox-content" role="dialog" aria-labelledby="facebox-header" aria-describedby="facebox-description">
+    </div>
+    <button type="button" class="facebox-close js-facebox-close" aria-label="Close modal">
+      <svg aria-hidden="true" class="octicon octicon-x" height="16" version="1.1" viewBox="0 0 12 16" width="12"><path d="M7.48 8l3.75 3.75-1.48 1.48L6 9.48l-3.75 3.75-1.48-1.48L4.52 8 .77 4.25l1.48-1.48L6 6.52l3.75-3.75 1.48 1.48z"></path></svg>
+    </button>
+  </div>
+</div>
 
-%  ACKNOWLEDGMENTS
-\begin{acknowledgments}
-This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL).
-The source code for the analysis is implemented in the statistical programming language {\tt R} (\url{http://www.R-project.org}) and is available on GitHub at \url{http://github.com/eSACP/...}.
-\end{acknowledgments}
-
-%%  REFERENCE LIST AND TEXT CITATIONS
-% 5\bibliographystyle{../BibTeX/agufull08}
-\bibliographystyle{agufull08}
-\bibliography{ref.bib}
-% Please use ONLY \citet and \citep for reference citations.
-
-%\begin{thebibliography}{37}
-%%   Before submitting: copy all the contents into the .bbl LaTeX file here
-%%   and run latex again
-%\providecommand{\natexlab}[1]{#1}
-%\expandafter\ifx\csname urlstyle\endcsname\relax
-%  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
-%  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
-%  \urlstyle{rm}\Url}\fi
-
-%\end{thebibliography}
-
-%% ------------------------------------------------------------------------ %%
-%  END ARTICLE
-%% ------------------------------------------------------------------------ %%
-\end{article}
-
-%% Enter Figures and Tables here:
-
-\end{document}
+  </body>
+</html>

### 2016-10-11 submission/paper.tex
solving conflict

@@
     <link rel="assets" href="https://assets-cdn.github.com/">
-    <link rel="web-socket" href="wss://live.github.com/_sockets/MTQ5NDUyMTM6MDUyNDYzZDIwMmVhNjNkYzMwNWI5ZWU1M2Q3NmQ3MWQ6NmY5MTBmMGJhNDA4NDZkNzQ4MGRkNDBjODdkYjBmYTI4ZjBkODAzMWJiY2IzODI3YWE5MDI3MGRjZTc0MjBkMQ==--89d25bdd8d4ae8902ca35bc19bf81c83badb6ba1">
+    <link rel="web-socket" href="wss://live.github.com/_sockets/MTQ5NDUyMTM6MDUyNDYzZDIwMmVhNjNkYzMwNWI5ZWU1M2Q3NmQ3MWQ6OTEyMjExMTA3Zjc3ZTY5MDBmZTIyZGE3OGEzYTgzNzI0MDYwNjlmNmY5MGI1ZGZlZWUyMWI0MzQ4YWY4YjMwZA==--00f18813aa925cae31ac9371dfeeafec2e350fd1">
     <meta name="pjax-timeout" content="1000">
     <link rel="sudo-modal" href="/sessions/sudo_modal">
-    <meta name="request-id" content="9C740835:0C52:28EFDD9:57FCCA5A" data-pjax-transient>
+    <meta name="request-id" content="9C740835:0C55:4926F4F:57FCCBF1" data-pjax-transient>
 
@@
 
-<meta content="collector.githubapp.com" name="octolytics-host" /><meta content="github" name="octolytics-app-id" /><meta content="9C740835:0C52:28EFDD9:57FCCA5A" name="octolytics-dimension-request_id" /><meta content="14945213" name="octolytics-actor-id" /><meta content="thordislinda" name="octolytics-actor-login" /><meta content="25d4553f8e8b0f7d5f53a85da67d6ac37df8675bf0532ac0ed5d5f1cf648984f" name="octolytics-actor-hash" />
+<meta content="collector.githubapp.com" name="octolytics-host" /><meta content="github" name="octolytics-app-id" /><meta content="9C740835:0C55:4926F4F:57FCCBF1" name="octolytics-dimension-request_id" /><meta content="14945213" name="octolytics-actor-id" /><meta content="thordislinda" name="octolytics-actor-login" /><meta content="25d4553f8e8b0f7d5f53a85da67d6ac37df8675bf0532ac0ed5d5f1cf648984f" name="octolytics-actor-hash" />
 <meta content="/&lt;user-name&gt;/&lt;repo-name&gt;/blob/show" data-pjax-transient="true" name="analytics-location" />
@@
         <meta name="expected-hostname" content="github.com">
-      <meta name="js-proxy-site-detection-payload" content="Y2ExZjFiMDljMmQxNzBiYjUzNWQzYjgxYmI2OGM1NzY0Nzc1NGRmNWJhNGVmNDNjOTJkYjc0NmJlOTA5ZGU1MXx7InJlbW90ZV9hZGRyZXNzIjoiMTU2LjExNi44LjUzIiwicmVxdWVzdF9pZCI6IjlDNzQwODM1OjBDNTI6MjhFRkREOTo1N0ZDQ0E1QSIsInRpbWVzdGFtcCI6MTQ3NjE4NDY3MX0=">
+      <meta name="js-proxy-site-detection-payload" content="NmYzZGVlY2NiN2QxMTcyMGEyNWQ2YzAyZGFiMjMzODJlMmUzYTAxZmY4ZjBjYmQ0MWY2MjZlZGYzODA2Yzg5NHx7InJlbW90ZV9hZGRyZXNzIjoiMTU2LjExNi44LjUzIiwicmVxdWVzdF9pZCI6IjlDNzQwODM1OjBDNTU6NDkyNkY0Rjo1N0ZDQ0JGMSIsInRpbWVzdGFtcCI6MTQ3NjE4NTA3NH0=">
 
@@
   <meta content="14945390" name="octolytics-dimension-user_id" /><meta content="eSACP" name="octolytics-dimension-user_login" /><meta content="69018538" name="octolytics-dimension-repository_id" /><meta content="eSACP/SeaLevelDecisions" name="octolytics-dimension-repository_nwo" /><meta content="false" name="octolytics-dimension-repository_public" /><meta content="false" name="octolytics-dimension-repository_is_fork" /><meta content="69018538" name="octolytics-dimension-repository_network_root_id" /><meta content="eSACP/SeaLevelDecisions" name="octolytics-dimension-repository_network_root_nwo" />
-  <link href="https://github.com/eSACP/SeaLevelDecisions/commits/master.atom?token=AOQLvX3BJqaD2T7TXulS-nsUHBg8-hiIks62CLlvwA%3D%3D" rel="alternate" title="Recent Commits to SeaLevelDecisions:master" type="application/atom+xml">
+  <link href="https://github.com/eSACP/SeaLevelDecisions/commits/master.atom?token=AOQLvTof4iBfdh_bxi-mAYExYzZraD1Yks62CLsCwA%3D%3D" rel="alternate" title="Recent Commits to SeaLevelDecisions:master" type="application/atom+xml">
 
@@
 
-        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/logout" class="logout-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="3yXD4WZhwnWJytG55fL3EwgkMSCMjYlEpuhXXi1XrsllPkR2n1RGivgoBFTGcsqjYxx3Jj0qwtsKsr9Lfl+eRQ==" /></div>
+        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/logout" class="logout-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="g3xzxmHvQyOZcdduEA2nIbMpdP5YMK990mQfSMu9pM1TjktF+qOiz6MsCcmnYQ+Jt1cPiz7zttHuxK0uLsGXCQ==" /></div>
           <button type="submit" class="dropdown-item dropdown-signout" data-ga-click="Header, sign out, icon:logout">
@@
   <li>
-        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/notifications/subscribe" class="js-social-container" data-autosubmit="true" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="ow3eH8530MPPaY+NBuQQ2cKNOZgKM52lzUCopLpBByIcEQvPDDXmUb+PxjlwNE8m1o6yftRt4YK14tIqIWRMUQ==" /></div>      <input class="form-control" id="repository_id" name="repository_id" type="hidden" value="69018538" />
+        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/notifications/subscribe" class="js-social-container" data-autosubmit="true" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="hNtXU+QFowQO8xu8ntiXO7r1xdu0xBY4GIIFLQWB7Y6JYx/XEVgXHLnlsLMxDCsUXDu3vdNe3hrVOf/ZCR7rYQ==" /></div>      <input class="form-control" id="repository_id" name="repository_id" type="hidden" value="69018538" />
 
@@
 
-    <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/unstar" class="starred" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="8jsT4FwhRy7KoSwMyDFFxWSrBfQf9q6Lff0fzQ8MdEJMZInhum5hVH/n3iXzf6+xNASq1AZqI4hBSUEtSLrMxA==" /></div>
+    <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/unstar" class="starred" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="GtlnpB914Jk1PD4m5DjZFtsdzJjGqKhfPlnTw/rmDmISrKiyNaWrqDj4fIZ3t1Dfm8u/tXyRQe1rb3gYakhOow==" /></div>
       <button
@@
 </form>
-    <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/star" class="unstarred" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="ks2hkl+V8bZA/NAEAqFMMvvxD9L863yDs4o1AGfGlBRqFH/bqDT71HtOaVIWzDHnNEB4JjA7PZjbQqVS5B4k9w==" /></div>
+    <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/star" class="unstarred" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" data-remote="true" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="JTg/qUMW8crgT/1Wxronr7Df4zFgXi4/AcBQdn4HM8g9fuZv6zWaC+roC7DS+jxfKAkNyyzVbwM1Tmr974bDCg==" /></div>
       <button
@@
 
-          <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/branches" class="js-create-branch select-menu-item select-menu-new-item-form js-navigation-item js-new-item-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="DcMyzVBLpaCbk6kx2+mnhjRNC9PhE1BLNhUynTbqPElsNsRr3UzZrSuzCAJO6wgj6TKTs9w6R/B+N0T7ENYGjQ==" /></div>
+          <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/branches" class="js-create-branch select-menu-item select-menu-new-item-form js-navigation-item js-new-item-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="tFLJXyMYF5qjVGsC1skVetROtxEq1jGt+I7qMaqTbXk4eX+2eWczuOKNAgbqOkMj6YkPiqcTebby8Fc6dw9Mug==" /></div>
           <svg aria-hidden="true" class="octicon octicon-git-branch select-menu-item-icon" height="16" version="1.1" viewBox="0 0 10 16" width="10"><path d="M10 5c0-1.11-.89-2-2-2a1.993 1.993 0 0 0-1 3.72v.3c-.02.52-.23.98-.63 1.38-.4.4-.86.61-1.38.63-.83.02-1.48.16-2 .45V4.72a1.993 1.993 0 0 0-1-3.72C.88 1 0 1.89 0 3a2 2 0 0 0 1 1.72v6.56c-.59.35-1 .99-1 1.72 0 1.11.89 2 2 2 1.11 0 2-.89 2-2 0-.53-.2-1-.53-1.36.09-.06.48-.41.59-.47.25-.11.56-.17.94-.17 1.05-.05 1.95-.45 2.75-1.25S8.95 7.77 9 6.73h-.02C9.59 6.37 10 5.73 10 5zM2 1.8c.66 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2C1.35 4.2.8 3.65.8 3c0-.65.55-1.2 1.2-1.2zm0 12.41c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2zm6-8c-.66 0-1.2-.55-1.2-1.2 0-.65.55-1.2 1.2-1.2.65 0 1.2.55 1.2 1.2 0 .65-.55 1.2-1.2 1.2z"></path></svg>
@@
 
-        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/edit/master/submission/paper.tex" class="inline-form js-update-url-with-hash" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="JHAATQHJKo7GYe0P+sOvc4ri8m7QuFAOQBjGHnIR/IVCje5LFQxPY/7jS0hKVjDbuBVmyoHhmKRJQS4L7b7bMA==" /></div>
+        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/edit/master/submission/paper.tex" class="inline-form js-update-url-with-hash" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="sjcXEW9oK9etoucE/sg8MOSNquWSXWvPhhM/jB58oJiS7Gk4B62bRn8/K6+aSzGHLvp8q3NHn/DV+c8Z9zqmBw==" /></div>
           <button class="btn-octicon tooltipped tooltipped-nw" type="submit"
@@
           </button>
-</form>        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/delete/master/submission/paper.tex" class="inline-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="GKa51xlrlUKP7iYS4nEBuU41UNrSrGU/IsCaU7p1GOZljm23SsoIOnuLz6TcbAiZmWeku1jesTDXHdkYwL0JMg==" /></div>
+</form>        <!-- '"` --><!-- </textarea></xmp> --></option></form><form accept-charset="UTF-8" action="/eSACP/SeaLevelDecisions/delete/master/submission/paper.tex" class="inline-form" data-form-nonce="01c9606cf95d90aee2e58783cb1f27d1662fa6a0" method="post"><div style="margin:0;padding:0;display:inline"><input name="utf8" type="hidden" value="&#x2713;" /><input name="authenticity_token" type="hidden" value="UohHCkSZS8wL4Q+ZfdepYV6n/fcmZBwa9J2ho7rJyhhMkr0x3hgPcbMUhMl9KEoLHk0YZZgNoF3pja5PRrYVVg==" /></div>
           <button class="btn-octicon btn-octicon-danger tooltipped tooltipped-nw" type="submit"
@@
     <ul class="site-footer-links">
-      <li>&copy; 2016 <span title="0.17424s from github-fe164-cp1-prd.iad.github.net">GitHub</span>, Inc.</li>
+      <li>&copy; 2016 <span title="0.12379s from github-fe162-cp1-prd.iad.github.net">GitHub</span>, Inc.</li>
         <li><a href="https://github.com/site/terms" data-ga-click="Footer, go to terms, text:terms">Terms</a></li>

### 2016-10-13 submission/paper.tex
merging version of tex files
still merging tex files

--- 2016-10-10 submission/paper.tex
@@
 % Author names in capital letters:
-\authorrunninghead{1st AUTHOR ET AL.}
+\authorrunninghead{THORARINSDOTTIR ET AL.}
 
@@
 
-\altaffiltext{2}{Danish Technological University}
+\altaffiltext{2}{Technical University of Denmark}
 
@@
 Each climate model temperature projection yields a different local sea level projection.
+%Equation
 
@@
 This allows us, for example, to get a confidence band for the years when a given sea level rise is obtained.
+%equation
 
@@
 Our current climate models are not able to resolve the ice processes sufficiently to include such so called tipping points into the projections.
-Also, the IPCC scenarios do not include changes in water usage \citep{change}.
+Also, the IPCC scenarios \citep{change} do not include changes in water usage.
 
@@
 \subsection{Data {\color{blue} (PG)}}
+We use local tide gauge data from the Permanent Service for Mean Sea Level, UK, which is the worldwide repository for national sea level data.
+Climate projections are from the fifth climate model intercomparison project, CMIP5.
 

### 2016-10-13 submission/paper.tex
Minor edits.
Added equations

@@
 
-\title{I don't know, are you sure you want to do this?}
+\title{I don't know, are you sure we want to do this?\
+Sea level adaptation decisions under uncertainty}
 
@@
 We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
-
 %Figure showing historical relationship?
@@
 We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}).
+For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
+
+\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t}\]
+where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level on observed global temperature and $\varsigma_t$ the integrated time series regression error.
+
 %Figure with sea level projections & IPCC projections?
@@
 Each climate model temperature projection yields a different local sea level projection.
-%Equation
+The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
+
+\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t}\]
+where $\gamma$ ie the annual land rise rate, t and $ {\hat b} $ is the regression coefficient relating global to local sea level.
 
@@
 This allows us, for example, to get a confidence band for the years when a given sea level rise is obtained.
-%equation
-
 \begin{figure}
@@
 The main assumption is using historical relationships in statistical projections of the type used in this paper is that there is no major change in how temperature relates to sea level, globally and locally.
-Among the factors that may invalidate this approach are changes in water storage on land (removing water from the oceans), excessive siphoning of groundwater (resulting in land subsidence), changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
+Among the factors that may invalidate this approach are changes in water storage on land (in essence removing water from the oceans), excessive siphoning of groundwater (resulting in land subsidence), changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
 For example, the rate of ice melt on Greenland may suddenly increase substantially due to intense warming of both air and sea water.

### 2016-10-13 code/Esbjerg.R
Code for Esbjerg analysis

@@ -1,0 +1,145 @@
+setwd("~/GitHub/SeaLevelDecisions/code")
+require(maps)
+require(maptools)
+require(ncdf4)
+require(forecast)
+require(excursions)
+require(Matrix)
+require(chron)
+require(zoo)
+source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
+source('~/GitHub/SeaLevelDecisions/code/ci_bergen.R')
+denmark=read.table("~/GitHub/SeaLevelProjections/denmark_sea_level/denmark.PSMSL.txt")#All Danish sea level stations with current data]
+#from PSMSL.org
+denmark[,1]=as.character(denmark[,1])
+
+denmarkurl=paste("http://www.psmsl.org/data/obtaining/rlr.monthly.data/",denmark[,2],".rlrdata",sep="")
+
+N=length(denmark[,2])
+denmarkdata=list(N,NULL)
+for(i in 1:N){
+
+	tmp=read.table(denmarkurl[i],na.strings="-99999",sep=";")
+	tmp=cbind(floor(tmp[,1]),round(12*(tmp[,1]-floor(tmp[,1]))+.5),tmp[,2:4])
+
+	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	if(tmp[N,2]<12) tmp=tmp[-((N+1-tmp[N,2]):N),]#delete final year if not full
+	tmp.rlr=tmp[,3]
+	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
+	tmp.mp=medpolish(tmp.rlr,na.rm=T)
+	tmp.mppred=(tmp.mp$overall+outer(tmp.mp$row,tmp.mp$col,"+"))
+	for(j in 1:12) tmp.rlr[which(is.na(tmp.rlr[,j])),j]=tmp.mppred[which(is.na(tmp.rlr[,j])),j]
+	tmp.ann=apply(tmp.rlr,1,"mean")/10 #cm
+	tmp.yrs=unique(floor(tmp[,1]))
+	tmp.ann=tmp.ann-mean(tmp.ann[which(is.element(tmp.yrs,1970:1999))],na.rm=T)
+	denmarkdata[i]=list(data.frame(years=unique(floor(tmp[,1])),data=tmp.ann))
+	}
+	
[…]

### 2016-10-14 code/Esbjerg.R
Header correction in Esbjerg figure

@@ -134,7 +134,7 @@ x = seq(from=0,to=50,length.out = 1000)
 year = 51
 
 
-plot(c(0,50), c(0,0.25), xlab="Anomaly (cm)",ylab="Density",main="Bergen 2050 RCP 8.5",type="n")
+plot(c(0,50), c(0,0.25), xlab="Anomaly (cm)",ylab="Density",main="Esbjerg 2050 RCP 8.5",type="n")
 v1=apply(matrix(unlist(cis$pred),ncol=101,byrow=T),2,median)[year]
 segments(v1,0,v1,0.25,lwd=2)
 hist(mu[,year],xlim=c(0,50), breaks=(0:51), ylim=c(0,0.50),freq=FALSE,add=T,lwd=2)

### 2016-10-24 submission/paper.tex
Code for Esbjerg analysis
Added text and results on Bergen case study

@@
 \usepackage{color}
+\usepackage{comment}
 
@@
 We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
-%Figure showing historical relationship?
 
@@
 
+We assume that damages related to sea level rise coincide with damages due to storm surges and that the accumulated annual damages are independent between years.
+More specifically, we model the distribution of annual damage, $F_{\textup{d}}$, by the three parameter Burr distribution \citep{Burr1942} with density
+\begin{linenomath*}
+  \begin{equation}\label{eq:Burr}
+  f_{\textup{d}} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
+  \end{equation}
+\end{linenomath*}
+for $x > 0$, where $\alpha$ and $\gamma$ are shape parameters with $\alpha, \gamma > 0$, and $\theta >0$ is a scale parameter.
+The Burr distribution has a heavy upper tail and is commonly used to model damage loss, see e.g.
+\cite{Klugman&2012}.
+
+Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d.
+values from the estimated damage distribution for $t_0 = 2015$, $\hat{F}_{\textup{d}, t_0}$.
+By repeating this process $J$ times, we obtain an empirial damage distribution for each future year $t_i$ given by
+\begin{linenomath*}
+  \[
+  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{d_{t_i}^{(j)} \leq x \}
+  \]
+\end{linenomath*}
+for $i = 1, \ldots, 85$. Similarly, we obtain an empirial distribution of the total damage over the period $2016-2100$ by considering $d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} d_{t_i}^{(j)}$.
+
+Now assume we are given a large sample of projections $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ for annual sea level rise anomalies compared to the 2015 value for the same time period.
+We model the relationship between the change in damage and change in sea level by a monotonic positive function $g(s| \boldsymbol \beta )$ with parameters $\boldsymbol \beta$, where $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$.
+We may further incorporate uncertainty in the shape of the function $g$ through the values of the parameters $\boldsymbol \beta$.
+An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and change in damage due to sea level rise is then given by
+\begin{linenomath*}
+  \begin{equation}\label{eq:FutureDamage}
+    \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}.
+  \end{equation}
+\end{linenomath*}
+
+The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures.
+In addition, we can incorporate an adaptation measure that protects against $K$ cm of increased sea level from year $t_k$ onwards.
+This results in a damage distribution given by
+\begin{linenomath*}
+  \[
+  \hat{F}_{\textup{d}, t_i} (x) =
+    \begin{cases}
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i < t_k \\
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i \geq t_k. \\
+    \end{cases}
+  \]
+\end{linenomath*}
+
+\begin{comment}
 \cite{Fankhauser&1999} describe a deterministic framework where an adaptation investment of $C^0$ now (at time $n=0$) leads to unmitigated damage of $d_0^0$ in period $0$, and a stream of partially mitigated damages $d_t^0$ in periods $t=1,2,\ldots$.
@@
   \end{linenomath*}
+\end{comment}
 
@@
 
+For the case study of sea level rise in Bergen, we use reported annual damage numbers from the Norwegian Natural Perils Pool which are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}.
+
 \subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
 
-A case study focusing on and comparing different cities in Norway.
+We estimate a damage distribution for Bergen in 2015 using annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP).
+The NPP data is aggregated to a county level and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion.
+Also, some of this discussion should probably be moved to the section on data.}.
+In order to improve the parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland.
+Rogaland has a similar geography to Hordaland and a slightly smaller population with its main city, Stavanger, also located on the coast.
+The damage data prior to 2015 is adjusted to the 2015 level using the consumer price index for Norway.
+After the adjustment, we assume stationarity over the period and fit a single Burr distribution to the data, cf.
+\eqref{eq:Burr}.
+The histogram of the observed annual damages and the estimated distribution are given in Figure~\ref{fig:BergenDamageDist}.
+The parameter estimates for the Burr distribution are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{DamageDistribution.png}
+\caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
+\label{fig:BergenDamageDist}
+\end{center}
+\end{figure}
+
+For modelling the relationship between the change in damage and change in sea level we use the results of \cite{Hallegatte&2013} for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseille, Naples, Porto, Rotterdam and Stockholm.
+\cite{Hallegatte&2013} estimate the mean annual flood damage in these cities in 2050 under no sea level change, 20 cm increase and 40 cm increase in sea level.
+We standardize their results for each city and consider the relative change for 20 and 40 cm increase.
+We then perform a linear extrapolation to obtain estimated relative changes in damage for a large interval of changes in sea level rise, see Figure~\ref{fig:ChangeInDamage}.
+Finally, the function $g$ is obtained by sampling from this ensemble of functions with all 15 ensemble members considered equally probable.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{EuropeanIncreaseLossExtrapolationUncertainty.png}
+\caption{ Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
+The median change and the corresponding extrapolation are indicated in red.}
+\label{fig:ChangeInDamage}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{AccumulatedFutureLoss.png}
+\caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and no adaptation measures.} 
+\label{fig:NoAction}
+\end{center}
+\end{figure}
 

### 2016-10-24 submission/paper.tex
Updated data section. Moved figures to data section—perhaps they should be in the opening sectoin of case study?

@@
 Again, we will use historical data to relate global sea level to isostatically corrected local sea level using a time series regression model. 
-Figure \ref{fig:obs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
-The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}.
-\begin{figure}
-\begin{center}
-\includegraphics[width=39pc]{bergenfit.png}
-\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
-The straight line is the time series regression line.}
-\label{fig:obs}
-\end{center}
-\end{figure}
-
 The local sea level projections are then obtained by first relating projected temperature to global sea level, and then relating the global sea level to the local one.
@@
 This allows us, for example, to get a confidence band for the years when a given sea level rise is obtained.
-\begin{figure}
-\begin{center}
-\includegraphics[width=39pc]{bergen_ci.png}
-\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
-The sea level data are shown in blue and end in 2015.
-The thin red lines are the projections without uncertainty based on each of the climate models.
-The dashed purple lines connect pointwise confidence intervals for each year.
-}
-\label{fig:ci}
-\end{center}
-\end{figure}
 
@@
 \subsection{Data {\color{blue} (PG)}}
+Historical global mean temperature is obtained from \citet{giss}.
+Climate projections of global mean temperature are from the fifth climate model intercomparison project, CMIP5 \citep{cmip5}.
+The global mean sea level series is obtained from \citet{csiro}.
 We use local tide gauge data from the Permanent Service for Mean Sea Level, UK, which is the worldwide repository for national sea level data.
-Climate projections are from the fifth climate model intercomparison project, CMIP5.
+Glacial isostatic adjustment for Bergen is obtained from \citet{Simpson2014}.
+
+Figure \ref{fig:obs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
+The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}.
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{bergenfit.png}
+\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
+The straight line is the time series regression line.}
+\label{fig:obs}
+\end{center}
+\end{figure}
+
+Figure \ref{fig:ci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999.
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{bergen_ci.png}
+\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
+The sea level data are shown in blue and end in 2015.
+The thin red lines are the projections without uncertainty based on each of the climate models.
+The dashed purple lines connect pointwise confidence intervals for each year.
+}
+\label{fig:ci}
+\end{center}
+\end{figure}
 

### 2016-10-29 submission/paper.tex
Updated details about data cleaning for Bergen

@@
 \subsection{Data {\color{blue} (PG)}}
-Historical global mean temperature is obtained from \citet{giss}.
+The historical global mean temperature series is obtained from \citet{giss}.
 Climate projections of global mean temperature are from the fifth climate model intercomparison project, CMIP5 \citep{cmip5}.
@@
 
+The Bergen monthly series is missing data for 62 months, including all of the years 1942--43.
+To deal with occasional short stretches of missing data (most one or two months) we use median polish replacement \citep{medpol}.
+For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other stations corrected by the average difference.
+
+For the case study of sea level rise in Bergen, we use reported annual damage numbers from the Norwegian Natural Perils Pool which are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}.
+
+\subsection{Sea level rise in Bergen}
+
 Figure \ref{fig:obs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
-The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}.
+The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17).
+The regression slope is 1.30 (0.12).
 \begin{figure}
@@
 
-Figure \ref{fig:ci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999.
+For the relationship between global annual mean temperature and global annual mean sea level rise we use the results from \citet{Bolin2014a}.
+Figure \ref{fig:ci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999 under scenario RCP 8.5, which is the scenario Norwegian authorities recommend for planning purposes.
 \begin{figure}
@@
 \end{figure}
-
-For the case study of sea level rise in Bergen, we use reported annual damage numbers from the Norwegian Natural Perils Pool which are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}.
 

### 2016-11-01 submission/agujournalsample.tex
Updated AGU style file

%% 2/18/2016
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalSample.tex: this sample file is for articles formatted with LaTeX
%
% This sample file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements.
%
% PLEASE DO NOT USE YOUR OWN MACROS
% DO NOT USE \newcommand, \renewcommand, or \def.
%
% FOR FIGURES, DO NOT USE \psfrag or \subfigure.
% DO NOT USE \psfrag or \subfigure commands.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% All questions should be e-mailed to latex@agu.org.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Step 1: Set the \documentclass
%
% There are two options for article format:
%
% 1) PLEASE USE THE DRAFT OPTION TO SUBMIT YOUR PAPERS.
% The draft option produces double spaced output.
% 
% 2) numberline will give you line numbers.

% Tip:
%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

%% To submit your paper:
\documentclass[linenumbers,draft]{agujournal}

% Now, type in the journal name: \journalname{<Journal Name>}
% ie,
\journalname{JGR-Atmospheres}

%% Choose from this list of Journals:
%
% JGR-Atmospheres
% JGR-Biogeosciences
% JGR-Earth Surface
% JGR-Oceans
% JGR-Planets
% JGR-Solid Earth
% JGR-Space Physics
% Global Biochemical Cycles
% Geophysical Research Letters
% Paleoceanography
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resource Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science

%% ------------------------------------------------------------------------ %%
%
%  ENTER Title Page commands:
%
%% ------------------------------------------------------------------------ %%

% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general terms then specific results are optimized in
% searches)

% Example: \title{This is a test title}

% (List authors by first name or initial followed by last name and
% separated by commas. Use \affil{} to number affiliations, and
% \thanks{} for author notes.  
% Additional author notes should be indicated with \thanks{} (for
% example, for current addresses).

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% \affiliation{1}{First Affiliation}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

%% Keypoints, final entry on title page.
% Example: 
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 100 characters or less with no special
% characters or punctuation 
% \end{keypoints}

%% \begin{abstract} begins second page

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Track Changes:
% To add words, \added{<word added>}
% To delete words, \deleted{<word deleted>}
% To replace words, \replace{<word to be replaced>}{<replacement word>}
% To explain why change was made: \explain{<explanation>}

% At the end of the document, use \listofchanges, which will list the
% changes and the page and line number where the change was made.

% When final version, \listofchanges will not produce anything,
% \added{} word will be printed, \deleted{} will take away the word,
% \replaced{}{} will print only the 2nd argument.
% \explain will not print anything.

% Optional argument:
% You can also add additional information to be printed with the list
% of changes, to indicate the initials of the person changing the text,
% and the time and/or date of the change, or any other comment by using
% the optional [] argument:
% \added[AH, 3:30pm, Feb 18, 2016]{added term}
% will yield 
% [AH, 3:30pm, Feb 18, 2016] added term on page...
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% ------------------------------------------------------------------------ %%
%
%  TITLE
%
%% ------------------------------------------------------------------------ %%

\title{Trends in the predictive performance of raw ensemble weather forecasts}

%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%

 \authors{A. B. Smith\affil{1}\thanks{Current address, McMurdo Station,
 Antartica},
 Eric Brown\affil{1,2}, Rick Williams\affil{3},
 John B. McDougall\affil{4}, and S. Visconti\affil{5}\thanks{Also
 funded by Monsanto.}}

\affiliation{1}{Department of Hydrology and Water Resources,
University of Arizona, Tucson, Arizona, USA.}
\affiliation{2}{Department of Geography, Ohio State University,
Columbus, Ohio, USA.}
\affiliation{3}{Department of Space Sciences, University of
Michigan, Ann Arbor, Michigan, USA.}
\affiliation{4}{Division of Hydrologic Sciences, Desert Research
Institute, Reno, Nevada, USA.}
\affiliation{5}{Dipartimento di Idraulica, Trasporti ed
Infrastrutture Civili, Politecnico di Torino, Turin, Italy.}

%% Corresponding Author
%(include name and email addresses of the corresponding author.  More
%than one corresponding author is allowed in this Word file and for
%publication; but only one corresponding author is allowed in our
%editorial system.)

\correspondingauthor{A. B. Smith}{email@address.edu}

%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 100 characters or less with no special characters or punctuation

\begin{keypoints}
\item Evolution of raw ensemble forecast \replaced{skill}{skills}
\item Future benefits from statistical post-processing
\item Global distribution of forecast skill development
\end{keypoints}

%% ------------------------------------------------------------------------ %%
%
%  ABSTRACT
%
%% ------------------------------------------------------------------------ %%

\begin{abstract}
This study applies statistical post-processing to ensemble
forecasts of near-surface temperature, 24-hour 
%% below are examples of track changes:
\deleted[date/time, etc.]{precipitation},
\added{all} totals,
and near-surface wind speed from the global model of the European
Centre for Medium-Range Weather Forecasts (ECMWF). The main objective
is to evaluate the evolution of the difference in skill between the
raw ensemble and the post-processed forecasts. Reliability and
sharpness, and 
%% another example of track changes.
\replaced{hence}{therefore} skill,
 of the former is expected to improve over
time. Thus, the gain by post-processing is expected to decrease. Based
on ECMWF forecasts from January 2002 to March 2014 and corresponding
observations from globally distributed stations we generate
post-processed forecasts by ensemble model output statistics
%% another example of track changes.
\added{abbreviated as} (EMOS)
for each station and variable. Given the higher average skill of the
post-processed forecasts, we analyse the evolution of the difference
in skill between raw ensemble and EMOS. This skill gap remains almost
constant over time indicating that post-processing will keep adding
skill in the foreseeable future.
\end{abstract}

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

%%% Suggested section heads:
% \section{Introduction}
% 
% The main text should start with an introduction. Except for short
% manuscripts (such as comments and replies), the text should be divided
% into sections, each with its own heading.

% Headings should be sentence fragments and do not begin with a
% lowercase letter or number. Examples of good headings are:

% \section{Materials and Methods}
% Here is text on Materials and Methods.

% \subsection{A descriptive heading about methods}
% More about Methods.
% 
% \section{Data} (Or section title might be a descriptive heading about data)
% 
% \section{Results} (Or section title might be a descriptive heading about the
% results)
% 
% \section{Conclusions}

\section{Introduction}
\label{sec:intro}
Over the last two decades the paradigm in weather forecasting has
shifted from being deterministic to probabilistic \citep[see
e.g.][]{Palm00,Hami00}. Accordingly, numerical weather prediction
(NWP) models have been run increasingly as ensemble forecasting
systems. The goal of such ensemble forecasts is to approximate the
forecast probability distribution by a finite sample of scenarios
\citet{Leit74}\footnote{See \citet{Leit74} for a more in-depth
description of these issues and their complex implications.}
Global ensemble forecast systems, like the European
Centre for Medium-Range Weather Forecasts (ECMWF) ensemble, are prone
to probabilistic biases, and are therefore not reliable. They
particularly tend to be underdispersive for surface weather parameters
\citet{Boug10, Park08}. In order to correct for forecast
underdispersion and bias in NWP ensembles different
statistical
post-processing methods have been developed, of which ensemble model
output statistics (EMOS) \citep{Gneit05} is among the most widely
applied. EMOS yields a parametric forecast distribution by linking its
parameters to ensemble statistics. Due to its simplicity and low
computational cost, we focus on EMOS for this study.

The ECMWF ensemble is under continuous development, and hence its
forecast skill improves over time \citep{Buiz98,Buiz07,Rich13,Haid14}.
Parts of these improvements may be due to a reduction of probabilistic
biases. From this we deduce the following hypothesis: As the raw
forecasts continuously improve, it is hypothesized that the gap in
skill between raw ensemble and post-processed forecasts narrows,
because systematic errors typically captured by post-processing are
reduced 
by those improvements. 
%%
%% Use \explain{} before change, so that it ends up on the right line.
\explain[Jon, 2/16/16]{Redundant sentence, better without it} 
\deleted{In other words, probabilistic biases,
which can be reduced by statistical post-processing methods, decrease
over time.}
Assuming that the raw ensemble forecasts continue to
improve in the future, the gap in skill may eventually be closed when
the raw ensemble forecasts become reliable and unbiased.  In this work
we analyse the evolution of the global performance of the operational
ECMWF raw ensemble and the corresponding post-processed EMOS forecasts
for 2 metre temperature (T2M), 24-hour precipitation (PPT24), and 10-m
wind speed (V10). We verify the forecasts against globally distributed
surface synoptic observations (SYNOP) data over a period of about 10
years. We firstly evaluate the monthly average skill in terms of CRPS
for both the raw and the EMOS forecasts. In order to assess the extent
to which the results depend on the choice of the post-processing
method, Bayesian model averaging (BMA), \citet{Raft05,Fra10} is
additionally applied to the T2M raw ensemble forecasts. We will use
the negatively oriented (i.e. the lower the value the higher the
skill) continuous ranked probability score (CRPS) \citep{Math76} as a
measure of skill. As the CRPS assesses both reliability and sharpness
and is a proper score \citep{Gneit07b}, we rely on it for model fitting
and verification throughout this study. Note that skill and
reliability are linked in that given constant sharpness an improvement
in reliability leads to an improvement in skill and vice versa. We
finally analyse the evolution of the gap in CRPS between raw ensemble
and post-processed forecasts.

After presenting the dataset in section~\ref{sec:intro} we summarize
the methods for post-processing and for the assessment of the global
skill evolution in section~\ref{sec:meth}. In section~\ref{sec:res}
the results are shown.  This is followed by a discussion in
section~\ref{sec:disc} along with some concluding
remarks. These
analyses have been performed using the statistical software \texttt{R}
\citep{R2013}.

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
%
%----------------
% EXAMPLE FIGURE
%
% \begin{figure}
% \end{figure}

\begin{figure}[ht!]
% \centering
% when using pdflatex, use pdf file:
% \includegraphics[width=20pc]{figsamp.pdf}
%
% when using dvips, use .eps file:
% \includegraphics[width=20pc]{figsamp.eps}
%
% If you don't specify the file extension, 
% \includegraphics will insert the right file:
\centerline{\includegraphics[height=1.5in]{figsamp}}
\caption{Short caption}
\label{figone}
\end{figure}

\begin{figure}[ht!]
\centerline{\includegraphics[angle=45,height=2in]{figsamp}}
\caption{The figure caption should begin with an overall descriptive
statement of the figure followed by additional text. They should be
immediately after each figure.  Figure parts are indicated with
lower-case letters ({\bf a, b, c}\ldots).  For initial submission, please place
both the figures and captions in the text near where they are cited.}
\label{fig:mean_and_slope}
\end{figure}

\begin{table}
\caption{Start this caption with a short description of your
table.
Large tables
especially presenting rich data should be presented as separate excel
or .cvs files, not as part of the main text.}
== Table Here ==
\end{table}

%
% ---------------
% EXAMPLE TABLE
%
\begin{table}
\caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
\centering
\begin{tabular}{l c}
\hline
 Run  & Time (min)  \\
\hline
  $l1$  & 260   \\
  $l2$  & 300   \\
  $l3$  & 340   \\
  $h1$  & 270   \\
  $h2$  & 250   \\
  $h3$  & 380   \\
  $r1$  & 370   \\
  $r2$  & 390   \\
\hline
\multicolumn{2}{l}{$^{a}$Table note text here.}
\end{tabular}
\end{table}

% See below for how to make sideways figures or tables.

% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
 \begin{sidewaysfigure}
 \includegraphics[width=20pc]{figsamp}
 \caption{Caption here}
 \label{newfig}
 \end{sidewaysfigure}
%
 \begin{sidewaystable}
 \caption{Caption here}
\label{tab:signif_gap_clos}
 \begin{tabular}{ccc}
one&two&three\\
four&five&six
 \end{tabular}
 \end{sidewaystable}

\clearpage

\section{Methods}
\label{sec:meth}

\subsection{Post-processing Using EMOS}
Post-processing using EMOS converts a raw ensemble of discrete
forecasts into a probability distribution. Let $y$ be the variable to
be forecast (here: T2M, PPT24 or V10) and let ${f}=(f_{1}, f_{2},
\dots, f_{K})^{T}$ be the vector of the $K$ member raw ensemble
forecasts (here: HRES, ENS, and CTRL). Then the EMOS
\added{predictive}
density can be written as:
\begin{linenomath*}
\begin{equation}
y|{f} \sim g(m, \sigma),
\end{equation}
\end{linenomath*}
where $g(\cdot)$ is a parametric density function with location and
scale parameters $m$ and $\sigma$, respectively, which depend on the
raw ensemble.

\subsubsection{Temperature}
For T2M forecasts $g(\cdot)$ is a normal density distribution with
mean $m$ and variance $\sigma^{2}$. Here, we use a variant of the
original EMOS approach similar to the one proposed by
\citet{ScheueBue13} where the departures of observed temperatures from
their climatological means are related to those of the forecasts.
Specifically, let $T=\{t_{1},\ldots,t_{n}\}$ be a training period of
$n$ days preceding the forecast initialization and denote by $f_{tk}$
the forecast of the k-th ensemble member and by $y_{t}$ the
observation on day $t\in T$. As a first step, we fit a regression
model
%%% \begin{linenomath*} ...\end{linenomath*} puts a line number in equation:
\begin{linenomath*}
\begin{equation}
y_{t_{j}} = c_{0} + c_{1}\sin\left(\frac{2\pi j}{365}\right) +
c_{2}\cos\left(\frac{2\pi j}{365}\right) + \varepsilon_{t_{j}}, \quad
j=1,\ldots,n
\end{equation}
\end{linenomath*}
which captures the seasonal variation of T2M. The residual terms
$\varepsilon_{t_{j}}$ are likely correlated over time, but for
simplicity an ordinary least squares fit is performed. We denote by
$\tilde{y}_{t}$ the fitted value of this periodic regression model on
day $t$ and interpret it as the climatological mean temperature on
this day. This model can easily be extrapolated to future days
$t_{d+1}, t_{d+2}, \ldots$ The above regression includes both a sine
and a cosine term which is equivalent to a cosine model with variable
phase and amplitude. Since $j=1,\ldots,n$ is just a numbering of the
days in $T$, different training periods have different phase
parameters and hence $c_{1}$ and $c_{2}$ evolve over the calendar
year. We fit the same type of model also to the ensemble mean,
control, and high resolution run and obtain climatological means
$\tilde{f}_{\overline{\mathrm{ENS}},t}, \tilde{f}_{\mathrm{CTRL},t}$,
and $\tilde{f}_{\mathrm{HRES},t}$. The mean of the forecast
distribution is then:
\begin{linenomath*}
\begin{equation}
m = \tilde{y} + a_{1}(f_{\mathrm{HRES}} - \tilde{f}_{\mathrm{HRES}}) +
a_{2}(f_{\mathrm{CTRL}} - \tilde{f}_{\mathrm{CTRL}}) +
a_{3}(f_{\overline{\mathrm{ENS}}} -
\tilde{f}_{\overline{\mathrm{ENS}}}).
\end{equation}
\end{linenomath*}
The variance of the forecast distribution is linked to the raw
ensemble by: 
\begin{linenomath*}
\begin{equation}
\sigma^{2} = b_{0} + b_{1}s^{2},
\end{equation}
\end{linenomath*}
where $s^{2} = \frac{1}{K}\sum_{k=1}^{K}(f_{k} -
\frac{1}{K}\sum_{k=1}^{K}f_{k})^{2}$. The parameters
${\theta}_{T2M}=(a_{1}, a_{2}, a_{3}, b_{0}, b_{1})^{T}$ are
constrained to be non-negative, and hence $a_{k}/\sum_{k=1}^{K} a_{k}$
can be understood as the weight of model $k$.

\subsubsection{Precipitation}
For PPT24 we use the EMOS approach proposed by \citet{Scheu14}, where
$g(\cdot)$ is a left-censored (at zero) generalized extreme value
(GEV) distribution. While the shape parameter $\xi$ of the GEV is kept
constant ($\xi=0.2$), the location and the scale parameters $m$ and
$\sigma$ are linked to the raw ensemble via:
\begin{linenomath*}
\begin{eqnarray}
m &=& a_{0} + a_{1} f_{\mathrm{HRES}} + a_{2} f_{\mathrm{CTRL}} +
a_{3} f_{\overline{\mathrm{ENS}}} + a_{4} \pi_{0},\nonumber\\
\sigma &=& b_{0} + b_{1} \mathrm{MD}_{f},
\end{eqnarray}
\end{linenomath*}
where $\pi_{0}$ is the fraction of ensemble members predicting zero
precipitation and $\mathrm{MD}_{f}:= K^{-2}$\break
$\sum_{k, k' = 1}^{K} |
f_{k} - f_{k'} |$ is the ensemble mean difference. Again, the
parameters are denoted by $\theta_{PPT24}=(a_{0}, \dots, a_{4}, b_{0},
b_{1})^{T}$. The parameters $a_{1}, a_{2}, a_{3}, b_{0}, b_{1}$ are
constrained to be non-negative, and hence the normalized parameters
$a_{1}$ to $a_{3}$ can be understood as weights.

\subsection{Global CRPS Analysis}
 \label{sec:meth_glob_crps}

As stated in the introduction, the main objective of this study is to
analyse whether the gap in CRPS between the raw ensemble and the
post-processed forecast narrows over time. This is assessed
station-wise using both a parametric and a non-parametric approach.
For the former, we fit the following regression model to the monthly
time series of CRPS differences ($\Delta \mathrm{CRPS}_{t} =
\mathrm{CRPS}_{\mathrm{raw},t} - \mathrm{CRPS}_{\mathrm{EMOS}, t}$):
\begin{linenomath*}
\begin{equation}
\Delta \mathrm{CRPS}_{t} = \beta_{0} + \beta_{1} t + \beta_{2} \sin\left(\frac{2\pi t}{12}\right) + \beta_{3} \cos\left(\frac{2\pi t}{12} \right) + \epsilon, \phantom{bla} \epsilon \sim \mathcal{N}(0, \sigma^{2}) \label{eq:linseas}
\end{equation}
\end{linenomath*}
where $\Delta \mathrm{CRPS}_{t}$ is the predictand, $t$ is now the
time in months, and $\sigma^{2}$ denotes the error variance. For the
latter, we use Kendall's $\tau$ correlation coefficient and the
associated test statistics \citep{Mann45} as implemented in the
\texttt{R} package \texttt{Kendall} \citep{Kendall}. In order to
correct for seasonal effects, we calculate the $\tau$ statistics using
the residuals of the following model:
\begin{linenomath*}
\begin{equation}
\Delta \mathrm{CRPS}_{t} = \gamma_{0} + \gamma_{1} \sin\left(\frac{2\pi t}{12}\right) + \gamma_{2} \cos\left(\frac{2\pi t}{12} \right) + \epsilon, \phantom{bla} \epsilon \sim \mathcal{N}(0, \sigma^{2})
\label{eq:tauseas}
\end{equation}
\end{linenomath*}
Note that negative $\tau$ values indicate a negative trend and
positive values a positive one. Figure~\ref{figone} a) and b)
show the regression lines estimated by model (\ref{eq:linseas}) for
monthly averages of $\Delta \mathrm{CRPS}$ and the corresponding
Kendall's $\tau$ test statistics for an example with decreasing and
increasing gap.

\section{Results}
\label{sec:res}

\subsection{Are There Any Significant Temporal Trends?}

The above results indicate a tendency of a decrease in $\Delta$CRPS
over time at least for T2M and PPT24. In the following we check the
percentages of stations with decreasing, an absence of, or increasing
trend in $\Delta$CRPS over time at a significance level of 0.05. In
order to be more confident about the results this analysis is
performed using both the parametric regression model and the
non-parametric Kendall's $\tau$ correlation coefficient test. As
already mentioned both approaches correct for seasonal effects.
Furthermore, in case of T2M the same analysis has been performed
additionally using BMA instead of EMOS in order to relax the
dependence on one particular post-processing method. As shown in
Table~\ref{tab:signif_gap_clos} the stations with no significant trend
outnumber the stations with either negative or positive trend for all
three variables and lead times considered. Note that the percentage of
stations without any significant trend increases with increasing lead
time. In line with the results shown in
Figure~\ref{fig:mean_and_slope}, significantly negative trends are
more common than positive ones for T2M and PPT24. The difference
between the number of stations with negative and those with positive
trend reduces with increasing lead time, but is still greater than
zero for a 10 day forecast. Note that the high number of
non-significant stations in case of PPT24 is likely to be due to the
high variability of precipitation amounts, and hence variability of
CRPS values, which leads to a large residual standard error in case of
the parametric regression model and to a lot of pairs (a pair denotes
here a value of $\Delta$CRPS and its associated time stamp) opposite
to the estimated direction in case of the $\tau$ test statistics. In
case of V10 the stations with a negative trend and those with a
positive trend are almost equally frequent regardless of the lead
time. Figures of the global distributions of stations with no,
significantly negative, and significantly positive trend in
$\Delta$CRPS are available as supplemental material to this paper.

\section{Discussion and Conclusions}
\label{sec:disc}

According to the above analyses the gap in CRPS between the raw
ensemble and the EMOS forecasts remains almost constant over time. For
T2M and PPT24 $\Delta$CRPS shows a slightly decreasing tendency. The
higher the lead time the less accentuated is this tendency. For V10
such a tendency cannot be detected. The parametric regression model
and the non-parametric $\tau$ test yield similar results. Hence, a
linear model that is overlaid by seasonal fluctuations seems to be
reasonable. Note that the skill of the raw ensemble and the EMOS
forecasts may sometimes be negatively affected by upgrades to the
atmospheric model. Model upgrades may deteriorate raw ensemble skill
at some individual stations. For instance, a resolution increase may
introduce new issues with statistical downscaling of the forecasts to
some specific observation sites. But more importantly, the skill of
the post-pro\-cessed forecasts can be lowered dramatically if a model
update happens between the training and the verification period. These
issues may result in positive trends in $\Delta$CRPS. Ideally,
post-processing would be based on a cascade of reforecasts. That is,
for each atmospheric model version, training of the post-processing
model would be done using a corresponding time series of reforecasts
made with that same model version. Furthermore, the observations may
be affected by measurement errors. If these errors change over time,
they may also influence the estimates of the trends in $\Delta$CRPS.
As the problems introduced by statistical downscaling may be mitigated
by verifying against model analysis, a similar study that replaces
observations by model analysis, as proposed by \citet{Ghel00} and
\citet{Papp09}, may give further insights.

From the above we conclude that the probabilistic skill of both the
raw ensembles and the EMOS forecasts improves over time. The fact that
the gap in skill has remained almost \textbf{constant}, especially for
V10, suggests that improvements to the atmospheric model have an
effect quite different from what calibration by statistical
post-processing is doing. That is, they are increasing potential
skill. Thus this study indicates that (a) further model development is
important even if one is just interested in point forecasts, and (b)
statistical post-processing is important because it will keep adding
skill in the foreseeable future.

%% ------------------------------------------------------------------------ %%
\section*{Citations}

% Please use ONLY \citet and \citep for reference citations.
% DO NOT use other cite commands (e.g., \cite, \citeyear, \nocite, \citealp, etc.).

\subsection*{Cites made with \tt\string\citet\string{\string}}
...as shown by \citet{Boug10}, \citet{Buiz07}, \citet{Fra10},
\citet{Ghel00}, and \citet{Leit74}.

\subsection*{Cites made with \tt\string\citep\string{\string}}
...as shown by \citep{Boug10}, \citep{Buiz07}, \citep{Fra10},
\citep{Ghel00, Leit74}.

...has been shown \citep [e.g.,][]{Boug10,Buiz07,Fra10}.

%%% End of body of article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendix goes here
%
% \appendix resets counters and redefines section heads
% but doesn't print anything.
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% A: Here Is Appendix Title
%
\appendix
\section{Here is a sample appendix}
This is an Appendix section.

\subsection{subsection}
This is an Appendix subsection.

\subsubsection{subsubsection}
This is an Appendix subsubsection.
\begin{linenomath*}
\begin{equation}asdf\end{equation}
\end{linenomath*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Optional Glossary, Notation or Acronym section goes here:
%
%%%%%%%%%%%%%%
% Glossary is only allowed in Reviews of Geophysics
\begin{glossary}
\term{Term}
 Term Definition here
\term{Term}
 Term Definition here
\term{Term}
 Term Definition here
\end{glossary}

%
%%%%%%%%%%%%%%
% Acronyms
 \begin{acronyms}
 \acro{Acronym}
 Definition here
 \acro{EMOS}
 Ensemble model output statistics 
 \acro{ECMWF}
 Centre for Medium-Range Weather Forecasts
 \end{acronyms}

%
%%%%%%%%%%%%%%
% Notation 
 \begin{notation}
 \notation{$a+b$} Notation Definition here
 \notation{$e=mc^2$} 
 Equation in German-born physicist Albert Einstein's theory of special
relativity that showed that the increased relativistic mass ($m$) of a
body comes from the energy of motion of the body—that is, its kinetic
energy ($E$)—divided by the speed of light squared ($c^2$).
 \end{notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  ACKNOWLEDGMENTS

\acknowledgments
The forecast data used in this study can be made available subject to
a handling charge. SYNOP observations are not available through ECMWF.
We are grateful to F. Rabier, D. S. Richardson, D. Lavers, and other
colleagues at ECMWF for helpful discussions and inputs. We thank David
M\"{o}ller of the University of Heidelberg for his contribution.
Furthermore, we like to thank T. Gneiting of the Heidelberg Institute
for Theoretical Studies and of the Institute for Stochastics at
Karlsruhe Institute of Technology for valuable comments, and the
colleagues at the Heidelberg Institute for Theoretical Studies.
Additionally, we are grateful to two anonymous reviewers for their
helpful comments. We are indebted to the ECMWF for granting access to
their products.

The Editor thanks Pierre Pinson and an anonymous reviewer for their assistance in evaluating this paper.

%%  REFERENCE LIST AND TEXT CITATIONS
%
% Either type in your references using
%
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% Or, to use BibTeX:
%
% Follow these steps
%
% 1. Type in \bibliography{<name of your .bib file>} 
%    Run LaTeX on your LaTeX file.
%
% 2. Run BiBTeX on your LaTeX file.
%
% 3. Open the new .bbl file containing the reference list and
%   copy all the contents into your LaTeX file here.
%
% 4. Run LaTeX on your new file which will produce the citations.
%
% AGU does not want a .bib or a .bbl file. Please copy in the contents of your .bbl file here.

\begin{thebibliography}{}

\bibitem[{\textit{Bell and Munoz}}(2008)]{Boug10} Bell, A.~H., and
Munoz, D.~P.  (2008). Activity in the superior colliculus reflects
dynamic interactions between voluntary and involuntary influences on
orienting behaviour. \textit{Eur. J. Neurosci.} 28, 1654--1660.

\bibitem[{\textit{Corbetta et~al.}}(1991)]{Buiz07} Corbetta, M.,
Miezin, F.~M., Dobmeyer, S., Shulman, G.~L., and Petersen, S.~E.
(1991). Selective and divided attention during visual discriminations
of shape, color, and speed: functional anatomy by positron emission
tomography. \textit{J.~Neurosci.} 11, 2383--2402.

\bibitem[{\textit{Borra et~al.}}(2014)]{Buiz98} Borra, E., Gerbella,
M., Rozzi, S., Tonelli, S., and Luppino, G.  (2014). Projections to
the superior colliculus from inferior parietal, ventral premotor, and
ventrolateral prefrontal areas involved in controlling goal-directed
hand actions in the macaque. \textit{Cereb. Cortex} 24, 1054--1065.

\bibitem[{\textit{Dorris et~al.}}(1997)]{Fra10}
 Dorris, M.~C.,
Par\'{e}, M., and Munoz, D.~P.  (1997). Neuronal activity in monkey
superior colliculus related to the initiation of saccadic eye
movements. \textit{J.~Neurosci.} 17, 8566--8579.

\bibitem[{\textit{Elsabbagh et~al.}}(2009)]{Ghel00} 
Elsabbagh, M.,
Volein, A., Holmboe, K., Tucker, L., Csibra, G., Baron-Cohen, S.,
et~al.  (2009). Visual orienting in the early broader autism
phenotype: disengagement. \textit{J.~Child Psychol. Psychiatry} 50,
637--642.

\bibitem[{\textit{Fortin et~al.}}(1999)]{Gneit05} Fortin, S., Chabli,
A., Dumont, I., Shumikhina, S., Itaya, S.~K., and Molotchnikoff, S.
(1999). Maturation of visual receptive field properties in the rat
superior colliculus. \textit{bibain Res. Dev. Brain Res.} 1112,
55--64.

\bibitem[{\textit{Felsen and Mainen}}(2008)]{Gneit07b} Felsen, G., and
Mainen, Z.~F.  (2008). Neural substrates of sensory-guided locomotor
decisions in the rat superior colliculus. \textit{Neuron} 60,
137--148.

\bibitem[{\textit{Gattass and Desimone}}(1996)]{Haid14} Gattass, R.,
and Desimone, R.  (1996). Responses of cells in the superior
colliculus during performance of a spatial attention task in the
macaque. \textit{Rev. Bras. Biol.} 56, 257--279.

\bibitem[{\textit{Goldberg and Wurtz}}(1972)]{Hami00} Goldberg, M.~E.,
and Wurtz, R.~H.  (1972). Activity of superior colliculus in behaving
monkey. II. Effect of attention on neuronal responses.
\textit{J.~Neurophysiol.} 35, \hbox{560--574}.

\bibitem[{\textit{Krauzlis}}(2003)]{Kendall} Krauzlis, R.~J.  (2003).
Neuronal activity in the rostral superior colliculus related to the
initiation of pursuit and saccadic eye movements.
\textit{J.~Neurosci.} 23, 4333--4344.

\bibitem[{\textit{Heesy}}(2009)]{Leit74} 
Heesy, C.~P.  (2009). Seeing
in stereo: the ecology and evolution of primate binocular vision and
stereopsis. \textit{Evol. Anthropol.} 18, 21--35.

\bibitem[{\textit{Hilbig et~al.}}(2000)]{Mann45} Hilbig, H., Bidmon,
H.~J., Ettrich, P., and M\"{u}ller, A.  (2000). Projection neurons in
the superficial layers of the superior colliculus in the rat: a
topographic and quantitative morphometric analysis.
\textit{Neuroscience} 96, 109--119.

\bibitem[{\textit{Ignashchenkova et~al.}}(2004)]{Math76}
Ignashchenkova, A., Dicke, P.~W., Haarmeier, T., and Their, P.
(2004). Neuron-specific contribution of the superior colliculus to
overt and covert shifts of attention. \textit{Nat. Neurosci.} 7,
56--64.

\bibitem[{\textit{Krauzlis et~al.}}(2013)]{Palm00} Krauzlis, R.~J.,
Lovejoy, L.~P., and Z\'{e}non, A.  (2013). Superior colliculus and
visual spatial attention. \textit{Annu. Rev. Neurosci.} 36, 165--182.

\bibitem[{\textit{Kustov and Robinson}}(1996)]{Papp09} Kustov, A.~A.,
and Robinson, D.~L.  (1996). Shared neural control of attentional
shifts and eye movements. \textit{Nature} 384, 74--77.

\bibitem[{\textit{Landry and Bryson}}(2004)]{Park08} Landry, R., and
Bryson, S.~E.  (2004). Impaired disengagement of attention in young
children with autism. \textit{J.~Child Psychol. Psychiatry} 45,
1115--1122.

\bibitem[{\textit{Kobayashi et~al.}}(2003)]{R2013} Kobayashi, T.,
Tran, A.~H., Nishijo, H., Ono, T., and Matsumoto, G.  (2003).
Contribution of hippocampal place cell activity to learning and
formation of goal-directed navigation in rats. \textit{Neuroscience}
117, 1025--1035.

\bibitem[{\textit{McHaffie et~al.}}(2005)]{Raft05} McHaffie, J.~G.,
Stanford, T.~R., Stein, B.~E., Coizet, V., and Redgrave, P.  (2005).
Subcortical loops through the basal ganglia. \textit{Trends Neurosci.}
28, 401--407.

\bibitem[{\textit{McPeek and Keller}}(2004)]{Rich13} McPeek, R.~M.,
and Keller, E.~L.  (2004). Deficits in saccade target selection after
inactivation of superior colliculus. \textit{Nat. Neurosci.} 7,
757--763.

\bibitem[{\textit{M\"{u}ller et~al.}}(2005)]{Scheu14} M\"{u}ller,
J.~R., Philiastides, M.~G., and Newsome, W.~T.  (2005).
Microstimulation of the superior colliculus focuses attention without
moving the eyes. \textit{Proc. Natl. Acad. Sci. U.S.A.} 102, 524--529.

\bibitem[{\textit{Munoz and Istvan}}(1998)]{ScheueBue13} Munoz, D.~P.,
and Istvan, P.~J.  (1998). Lateral inhibitory interactions in
the intermediate layers of the monkey superior colliculus.
\textit{J.~Neurophysiol.} 79, 1193--1209.

\end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Track Changes:
% To add words, \added{<word added>}
% To delete words, \deleted{<word deleted>}
% To replace words, \replace{<word to be replaced>}{<replacement word>}

% At the end of the document, use \listofchanges, which will list the
% changes and the page and line number where the change was made.

% When final version, \listofchanges will not produce anything,
% \added{} word will be printed, \deleted{} will take away the word,
% \replaced{}{} will print only the 2nd argument.

%%%
\listofchanges
%%%

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supporting Information
%% (Optional) See AGUSuppInfoSamp.tex/pdf for requirements 
%% for Supporting Information.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

More Information and Advice:

%% ------------------------------------------------------------------------ %%
%
%  SECTION HEADS
%
%% ------------------------------------------------------------------------ %%

% Capitalize the first letter of each word (except for
% prepositions, conjunctions, and articles that are
% three or fewer letters).

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%% ------------------------------------------------------------------------ %%
%
%  IN-TEXT LISTS
%
%% ------------------------------------------------------------------------ %%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%% ------------------------------------------------------------------------ %%
%
%  EQUATIONS
%
%% ------------------------------------------------------------------------ %%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%% ------------------------------------------------------------------------ %%
%
%  EQUATION NUMBERING: COUNTER
%
%% ------------------------------------------------------------------------ %%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

% If you are using line numbers, remember to surround
% equations with \begin{linenomath*}...\end{linenomath*}

%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

### 2016-11-01 submission/agusuppinfosample.tex
Updated AGU style file

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supporting Information
%% (Optional)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% OVERVIEW
%
% Please note that all supporting information will be peer reviewed with your manuscript.
% In general, the purpose of the supporting information is to enable
% authors to provide and archive auxiliary information such as data
% tables, method information, figures, video, or computer software,
% in digital formats so that other scientists can use it.

% The key criteria are that the data:
% 1. supplement the main scientific conclusions of the paper but are not essential to the conclusions (with the exception of
%    including data so the experiment can be reproducible);
% 2. are likely to be usable or used by other scientists working in the field;
% 3. are described with sufficient precision that other scientists can understand them, and
% 4. are not exe files.
%

% All Supporting text and figures should be included in this document.

% Data sets, large tables, movie files,
% and audio files should be uploaded separately, following AGU naming
% conventions. Include their captions in this document and list the
% file name with the caption. You will be prompted to upload these
% files on the Upload Files tab during the submission process, using
% file type “Supporting Information (SI)”

\documentclass[draft]{agujournal}

% Please type in the journal name: \journalname{<Journal Name>}
% ie,
\journalname{Journal of Geophysical Research}

%% Choose from this list of Journals:
%
% Journal of Geophysical Research
% JGR-Biogeosciences
% JGR-Earth Surface
% JGR-Planets
% JGR-Solid Earth
% JGR-Space Physics
% Global Biochemical Cycles
% Geophysical Research Letters
% Paleoceanography
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resource Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science

\begin{document}

%% This command needs article title as argument to \supportinginfo{}:
\supportinginfo{Trends in the predictive performance of raw ensemble weather forecasts}

 \authors{A. B. Smith\affil{1}\thanks{Current address, McMurdo Station,
 Antartica},
 Eric Brown\affil{1,2}, Rick Williams\affil{3},
 John B. McDougall\affil{4}, and S. Visconti\affil{5}\thanks{Also
 funded by Monsanto.}}

\affiliation{1}{Department of Hydrology and Water Resources,
University of Arizona, Tucson, Arizona, USA.}
\affiliation{2}{Department of Geography, Ohio State University,
Columbus, Ohio, USA.}
\affiliation{3}{Department of Space Sciences, University of
Michigan, Ann Arbor, Michigan, USA.}
\affiliation{4}{Division of Hydrologic Sciences, Desert Research
Institute, Reno, Nevada, USA.}
\affiliation{5}{Dipartimento di Idraulica, Trasporti ed
Infrastrutture Civili, Politecnico di Torino, Turin, Italy.}

%% Corresponding Author
%(include name and email addresses of the corresponding author.  More
%than one corresponding author is allowed in this Word file and for
%publication; but only one corresponding author is allowed in our
%editorial system.)

\correspondingauthor{A. B. Smith}{email@address.edu}

%% ------------------------------------------------------------------------ %%
%
%  TEXT
%
%% ------------------------------------------------------------------------ %%

\section*{Contents}
%%%Remove or add items as needed%%%
\begin{enumerate}
\item Text S1 to Sx
\item Figures S1 to Sx
\item Tables S1 to Sx
\end{enumerate}

\section*{Additional Supporting Information (Files uploaded separately)}

\begin{enumerate}
\item Captions for Datasets S1 to Sx
\item Captions for large Tables S1 to Sx (if larger than 1 page, upload as separate excel file)
\item Captions for Movies S1 to Sx
\item Captions for Audio S1 to Sx
\end{enumerate}

\section*{Introduction}

%Delete all unused file types below. 
%Copy/paste for multiples of each file type as needed.

\section*{Text S1.}

Type or paste text here. This should be additional explanatory text,
such as: extended descriptions of results, full details of models,
extended lists of acknowledgements etc.  It should not be
additional discussion, analysis, interpretation or critique. It
should not be an additional scientific experiment or paper.

Repeat for any additional Supporting Text

%%Enter Data Set, Movie, and Audio captions here
%%EXAMPLE CAPTIONS

\section*{Data Set S1.} %Type or paste caption here.

Upload your dataset(s) to AGU's journal submission site and select
"Supporting Information (SI)" as the file type. Following naming
convention: ds01.

Repeat for any additional Supporting data sets

\section*{Movie S1.}

Type or paste caption here.

Upload your movie(s) to AGU's journal submission site and select,
"Supporting Information (SI)" as the file type. Following naming convention: ms01.

Repeat any additional Supporting movies

\section*{Audio S1.}

Type or paste caption here.

Upload your audio file(s) to AGU's journal submission site and select
Supporting Information ``(SI)" as the file type. Following naming
convention: auds01.

Repeat for any additional Supporting audio files

\end{document}

### 2016-11-02 submission/paper.tex
Updated AGU style file
Corrected the Bergen code, updated figures, edited some text

@@
-%\documentclass[wrr]{agutex}
-\documentclass[wrr, draft]{agutex}
-% Author names in capital letters:
-\authorrunninghead{THORARINSDOTTIR ET AL.}
+%% To submit your paper:
+\documentclass[draft,linenumbers]{agujournal}
+\draftfalse
 
-% Shorter version of title entered in capital letters:
-\titlerunninghead{ADAPTING TO UNCERTAIN SEA LEVEL RISE}
+%% For final version.
+% \documentclass{agujournal}
 
-%Corresponding author mailing address and e-mail address:
-\authoraddr{Corresponding author: Name, Address. (email)}
+\journalname{Water Resources Research}
 
-\usepackage{bbm, amsmath}
-%\usepackage[dvips]{graphicx}
-\usepackage{graphicx}
-\setkeys{Gin}{draft=false}
-\usepackage{paralist, booktabs}
-\usepackage{url}
-\usepackage{color}
-\usepackage{comment}
-
-\usepackage{lineno}
- \linenumbers*[1]
-%  To add line numbers to lines with equations:
+\usepackage{url, bbm, comment}
 
 \begin{document}
-
-%% ------------------------------------------------------------------------ %%
-%  TITLE
-%% ------------------------------------------------------------------------ %%
 
@@
 
-%% ------------------------------------------------------------------------ %%
-%  AUTHORS AND AFFILIATIONS
-%% ------------------------------------------------------------------------ %%
+\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, and K. de Bruin\affil{3,4}}
 
-% \altaffilmark will produce footnote;
-% matching \altaffiltext will appear at bottom of page.
+\affiliation{1}{Norwegian Computing Center, Oslo, Norway}
+\affiliation{2}{Technical University of Denmark, Copenhagen, Denmark}
+\affiliation{3}{Center for International Climate and Environmental Research, Oslo, Norway}
+\affiliation{4}{Wageningen Environmental Research, Wageningen, The Netherlands}
 
-\authors{T. Thorarinsdottir\altaffilmark{1}, P. Guttorp\altaffilmark{1}, M. Drews\altaffilmark{2}}
+\correspondingauthor{T. L. Thorarinsdottir}{thordis@nr.no}
 
-\altaffiltext{1}{Norwegian Computing Centere}
-
-\altaffiltext{2}{Technical University of Denmark}
-
-\altaffiltext{3}{Affiliation three}
-
-%% ------------------------------------------------------------------------ %%
-%  ABSTRACT
-%% ------------------------------------------------------------------------ %%
+\begin{keypoints}
+\item Keypoint 1
+\item Keypoint 2
+\item Keypoint 3
+\end{keypoints}
 
@@
 %% ------------------------------------------------------------------------ %%
-
-\begin{article}
 
@@
 \[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t}\]
-where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level on observed global temperature and $\varsigma_t$ the integrated time series regression error.
-
-%Figure with sea level projections & IPCC projections?
+where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.
 
@@
 The Bergen monthly series is missing data for 62 months, including all of the years 1942--43.
-To deal with occasional short stretches of missing data (most one or two months) we use median polish replacement \citep{medpol}.
-For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other stations corrected by the average difference.
+To deal with occasional short stretches of missing data (most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages.
+For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other Norwegian stations corrected by the average difference.
+
+The Esbjerg monthly series is missing data for 19 months. Here, too, we use median polish to fill in missing data and then compute annual averages.
 
@@
 
-\subsection{Sea level rise in Bergen}
-
-Figure \ref{fig:obs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
-The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17).
-The regression slope is 1.30 (0.12).
-\begin{figure}
-\begin{center}
-\includegraphics[width=39pc]{bergenfit.png}
-\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
-The straight line is the time series regression line.}
-\label{fig:obs}
-\end{center}
-\end{figure}
-
-For the relationship between global annual mean temperature and global annual mean sea level rise we use the results from \citet{Bolin2014a}.
-Figure \ref{fig:ci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999 under scenario RCP 8.5, which is the scenario Norwegian authorities recommend for planning purposes.
-\begin{figure}
-\begin{center}
-\includegraphics[width=39pc]{bergen_ci.png}
-\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
-The sea level data are shown in blue and end in 2015.
-The thin red lines are the projections without uncertainty based on each of the climate models.
-The dashed purple lines connect pointwise confidence intervals for each year.
-}
-\label{fig:ci}
-\end{center}
-\end{figure}
-
-\subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
-
-We estimate a damage distribution for Bergen in 2015 using annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP).
-The NPP data is aggregated to a county level and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion.
+We estimate a damage distribution for Bergen in 2015 using annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
+The NPP data are aggregated to a county level, and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion.
 Also, some of this discussion should probably be moved to the section on data.}.
@@
 \end{figure}
+
+\subsection{Sea level rise in Bergen and Esbjerg}
+
+Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
+The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17).
+The regression slope is 1.30 (0.12).
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{bergenfit.png}
+\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
+The straight line is the time series regression line.}
+\label{fig:bergenobs}
+\end{center}
+\end{figure}
+
+For the relationship between global annual mean temperature and global annual mean sea level rise we use the results from \citet{Bolin2014a}.
+Figure \ref{fig:bergenci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999 under scenario RCP 8.5, which is the scenario Norwegian authorities recommend for planning purposes.
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{bergen_ci.png}
+\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
+The sea level data are shown in blue and end in 2015.
+The thin red lines are the projections without uncertainty based on each of the climate models.
+The dashed purple lines connect pointwise confidence intervals for each year.
+}
+\label{fig:bergenci}
+\end{center}
+\end{figure}
+
+For Esbjerg, the time series regression model relating local to global sea level is an MA(1) model with parameter 0.17 (0.09).
+The regression slope is 1.02 (0.06).
+Figure \ref{fig:esbjergci} shows the simultaneous 90\% confidence region for sea level rise relative to 1999 under scenario RCP 8.5.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=39pc]{esbjerg_ci.png}
+\caption{Simultaneous 90\% confidence set (thick black lines) for Esbjerg sea level projections for the years 2000-2100 using RCP8.5.
+The sea level data are shown in blue and end in 2015.
+The thin red lines are the projections without uncertainty based on each of the climate models.
+The dashed purple lines connect pointwise confidence intervals for each year.
+}
+\label{fig:esbjergci}
+\end{center}
+\end{figure}
+
+\subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
 
@@
 \includegraphics[width=39pc]{AccumulatedFutureLoss.png}
-\caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and no adaptation measures.} 
+\caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 
 \label{fig:NoAction}
@@
 % 5\bibliographystyle{../BibTeX/agufull08}
-\bibliographystyle{agufull08}
+
 \bibliography{ref.bib}
@@
 
-%% ------------------------------------------------------------------------ %%
-%  END ARTICLE
-%% ------------------------------------------------------------------------ %%
-\end{article}
-
 %% Enter Figures and Tables here:

### 2016-11-02 code/Bergen.R

@@ -25,7 +25,8 @@ for (i in 1:N){
 
 	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
 	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	if(tmp[N,2]<12) tmp=tmp[-((N+1-tmp[N,2]):N),]#delete final year if not full
+	NN=dim(tmp)[1]
+	if(tmp[NN,2]<12) tmp=tmp[-((NN+1-tmp[NN,2]):NN),]#delete final year if not full
 	tmp.rlr=tmp[,3]
 	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
 	tmp.mp=medpolish(tmp.rlr,na.rm=T)
@@ -34,29 +35,29 @@ for (i in 1:N){
 	tmp.ann=apply(tmp.rlr,1,"mean")/10 #cm
 	tmp.yrs=unique(floor(tmp[,1]))
 	tmp.ann=tmp.ann-mean(tmp.ann[which(is.element(tmp.yrs,1970:1999))],na.rm=T)
-	norwaydata[i]=list(data.frame(years=unique(floor(tmp[,1])),data=tmp.ann))
+	norwaydata[i]=list(data.frame(years=tmp.yrs,data=tmp.ann))
 	}
 	
 
-#deal with Bergen missing data 1941-42
+#deal with Bergen missing data 1942-43
 #use average difference between Bergen and all stations
-#in 1940 and 1943 to estimate vaues for 1941 and 1942
+#in 1941 and 1944 to estimate vaues for 1942 and 1943
 
 tmp0=tmp1=tmp2=tmp3=NULL
 for (i in 1:N)
 {
-	tmp0=c(tmp0,norwaydata[[i]]$data[which(is.element(norwaydata[[i]]$years,1940))])
-tmp1=c(tmp1,norwaydata[[i]]$data[which(is.element(norwaydata[[i]]$years,1941))])
-tmp2=c(tmp2,norwaydata[[i]]$data[which(is.element(norwaydata[[i]]$years,1942))])
-tmp3=c(tmp3,norwaydata[[i]]$data[which(is.element(norwaydata[[i]]$years,1943))])
+	tmp0=c(tmp0,norwaydata[[i]]$data[which(is.element(norwaydata[[i]]$years,1941))])
[…]

### 2016-11-02 code/Esbjerg.R

@@ -23,8 +23,9 @@ for(i in 1:N){
 	tmp=cbind(floor(tmp[,1]),round(12*(tmp[,1]-floor(tmp[,1]))+.5),tmp[,2:4])
 
 	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	if(tmp[N,2]<12) tmp=tmp[-((N+1-tmp[N,2]):N),]#delete final year if not full
+	#if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	NN=dim(tmp)[1]
+	if(tmp[NN,2]<12) tmp=tmp[-((NN+1-tmp[NN,2]):N),]#delete final year if not full
 	tmp.rlr=tmp[,3]
 	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
 	tmp.mp=medpolish(tmp.rlr,na.rm=T)

### 2016-11-02 submission/paper.tex
Joining the two confidence bands in one figure

@@
 Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
-%Gravitaional effects
+%Gravitational effects
 Another major effect in Fennoskandia is the land rise due to isostatic rebound from the glaciers of the last ice age. 
@@
 \begin{center}
-\includegraphics[width=39pc]{DamageDistribution.png}
+\includegraphics[width=\linewidth]{DamageDistribution.png}
 \caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
@@
 \begin{center}
-\includegraphics[width=39pc]{bergenfit.png}
+\includegraphics[width=\linewidth]{bergenfit.png}
 \caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
@@
 For the relationship between global annual mean temperature and global annual mean sea level rise we use the results from \citet{Bolin2014a}.
-Figure \ref{fig:bergenci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999 under scenario RCP 8.5, which is the scenario Norwegian authorities recommend for planning purposes.
+The left panel of figure \ref{fig:ci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999 under scenario RCP 8.5, which is the scenario Norwegian authorities recommend for planning purposes.
 \begin{figure}
 \begin{center}
-\includegraphics[width=39pc]{bergen_ci.png}
-\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen sea level projections for the years 2000-2100 using RCP8.5.
+\begin{minipage}{.5\textwidth}
+
+ \includegraphics[width=\linewidth]{bergen_ci.png}
+ % \label{fig:bergenci}
+
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+
+    \includegraphics[width=\linewidth]{esbjerg_ci.png}
+ % \label{fig:esbjergci}
+
+\end{minipage}
+\caption{Simultaneous 90\% confidence set (thick black lines) for Bergen (left) and Esbjerg( right) sea level projections for the years 2000-2100 using RCP8.5.
 The sea level data are shown in blue and end in 2015.
@@
 }
-\label{fig:bergenci}
+\label{fig:ci}
 \end{center}
@@
 The regression slope is 1.02 (0.06).
-Figure \ref{fig:esbjergci} shows the simultaneous 90\% confidence region for sea level rise relative to 1999 under scenario RCP 8.5.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=39pc]{esbjerg_ci.png}
-\caption{Simultaneous 90\% confidence set (thick black lines) for Esbjerg sea level projections for the years 2000-2100 using RCP8.5.
-The sea level data are shown in blue and end in 2015.
-The thin red lines are the projections without uncertainty based on each of the climate models.
-The dashed purple lines connect pointwise confidence intervals for each year.
-}
-\label{fig:esbjergci}
-\end{center}
-\end{figure}
+The right panel of figure \ref{fig:ci} shows the simultaneous 90\% confidence region for sea level rise relative to 1999 under scenario RCP 8.5.
 

### 2016-11-03 submission/paper.tex
Some text in introduction

@@
 \section{Introduction}\label{sec:intro}
+The potential impact of climate change on local sea level, yielding effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats, requires city planners to make decisions in the presence of substantial uncertainty.
+
+In this paper we illustrate the uncertainty of sea level projections, and combine this with the a simplified decision problem where planners want to know how early they should implement costly adaptation measures.
 
@@
 In the simplest case, there is no change in investment costs ($C^0 = C^1 = C$) and the delay has no lasting effects beyond period 1 ($d_t^1 = d_t^0$ for $t > 1$).
-In this case, the comparison is between the expected return $r$ earned on the captial while implementation is delayed and one addtional time period of unmitigated damage,
+In this case, the comparison is between the expected return $r$ earned on the capital while implementation is delayed and one additional time period of unmitigated damage,
 \begin{linenomath*}
@@
 We use local tide gauge data from the Permanent Service for Mean Sea Level, UK, which is the worldwide repository for national sea level data.
-Glacial isostatic adjustment for Bergen is obtained from \citet{Simpson2014}.
+Glacial isostatic adjustment for Bergen is obtained from \citet{Simpson2014}, and for Esbjerg in personal communication from Peter Thejll at the Danish Meteorological Institute.
 
@@
 
-For the case study of sea level rise in Bergen, we use reported annual damage numbers from the Norwegian Natural Perils Pool which are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}.
-
-We estimate a damage distribution for Bergen in 2015 using annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
+We estimate a damage distribution for Bergen in 2015 using reported annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
 The NPP data are aggregated to a county level, and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion.
@@
 \eqref{eq:Burr}.
-The histogram of the observed annual damages and the estimated distribution are given in Figure~\ref{fig:BergenDamageDist}.
-The parameter estimates for the Burr distribution are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=\linewidth]{DamageDistribution.png}
-\caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
-\label{fig:BergenDamageDist}
-\end{center}
-\end{figure}
-
 \subsection{Sea level rise in Bergen and Esbjerg}
@@
 Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
+The glacial isostatic adjustment is 0.26 (0.07) cm/yr.
 The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17).
@@
 
- \includegraphics[width=\linewidth]{bergen_ci.png}
+ \includegraphics[width=\linewidth]{bergen_ci.pdf}
  % \label{fig:bergenci}
@@
 
-    \includegraphics[width=\linewidth]{esbjerg_ci.png}
+    \includegraphics[width=\linewidth]{esbjerg_ci.pdf}
  % \label{fig:esbjergci}
@@
 
-For Esbjerg, the time series regression model relating local to global sea level is an MA(1) model with parameter 0.17 (0.09).
+For Esbjerg, the glacial isostatic adjustment is 0.06 (0.03) cm/yr.
+The time series regression model relating gia-corrected local to global sea level is an MA(1) model with parameter 0.17 (0.09).
 The regression slope is 1.02 (0.06).
@@
 \subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
+
+The histogram of the observed annual damages and the estimated distribution are given in Figure~\ref{fig:BergenDamageDist}.
+The parameter estimates for the Burr distribution are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\linewidth]{DamageDistribution.png}
+\caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
+\label{fig:BergenDamageDist}
+\end{center}
+\end{figure}
 
@@
 \begin{center}
-\includegraphics[width=39pc]{EuropeanIncreaseLossExtrapolationUncertainty.png}
+\includegraphics[width=\linewidth]{EuropeanIncreaseLossExtrapolationUncertainty.png}
 \caption{ Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
@@
 \begin{center}
-\includegraphics[width=39pc]{AccumulatedFutureLoss.png}
+\includegraphics[width=\linewidth]{AccumulatedFutureLoss.png}
 \caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 

### 2016-11-04 submission/paper.tex
Started work on introduction and conclusion sections

@@
 
-\title{I don't know, are you sure we want to do this?\
+\title{I don't know, are you sure we want to do this?\\
 Sea level adaptation decisions under uncertainty}
@@
 
-In this paper we illustrate the uncertainty of sea level projections, and combine this with the a simplified decision problem where planners want to know how early they should implement costly adaptation measures.
+In this paper we illustrate the uncertainty of sea level projections, and combine this with a simplified decision problem where planners want to know how early they should implement costly adaptation measures.
+
+Section \ref{sealevelproj} describes our approach to projecting sea level.
+In section \ref{decision_tools} we describe the type of decision problem that we are going to attack.
+We apply these tools to a sea level projection for Bergen, Norway, %and for Esbjerg, Denmark
+in section \ref{cases}.  In the final section \ref{unc}  we demonstrate the consequences of ignoring the uncertainty in the projections.
 
 \section{Sea level projections {\color{blue} (PG)}}
+
+\label{sealevelproj}
 
@@
 
-We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies \citep{giss} to global mean sea level anomalies \citep{csiro}.
+We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies to global mean sea level anomalies.
 
@@
 \section{Decision tools {\color{blue} (KdB, MD, TT)}}
+\label{decision_tools}
 
@@
 \section{Case studies}
+\label{cases}
 
@@
 
-\section{Conclusions}
+\section{The value of including uncertainty}
+\label{unc}
+
+In many cases sea level rise projections are given as a single number for each scenario, usually the mean or median of the ensemble of projections from different climate models (e.g.
+\citet{climateimpactgroup}).
+Sometimes the spread of the ensemble is used to assess the uncertainty in the projections (e.g., the Norwegian Environmental Agency recommends using the upper ensemble value for RCP 8.5 as the basis for planning decision, pers.
+comm.
+from Even Nilsson, Norwegian Mapping Authority).
+In our analysiis there are two more sources of uncertainty, namely the two regression models.
+Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=\linewidth]{unc.pdf}
+\caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5.
+The black line is the median projection (with no uncertainty), while the grey histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
+}
+\label{fig:unc}
+\end{center}
+\end{figure}
 

### 2016-11-04 code/Esbjerg.R

@@ -40,7 +40,7 @@ for(i in 1:N){
 	gia.DK=giase.DK=rep(NA,N)
 	index=c(1,2,4,5,6,14) #with gia 
 	gia.DK[index]=c(-.012,-.306,-.028,-.009,-.010,0.06)
-	giase.DK[index]=c(.01,.14,.008,.01,.01,.02)
+	giase.DK[index]=c(.01,.14,.008,.01,.01,.03)
 
 #Get global model, data and CMIP5 output 
 load('~/GitHub/SeaLevelProjections/data&global.Rdata')

### 2016-11-07 code/BergenDecisions/BergenDamageData.R
Add code for Bergen decision-making

@@ -1,0 +1,206 @@
+rm(list=ls())
+A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
+A <- A[-37,] ## remove 2016 as data not complete
+V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
+V <- V[-37,] ## remove 2016 as data not complete
+CPI <- read.csv("KPI.csv", header=TRUE)
+CPI <- CPI[-1,] ## remove 2016 as data not yet available
+CPI$CPI <- CPI$CPI/CPI$CPI[1] ## normalize according to 2015 
+
+## plot data for Hordaland and Rogaland
+png(file="DamagesAmounts.png", width=720, height=480, points=18)
+par(mfrow=c(2,3),mex=0.75)
+plot(A$Hordaland, V$Hordaland/CPI$CPI[1:36], main="Hordaland 1980-2015", xlab="Nr. of damages", ylab="Amount (1000 NOK)", ylim=c(0,65000), xlim=c(0,500))
+hist(A$Hordaland, main="Hordaland 1980-2015", xlab="Nr. of damages", col="gray60", breaks=seq(0,500,by=10), xlim=c(0,500), ylim=c(0,18))
+hist(V$Hordaland/CPI$CPI[1:36], main="Hordaland 1980-2015", xlab="Amount (1000 NOK)", col="gray60", breaks=50, xlim=c(0,65000), ylim=c(0,22)) 
+
+plot(A$Rogaland, V$Rogaland/CPI$CPI[1:36], main="Rogaland 1980-2015", xlab="Nr. of damages", ylab="Amount (1000 NOK)", ylim=c(0,65000), xlim=c(0,500))
+hist(A$Rogaland, main="Rogaland 1980-2015", xlab="Nr. of damages", col="gray60", breaks=seq(0,500,by=10), xlim=c(0,500), ylim=c(0,18))
+hist(V$Rogaland/CPI$CPI[1:36], main="Rogaland 1980-2015", xlab="Amount (1000 NOK)", col="gray60", breaks=50, xlim=c(0,65000), ylim=c(0,22)) 
+dev.off()
+
+### Combine Hordaland and Rogaland data
+sf.nr <- c(A$Hordaland, A$Rogaland)
+sf.damage <- c(V$Hordaland/CPI$CPI[1:36], V$Rogaland/CPI$CPI[1:36])
+x11()
[…]

### 2016-11-08 code/BergenDecisions/BergenDamageData.R
updates to Bergen example

@@ -34,7 +34,7 @@ sf.damage[ind] <- 0.1
 
 log.lik <- function(theta)
 {
-    res <- sum(dburr(sf.damage, shape1=theta[1], shape2=theta[2], rate=theta[3], log=TRUE))
+    res <- sum(dburr(sf.damage/1e3, shape1=theta[1], shape2=theta[2], rate=theta[3], log=TRUE))
     return(-res)
 }
 
@@ -43,11 +43,11 @@ theta.start <- c(0.2, 1.5, 1.5)
 optim.out <- optim(theta.start, fn=log.lik, method="BFGS", control=list(maxit=1000))
 
 ## plot data and fitted distribution
-png(file="DamageDistribution.png", width=480, height=480, points=12)
-# pdf(file="DamageDistribution.pdf", width=5, height=5, points=12)
+## png(file="../../submission/DamageDistribution.png", width=480, height=480, points=12)
+pdf(file="../../submission/DamageDistribution.pdf", width=5, height=5, points=12)
 par(mex=0.75)
-hist(sf.damage, main="", xlab="Amount (1000 NOK)", col="gray60", breaks=50, xlim=c(0,65000), freq=FALSE)
-x <- seq(0, 65000, by=10)
+hist(sf.damage/1e3, main="", xlab="Amount (million NOK)", col="gray60", breaks=50, xlim=c(0,65), freq=FALSE)
+x <- seq(0, 65, by=0.01)
 y <- dburr(x, shape1=optim.out$par[1], shape2=optim.out$par[2], rate=optim.out$par[3])
 lines(x,y,col="red",lwd=2)
 dev.off()
@@ -83,10 +83,11 @@ lines(z, res$coef[1] + z * res$coef[2], col="red", lwd=2)
 dev.off()
 
 ## Uncertainty assessment by performing linear extrapolation per city
-png(file="EuropeanIncreaseLossExtrapolationUncertainty.png", width=480, height=480, points=14)
+pdf(file="../../submission/EuropeanIncreaseLossExtrapolationUncertainty.pdf", width=5, height=5, points=12)
+## png(file="EuropeanIncreaseLossExtrapolationUncertainty.png", width=480, height=480, points=14)
 par(mex=0.75)
 x <- c(0,20,40)
[…]

### 2016-11-09 code/Esbjerg.R
Esbjerg path simulation added

@@ -77,6 +77,13 @@ sta.un=sta$data #uncorrected station series
 stafit=arsta(seal,sta)
 
 summary(stafit$ar)
+# Series: sta$data[sta.common.years] 
+# ARIMA(0,0,1) with zero mean     
+
+# Coefficients:
+         # ma1  as.matrix(seal$sealevel[sea.common.years])
+      # 0.1746                                      1.0228
+# s.e.  0.0939                                      0.0638
 
 plot(seal$sealevel[sea.common.years],stafit$data[sta.common.years],xlab="Global sea level anomaly (cm)",ylab="Local sea level anomaly (cm)")
 
@@ -144,3 +151,10 @@ lines(x,Fmix.pdf(x,matrix(unlist(cis$pred),ncol=101,byrow=T)[,year],cis$sd.g[yea
 lines(x,Fmix.pdf(x,matrix(unlist(cis$pred),ncol=101,byrow=T)[,year],cis$sd[year]),col="blue",lwd=2)
 dev.off()
 
+ #Simulate paths from model
+nsample = 10000
+k=sample(1:K,nsample,replace=T) #vector of nsample climate models
+sim=matrix(nrow=nsample,ncol=101)
+for (i in 1:nsample)
+sim[i,]=rnorm(101,mu[k[i],],sd)
+save(sim,file="EsbjergSim.Rdata")

### 2016-11-09 submission/paper.tex
Minor edit in sec 5
Add code for Bergen decision-making
updates to Bergen example
Added  first stab at key terms

@@
 \begin{keypoints}
-\item Keypoint 1
-\item Keypoint 2
-\item Keypoint 3
+\item Decisions on adaptation measures need to take careful account of uncertanties
+\item More careful models are needed for increased costs of the effects of sea level rise
+\item Modelling local sea level rise is essential
 \end{keypoints}
@@
 The regression slope is 1.30 (0.12).
-\begin{figure}
+\begin{figure}[!hbpt]
 \begin{center}
@@
 The left panel of figure \ref{fig:ci} shows the simultaneous 90 \% confidence region for Bergen sea level rise relative to 1999 under scenario RCP 8.5, which is the scenario Norwegian authorities recommend for planning purposes.
-\begin{figure}
+\begin{figure}[!hbpt]
 \begin{center}
@@
 \begin{center}
-\includegraphics[width=\linewidth]{DamageDistribution.png}
+\includegraphics[width=0.5\linewidth]{DamageDistribution.pdf}
 \caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
@@
 
-\begin{figure}
+\begin{figure}[!hbpt]
 \begin{center}
-\includegraphics[width=\linewidth]{EuropeanIncreaseLossExtrapolationUncertainty.png}
+\includegraphics[width=0.5\linewidth]{EuropeanIncreaseLossExtrapolationUncertainty.pdf}
 \caption{ Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
@@
 
-\begin{figure}
+\begin{figure}[!hbpt]
 \begin{center}
@@
 Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
+We see that the ensemble range is about 16 cm, whereas the overall uncertainty range is about 40 cm.
 
-\begin{figure}
+\begin{figure}[!hbpt]
 \begin{center}

### 2016-11-10 code/BergenDecisions/BergenDamageData.R
small updates to Bergen example

@@ -212,8 +212,13 @@ add.damage <- yearly.damage - constant.damage
 cumsum.add.damage <- array(NA, dim=dim(add.damage))
 for(i in 1:dim(add.damage)[1]) 
     cumsum.add.damage[i,] <- cumsum(add.damage[i,])
-upper.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.9)
-lower.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.1)
+upper.95.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.95)
+lower.5.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.05)
+upper.90.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.9)
+lower.10.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.1)
+median.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.5)
+
+discount.rate <- c( rep(1.04, 40), rep(1.03, 35), rep(1.02, 10)) 
 
 png(file="CumulatedFutureAdditionalLoss5_95.png", width=480, height=480, points=14)
 par(mex=0.75)
@@ -222,17 +227,19 @@ plot(2016:2100, upper.add.cumsum/1e6/(1:85),type="l", ylim=c(0,15), xlab="Year",
 lines(2016:2100, lower.add.cumsum/1e6/(1:85))
 dev.off()
 
-png(file="CumulatedFutureAdditionalLoss5_95.png", width=480, height=480, points=14)
+png(file="CumulativeAdditionalDamageCosts.png", width=480, height=480, points=14)
 par(mex=0.75)
-x11()
-plot(2016:2100, upper.add.cumsum/1e3/1.02^(1:85),type="l", ylim=c(0,15), xlab="Year",
-      ylab="Accumulated damage in billion NOK", main="")
-lines(2016:2100, lower.add.cumsum/1e3/1.02^(1:85))
-lines(2016:2100, 1.1/1.02^(1:85), col="red")
+plot(2016:2100, upper.90.add.cumsum/1e3/cumprod(discount.rate),type="l", ylim=c(0,7), xlab="Year",
+      ylab="Accumulated damage (billion NOK)", main="", lty=2)
+lines(2016:2100, lower.10.add.cumsum/1e3/cumprod(discount.rate), lty=2)
[…]

### 2016-11-10 code/Bergen.R
Prob exceed 25cm

@@ -138,17 +138,16 @@ lines(2000:2100,cis$res$a,lwd=2)
 lines(2000:2100,cis$res$a.marginal,col="purple",lty=3)
 lines(2000:2100,cis$res$b.marginal,col="purple",lty=3)
 
-F.20 = F.40 = rep(0,101)
+F.25 =  rep(0,101)
 for(k in 1:K){
-	F.20 = F.20 + cis$res.20[[k]]$F/K
-	F.40 = F.40 + cis$res.40[[k]]$F/K
-}
+	F.25 = F.25 + cis$res.25[[k]]$F/K
+	}
 
-plot(range(2000,2100),c(0,1),type="n",col=1,xlab="Year",ylab="Probability",main="P(Y(s)<20, s<t)")
-lines(2000:2100,F.20,col=4,lwd=2)
-lines(2000:2100,F.40,col=2,lwd=2)
-legend(2000,0, c("20 cm","40 cm"), 
-lty=c(1,1), lwd=c(2.5,2.5),col=c(2,4),xjust=0,yjust=0) 
+plot(range(2000,2100),c(0,1),type="n",col=1,xlab="Year",ylab="Probability",main="P(Y(s)<25, s<t)")
+lines(2000:2100,F.25,col=4,lwd=2)
+#lines(2000:2100,F.40,col=2,lwd=2)
+#legend(2000,0, c("25 cm"), 
+#lty=c(1,1), lwd=c(2.5,2.5),col=c(2,4),xjust=0,yjust=0) 
 
 
 mu=NULL

### 2016-11-10 code/ci_bergen.R

@@ -78,16 +78,16 @@ for(k in 1:K){
 res <- simconf.mixture(alpha=alpha, mu = mu.pred, Q = Q, w = w)
 
 
-print(c("8.5","20cm",min((2000:2100)[res$b>20]),min((2000:2100)[apply(matrix(unlist(mu.pred),ncol=101,byrow=T), 2, mean, na.rm=T)>20]),min((2000:2100)[res$a>20])))
-print(c("8.5","40cm",min((2000:2100)[res$b>40]),min((2000:2100)[apply(matrix(unlist(mu.pred),ncol=101,byrow=T), 2, mean, na.rm=T)>40]),min((2000:2100)[res$a>40])))
+print(c("8.5","25cm",min((2000:2100)[res$b>25]),min((2000:2100)[apply(matrix(unlist(mu.pred),ncol=101,byrow=T), 2, mean, na.rm=T)>25]),min((2000:2100)[res$a>25])))
+#print(c("8.5","40cm",min((2000:2100)[res$b>40]),min((2000:2100)[apply(matrix(unlist(mu.pred),ncol=101,byrow=T), 2, mean, na.rm=T)>40]),min((2000:2100)[res$a>40])))
 
- res.20 = res.40 = NULL
+ res.25= NULL
 for(k in 1:K){
 	
-	res.20[[k]] = excursions(mu = mu.pred[[k]],Q=Q.pred,ind = 1:101,
-						   u=20, alpha=1,type='<')
-		res.40[[k]] = excursions(mu = mu.pred[[k]],Q=Q.pred,ind = 1:101,
-						   u=40, alpha=1,type='<')
+	res.25[[k]] = excursions(mu = mu.pred[[k]],Q=Q.pred,ind = 1:101,
+						   u=25, alpha=1,type='<')
+		# res.40[[k]] = excursions(mu = mu.pred[[k]],Q=Q.pred,ind = 1:101,
+						   # u=40, alpha=1,type='<')
 }
 #Calculate standard deviations for global sea level
 Sigma.g = a4^2*Sigma.i2
@@ -101,6 +101,6 @@ sd.g = sqrt(diag(Sigma.g.pred))
 
 
 
-return(list(pred=mu.pred,res=res,res.20=res.20,res.40=res.40,sd=sd,sd.g=sd.g))
+return(list(pred=mu.pred,res=res,res.25=res.25,sd=sd,sd.g=sd.g))
 }
 

### 2016-11-10 submission/paper.tex
Prob exceed 25cm
Abstract (still needs findings)

@@
 \begin{abstract}
-...
+Sea level rise has serious consequences for harbor infrastructure, storm drain and sewer systems, and many other issues.
+Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (inluding no action), and assessing at what point in time the chosen strategy should be implemented.
+All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, in the cost of no action.
+We develop two illustrative examples: for Bergen on Noirway's west coast and for Esbjerg on the west coast of Denmark.
+Different comonents of uncertainty are visualized.
+{\color{blue} Findings go here}
 \end{abstract}

### 2016-11-13 code/BergenDecisions/BergenDamageData.R
cleaning up Bergen code

@@ -1,13 +1,117 @@
 rm(list=ls())
-A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
-A <- A[-37,] ## remove 2016 as data not complete
-V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
-V <- V[-37,] ## remove 2016 as data not complete
-CPI <- read.csv("KPI.csv", header=TRUE)
-CPI <- CPI[-1,] ## remove 2016 as data not yet available
-CPI$CPI <- CPI$CPI/CPI$CPI[1] ## normalize according to 2015 
 
-## plot data for Hordaland and Rogaland
+### Read in damage data. 
+A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
+A <- A[-37,] ## Remove 2016 as data not complete
+V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
+V <- V[-37,] ## Remove 2016 as data not complete
+
+### Read in consumer price index (CPI).
+CPI <- read.csv("KPI.csv", header=TRUE)
+CPI <- CPI[-1,] ## Remove 2016 as data not yet available
+CPI$CPI <- CPI$CPI/CPI$CPI[1] ## Normalize to 2015 
+
+### Combine Hordaland and Rogaland data; standardize costs to 2015 values.
+sf.nr <- c(A$Hordaland, A$Rogaland)
+sf.damage <- c(V$Hordaland/CPI$CPI[1:36], V$Rogaland/CPI$CPI[1:36])
+sf.damage <- sf.damage/1e3 ## Normalize to million NOK. 
+
+### Fit Burr distribution to damage data.  
+library(actuar)
+ind <- which(sf.damage==0)
+sf.damage[ind] <- 1e-04 ## Improves numerical stability of estimates. 
+log.lik <- function(theta)
+{
+  res <- sum(dburr(sf.damage, 
+                   shape1=theta[1], 
+                   shape2=theta[2], 
+                   rate=theta[3], 
+                   log=TRUE))
+  return(-res)
+}
+theta.start <- c(5, 0.5, 0.05)
+optim.out <- optim(theta.start, 
+                   fn=log.lik, 
+                   method="BFGS", 
+                   control=list(maxit=1000))
+
+### Get Hallegatte et al. (2010) data on relation between change in damage 
+### cost and change in sea level.
+E <- read.csv("EuropeanLossData.csv", header=TRUE)
[…]

### 2016-11-20 submission/paper.tex
Minor fixes.

--- 2016-11-09 submission/paper.tex
@@
 \begin{abstract}
-...
+Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues.
+Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented.
+All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, in the cost of no action.
+We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark.
+Different components of uncertainty are visualized.
+{\color{blue} Findings go here}
 \end{abstract}
@@
 In section \ref{decision_tools} we describe the type of decision problem that we are going to attack.
-We apply these tools to a sea level projection for Bergen, Norway, %and for Esbjerg, Denmark
+We apply these tools to  sea level projections for Bergen, Norway, and for Esbjerg, Denmark
 in section \ref{cases}.  In the final section \ref{unc}  we demonstrate the consequences of ignoring the uncertainty in the projections.
@@
 \[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t}\]
-where $\gamma$ ie the annual land rise rate, t and $ {\hat b} $ is the regression coefficient relating global to local sea level.
+where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.
 
@@
 \begin{center}
-\includegraphics[width=\linewidth]{AccumulatedFutureLoss.png}
+\includegraphics[width=0.5\linewidth]{AccumulatedFutureLoss.png}
 \caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 
@@
 In our analysiis there are two more sources of uncertainty, namely the two regression models.
-Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
+Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
 We see that the ensemble range is about 16 cm, whereas the overall uncertainty range is about 40 cm.
@@
 \begin{center}
-\includegraphics[width=\linewidth]{unc.pdf}
+\includegraphics[width=0.5\linewidth]{unc.pdf}
 \caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5.
@@
 This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL).
-The source code for the analysis is implemented in the statistical programming language {\tt R} (\url{http://www.R-project.org}) and is available on GitHub at \url{http://github.com/eSACP/...}.
+The source code for the analysis is implemented in the statistical programming language {\tt R} (\url{http://www.R-project.org}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions}.
 \end{acknowledgments}

### 2016-11-21 submission/paper.tex
Added text about Bergen and Esbjerg in the introduction.

@@
 The potential impact of climate change on local sea level, yielding effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats, requires city planners to make decisions in the presence of substantial uncertainty.
+
+The Norwegian city of Bergen is the capital of Hordaland County.
+The city center is located on Byfjorden, and is surrounded by mountains.
+It has the largest port in Norway, both in terms of freight and passengers.
+The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UNESCO World Heritage site (\url{http://whc.unesco.org/en/list/59}).
+Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become a major problem in other parts of Bergen as well \citep{bergenreport}.
+
+Esbjerg, on the west coast of Jutland,  has the second largest harbour in Denmark.
+It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.
 
@@
 \caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5.
-The black line is the median projection (with no uncertainty), while the grey histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
+The black vertical line is the median projection (with no uncertainty), while the grey histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
 }

### 2016-11-21 submission/paper.tex
Updated sea level section

@@
 Among the factors that may invalidate this approach are changes in water storage on land (in essence removing water from the oceans), excessive siphoning of groundwater (resulting in land subsidence), changes in the rates of glacial and land ice melt, and changes in Earth's gravitational field due to transfer of mass from land ice to ocean water.
-For example, the rate of ice melt on Greenland may suddenly increase substantially due to intense warming of both air and sea water.
+For example, the rate of ice melt on Greenland may suddenly increase substantially due to intense warming of both air and sea water \citep{bamber2013}.
+A recent paper \citep{jevrejeva2016} indicates that the upper tails of sea level rise may be substantially higher when taking into account expert assessment of land ice melting.
 Our current climate models are not able to resolve the ice processes sufficiently to include such so called tipping points into the projections.
-Also, the IPCC scenarios \citep{change} do not include changes in water usage.
+Also, the IPCC scenarios \citep{change} do not include changes in water usage (cf.
+\citet{wada2012}).
 

### 2016-11-22 submission/paper.tex
Minor changes to equation in sea level predictions

@@
 \[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t}\]
+
 where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.

### 2016-11-25 code/BergenDecisions/BergenDamageData.R
Bergen code example

@@ -1,6 +1,7 @@
 rm(list=ls())
 
-### Read in damage data. 
+### Read in damage data obtained from Finance Norway's website:
+### www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/
 A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
 A <- A[-37,] ## Remove 2016 as data not complete
 V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
@@ -11,12 +12,13 @@ CPI <- read.csv("KPI.csv", header=TRUE)
 CPI <- CPI[-1,] ## Remove 2016 as data not yet available
 CPI$CPI <- CPI$CPI/CPI$CPI[1] ## Normalize to 2015 
 
-### Combine Hordaland and Rogaland data; standardize costs to 2015 values.
+### Combine Hordaland and Rogaland data.
+### Standardize costs to 2015 values using CPI. 
 sf.nr <- c(A$Hordaland, A$Rogaland)
 sf.damage <- c(V$Hordaland/CPI$CPI[1:36], V$Rogaland/CPI$CPI[1:36])
 sf.damage <- sf.damage/1e3 ## Normalize to million NOK. 
 
-### Fit Burr distribution to damage data.  
+### Fit Burr distribution to damage data (in MNOK)  
 library(actuar)
 ind <- which(sf.damage==0)
 sf.damage[ind] <- 1e-04 ## Improves numerical stability of estimates. 
@@ -35,12 +37,13 @@ optim.out <- optim(theta.start,
                    method="BFGS", 
                    control=list(maxit=1000))
 
-### Get Hallegatte et al. (2010) data on relation between change in damage 
-### cost and change in sea level.
+### Get Hallegatte et al. (2013) data on relation between change in damage 
+### cost and change in sea level (csv files copied manually from
+### online supplementary information).   
 E <- read.csv("EuropeanLossData.csv", header=TRUE)
 E[,2:4] <- E[,2:4]/E[,2] 
 
-### Hallegatte et al. (2010) provide data for 0, 20, 40 cm sea level rise.
+### Hallegatte et al. (2013) provide data for 0, 20, 40 cm sea level rise.
 ### Conservative extrapolation: Linear extrapolation of [20,40] beyond 40
 ### and [0,20] below 0. 
 x <- c(0, 20, 40)
[…]

### 2016-12-12 submission/paper.tex
Added material to introduction

@@
 \section{Introduction}\label{sec:intro}
-The potential impact of climate change on local sea level, yielding effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats, requires city planners to make decisions in the presence of substantial uncertainty.
+Long-term planning and decision-making regarding societal infrastructure along coastal areas must account for a changing climate.
+Especially as the potential impact of climate change on local sea level can yield numerous effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats (add REF).
+Considerable challenges continue to exist in understanding  the impacts of a combination of sea level rise and storm surges and the implementation of potential adaptation options to reduce these impacts.
+Severe inherent uncertainties persist along all parts of the processing chain from climate projections to adaptation assessments, and it is critical that decision-making appropriately accounts for this.
 
+As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012} adaptive planning methods need to support decisions in the short term, while considering long-term developments.
+Challenges of adaptation decision-making under uncertainty relate to the incorporation of spatial, inter-temporal and flexibility aspects of adaptation priorities \citep{FankhauserSoare2013}, and the linkage with specific characteristics of sectors and contexts \citep{BisaroSwartHinkel2016, HinkelBisaro2016}.
+Several economic decision support tools and methods exist for adaptation assessment under uncertainty \citep[e.g.][]{Chambwera&2014, WilbyDessai2010, WalkerHaasnootKwakkel2013}.
+However, \citet{Watkiss&2015} conclude that these tools are very resource intensive and complex in the context of long-term adaptation investment decisions and call for the development of ``light touch'' approaches to better support local adaptation making.
+
+In this paper we combine the uncertainty of sea level projections with a simplified decision framework under uncertainty where planners want to know how early they should implement costly adaptation measures.
+We discuss the advantages of using a fully probabilistic approach in a light touch decision framework that accounts for uncertainty both in climate projections and damage assessments, where information on uncertainty is turned into a driver of change rather than a barrier.
+
+We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark.
 The Norwegian city of Bergen is the capital of Hordaland County.
 The city center is located on Byfjorden, and is surrounded by mountains.
-It has the largest port in Norway, both in terms of freight and passengers.
-The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UNESCO World Heritage site (\url{http://whc.unesco.org/en/list/59}).
+It has the largest port in Nor way, both in terms of freight and passengers.
+The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UN-ESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}.
 Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become  a major problem in other parts of Bergen as well \citep{bergenreport}.
-
 Esbjerg, on the west coast of Jutland, has the second largest harbour in Denmark.
@@
 
-In this paper we illustrate the uncertainty of sea level projections, and combine this with a simplified decision problem where planners want to know how early they should implement costly adaptation measures.
-
 Section \ref{sealevelproj} describes our approach to projecting sea level.
-In section \ref{decision_tools} we describe the type of decision problem that we are going to attack.
-We apply these tools to  sea level projections for Bergen, Norway, and for Esbjerg, Denmark
-in section \ref{cases}.  In the final section \ref{unc}  we demonstrate the consequences of ignoring the uncertainty in the projections.
+In section \ref{decision_tools} we describe the decision framework focused on the timing of proactive adaptation decision.
+We apply this framework to sea level projections and adaptation for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
+In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.
 

### 2016-12-12 code/BergenDecisions/.#BergenDamageData.R
New plots for Bergen example

@@ -1,0 +1,1 @@
+disa@disa-VirtualBox.15425:1481529181

### 2016-12-12 code/BergenDecisions/BergenDamageData.R

@@ -51,19 +51,20 @@ low.slope <- rep(NA, 15)
 high.slope <- rep(NA, 15)
 for(i in 1:15)
 {
-  res <- lm(apply(E[i,2:3],2,mean)~x[1:2])
-  low.slope[i] <- res$coef[2]
-  res <- lm(apply(E[i,3:4],2,mean)~x[2:3])
-  high.slope[i] <- res$coef[2]
+    y <- as.double(E[i,2:4])
+    res <- lm(y[1:2]~x[1:2])
+    low.slope[i] <- res$coef[2]
+    res <- lm(y[2:3]~x[2:3])
+    high.slope[i] <- res$coef[2]
 }
 
 ### Function that calculates the multiplicative change in damage for sea level x 
 ### and Hallegatte et al. (2013) estimates for city k.  
 get.damage.mult <- function(x, k)
 {
-  if(x < 0) return(-1/(x * low.slope[k]))
-  else if(x < 20) return(x * low.slope[k])
-  else return(E[k,3] + x * high.slope[k])
+  if(x < 0) return(1/(1 + abs(x) * low.slope[k]))
+  else if(x < 20) return(1 + x * low.slope[k])
+  else return(as.double(E[k,3]) + x * high.slope[k])
 }
 
 ### Read in sea level rise projections for Bergen. 
@@ -156,8 +157,6 @@ for(i in 1:dim(add.damage)[1])
 }
 upper.95.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.95)
 lower.5.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.05)
-upper.90.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.9)
-lower.10.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.1)
 median.add.cumsum <- apply(cumsum.add.damage,2,quantile,0.5)
 upper.95.yearly.cumsum <- apply(cumsum.yearly.damage,2,quantile,0.95)
 lower.5.yearly.cumsum <- apply(cumsum.yearly.damage,2,quantile,0.05)
@@ -184,32 +183,101 @@ for(i in 1:I)
 ### adaptation options
 adapt.median <- apply(cumsum.adapt.damage[,,85],1,median)
 adapt.upper.95 <- apply(cumsum.adapt.damage[,,85],1,quantile,0.95)
[…]

### 2016-12-13 submission/paper.tex
Minor changes in acknowledgements.

@@
 Different components of uncertainty are visualized.
-{\color{blue} Findings go here}
+We show that failing to take uncertainty into account can increase the damage costs by an order of magnitude.
 \end{abstract}
@@
 from Even Nilsson, Norwegian Mapping Authority).
-In our analysiis there are two more sources of uncertainty, namely the two regression models.
+In our analysis there are two more sources of uncertainty, namely the two regression models.
 Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
@@
 This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL).
-The source code for the analysis is implemented in the statistical programming language {\tt R} (\url{http://www.R-project.org}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions}.
+The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions/Code}.
 \end{acknowledgments}

### 2016-12-13 code/BergenDecisions/.#BergenDamageData.R
Updated Bergen analysis

@@ -1,1 +1,0 @@
-disa@disa-VirtualBox.15425:1481529181

### 2016-12-13 code/BergenDecisions/BergenDamageData.R

@@ -1,3 +1,7 @@
+### Decision analysis for Bergen. All the data files are located in the same
+### directory as this script. The R working directory should thus be set
+### to the current directory. 
+
 rm(list=ls())
 
 ### Read in damage data obtained from Finance Norway's website:
@@ -368,8 +372,8 @@ dev.off()
 
 ## Plot comparing distributions of total damage for various uncertainty settings
 pdf(file="../../submission/Uncertainty.pdf", width=10, height=5, points=12)
-par(mex=0.75)
-hist(log(total.damage/1e3, freq=FALSE), ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
+par(mex=0.75, mar=c(5,4,2,2)+0.1)
+hist(total.damage/1e3, freq=FALSE, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
      col="black", border="black", xlab="Total damage 2016-2100 (billion NOK)",
      ylab="Density", xaxs="i", yaxs="i", main="") # black
 hist(total.slr.damage/1e3, freq=FALSE, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
@@ -383,35 +387,33 @@ abline(v=median(total.slr.damage/1e3), col="#7D26CD", lwd=3)
 abline(v=median(total.mf.damage/1e3), col="#4876FF", lwd=3)
 abline(v=median(total.dam.damage/1e3), col="#008B45", lwd=3)
 abline(v=sum(yearly.median.damage/1e3), col="yellow", lwd=3)
+legend("topright", legend=c("Full uncertainty","SLR uncertainty","Effect uncertainty","Damage uncertainty", "No uncertainty"),
+       col=c("black", "#7D26CD", "#4876FF", "#008B45", "yellow"), lty=1, lwd=2)
 box()
 dev.off()
 
+### Same as above but on a log-log scale
 pdf(file="../../submission/UncertaintyLog.pdf", width=10, height=5, points=12)
-par(mex=0.75)
-buckets <- seq(-7,5.2, by=0.05)
-my.hist <- hist(log(total.damage/1e3), breaks=buckets, plot=FALSE)
-, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
-                col="black", border="black", xlab="Total damage 2016-2100 (billion NOK)",
[…]

### 2016-12-14 submission/paper.tex
Updates to Bergen example in paper

@@
 
-\section{Sea level projections {\color{blue} (PG)}}
+\section{Sea level projections}
 
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{DamageDistribution.pdf}
-\caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
+\includegraphics[width=\linewidth]{DecisionAnalysisBergen.pdf}
+\caption{Damage data used in the decision analysis for Bergen: (a) Estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars); (b) Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
+The median change and the corresponding extrapolation are indicated in red.}
 \label{fig:BergenDamageDist}
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{EuropeanIncreaseLossExtrapolationUncertainty.pdf}
-\caption{ Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
-The median change and the corresponding extrapolation are indicated in red.}
-\label{fig:ChangeInDamage}
-\end{center}
-\end{figure}
-
-\begin{figure}[!hbpt]
-\begin{center}
-\includegraphics[width=0.5\linewidth]{AccumulatedFutureLoss.png}
-\caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 
+\includegraphics[width=0.5\linewidth]{CumDamageCostsBergen.pdf}
+\caption{Distribution of accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 
 \label{fig:NoAction}

### 2016-12-17 code/BergenDecisions/BergenDamageData.R
Uncertainty in Bergen

@@ -402,17 +402,25 @@ my.slr.hist <- hist(total.slr.damage, breaks=buckets, plot=FALSE)
 my.mf.hist <- hist(total.mf.damage, breaks=buckets, plot=FALSE)
 my.dam.hist <- hist(total.dam.damage, breaks=buckets, plot=FALSE)
 plot(log(my.slr.hist$breaks[-1]), log(my.slr.hist$counts), type="h", col="#7D26CD",
-     main="", ylab="Log frequency", xlab="Total damage 2016-2100 (log million NOK)", lwd=2)
+     main="", ylab="Log frequency", xlab="Total damage 2016-2100 (million NOK)",
+     lwd=2, axes=FALSE)
+ticks <- c(50, 150, 500, 1500, 5000, 15000, 50000, 150000)
+axis(1, at = log(ticks), labels=ticks)
+axis(2)
+box()
 lines(log(my.hist$breaks[-1])+0.02, log(my.hist$counts), col="black", type="h", lwd=2)
 lines(log(my.mf.hist$breaks[-1])+0.04, log(my.mf.hist$counts), col="orange", type="h", lwd=2)
 lines(log(my.dam.hist$breaks[-1])+0.06, log(my.dam.hist$counts), col="#008B45", type="h", lwd=2)
-# abline(v=log(median(total.damage)), col="black", lty=2)
-# abline(v=log(median(total.slr.damage)), col="#7D26CD", lty=2)
-# abline(v=log(median(total.mf.damage)), col="orange", lty=2)
-# abline(v=log(median(total.dam.damage)), col="#008B45", lty=2)
-# abline(v=log(sum(yearly.median.damage)), col="gray50", lty=2)
-legend("topright", legend=c("Full uncertainty","SLR uncertainty","Effect uncertainty","Damage uncertainty"),
-       col=c("black", "#7D26CD", "orange", "#008B45"), lty=1, lwd=2)
+abline(v=log(sum(yearly.median.damage)),  col="gray50", lwd=2)
+points(log(median(total.damage)), 7.8, col="black", pch=16)
+points(log(median(total.slr.damage)),7.8,  col="#7D26CD", pch=16)
+points(log(median(total.dam.damage)),7.8,  col="#008B45", pch=16)
+points(log(sum(yearly.median.damage)),7.8,  col="gray50", pch=16)
+points(log(median(total.mf.damage)),7.8,  col="orange", pch=16)
+legend("topright",
+       legend=c("Full uncertainty","SLR uncertainty","Effect uncertainty",
[…]

### 2016-12-29 submission/paper.tex
updating paper

@@
 \begin{keypoints}
-\item Decisions on adaptation measures need to take careful account of uncertanties
+\item Decisions on adaptation measures need to take careful account of uncertainties
 \item More careful models are needed for increased costs of the effects of sea level rise
-\item Modelling local sea level rise is essential
+\item Modeling local sea level rise is essential
 \end{keypoints}
@@
 Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented.
-All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, in the cost of no action.
+All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action.
 We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark.
 Different components of uncertainty are visualized.
-We show that failing to take uncertainty into account can increase the damage costs by an order of magnitude.
+We show that failing to take uncertainty into account can increase the projected damage costs by an order of magnitude.
 \end{abstract}
@@
 Long-term planning and decision-making regarding societal infrastructure along coastal areas must account for a changing climate.
-Especially as the potential impact of climate change on local sea level can yield numerous effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats (add REF).
+In particular, potential changes in local sea level can yield numerous effects such as frequent flooding, inundation and back flow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats.
 Considerable challenges continue to exist in understanding  the impacts of a combination of sea level rise and storm surges and the implementation of potential adaptation options to reduce these impacts.
@@
 The city center is located on Byfjorden, and is surrounded by mountains.
-It has the largest port in Nor way, both in terms of freight and passengers.
+It has the largest port in Norway, both in terms of freight and passengers.
 The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UN-ESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}.
 Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become a major problem in other parts of Bergen as well \citep{bergenreport}.
-Esbjerg, on the west coast of Jutland, has the second largest harbour in Denmark.
+
+The municipality of Bergen has, in cooperation with private actors, analyzed several possible adaptation measures against sea level rise.
+The measures range from an outer barrier that would protect the entire metropolitan area to various protection measures of limited areas in the inner harbor \citep{bergenreport}.
+While the viability of the constructions and the associated construction costs have been carefully analyzed, the optimal timing of potential adaptation measures and the effects of the associated uncertainties have yet to be investigated.
+We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs and the effects of sea level rise on changes in damage costs.
+
+Esbjerg, on the west coast of Jutland, has the second largest harbor in Denmark.
 It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.
 
+The remainder of the paper is organized as follows.
 Section \ref{sealevelproj} describes our approach to projecting sea level.
-In section \ref{decision_tools} we describe the decision framework focused on the timing of proactive adaptation decision.
+In section \ref{decision_tools_PartI} we describe the decision framework focused on the timing of proactive adaptation decision.
 We apply this framework to sea level projections and adaptation for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
@@
 \section{Sea level projections}
+\label{sealevelproj}
 
-\label{sealevelproj}
+We project local sea level changes by Modeling two processes:
+\begin{enumerate}
+\item The relationship between global temperature and global sea level;
+\item The relationship between global sea level and local sea level.
+\end{enumerate}
 
@@
 For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
-
-\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t}\]
+\begin{linenomath*}
+\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
+\end{linenomath*}
 where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.
@@
 \subsection{Local sea level}
-In order to get from global sea level projections to local ones, it is important to note that sea level rise not is uniform over the globe.
+In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe.
 Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
@@
 The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
-
-\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t}\]
-
+\begin{linenomath*}
+\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
+\end{linenomath*}
 where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.
@@
 
-\section{Decision tools {\color{blue} (KdB, MD, TT)}}
-\label{decision_tools}
+\section{Timing of adaptation measures}
+\label{decision_tools_PartI}
 
-\subsection{Timing of adaptation measures}
+To assess the optimal timing of an adaptation measure, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options.
+The probabilistic distribution is constructed by considering uncertainty
+\begin{itemize}
+\item in the local sea level projections;
+\item in the annual damage costs;
+\item in the effect of changes in sea level on the annual damage costs. 
+\end{itemize}
 
-We consider adaptation decision making related to the timing of proactive adaptation measures.
-That is, the goal is to adapt to sea level rise before major damages occur.
-In a cost-benefit framework, an investment should be delayed as long as the benefits of delay (avoided investment costs) are greater than the associated costs (higher climate change damages) \citep{Fankhauser&1999}.
+\subsection{Annual damage costs}
 
-We assume that damages related to sea level rise coincide with damages due to storm surges and that the accumulated annual damages are independent between years.
-More specifically, we model the distribution of annual damage, $F_{\textup{d}}$, by the three parameter Burr distribution \citep{Burr1942} with density
+We model the distribution of annual damage, $F_{\textup{d}, t_0}$, for the year $t_0 = 2015$ by the three parameter Burr distribution \citep{Burr1942} with density
 \begin{linenomath*}
   \begin{equation}\label{eq:Burr}
-  f_{\textup{d}} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
+  f_{\textup{d}, t_0} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
   \end{equation}
@@
 \cite{Klugman&2012}.
+The parameters of the distribution are estimated using historical data for annual storm surge damage.
+Data prior to 2015 are adjusted to the 2015 level using the consumer price index.
+After adjustment, we assume stationarity over the period and independence between years.
 
 Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d.
-values from the estimated damage distribution for $t_0 = 2015$, $\hat{F}_{\textup{d}, t_0}$.
-By repeating this process $J$ times, we obtain an empirial damage distribution for each future year $t_i$ given by
+values from the estimated distribution $\hat{F}_{\textup{d}, t_0}$.
+By repeating this process $J$ times, we obtain an empirical damage distribution for each future year $t_i$ given by
 \begin{linenomath*}
   \[
-  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{d_{t_i}^{(j)} \leq x \}
+  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l}) } \leq x \Bigg\}
   \]
 \end{linenomath*}
-for $i = 1, \ldots, 85$. Similarly, we obtain an empirial distribution of the total damage over the period $2016-2100$ by considering $d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} d_{t_i}^{(j)}$.
+for $i = 1, \ldots, 85$, where $r_{t_i}$ is the discount rate for year $t_i$. Alternatively, we obtain an empirical distribution of the total damage over the period $2016-2100$ by considering
+\begin{linenomath*}
+  \[
+  d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})}
+  \]
+\end{linenomath*}
+and similarly for the cumulative damage.
 
-Now assume we are given a large sample of projections $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ for annual sea level rise anomalies compared to the 2015 value for the same time period.
-We model the relationship between the change in damage and change in sea level by a monotonic positive function $g(s| \boldsymbol \beta )$ with parameters $\boldsymbol \beta$, where $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$.
-We may further incorporate uncertainty in the shape of the function $g$ through the values of the parameters $\boldsymbol \beta$.
-An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and change in damage due to sea level rise is then given by
+\subsection{Effect of changes in sea level}
+
+We assume that changes in sea level have a multiplicative effect on the annual damage cost.
+That is, for a sea level anomaly $s_{t_i}$ in year $t_i > 2015$ compared to the 2015 level, the annual damage cost becomes
+\begin{linenomath*}
+  \[
+  g(s_{t_i} | \boldsymbol \beta ) d_{t_i},
+  \]
+\end{linenomath*}
+where $g( \cdot | \boldsymbol \beta )$ is a monotonic positive function with parameter vector  $\boldsymbol \beta$ such that $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$.
+\cite{Hallegatte&2013} estimate a similar effect function valid in 2050 for $s \in \{0,20,40\}$ for 136 coastal cities.
+Here, we use their results for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseilles, Naples, Porto, Rotterdam and Stockholm.
+To obtain a city-specific effect function for a large range of sea level anomalies we employ a linear extrapolation as shown in Figure~\ref{fig:EffectFct}.
+We then obtain a sample of effect functions $\{g(\cdot|\boldsymbol \beta^{(j)})\}_{j=1}^J$ by sampling with replacement from this ensemble of trajectories with all 15 ensemble members considered equally probable.
+
+\begin{figure}[!hbpt]
+\begin{center}
+\includegraphics[width=0.5\linewidth]{DecisionAnalysisBergenPartII.pdf}
+\caption{Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
+The median change and the corresponding extrapolation are indicated in red.}
+\label{fig:EffectFct}
+\end{center}
+\end{figure}
+
+Let further $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ denote a sample of projections for annual sea level anomalies compared to the 2015 value.
+An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and its effect on the damage is then given by
 \begin{linenomath*}
   \begin{equation}\label{eq:FutureDamage}
-    \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}.
+    \hat{F}_{\textup{d}, t_i}^\textup{s} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}.
   \end{equation}
@@
 The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures.
-In addition, we can incorporate an adaptation measure that protects against $K$ cm of increased sea level from year $t_k$ onwards.
+In addition, we can incorporate an adaptation measure of cost $C$ that protects against $K$ cm of increased sea level from year $t_k$ onward.
 This results in a damage distribution given by
@@
   \[
-  \hat{F}_{\textup{d}, t_i} (x) =
+  \hat{F}_{\textup{d}, t_i}^{\textup{s}, \textup{a}_k} (x) =
     \begin{cases}
-      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i < t_k \\
-      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i \geq t_k. \\
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i < t_k \\[2ex]
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} + C}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i = t_k \\[2ex]
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i > t_k. \\
     \end{cases}
@@
 
-\begin{comment}
-\cite{Fankhauser&1999} describe a deterministic framework where an adaptation investment of $C^0$ now (at time $n=0$) leads to unmitigated damage of $d_0^0$ in period $0$, and a stream of partially mitigated damages $d_t^0$ in periods $t=1,2,\ldots$.
-If $r$ is the discount rate, the net present value damage, $D^0$, associated with this investment is
-\begin{linenomath*}
-\begin{equation}\label{eq:deterministic damage}
-D^0 = C^0 + d_0^0 + \frac{d_1^0}{1+r} + \frac{d_2^0}{(1+r)^2} + \cdots  
-\end{equation}
-\end{linenomath*}
-In comparison, postponing the adaptation investment to time period $n=1$ would lead to unmitigated damages in periods $0$ and $1$, and partially mitigated damages, $d_t^1$, thereafter.
-The delay would be preferable if
-\begin{linenomath*}
-\[
-C^0 - \frac{C^1}{(1+r)} > (d_0^1 - d_0^0) + \frac{d_1^1 - d_1^0}{1+r} + \frac{d_2^1 - d_2^0}{(1+r)^2} + \cdots
-\]
-\end{linenomath*}
-Here, the expression on the left describes the benefits of the delay while the expression on the right describes the cost of the delay.
-In the simplest case, there is no change in investment costs ($C^0 = C^1 = C$) and the delay has no lasting effects beyond period 1 ($d_t^1 = d_t^0$ for $t > 1$).
-In this case, the comparison is between the expected return $r$ earned on the capital while implementation is delayed and one additional time period of unmitigated damage,
-\begin{linenomath*}
-  \[
-  r C > d_1^1 - d_1^0.
-  \]
-  \end{linenomath*}
-\end{comment}
+\subsection{Limitations of the decision framework}
+The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs.
+In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise.
+However, with only two data points, extrapolation approaces such as a power law or exponential growth seem unfeasible.
+Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level.
+Additionally, we have not specifically accounted for potential changes in storm surge patterns.
 
-\subsection{Limitations of the decision framework}
+\section{Danish decision framework}
 
@@
 
-\subsection{Data {\color{blue} (PG)}}
+\subsection{Data}
 The historical global mean temperature series is obtained from \citet{giss}.
@@
 The Bergen monthly series is missing data for 62 months, including all of the years 1942--43.
-To deal with occasional short stretches of missing data (most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages.
+To deal with occasional short stretches of missing data (at most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages.
 For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other Norwegian stations corrected by the average difference.
@@
 
-We estimate a damage distribution for Bergen in 2015 using reported annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
-The NPP data are aggregated to a county level, and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion.
-Also, some of this discussion should probably be moved to the section on data.}.
-In order to improve the parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland.
-Rogaland has a similar geography to Hordaland and a slightly smaller population with its main city, Stavanger, also located on the coast.
-The damage data prior to 2015 is adjusted to the 2015 level using the consumer price index for Norway.
-After the adjustment, we assume stationarity over the period and fit a single Burr distribution to the data, cf.
-\eqref{eq:Burr}.
+Annual damage costs for the Bergen case study are obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
+The NPP data are available for the period 1980-2015 and are aggregated to a county level.
+For improved parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland and with similar characteristics.
+We use a dicsount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf.
+Section 5.8 of \citet{DiscountRate}).
+
 \subsection{Sea level rise in Bergen and Esbjerg}
@@
 The regression slope is 1.30 (0.12).
+
 \begin{figure}[!hbpt]
 \begin{center}
-\includegraphics[width=\linewidth]{bergenfit.png}
-\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
+\includegraphics[width=0.75\linewidth]{bergenfit_edit.png}
+\caption{ The left figure shows raw (black) and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
 The straight line is the time series regression line.}
@@
 
-For modelling the relationship between the change in damage and change in sea level we use the results of \cite{Hallegatte&2013} for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseille, Naples, Porto, Rotterdam and Stockholm.
-\cite{Hallegatte&2013} estimate the mean annual flood damage in these cities in 2050 under no sea level change, 20 cm increase and 40 cm increase in sea level.
-We standardize their results for each city and consider the relative change for 20 and 40 cm increase.
-We then perform a linear extrapolation to obtain estimated relative changes in damage for a large interval of changes in sea level rise, see Figure~\ref{fig:ChangeInDamage}.
-Finally, the function $g$ is obtained by sampling from this ensemble of functions with all 15 ensemble members considered equally probable.
-
 \begin{figure}[!hbpt]
@@
 \caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5.
-The black vertical line is the median projection (with no uncertainty), while the grey histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
+The black vertical line is the median projection (with no uncertainty), while the gray histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
 }

### 2016-12-29 code/BergenDecisions/BergenDamageData.R

@@ -146,7 +146,7 @@ for(k in 1:85)
 }
 
 # save(orig.damage, damage.scenario, yearly.damage, constant.damage, adapted.damage, file="calculatedDamage.RData")
-# load("calculatedDamage.RData")
+#  load("calculatedDamage.RData")
 
 ### Cumulated additional damage due to sea level rise.
 add.damage <- yearly.damage - constant.damage
@@ -288,18 +288,22 @@ total.damage <- apply(yearly.damage, 1, sum)
 #######################################################################
 
 ## Plots showing the underlying data used in the adaptation decision analysis
-pdf(file="../../submission/DecisionAnalysisBergen.pdf", width=10, height=5, points=12)
+pdf(file="../../submission/DecisionAnalysisBergenPartI.pdf", width=5, height=5, points=12)
 ### (a) Plot of damage data and fitted distribution
-par(mex=0.75, mfrow=c(1,2))
-hist(sf.damage, main="(a)", xlab="Amount (million NOK)", col="gray60", breaks=50, xlim=c(0,65),
+## par(mex=0.75, mfrow=c(1,2))
+par(mex=0.75, mar=c(5,4,2,2)+0.1)
+hist(sf.damage, main="", xlab="Amount (million NOK)", col="gray60", breaks=50,  xlim=c(0,65),
      freq=FALSE)
 x <- seq(0, 65, by=0.01)
 y <- dburr(x, shape1=optim.out$par[1], shape2=optim.out$par[2], rate=optim.out$par[3])
 lines(x,y,col="red",lwd=2)
+dev.off()
+pdf(file="../../submission/DecisionAnalysisBergenPartII.pdf", width=5, height=5, points=12)
+par(mex=0.75, mar=c(5,4,2,2)+0.1)
 ### (b) Plot of extrapolated damage change profiles 
 x <- c(0,20,40)
 plot(x, E[1, 2:4], type="o", col="black", xlim=c(-100,130), ylim=c(-180,1000),
-     xlab="Sea level anomaly (cm)", ylab="Relative mean annual damage", main="(b)", axes=FALSE)
+     xlab="Sea level anomaly (cm)", ylab="Relative mean annual damage", main="", axes=FALSE)
 box()
 axis(1)
 axis(2, at=c(-200, 1, 200, 400, 600, 800, 1000), labels=c(0.005, 1, 200, 400, 600, 800, 1000))

### 2016-12-29 submission/paper.tex
more results on Bergen

@@
 
-We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark.
+\begin{figure}[!hbpt]
+\begin{center}
+  \includegraphics[width=0.45\linewidth]{BergenMap.png}
+  \includegraphics[width=0.45\linewidth]{EsbjergMap.png}
+\caption{Terrain maps of central Bergen, Norway (left) and Esbjerg, Denmark (right).}
+\label{fig:CityMaps}
+\end{center}
+\end{figure}
+
+We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark, see Figure~\ref{fig:CityMaps}.
 The Norwegian city of Bergen is the capital of Hordaland County.
@@
 
-We project local sea level changes by Modeling two processes:
-\begin{enumerate}
-\item The relationship between global temperature and global sea level;
-\item The relationship between global sea level and local sea level.
-\end{enumerate}
+We project local sea level changes by Modeling two processes, the relationship between global temperature and global sea level, and the relationship between global sea level and local sea level.
 
@@
 To assess the optimal timing of an adaptation measure, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options.
-The probabilistic distribution is constructed by considering uncertainty
-\begin{itemize}
-\item in the local sea level projections;
-\item in the annual damage costs;
-\item in the effect of changes in sea level on the annual damage costs. 
-\end{itemize}
+The probabilistic distribution is constructed by considering uncertainty in the local sea level projections, in the annual damage costs, and in the effect of changes in sea level on the annual damage costs.
 
@@
 Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level.
+Our framework also simplifies the cost and effect of an adaptation option during construction in that we assume no effect until the construction is finished with all the construction cost falling in the last year of the construction.
+Especially for larger constructions, these assumptions might need to be modified.
 Additionally, we have not specifically accounted for potential changes in storm surge patterns.
@@
 
-\subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
+\subsection{Timing of adaptation measures in Bergen}
 
-The histogram of the observed annual damages and the estimated distribution are given in Figure~\ref{fig:BergenDamageDist}.
-The parameter estimates for the Burr distribution are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
+\begin{figure}[!hbpt]
+\begin{center}
+\includegraphics[width=0.5\linewidth]{DecisionAnalysisBergenPartI.pdf}
+\caption{Estimated distribution of annual damage costs in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland counties 1980-2015 (gray bars).}
+\label{fig:BergenDamageDist}
+\end{center}
+\end{figure}
 
-\begin{figure}
+Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution.
+The parameter estimates are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
+\citet{bergenreport} discuss several different adaptation options for Bergen.
+In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen.
+The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level).
+
+\begin{figure}[!hbpt]
 \begin{center}
-\includegraphics[width=\linewidth]{DecisionAnalysisBergen.pdf}
-\caption{Damage data used in the decision analysis for Bergen: (a) Estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars); (b) Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
-The median change and the corresponding extrapolation are indicated in red.}
-\label{fig:BergenDamageDist}
+\includegraphics[width=\linewidth]{TotalDamageCostsAdaptation.pdf}
+\caption{Projected total damage costs in Bergen for the time period 2016-2100 as a function of the timing of an adaptation measure consisting of the construction of two inner barriers.
+The median projection under each adaptation scenario is indicated in red with gray bars denoting the 90\% projection intervals.
+The median projected total damage cost under no action is shown with a black line with the corresponding 90\% projection interval indicated by dotted lines.
+}
+\label{fig:TotalDamageBergen}
 \end{center}
@@
 \includegraphics[width=0.5\linewidth]{CumDamageCostsBergen.pdf}
-\caption{Distribution of accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 
+\caption{Median projected cumulative damage costs in Bergen under constant sea level (gray line), under sea level rise according to RCP 8.5 with no adaptation action (black line) and with the construction of two inner barriers in 2047 (red line).
+The shaded gray area denotes the 90\% projection interval under constant sea level.
+Dotted lines indicate the 90\% projection intervals with sea level rise according to RCP 8.5.
+}
 \label{fig:NoAction}

### 2016-12-29 code/BergenDecisions/BergenDamageData.R

@@ -427,6 +427,17 @@ legend("topright",
        col=c("black", "#7D26CD", "orange", "#008B45", "gray50"), lty=1, lwd=2)
 dev.off()
 
+### Map of Bergen
+library(RgoogleMaps)
+MyMap = GetMap(center=c(60.3913, 5.3221), zoom=13, destfile="../../submission/BergenMap.png", size=c(500,500), maptype="terrain", GRAYSCALE=TRUE)
+MyMap2 = GetMap(center=c(55.4765, 8.4594), zoom=10, destfile="../../submission/EsbjergMap.png", size=c(500,500), maptype="terrain", GRAYSCALE=TRUE)
+
+
+## mymarkers <- cbind.data.frame(lat=c(60.383723, 60.400241), lon=c(5.316847, 5.312512))
+## pdf(file="../../submission/BergenMap.pdf", width=5, height=5, points=12)
+## tmp = PlotOnStaticMap(MyMap, lat=mymarkers[,"lat"], lon=mymarkers[,"lon"], cex=2.5, pch=20, col="red", add=T) 
+## dev.off()
+
 #######################################################################
 ### Other plots 
 #######################################################################

### 2016-12-29 submission/paper.tex
input from Martin and Karianne

@@
 \section{Introduction}\label{sec:intro}
-Long-term planning and decision-making regarding societal infrastructure along coastal areas must account for a changing climate.
-In particular, potential changes in local sea level can yield numerous effects such as frequent flooding, inundation and back flow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats.
-Considerable challenges continue to exist in understanding  the impacts of a combination of sea level rise and storm surges and the implementation of potential adaptation options to reduce these impacts.
-Severe inherent uncertainties persist along all parts of the processing chain from climate projections to adaptation assessments, and it is critical that decision-making appropriately accounts for this.
+The potential impact of climate change on local sea level, yielding effects such as frequent flooding, inundation and backflow of storm drainage and sewer systems, destructive erosion and contamination of wetlands and other habitats, requires city planners to make decisions in the presence of substantial uncertainty.
 
@@
 
-In this paper we combine the uncertainty of sea level projections with a simplified decision framework under uncertainty where planners want to know how early they should implement costly adaptation measures.
-We discuss the advantages of using a fully probabilistic approach in a light touch decision framework that accounts for uncertainty both in climate projections and damage assessments, where information on uncertainty is turned into a driver of change rather than a barrier.
+In this paper we employ light touch decision tools to demonstrate the importance of combining projections of sea level rise and flood damages alongside a detailed quantification of both hydrologic and economic uncertainties in the context of real-life decision-problems experienced by stakeholders and authorities in two northern European cities, Bergen in Norway and Esbjerg in Denmark, see Figure~\ref{fig:CityMaps}.
+Based on communications with local end-users we highlight the value of taking into account uncertainty through two simplified and complementary case studies, where in the first one  planners want to know how early they should implement costly adaptation measures, whereas in the second case the aim is to highlight the risk of flooding in coastal areas, e.g.
+in order to prioritize future adaptation actions and investments.
+In both cases we show that embracing the uncertainties derived from economic and hydrologic models is absolutely crucial in order to answer the question of "are we sure we want to do this?".
 
@@
 
-We illustrate the sea level rise projections and timing of adaptation measures for two case studies in Northern Europe, namely the city of Bergen in Norway and Esbjerg in Denmark, see Figure~\ref{fig:CityMaps}.
 The Norwegian city of Bergen is the capital of Hordaland County.
@@
 It has the largest port in Norway, both in terms of freight and passengers.
-The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UN-ESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}.
+The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UNESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}.
 Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become a major problem in other parts of Bergen as well \citep{bergenreport}.
@@
 
-Esbjerg, on the west coast of Jutland, has the second largest harbor in Denmark.
-It can be subjected to substantial storm surges, which, in conjunction with heavy rainfall, can flood substantial parts of the town.
+Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region.
+The city hosts one of the largest harbours  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas.
+As a result critical infrastructures and commercial buildings figure prominently in the coastal zone.
+Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbour and the city.
+The highest ever since 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached a record-high 433 cm above the norm, causing massive economic losses.
+More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.
+
+As in the case of Bergen, sea level rise caused by climate change is expected to compound these risks, alongside parallel threats caused by increased risks of pluvial flooding and rising ground water levels in Esbjerg.
+The municipality recently adopted its climate adaptation plan, which in its first phase is aimed at identifying present and future flood-prone areas, e.g.
+to avoid urban development into such areas, to limit damages to buildings of high societal or cultural value, and to pave the way for implementing cost-effective adaptation measures in the second phase of the plan.
 
@@
 Section \ref{sealevelproj} describes our approach to projecting sea level.
-In section \ref{decision_tools_PartI} we describe the decision framework focused on the timing of proactive adaptation decision.
-We apply this framework to sea level projections and adaptation for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
+In sections \ref{decision_tools_PartI} and \ref{decision_tools_PartII} we describe the type of decision problems that we are going to attack.
+We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
 In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.
@@
 
-To assess the optimal timing of an adaptation measure, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options.
+There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes adaptation is moving from theory to practice and practitioners try to deal with how to begin adapting.
+This leads to an increasing need and interest in the appraisal of options.
+
+Several economic decision support tools and methods exist for adaptation assessment under uncertainty.
+Robust decision-making approaches are able to better incorporate uncertainty and a broad range of climate scenarios to capture as much of the uncertainty on future climates as possible (Dittrich et al.
+2016).
+These approaches can be classified according to a science-first or policy-first approach.
+The former has a ‘predict-then-act’ foundation, which starts with for example probabilistic climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
+The latter starts out with the formulated adaptation plans and not impacts and their functioning is tested against different future projections (Dittrich et al.
+2016).
+
+We take a policy-first approach, in which we test a current adaptation plan against different sea level and damage projections and the inclusion of different sources of uncertainty.
+We focus on what this implies for the timing of adaptation measures and the implications of including uncertainty.
+In particular, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options.
 The probabilistic distribution is constructed by considering uncertainty in the local sea level projections, in the annual damage costs, and in the effect of changes in sea level on the annual damage costs.
@@
 However, with only two data points, extrapolation approaces such as a power law or exponential growth seem unfeasible.
+
 Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level.
-Our framework also simplifies the cost and effect of an adaptation option during construction in that we assume no effect until the construction is finished with all the construction cost falling in the last year of the construction.
+The elements of such a framework might include an appropriate social discount rate, valuing environmental goods in monetary terms, incorporate socio-economic assumptions and long-term policy goals of decision makers, as well as that climate change is often not the only driver that decision makers should consider, therefore costs and benefits should be studied in a wider context (Dittrich et al, 2016).
+
+Our framework simplifies the cost and effect of an adaptation option during construction in that we assume no effect until the construction is finished with all the construction cost falling in the last year of the construction.
 Especially for larger constructions, these assumptions might need to be modified.
@@
 \section{Danish decision framework}
+\label{decision_tools_PartI}
 
@@
 Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution.
-The parameter estimates are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
+The parameter estimates are $\hat{\alpha} = 7.84$, $\hat{\gamma} = 0.40$ and $\hat{\theta} = 0.007$.
 \citet{bergenreport} discuss several different adaptation options for Bergen.
@@
 
-\subsection{Selection of adaptation measures(?) {\color{blue} (MD)}}
+\subsection{Identifying flood-prone areas}
 
-A case study focusing on Denmark.
+In 2014 the municipality of Esbjerg adopted its climate adaptation plan {\color{blue}(ref)}, which will cover the period from 2014-2026, and which aims to reduce the risk of flooding caused by storm surges, heavy rainfall and rising ground water levels, respectively.
+In the following and based on interviews with end-users from the municipality we will focus on the harbour of Esbjerg and the nearby coastal areas, and we will only consider the hazard of coastal flooding, though evidently in some areas the flood risk is compounded.
+
+The initial scoping of the climate adaptation plan in Esbjerg was completed in 2014 and includes a preliminary value and risk mapping, considering critical infrastructure and buildings of high cultural and societal value as identified by the municipality, while informed by spatial floods maps for different scenarios corresponding to each of the three different kinds of hydrological events (sea level rise/storm surges, pluvial flooding and rising ground water levels).
+In terms of coastal floods the mapping considers only one type of storm surge, corresponding to a 20-year return event (based on historical storm surge statistics), and increased sea level due to climate change was not accounted for.
+
+The risk for any given map area as the probability of, e.g.
+a certain flood depth, is derived from a hydrological flood model (not including urban drainage system) times a valuation of the consequences, which - similar to the Bergen case - is essentially a loss function associating the flood depth with a measure of economic cost.
 
@@
 \label{unc}
+
+\subsection{Sea level projections}
 
@@
 \end{figure}
+
+\subsection{Damage costs}
+
+\subsection{...}
+
+\section{Conclusions and discussion}
+
+Important to continue an iterative and interactive with end-users, as the decision framework provides actionable information to decision-makers, who will then make their own decisions.
+These decision are then incorporated into the current adaptation strategy.
+Continued loop to identify potential vulnerabilities of the approaches across a wide-range of possible futures.
+
+Take home message
+
+\begin{itemize}
+\item It is possible to account for uncertainty even if you do a light touch decision-making approach -- Explain what it implies if you don't account and if you do
+\item Proof of concept -- connect in a probabilistic way (normally done in a discrete way) -- consider uncertainty in both ways
+\item Improve the accessibility and demonstrating the general applicability of policy first approaches -- towards more generic toolkit
+\end{itemize}
 

### 2016-12-29 submission/paperMD.tex
input from Martin and Karianne

--- 2016-12-29 submission/paper.tex
@@
 
-\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, and K. de Bruin\affil{3,4}}
+\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, K. de Bruin\affil{3,4}, and P. S. Kaspersen\affil{2}}
 
 \affiliation{1}{Norwegian Computing Center, Oslo, Norway}
-\affiliation{2}{Technical University of Denmark, Copenhagen, Denmark}
+\affiliation{2}{Technical University of Denmark, Kgs. Lyngby, Denmark}
 \affiliation{3}{Center for International Climate and Environmental Research, Oslo, Norway}
@@
 \begin{keypoints}
-\item Decisions on adaptation measures need to take careful account of uncertainties
+\item Decisions on adaptation measures need to take careful account of uncertanties
 \item More careful models are needed for increased costs of the effects of sea level rise
-\item Modeling local sea level rise is essential
+\item Modelling local sea level rise is essential
 \end{keypoints}
@@
 Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented.
-All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action.
+All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, in the cost of no action.
 We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark.
 Different components of uncertainty are visualized.
-We show that failing to take uncertainty into account can increase the projected damage costs by an order of magnitude.
+{\color{blue} Findings go here}
 \end{abstract}
@@
 
-As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012} adaptive planning methods need to support decisions in the short term, while considering long-term developments.
-Challenges of adaptation decision-making under uncertainty relate to the incorporation of spatial, inter-temporal and flexibility aspects of adaptation priorities \citep{FankhauserSoare2013}, and the linkage with specific characteristics of sectors and contexts \citep{BisaroSwartHinkel2016, HinkelBisaro2016}.
-Several economic decision support tools and methods exist for adaptation assessment under uncertainty \citep[e.g.][]{Chambwera&2014, WilbyDessai2010, WalkerHaasnootKwakkel2013}.
-However, \citet{Watkiss&2015} conclude that these tools are very resource intensive and complex in the context of long-term adaptation investment decisions and call for the development of ``light touch'' approaches to better support local adaptation making.
-
-In this paper we employ light touch decision tools to demonstrate the importance of combining projections of sea level rise and flood damages alongside a detailed quantification of both hydrologic and economic uncertainties in the context of real-life decision-problems experienced by stakeholders and authorities in two northern European cities, Bergen in Norway and Esbjerg in Denmark, see Figure~\ref{fig:CityMaps}.
+In this paper we demonstrate the importance of combining projections of sea level rise and flood damages alongside a detailed quantification of both hydrologic and economic uncertainties in the context of real-life decision-problems experienced by stakeholders and authorities in Bergen and Esbjerg.
 Based on communications with local end-users we highlight the value of taking into account uncertainty through two simplified and complementary case studies, where in the first one  planners want to know how early they should implement costly adaptation measures, whereas in the second case the aim is to highlight the risk of flooding in coastal areas, e.g.
@@
 
-\begin{figure}[!hbpt]
-\begin{center}
-  \includegraphics[width=0.45\linewidth]{BergenMap.png}
-  \includegraphics[width=0.45\linewidth]{EsbjergMap.png}
-\caption{Terrain maps of central Bergen, Norway (left) and Esbjerg, Denmark (right).}
-\label{fig:CityMaps}
-\end{center}
-\end{figure}
-
 The Norwegian city of Bergen is the capital of Hordaland County.
@@
 It has the largest port in Norway, both in terms of freight and passengers.
-The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UNESCO World Heritage site\footnote{See \url{http://whc.unesco.org/en/list/59}.}.
+The historic harbor area, Bryggen, is the only Hanseatic trade center remaining in its original style, and has been declared a UNESCO World Heritage site (\url{http://whc.unesco.org/en/list/59}).
 Bryggen is regularly flooded at extreme tides, and it is feared that as sea levels rise, floods will become a major problem in other parts of Bergen as well \citep{bergenreport}.
-
-The municipality of Bergen has, in cooperation with private actors, analyzed several possible adaptation measures against sea level rise.
-The measures range from an outer barrier that would protect the entire metropolitan area to various protection measures of limited areas in the inner harbor \citep{bergenreport}.
-While the viability of the constructions and the associated construction costs have been carefully analyzed, the optimal timing of potential adaptation measures and the effects of the associated uncertainties have yet to be investigated.
-We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs and the effects of sea level rise on changes in damage costs.
 
@@
 More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.
-
 As in the case of Bergen, sea level rise caused by climate change is expected to compound these risks, alongside parallel threats caused by increased risks of pluvial flooding and rising ground water levels in Esbjerg.
@@
 
-The remainder of the paper is organized as follows.
 Section \ref{sealevelproj} describes our approach to projecting sea level.
-In sections \ref{decision_tools_PartI} and \ref{decision_tools_PartII} we describe the type of decision problems that we are going to attack.
-We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
-In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.
+In section \ref{decision_tools} we describe the type of decision problem that we are going to attack.
+We apply these tools to  sea level projections for Bergen, Norway, and for Esbjerg, Denmark
+in section \ref{cases}.  In the final section \ref{unc}  we demonstrate the consequences of ignoring the uncertainty in the projections.
 
-\section{Sea level projections}
+\section{Sea level projections {\color{blue} (PG)}}
+
 \label{sealevelproj}
-
-We project local sea level changes by Modeling two processes, the relationship between global temperature and global sea level, and the relationship between global sea level and local sea level.
 
@@
 For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
-\begin{linenomath*}
-\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
-\end{linenomath*}
+
+\[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t}\]
 where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.
@@
 \subsection{Local sea level}
-In order to get from global sea level projections to local ones, it is important to note that sea level rise is not uniform over the globe.
+In order to get from global sea level projections to local ones, it is important to note that sea level rise not is uniform over the globe.
 Glacial and land ice melting affect the local sea level differently depending on where the melted ice is located.
@@
 The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
-\begin{linenomath*}
-\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
-\end{linenomath*}
+
+\[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t}\]
+
 where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.
@@
 
-\section{Timing of adaptation measures}
-\label{decision_tools_PartI}
+\section{Decision tools {\color{blue} (KdB, MD, TT)}}
+\label{decision_tools}
 
-There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes adaptation is moving from theory to practice and practitioners try to deal with how to begin adapting.
-This leads to an increasing need and interest in the appraisal of options.
+\subsection{Timing of adaptation measures}
 
-Several economic decision support tools and methods exist for adaptation assessment under uncertainty.
-Robust decision-making approaches are able to better incorporate uncertainty and a broad range of climate scenarios to capture as much of the uncertainty on future climates as possible (Dittrich et al.
-2016).
-These approaches can be classified according to a science-first or policy-first approach.
-The former has a ‘predict-then-act’ foundation, which starts with for example probabilistic climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
-The latter starts out with the formulated adaptation plans and not impacts and their functioning is tested against different future projections (Dittrich et al.
-2016).
+We consider adaptation decision making related to the timing of proactive adaptation measures.
+That is, the goal is to adapt to sea level rise before major damages occur.
+In a cost-benefit framework, an investment should be delayed as long as the benefits of delay (avoided investment costs) are greater than the associated costs (higher climate change damages) \citep{Fankhauser&1999}.
 
-We take a policy-first approach, in which we test a current adaptation plan against different sea level and damage projections and the inclusion of different sources of uncertainty.
-We focus on what this implies for the timing of adaptation measures and the implications of including uncertainty.
-In particular, we employ a probabilistic extension of the framework described by \citep{Fankhauser&1999} in which we obtain a probabilistic distribution for the net present value damage in a given year for various adaptation options.
-The probabilistic distribution is constructed by considering uncertainty in the local sea level projections, in the annual damage costs, and in the effect of changes in sea level on the annual damage costs.
-
-\subsection{Annual damage costs}
-
-We model the distribution of annual damage, $F_{\textup{d}, t_0}$, for the year $t_0 = 2015$ by the three parameter Burr distribution \citep{Burr1942} with density
+We assume that damages related to sea level rise coincide with damages due to storm surges and that the accumulated annual damages are independent between years.
+More specifically, we model the distribution of annual damage, $F_{\textup{d}}$, by the three parameter Burr distribution \citep{Burr1942} with density
 \begin{linenomath*}
   \begin{equation}\label{eq:Burr}
-  f_{\textup{d}, t_0} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
+  f_{\textup{d}} (x) = \frac{\alpha \gamma (x/ \theta)^{\gamma}}{x [ 1 + (x/ \theta)^{\gamma}]^{\alpha + 1}}
   \end{equation}
@@
 \cite{Klugman&2012}.
-The parameters of the distribution are estimated using historical data for annual storm surge damage.
-Data prior to 2015 are adjusted to the 2015 level using the consumer price index.
-After adjustment, we assume stationarity over the period and independence between years.
 
 Under a constant sea level, we can obtain a sample trajectory $\{d_{t_1}, d_{t_2}, \ldots, d_{t_{85}}\}$ of future annual damages for $t_1 = 2016, \ldots, t_{85} = 2100$ by drawing $85$ i.i.d.
-values from the estimated distribution $\hat{F}_{\textup{d}, t_0}$.
-By repeating this process $J$ times, we obtain an empirical damage distribution for each future year $t_i$ given by
+values from the estimated damage distribution for $t_0 = 2015$, $\hat{F}_{\textup{d}, t_0}$.
+By repeating this process $J$ times, we obtain an empirial damage distribution for each future year $t_i$ given by
 \begin{linenomath*}
   \[
-  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l}) } \leq x \Bigg\}
+  \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{d_{t_i}^{(j)} \leq x \}
   \]
 \end{linenomath*}
-for $i = 1, \ldots, 85$, where $r_{t_i}$ is the discount rate for year $t_i$. Alternatively, we obtain an empirical distribution of the total damage over the period $2016-2100$ by considering
-\begin{linenomath*}
-  \[
-  d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} \frac{d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})}
-  \]
-\end{linenomath*}
-and similarly for the cumulative damage.
+for $i = 1, \ldots, 85$. Similarly, we obtain an empirial distribution of the total damage over the period $2016-2100$ by considering $d_{\textup{total}}^{(j)} = \sum_{i=1}^{85} d_{t_i}^{(j)}$.
 
-\subsection{Effect of changes in sea level}
-
-We assume that changes in sea level have a multiplicative effect on the annual damage cost.
-That is, for a sea level anomaly $s_{t_i}$ in year $t_i > 2015$ compared to the 2015 level, the annual damage cost becomes
-\begin{linenomath*}
-  \[
-  g(s_{t_i} | \boldsymbol \beta ) d_{t_i},
-  \]
-\end{linenomath*}
-where $g( \cdot | \boldsymbol \beta )$ is a monotonic positive function with parameter vector  $\boldsymbol \beta$ such that $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$.
-\cite{Hallegatte&2013} estimate a similar effect function valid in 2050 for $s \in \{0,20,40\}$ for 136 coastal cities.
-Here, we use their results for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseilles, Naples, Porto, Rotterdam and Stockholm.
-To obtain a city-specific effect function for a large range of sea level anomalies we employ a linear extrapolation as shown in Figure~\ref{fig:EffectFct}.
-We then obtain a sample of effect functions $\{g(\cdot|\boldsymbol \beta^{(j)})\}_{j=1}^J$ by sampling with replacement from this ensemble of trajectories with all 15 ensemble members considered equally probable.
-
-\begin{figure}[!hbpt]
-\begin{center}
-\includegraphics[width=0.5\linewidth]{DecisionAnalysisBergenPartII.pdf}
-\caption{Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
-The median change and the corresponding extrapolation are indicated in red.}
-\label{fig:EffectFct}
-\end{center}
-\end{figure}
-
-Let further $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ denote a sample of projections for annual sea level anomalies compared to the 2015 value.
-An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and its effect on the damage is then given by
+Now assume we are given a large sample of projections $\{s_{t_1}^{(j)}, \ldots, s_{t_{85}}^{(j)} \}_{j=1}^J$ for annual sea level rise anomalies compared to the 2015 value for the same time period.
+We model the relationship between the change in damage and change in sea level by a monotonic positive function $g(s| \boldsymbol \beta )$ with parameters $\boldsymbol \beta$, where $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$.
+We may further incorporate uncertainty in the shape of the function $g$ through the values of the parameters $\boldsymbol \beta$.
+An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and change in damage due to sea level rise is then given by
 \begin{linenomath*}
   \begin{equation}\label{eq:FutureDamage}
-    \hat{F}_{\textup{d}, t_i}^\textup{s} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}.
+    \hat{F}_{\textup{d}, t_i} (x) = \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}.
   \end{equation}
@@
 The distribution in \eqref{eq:FutureDamage} describes the projected damage distribution with no adaptation measures.
-In addition, we can incorporate an adaptation measure of cost $C$ that protects against $K$ cm of increased sea level from year $t_k$ onward.
+In addition, we can incorporate an adaptation measure that protects against $K$ cm of increased sea level from year $t_k$ onwards.
 This results in a damage distribution given by
@@
   \[
-  \hat{F}_{\textup{d}, t_i}^{\textup{s}, \textup{a}_k} (x) =
+  \hat{F}_{\textup{d}, t_i} (x) =
     \begin{cases}
-      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i < t_k \\[2ex]
-      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} + C}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i = t_k \\[2ex]
-      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\Bigg\{ \frac{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)}}{\prod_{l \leq i} (1 + r_{t_l})} \leq x \Bigg\}, & t_i > t_k. \\
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)}| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i < t_k \\
+      \frac{1}{J} \sum_{j=1}^J \mathbbm{1}\{g(s_{t_i}^{(j)} - K| \boldsymbol \beta^{(j)}) d_{t_i}^{(j)} \leq x \}, & t_i \geq t_k. \\
     \end{cases}
@@
 
+\begin{comment}
+\cite{Fankhauser&1999} describe a deterministic framework where an adaptation investment of $C^0$ now (at time $n=0$) leads to unmitigated damage of $d_0^0$ in period $0$, and a stream of partially mitigated damages $d_t^0$ in periods $t=1,2,\ldots$.
+If $r$ is the discount rate, the net present value damage, $D^0$, associated with this investment is
+\begin{linenomath*}
+\begin{equation}\label{eq:deterministic damage}
+D^0 = C^0 + d_0^0 + \frac{d_1^0}{1+r} + \frac{d_2^0}{(1+r)^2} + \cdots  
+\end{equation}
+\end{linenomath*}
+In comparison, postponing the adaptation investment to time period $n=1$ would lead to unmitigated damages in periods $0$ and $1$, and partially mitigated damages, $d_t^1$, thereafter.
+The delay would be preferable if
+\begin{linenomath*}
+\[
+C^0 - \frac{C^1}{(1+r)} > (d_0^1 - d_0^0) + \frac{d_1^1 - d_1^0}{1+r} + \frac{d_2^1 - d_2^0}{(1+r)^2} + \cdots
+\]
+\end{linenomath*}
+Here, the expression on the left describes the benefits of the delay while the expression on the right describes the cost of the delay.
+In the simplest case, there is no change in investment costs ($C^0 = C^1 = C$) and the delay has no lasting effects beyond period 1 ($d_t^1 = d_t^0$ for $t > 1$).
+In this case, the comparison is between the expected return $r$ earned on the capital while implementation is delayed and one additional time period of unmitigated damage,
+\begin{linenomath*}
+  \[
+  r C > d_1^1 - d_1^0.
+  \]
+  \end{linenomath*}
+\end{comment}
+
 \subsection{Limitations of the decision framework}
-The main limitation of this light touch decision framework is that we have significantly simplified the assessment of the effect of sea level rise on the damage costs.
-In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise.
-However, with only two data points, extrapolation approaces such as a power law or exponential growth seem unfeasible.
-
-Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level.
-The elements of such a framework might include an appropriate social discount rate, valuing environmental goods in monetary terms, incorporate socio-economic assumptions and long-term policy goals of decision makers, as well as that climate change is often not the only driver that decision makers should consider, therefore costs and benefits should be studied in a wider context (Dittrich et al, 2016).
-
-Our framework simplifies the cost and effect of an adaptation option during construction in that we assume no effect until the construction is finished with all the construction cost falling in the last year of the construction.
-Especially for larger constructions, these assumptions might need to be modified.
-Additionally, we have not specifically accounted for potential changes in storm surge patterns.
-
-\section{Danish decision framework}
-\label{decision_tools_PartI}
 
@@
 
-\subsection{Data}
+\subsection{Data {\color{blue} (PG)}}
 The historical global mean temperature series is obtained from \citet{giss}.
@@
 The Bergen monthly series is missing data for 62 months, including all of the years 1942--43.
-To deal with occasional short stretches of missing data (at most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages.
+To deal with occasional short stretches of missing data (most one or two months) we use median polish replacement \citep{medpol} and then compute annual averages.
 For the years 1942-43, we use use the average difference between Bergen and the average of all other Norwegian stations in 1940 and 1943 to estimate values for 1941 and 1942, using the average of all other Norwegian stations corrected by the average difference.
@@
 
-Annual damage costs for the Bergen case study are obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
-The NPP data are available for the period 1980-2015 and are aggregated to a county level.
-For improved parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland and with similar characteristics.
-We use a dicsount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf.
-Section 5.8 of \citet{DiscountRate}).
-
+We estimate a damage distribution for Bergen in 2015 using reported annual data on storm surge damage from Hordaland and Rogaland counties over the period 1980-2015 obtained from the Norwegian Natural Perils Pool (NPP;  data are available at \url{https://www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/}).
+The NPP data are aggregated to a county level, and while only about 55\% of the total population of Hordaland lives in Bergen, we do not correct for this as a substantial part of the remaining population lives away from the coast\footnote{TLT: Need to improve this discussion.
+Also, some of this discussion should probably be moved to the section on data.}.
+In order to improve the parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland.
+Rogaland has a similar geography to Hordaland and a slightly smaller population with its main city, Stavanger, also located on the coast.
+The damage data prior to 2015 is adjusted to the 2015 level using the consumer price index for Norway.
+After the adjustment, we assume stationarity over the period and fit a single Burr distribution to the data, cf.
+\eqref{eq:Burr}.
 \subsection{Sea level rise in Bergen and Esbjerg}
@@
 The regression slope is 1.30 (0.12).
-
 \begin{figure}[!hbpt]
 \begin{center}
-\includegraphics[width=0.75\linewidth]{bergenfit_edit.png}
-\caption{ The left figure shows raw (black) and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
+\includegraphics[width=\linewidth]{bergenfit.png}
+\caption{ The left figure shows raw (black)and gia-corrected (red) sea level data from Bergen, The right figure relates the gia-corrected Bergen sea level to the global sea level series of \citet{csiro}.
 The straight line is the time series regression line.}
@@
 
-\subsection{Timing of adaptation measures in Bergen}
+\subsection{Timing of adaptation measures {\color{blue} (KdB, TT)}}
 
-\begin{figure}[!hbpt]
+The histogram of the observed annual damages and the estimated distribution are given in Figure~\ref{fig:BergenDamageDist}.
+The parameter estimates for the Burr distribution are $\hat{\alpha} = 1.27$, $\hat{\gamma} = 0.51$ and $\hat{\theta} = 0.002$.
+
+\begin{figure}
 \begin{center}
-\includegraphics[width=0.5\linewidth]{DecisionAnalysisBergenPartI.pdf}
-\caption{Estimated distribution of annual damage costs in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland counties 1980-2015 (gray bars).}
+\includegraphics[width=0.5\linewidth]{DamageDistribution.pdf}
+\caption{ The estimated distribution of annual storm surge damage in Bergen for 2015 (red) based on observed annual damage in Hordaland and Rogaland 1980-2015 (gray bars). }
 \label{fig:BergenDamageDist}
@@
 
-Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution.
-The parameter estimates are $\hat{\alpha} = 7.84$, $\hat{\gamma} = 0.40$ and $\hat{\theta} = 0.007$.
-\citet{bergenreport} discuss several different adaptation options for Bergen.
-In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen.
-The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level).
+For modelling the relationship between the change in damage and change in sea level we use the results of \cite{Hallegatte&2013} for 15 European cites: Amsterdam, Athens, Barcelona, Dublin, Glasgow, Hamburg, Helsinki, Copenhagen, Lisbon, London, Marseille, Naples, Porto, Rotterdam and Stockholm.
+\cite{Hallegatte&2013} estimate the mean annual flood damage in these cities in 2050 under no sea level change, 20 cm increase and 40 cm increase in sea level.
+We standardize their results for each city and consider the relative change for 20 and 40 cm increase.
+We then perform a linear extrapolation to obtain estimated relative changes in damage for a large interval of changes in sea level rise, see Figure~\ref{fig:ChangeInDamage}.
+Finally, the function $g$ is obtained by sampling from this ensemble of functions with all 15 ensemble members considered equally probable.
 
@@
 \begin{center}
-\includegraphics[width=\linewidth]{TotalDamageCostsAdaptation.pdf}
-\caption{Projected total damage costs in Bergen for the time period 2016-2100 as a function of the timing of an adaptation measure consisting of the construction of two inner barriers.
-The median projection under each adaptation scenario is indicated in red with gray bars denoting the 90\% projection intervals.
-The median projected total damage cost under no action is shown with a black line with the corresponding 90\% projection interval indicated by dotted lines.
-}
-\label{fig:TotalDamageBergen}
+\includegraphics[width=0.5\linewidth]{EuropeanIncreaseLossExtrapolationUncertainty.pdf}
+\caption{ Relative change in mean annual damage as a function of sea level rise for 15 European cites as estimated by \cite{Hallegatte&2013} (black circles) with linearly extrapolated values indicated by gray lines.
+The median change and the corresponding extrapolation are indicated in red.}
+\label{fig:ChangeInDamage}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{CumDamageCostsBergen.pdf}
-\caption{Median projected cumulative damage costs in Bergen under constant sea level (gray line), under sea level rise according to RCP 8.5 with no adaptation action (black line) and with the construction of two inner barriers in 2047 (red line).
-The shaded gray area denotes the 90\% projection interval under constant sea level.
-Dotted lines indicate the 90\% projection intervals with sea level rise according to RCP 8.5.
-}
+\includegraphics[width=0.5\linewidth]{AccumulatedFutureLoss.png}
+\caption{Distribution of log  accumulated future loss 2016-2100 under RCP 8.5 and assuming that no adaptation measures are implemented.} 
 \label{fig:NoAction}
@@
 
-\subsection{Identifying flood-prone areas}
+\subsection{Identifying flood-prone areas {\color{blue} (MD)}}
 
@@
 
-\subsection{Sea level projections}
-
 In many cases sea level rise projections are given as a single number for each scenario, usually the mean or median of the ensemble of projections from different climate models (e.g.
@@
 from Even Nilsson, Norwegian Mapping Authority).
-In our analysis there are two more sources of uncertainty, namely the two regression models.
+In our analysiis there are two more sources of uncertainty, namely the two regression models.
 Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
@@
 \caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5.
-The black vertical line is the median projection (with no uncertainty), while the gray histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
+The black vertical line is the median projection (with no uncertainty), while the grey histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
 }
@@
 
-\subsection{Damage costs}
-
-\subsection{...}
-
-\section{Conclusions and discussion}
-
-Important to continue an iterative and interactive with end-users, as the decision framework provides actionable information to decision-makers, who will then make their own decisions.
-These decision are then incorporated into the current adaptation strategy.
-Continued loop to identify potential vulnerabilities of the approaches across a wide-range of possible futures.
-
-Take home message
-
-\begin{itemize}
-\item It is possible to account for uncertainty even if you do a light touch decision-making approach -- Explain what it implies if you don't account and if you do
-\item Proof of concept -- connect in a probabilistic way (normally done in a discrete way) -- consider uncertainty in both ways
-\item Improve the accessibility and demonstrating the general applicability of policy first approaches -- towards more generic toolkit
-\end{itemize}
-
 %  ACKNOWLEDGMENTS
@@
 This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL).
-The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions/Code}.
+The source code for the analysis is implemented in the statistical programming language {\tt R} (\url{http://www.R-project.org}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions}.
 \end{acknowledgments}

### 2016-12-29 submission/paper.tex
sync

@@
 
-We project local sea level changes by Modeling two processes, the relationship between global temperature and global sea level, and the relationship between global sea level and local sea level.
+We project local sea level changes by modeling two processes, the relationship between global temperature and global sea level, and the relationship between global sea level and local sea level.
 
@@
 \subsection{Damage costs}
+A simplistic analysis of projected damage costs, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage factor at 2100 at the median sea level rise projected for 2100.
 

### 2016-12-29 submission/paper.tex
Included Esbjerg analysis and uncertainty component analysis for Bergen

@@
 \begin{keypoints}
-\item Decisions on adaptation measures need to take careful account of uncertainties
-\item More careful models are needed for increased costs of the effects of sea level rise
+\item Decisions on adaptation measures need to take careful account of all uncertainties
 \item Modeling local sea level rise is essential
+\item Decisions under uncertainty do not have to be be based on precise costs
 \end{keypoints}
@@
 We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
-In the final section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections.
+In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}..
 
@@
 \end{linenomath*}
-where $\gamma$ is the annual land rise rate, $t$ denotes year, and $ {\hat b} $ is the regression coefficient relating global to local sea level.
+where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors..
 
@@
 These approaches can be classified according to a science-first or policy-first approach.
-The former has a ‘predict-then-act’ foundation, which starts with for example probabilistic climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
+The former has a €``predict-then-act''€™ foundation, which starts with for example probabilistic climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
 The latter starts out with the formulated adaptation plans and not impacts and their functioning is tested against different future projections (Dittrich et al.
@@
 \section{Danish decision framework}
-\label{decision_tools_PartI}
+\label{decision_tools_PartII}
+In 2014 the municipality of Esbjerg adopted its climate adaptation plan {\color{blue}(ref)}, which will cover the period from 2014-2026, and which aims to reduce the risk of flooding caused by storm surges, heavy rainfall and rising ground water levels, respectively.
+In the following (and based on interviews with end-users from the municipality) we will focus on the harbour of Esbjerg and the nearby coastal areas, and we will only consider the hazard of coastal flooding, though evidently in some areas the flood risk is compounded.
+
+The initial scoping of the climate adaptation plan in Esbjerg includes a preliminary value and risk mapping, considering critical infrastructure and buildings of high cultural and societal value as identified by the municipality, while informed by spatial floods maps for different scenarios corresponding to each of the three different kinds of hydrological events (sea level rise/storm surges, pluvial flooding and rising ground water levels).
+In terms of coastal floods the mapping considers only one type of storm surge, corresponding to a 20-year return event (based on historical storm surge statistics), and increased sea level due to climate change was not accounted for.
+We will extend these flood maps to also deal with 100-year return events and quantiles of the projected sea level rise.
+
+The risk for any given map area as the probability of, e.g., a certain flood depth, is derived from a hydrological flood model (not including urban drainage system) times a valuation of the consequences, which--similar to the Bergen case--is essentially a loss function associating the flood depth with a measure of cost.
 
@@
 Figure~\ref{fig:BergenDamageDist} shows the histogram of observed annual damage costs for Bergen and the associated Burr distribution.
-The parameter estimates are $\hat{\alpha} = 7.84$, $\hat{\gamma} = 0.40$ and $\hat{\theta} = 0.007$.
+The parameter estimates are $\hat{\alpha} = 7.84$ (3.6), $\hat{\gamma} = 0.40$ (0.04) and $\hat{\theta} = 0.007$ (0.01).
 \citet{bergenreport} discuss several different adaptation options for Bergen.
@@
 \subsection{Identifying flood-prone areas}
+Table \ref{tab:stormsurge} contains the total projected storm surges for Esbjerg, corresponding to the 20-year and 100-year historical surges, with and without sea level rise.
+We see that the historical maximum 433 cm is very likely to be exceeded by 2100.
 
-In 2014 the municipality of Esbjerg adopted its climate adaptation plan {\color{blue}(ref)}, which will cover the period from 2014-2026, and which aims to reduce the risk of flooding caused by storm surges, heavy rainfall and rising ground water levels, respectively.
-In the following and based on interviews with end-users from the municipality we will focus on the harbour of Esbjerg and the nearby coastal areas, and we will only consider the hazard of coastal flooding, though evidently in some areas the flood risk is compounded.
+\begin{table}
+\centering
+\begin{tabular}[]{c | c| c c c}
+\hline
+&&&RCP 8.5&\\
+Storm surge&No sea level rise&5th percentile&median&95th percentile\\
+\hline
+RP 20&388 cm&431 cm&464 cm& 500 cm\\
+RP100&405 cm&448 cm& 481 cm& 517 cm\\
+\hline
+\end{tabular}
+\label{tab:stormsurge}
+\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20 years (RP20) and 100 years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
+\end{table}
 
-The initial scoping of the climate adaptation plan in Esbjerg was completed in 2014 and includes a preliminary value and risk mapping, considering critical infrastructure and buildings of high cultural and societal value as identified by the municipality, while informed by spatial floods maps for different scenarios corresponding to each of the three different kinds of hydrological events (sea level rise/storm surges, pluvial flooding and rising ground water levels).
-In terms of coastal floods the mapping considers only one type of storm surge, corresponding to a 20-year return event (based on historical storm surge statistics), and increased sea level due to climate change was not accounted for.
-
-The risk for any given map area as the probability of, e.g.
-a certain flood depth, is derived from a hydrological flood model (not including urban drainage system) times a valuation of the consequences, which - similar to the Bergen case - is essentially a loss function associating the flood depth with a measure of economic cost.
+Figure \ref{fig:esbjerg} shows flooding maps corresponding to the entries in Table \ref{tab:stormsurge}. These maps can be used to further develop adaptation policies for Esbjerg.
+\begin{figure}[!hbpt]
+\begin{center}
+\includegraphics[width=0.9\linewidth]{Floodmaps_Harbour.png}
+\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 90th percentile (d).
+}
+\label{fig:esbjerg}
+\end{center}
+\end{figure}
 
@@
 \subsection{Damage costs}
-A simplistic analysis of projected damage costs, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage factor at 2100 at the median sea level rise projected for 2100.
+A simplistic analysis of projected damage costs at the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100.
+This yields a damage cost of NOK 335 million (the grey vertical line in Figure \ref{fig:costunc}.
+Similar results obtain when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}).
+However, allowing only the damage cost to vary yields a median cost of 3.84 billion (green dot on top of Figure \ref{fig:costunc}).
+The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their 2100 distributions.
+This corresponds to a median cost of 3.00 billion NOK, a factor of 9 higher than the simplistic value.
+Over 99\% of the costs in our model are higher than the simplistic median.
 
-\subsection{...}
+\begin{figure}[!hbpt]
+\begin{center}
+\includegraphics[width=0.8\linewidth]{UncertaintyLog.pdf}
+\caption{Simulated distribution of total damage cost for 2100 without adaptation on log-log scale (black histogram).
+We also show the distributions of costs varying only one aspect of the uncertainty (sea level rise in blue, effect multiplier in yellow, and damage cost in green), holding the other two at their median values.
+The grey vertical line is the result of holding all three factors at their median value.
+The median of each distribution is shown as a dot on top of the figure.}
+\label{fig:costunc}
+\end{center}
+\end{figure}
+
+%\subsection{...}
 
 \section{Conclusions and discussion}
+\label{disc}
 

### 2016-12-29 code/BergenDecisions/BergenDamageData (Autosaved).R

@@ -1,0 +1,467 @@
+### Decision analysis for Bergen. All the data files are located in the same
+### directory as this script. The R working directory should thus be set
+### to the current directory. 
+
+rm(list=ls())
+
+### Read in damage data obtained from Finance Norway's website:
+### www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/
+A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
+A <- A[-37,] ## Remove 2016 as data not complete
+V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
+V <- V[-37,] ## Remove 2016 as data not complete
+
+### Read in consumer price index (CPI).
+CPI <- read.csv("KPI.csv", header=TRUE)
+CPI <- CPI[-1,] ## Remove 2016 as data not yet available
+CPI$CPI <- CPI$CPI/CPI$CPI[1] ## Normalize to 2015 
+
+### Combine Hordaland and Rogaland data.
+### Standardize costs to 2015 values using CPI. 
+sf.nr <- c(A$Hordaland, A$Rogaland)
+sf.damage <- c(V$Hordaland/CPI$CPI[1:36], V$Rogaland/CPI$CPI[1:36])
+sf.damage <- sf.damage/1e3 ## Normalize to million NOK. 
+
+### Fit Burr distribution to damage data (in MNOK)  
+library(actuar)
+ind <- which(sf.damage==0)
+sf.damage[ind] <- 1e-04 ## Improves numerical stability of estimates. 
+log.lik <- function(theta)
+{
+  res <- sum(dburr(sf.damage, 
+                   shape1=theta[1], 
+                   shape2=theta[2], 
+                   rate=theta[3], 
+                   log=TRUE))
+  return(-res)
+}
+theta.start <- c(5, 0.5, 0.05)
+optim.out <- optim(theta.start, 
+                   fn=log.lik, 
+                   method="BFGS", 
+                   control=list(maxit=1000))
+
+### Get Hallegatte et al. (2013) data on relation between change in damage 
+### cost and change in sea level (csv files copied manually from
+### online supplementary information).   
+E <- read.csv("EuropeanLossData.csv", header=TRUE)
+E[,2:4] <- E[,2:4]/E[,2] 
+
[…]

### 2016-12-30 submission/paper.tex
Added text to Bergen adaptation section. Minor corrections throughout

@@
 Different components of uncertainty are visualized.
-We show that failing to take uncertainty into account can increase the projected damage costs by an order of magnitude.
+We show that failing to take uncertainty into account can result in the median projected damage costs being an order of magnitude smaller.
 \end{abstract}
@@
 
-As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012} adaptive planning methods need to support decisions in the short term, while considering long-term developments.
+As adaptation decision-making is an ongoing process of weighing and choosing which measures should be taken at which moment in time \citep{Hallegatte&2012}, adaptive planning methods need to support decisions in the short term, while considering long-term developments.
 Challenges of adaptation decision-making under uncertainty relate to the incorporation of spatial, inter-temporal and flexibility aspects of adaptation priorities \citep{FankhauserSoare2013}, and the linkage with specific characteristics of sectors and contexts \citep{BisaroSwartHinkel2016, HinkelBisaro2016}.
@@
 in order to prioritize future adaptation actions and investments.
-In both cases we show that embracing the uncertainties derived from economic and hydrologic models is absolutely crucial in order to answer the question of "are we sure we want to do this?".
+In both cases we show that embracing the uncertainties derived from economic and hydrologic models is absolutely crucial in order to answer the question of ``are we sure we want to do this?''
 
@@
 While the viability of the constructions and the associated construction costs have been carefully analyzed, the optimal timing of potential adaptation measures and the effects of the associated uncertainties have yet to be investigated.
-We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs and the effects of sea level rise on changes in damage costs.
+We perform such an analysis where we consider uncertainty in projected sea level rise, damage costs, and the effect of sea level rise on changes in damage costs.
 
@@
 Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbour and the city.
-The highest ever since 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached a record-high 433 cm above the norm, causing massive economic losses.
+The highest since records began in 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.
 More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.
@@
 We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
-In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}..
+In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}.
 
@@
 Most climate models do not explicitly provide sea level as an output of the calculations.
-Rather, the IPCC AR5 report \citep[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age.
+Rather, the IPCC AR5 report \citep[ch.~13]{ipcc} combines the heat expansion of the ocean with temperature forced models for glacial melt, Greenland ice melt, and Antarctic ice melt and with land rise due to rebound from the last ice age and other tectonic effects.
 Judging from the supplementary material to \citet[ch.~13]{ipcc}, the uncertainty assessment is only based on the spread of the ensemble of temperature projections, not on the additional uncertainty in the ice models used.
@@
 We will instead use the empirical approach of Rahmstorf and collaborators \citep{Rahmstorf07,Rahmstorf11}, employing the statistical modeling of \citet{Bolin2014a} to relate global annual mean temperature anomalies to global mean sea level anomalies.
-
 We then apply the estimated historical relationship to projected temperatures from the CMIP5 experiment \citep{cmip5} to obtain projected global annual mean sea level, taking into account the uncertainty in the statistical model as well as the spread of the temperature projection ensemble (see subsection \ref{unc_ass}).
@@
 
-There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes adaptation is moving from theory to practice and practitioners try to deal with how to begin adapting.
+There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes, adaptation is moving from theory to practice, and practitioners try to deal with how to begin adapting.
 This leads to an increasing need and interest in the appraisal of options.
@@
 These approaches can be classified according to a science-first or policy-first approach.
-The former has a €``predict-then-act''€™ foundation, which starts with for example probabilistic climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
-The latter starts out with the formulated adaptation plans and not impacts and their functioning is tested against different future projections (Dittrich et al.
+The former has a ``predict-then-act'' foundation, which starts with climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
+The latter starts out with the formulated adaptation plans and not impacts, and their functioning is tested against different future projections (Dittrich et al.
 2016).
@@
 In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise.
-However, with only two data points, extrapolation approaces such as a power law or exponential growth seem unfeasible.
+However, with only two data points, extrapolation approaches such as a power law or exponential growth seem difficult to justify\`.
 
@@
 
+Storm surge data for Esbjerg ate obtained from the Danish Coastal Authority \citep{sealevel2012}.
+
 \subsection{Sea level rise in Bergen and Esbjerg}
@@
 Figure \ref{fig:bergenobs} shows uncorrected and corrected Bergen sea level data, and the relationship between the corrected Bergen data and the global sea level data.
-The glacial isostatic adjustment is 0.26 (0.07) cm/yr.
+The glacial isostatic adjustment is 0.26 (standard error 0.07) cm/yr.
 The time series regression uses an ARMA(1,1)-model \citep{boxjenkins}, with AR parameter 0.82 (0.13), and MA parameter --0.61 (0.17).
@@
 In Figure~\ref{fig:TotalDamageBergen} we consider the optimal timing of an adaptation option that includes two inner barriers at V{\aa}gen and Damg{\aa}rdssundet, that is, one on each side of central Bergen.
-The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level).
+The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level)\footnote{100 NOK is about 11 EUR.}.
+
+Applying the methodology from section \ref{decision_tools_PartI} we find that the optimal time of building the barriers is in 2046 (Figure \ref{fig:TotalDamageBergen}), and that by the year 2100 this decision will on average save about 1/3 of the median damage costs without adaptation (Figure \ref{fig:NoAction}).
+At the high end (or the 95th percentile of the damage distribution) adaptation saves even more, about 43\%.
 
@@
 \begin{table}
-\centering
-\begin{tabular}[]{c | c| c c c}
-\hline
+
+\begin{center}
+\begin{tabular}[]{c | c | c c c}
+
 &&&RCP 8.5&\\
@@
 \end{tabular}
+\end{center}
+\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20 years (RP20) and 100 years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
 \label{tab:stormsurge}
-\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20 years (RP20) and 100 years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
 \end{table}
@@
 A simplistic analysis of projected damage costs at the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100.
-This yields a damage cost of NOK 335 million (the grey vertical line in Figure \ref{fig:costunc}.
+This yields a damage cost of NOK 338 million (the grey vertical line in Figure \ref{fig:costunc}).
 Similar results obtain when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}).
-However, allowing only the damage cost to vary yields a median cost of 3.84 billion (green dot on top of Figure \ref{fig:costunc}).
+However, allowing only the damage cost to vary yields a median cost of 3.85 billion (green dot on top of Figure \ref{fig:costunc}).
 The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their 2100 distributions.
-This corresponds to a median cost of 3.00 billion NOK, a factor of 9 higher than the simplistic value.
-Over 99\% of the costs in our model are higher than the simplistic median.
+This corresponds to a median cost of 3.15 billion NOK, over 9 times higher than the simplistic value.
+Over 99\% of the costs in our simulation are higher than the simplistic median.
 

### 2016-12-30 submission/paper.tex
First draft of discussion section.

@@
 Table \ref{tab:stormsurge} contains the total projected storm surges for Esbjerg, corresponding to the 20-year and 100-year historical surges, with and without sea level rise.
-We see that the historical maximum 433 cm is very likely to be exceeded by 2100.
+We see that using our projections the historical maximum 433 cm is almost certain to be exceeded by 2100.
 
@@
 \begin{center}
-\includegraphics[width=0.9\linewidth]{Floodmaps_Harbour.png}
-\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 90th percentile (d).
+\includegraphics[width=\linewidth]{Floodmaps_Harbour.png}
+\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d).
 }
@@
 \label{disc}
+Our case studies demonstrate that it is possible to take uncertainty into account in deciding when and where to implement adaptation measures, even if one uses a light touch decision-making approach.
+If one fails to do so, bad scenarios, such as 95th percentiles, can be a order of magnitude worse than what the planners are expecting.
+It is likely worthwhile to be pessimistic in the planning and in the projections.
 
-Important to continue an iterative and interactive with end-users, as the decision framework provides actionable information to decision-makers, who will then make their own decisions.
-These decision are then incorporated into the current adaptation strategy.
-Continued loop to identify potential vulnerabilities of the approaches across a wide-range of possible futures.
+We consider our case studies proofs of concept, which will be first steps in a sequence of interactions with local planners and other end-users.This has to be an iterative and interactive process, as the decision framework provides actionable information to decision-makers, who will then make their own decisions.
+These decision can then be incorporated into the current adaptation strategy.
+Further simulation studies allow a continued loop to identify potential vulnerabilities of the approaches across a wide range of possible futures.
 
-Take home message
-
-\begin{itemize}
-\item It is possible to account for uncertainty even if you do a light touch decision-making approach -- Explain what it implies if you don't account and if you do
-\item Proof of concept -- connect in a probabilistic way (normally done in a discrete way) -- consider uncertainty in both ways
-\item Improve the accessibility and demonstrating the general applicability of policy first approaches -- towards more generic toolkit
-\end{itemize}
+Down the line we plan to develop a flexible and easy-to-use tool kit for decision-making under uncertainty regarding sea level rise.
+An initial step in this direction is the software used in this analysis, which is publicly available and using free software.
 

### 2016-12-30 submission/paper.tex
Minor correction.

@@
 \includegraphics[width=\linewidth]{Floodmaps_Harbour.png}
-\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d).
+\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d)).
 }

### 2016-12-31 submission/paper.tex
Added Per to the author list
final edits on submission

--- 2016-12-30 submission/paper.tex
@@
 
-\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, and K. de Bruin\affil{3,4}}
+\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, P. Skougaard Kaspersen\affil{2} and K. de Bruin\affil{3,4}}
 
@@
 Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues.
-Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented.
-All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action.
-We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark.
+Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing where and at what point in time the chosen strategy should be implemented.
+All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost and prioritization of adaptation actions, and in the implications of no action.
+Here we develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark, to highlight how technical efforts to understand and quantify uncertainties in hydrologic projections can be coupled with concrete decision-problems framed by the needs of the end-users using statistical formulations.
 Different components of uncertainty are visualized.
-We show that failing to take uncertainty into account can result in the median projected damage costs being an order of magnitude smaller.
+We demonstrate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller.
 \end{abstract}
@@
 Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region.
-The city hosts one of the largest harbours  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas.
+The city hosts one of the largest harbors  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas.
 As a result critical infrastructures and commercial buildings figure prominently in the coastal zone.
-Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbour and the city.
-The highest since records began in 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.
-More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.
+Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbor and the city.
+The highest since records began in 1874 was recorded in 1981, where the harbor was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.
+More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 meters have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.
 
@@
 to avoid urban development into such areas, to limit damages to buildings of high societal or cultural value, and to pave the way for implementing cost-effective adaptation measures in the second phase of the plan.
+In this study we expand the initial analysis of the flood risk, which was carried out by the municipality, considering the uncertainty in projected sea level rise and the potential implications of uncertainties related to damage estimates for the risk assessment.
 
@@
 We apply these tools to sea level projections for Bergen, Norway, and for Esbjerg, Denmark in section \ref{cases}.
-In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with a summary and discussion in section \ref{disc}.
+In section \ref{unc} we demonstrate the consequences of ignoring the uncertainty in the projections, and the paper is closed with conclusions in section \ref{disc}.
 
@@
 \end{linenomath*}
-where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors..
+where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors.
 
@@
 
-There is an increasing need for more detailed economic analysis, including simple methods and tools for assessment of options, especially since as Downing (2012) recognizes, adaptation is moving from theory to practice, and practitioners try to deal with how to begin adapting.
-This leads to an increasing need and interest in the appraisal of options.
+Climate change adaptation is currently moving from theory to practice with practitioners needing to decide how to begin adapting.
+This implies an increasing need for detailed economic analysis and appraisal of options   \citep{Downing2012}.
+Robust decision-making approaches can incorporate various types of uncertainty, including a broad range of climate scenarios \citep{Dittrich&2016}.
+These methods can be classified as science-first or policy-first approaches.
+The former type has a ``predict-then-act'' foundation, which starts with climate projections and impact assessments, not linked to any specific adaptation choices \citep{Jones&2014}.
+The latter starts out with the formulated adaptation plans, and their functioning is tested against different future projections \citep{Dittrich&2016}.
 
-Several economic decision support tools and methods exist for adaptation assessment under uncertainty.
-Robust decision-making approaches are able to better incorporate uncertainty and a broad range of climate scenarios to capture as much of the uncertainty on future climates as possible (Dittrich et al.
-2016).
-These approaches can be classified according to a science-first or policy-first approach.
-The former has a ``predict-then-act'' foundation, which starts with climate projections and impact assessments, not linked to any specific adaptation choices (Jones et al., 2014).
-The latter starts out with the formulated adaptation plans and not impacts, and their functioning is tested against different future projections (Dittrich et al.
-2016).
-
-We take a policy-first approach, in which we test a current adaptation plan against different sea level and damage projections and the inclusion of different sources of uncertainty.
+Here, we take a policy-first approach where we test a current adaptation plan against different sea level and damage projections and the inclusion of different sources of uncertainty.
 We focus on what this implies for the timing of adaptation measures and the implications of including uncertainty.
@@
 In particular, the linear extrapolation of the results reported in \citet{Hallegatte&2013} might provide a conservative estimate of the effect of extreme sea level rise.
-However, with only two data points, extrapolation approaches such as a power law or exponential growth seem difficult to justify\`.
+However, with only two data points, extrapolation approaches such as a power law or exponential growth seem difficult to justify.
 
 Alternatively, a modeling framework similar to that of \citet{Hallegatte&2013} could be applied directly to a larger range of potential changes in sea level.
-The elements of such a framework might include an appropriate social discount rate, valuing environmental goods in monetary terms, incorporate socio-economic assumptions and long-term policy goals of decision makers, as well as that climate change is often not the only driver that decision makers should consider, therefore costs and benefits should be studied in a wider context (Dittrich et al, 2016).
+The elements of such a framework might include an appropriate social discount rate, valuing environmental goods in monetary terms, incorporate socio-economic assumptions and long-term policy goals of decision makers, as well as that climate change is often not the only driver that decision makers should consider, therefore costs and benefits should be studied in a wider context \citep{Dittrich&2016}.
 
@@
 
-\section{Danish decision framework}
+\section{Risk mapping}
 \label{decision_tools_PartII}
-In 2014 the municipality of Esbjerg adopted its climate adaptation plan {\color{blue}(ref)}, which will cover the period from 2014-2026, and which aims to reduce the risk of flooding caused by storm surges, heavy rainfall and rising ground water levels, respectively.
-In the following (and based on interviews with end-users from the municipality) we will focus on the harbour of Esbjerg and the nearby coastal areas, and we will only consider the hazard of coastal flooding, though evidently in some areas the flood risk is compounded.
 
-The initial scoping of the climate adaptation plan in Esbjerg includes a preliminary value and risk mapping, considering critical infrastructure and buildings of high cultural and societal value as identified by the municipality, while informed by spatial floods maps for different scenarios corresponding to each of the three different kinds of hydrological events (sea level rise/storm surges, pluvial flooding and rising ground water levels).
-In terms of coastal floods the mapping considers only one type of storm surge, corresponding to a 20-year return event (based on historical storm surge statistics), and increased sea level due to climate change was not accounted for.
-We will extend these flood maps to also deal with 100-year return events and quantiles of the projected sea level rise.
+The initial scoping of climate adaptation in Esbjerg \citep{Esbjerg2014} has been informed by a limited set of floods maps representing different scenarios corresponding, respectively, to flooding caused by storm surges, heavy rainfall and rising ground water levels.
+The flood maps themselves are produced using simplified modelling approaches representing present day climate conditions, i.e.
+they do not consider the expected changes in e.g.
+sea level rise and rainfall characteristics caused by climate change.
+They also do include an explicit representation of the urban drainage system.
+Thus, in the context of making decisions on adaptation this preliminary mapping is primarily meant as a tool for identifying high-risk areas, e.g.
+in terms of avoiding future urban development into such areas, and as a precurser for much more detailed (and resource intensive) local hydrological and economical modelling efforts, e.g.
+to pave the way for deciding upon cost-effective adaptation measures in the second phase of the plan.
+As a result what is of relevance to the stakeholders at this stage is not the mapping of hydrological hazards by itself, but rather the mapping of risk, which compounds hazard with (economic) valuation of its consequences for any given map area or pixel.
+In general, this may be expressed as the probability, e.g.
+of a certain flood depth, derived from the urban flood model times a damage function associating the flood depth with the chosen measure of cost (e.g.
+\citet{Halsnaes&2015}).
+The damage function - as in the case of Bergen - is often expressed in economic terms, however, in Esbjerg a categorical approach has been preferred, wherein a value between 1 and 10 is allocated to each map object, cf.
+Table \ref{tab:valuemap}.
 
-The risk for any given map area as the probability of, e.g., a certain flood depth, is derived from a hydrological flood model (not including urban drainage system) times a valuation of the consequences, which--similar to the Bergen case--is essentially a loss function associating the flood depth with a measure of cost.
+\begin{table}
+\begin{center}
+\begin{tabular}[]{l | l | c | c }
+
+Map object&Description&Unit&Value\\
+\hline
+Important buildings & E.g. industry, hospitals and other public buildings & m2 & 10\\
+Other buildings & All other building types than the above & m2 & 8\\
+Critical infrastructure & Waste water treatment plant, railroads, etc. & m2 & 10\\
+Roads & Essentially all types of roads from main roads to & m2 & 6\\
+& minor paths & & \\
+Cultural heritage & Including cemetaries, historical landscape & m2 & 0.1-4\\
+Natural systems & Included protected areas and sports facilities & m2 & 0.1\\
+\hline
+\end{tabular}
+\end{center}
+\caption{Examples of values allocated to different map objects in Esbjerg, prescribed by the end-users \citep{Esbjerg2014}.}
+\label{tab:valuemap}
+\end{table}
+
+Correspondingly, risk is mapped categorically on a scale from 1-10, where 10 indicates the highest risk level.
+What we observe is that buildings and critical infrastructure within this simplistic decision-framework are always associated with a high risk of economic losses, whereas all other objects are in practice given a lower priority.
+As a result distinguishing specific high risk areas in the built environment becomes crucially dependent on the results of the hazard modelling, which again means that considering the uncertainties can play a large role in what areas are identified.
+
+In the following we will expand on the existing light touch approach used in Esbjerg to identify areas susceptible to coastal floods, to test the importance of considering uncertainties related to projections of sea level rise and storm surges, as well as uncertainties related to the probability of occurence.
+The latter will be done by comparing the consequences of storm surge events of different severity: a moderately likely 20-years and an extreme 100-years return event (return periods calculated from historical storm surge statistics).
 
@@
 For improved parameter estimation, we include the data from Rogaland county which is the county directly south of Hordaland and with similar characteristics.
-We use a dicsount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf.
+We use a discount rate of 4\% for the first 40 years of the analysis, a rate of 3\% for 40 to 75 years into the future and a rate of 2\% beyond 75 years (cf.
 Section 5.8 of \citet{DiscountRate}).
@@
 The combined construction cost of the two barriers for a protection against 75 cm sea level rise is estimated at 1.13 billion NOK (2015 level)\footnote{100 NOK is about 11 EUR.}.
+As the inner barriers only provide partial areal protection, we assume that they have a mitigating effect on 50\% of damages incurred by storm surges.
 
-Applying the methodology from section \ref{decision_tools_PartI} we find that the optimal time of building the barriers is in 2046 (Figure \ref{fig:TotalDamageBergen}), and that by the year 2100 this decision will on average save about 1/3 of the median damage costs without adaptation (Figure \ref{fig:NoAction}).
+\citet{bergenreport} conclude that adaptation measures need not be implemented before 2065.
+Applying the methodology from section \ref{decision_tools_PartI} we find that the optimal time of building the barriers is almost two decades earlier, in 2047 (Figure \ref{fig:TotalDamageBergen}), and that by the year 2100 this decision will on average save about 1/3 of the median damage costs without adaptation (Figure \ref{fig:NoAction}).
 At the high end (or the 95th percentile of the damage distribution) adaptation saves even more, about 43\%.
+
+Furthermore, \citet{bergenreport} consider a far-reaching protection measure consisting of an outer barrier that would protect the whole metropolitan area.
+The cost of this construction is estimated to be at least 34 billion NOK (2015 level).
+It is assessed that such a construction is not viable due to large environmental and economic consequences.
+Our estimates of the total damage (Figure 7) support this conclusion.
 
@@
 
-\subsection{Identifying flood-prone areas}
-Table \ref{tab:stormsurge} contains the total projected storm surges for Esbjerg, corresponding to the 20-year and 100-year historical surges, with and without sea level rise.
-We see that using our projections the historical maximum 433 cm is almost certain to be exceeded by 2100.
+\subsection{Identifying flood-prone areas in Esbjerg}
+Table \ref{tab:stormsurge} contains the total projected storm surges for Esbjerg, corresponding to 20-year and 100-year historical surges, with and without considering the projected sea level rise.
+The first column shows the historical storm surge statistics based on 139-years of observations in Esbjerg Harbour (\citep{sealevel2012}) , where the numbers in brackets indicate the standard deviation derived from a statistical analysis.
+The following columns indicate future storm surge water levels, constructed by adding the projected sea level rise corresponding to the 5th percentile, median, and 95th percentile of the distribution shown in Figure \ref{fig:ci} to the values inferred historical storm surge statistics.
+We therefore implicitly assume that, e.g.
+the statistics of a 20-years return event, will remain unchained in the projection period.
+We see that using our projections the historical maximum of 433 cm is almost certain to be exceeded by 2100.
 
@@
 \hline
-RP 20&388 cm&431 cm&464 cm& 500 cm\\
-RP100&405 cm&448 cm& 481 cm& 517 cm\\
+RP 20&362 (+/-11) cm&431 cm&464 cm& 500 cm\\
+RP100&405 (+/-16) cm&448 cm& 481 cm& 517 cm\\
 \hline
@@
 \end{center}
-\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20 years (RP20) and 100 years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
+\caption{Storm surge water levels \citep{sealevel2012} with a return period of 20-years (RP20) and 100-years (RP100) with no sea level rise  and with sea level rise corresponding to RCP 8.5 (5th percentile, median, and 95th percentile).}
 \label{tab:stormsurge}
@@
 
-Figure \ref{fig:esbjerg} shows flooding maps corresponding to the entries in Table \ref{tab:stormsurge}. These maps can be used to further develop adaptation policies for Esbjerg.
+Figure \ref{fig:esbjerg} shows flood maps corresponding to the entries in Table \ref{tab:stormsurge}, based on data provided by the Danish Geodata Agency\footnote{See \url{http://download.kortforsyningen.dk/content/havstigning-410-500-cm}.} The figure indicates the expected flood depth, derived using a simple hydrologically adapted topographical model\footnote{For more information see \url{http://www.klimatilpasning.dk/kommuner/kortlaegning-til-brug-for-klimatilpasning/den-kommunale-risikokortlaegning/kyst.aspx} (in Danish).}, similar to the maps employed in the first phase of the Esbjerg climate adaptation plan.
+The blue and red color scales show the results for the 20-years and 100-years return events, whereas the frames left through right corresponds, respectively, to (a) no sea level rise/present day conditions (b) 5th percentile of the sea level rise distribution, (c) the median and (d) the 95th percentile.
+
+The most remarkable feature in Figure \ref{fig:esbjerg} is the apparent dike breach, which results in massive flooding to the north-west of Esbjerg, when considering the 95th percentile both for a 20-years and 100-years return event, and which does not appear when considering only the median.
+Secondly, it is also clear that the differences between a 20-years and a 100-years return storm surge event both in terms of the areas flooded and the flood depth overall are found to be relatively small, implying that already a moderate storm surge may potentially have high impacts.
+When further accouting for the projected sea level rise as shown in Figure  \ref{fig:ci} such events are likely to appear considerably more frequent in the future.
+
+Lastly, in terms of the harbour and the coastal areas which are dominated by high-value objects such as critical infrastructure, industry and buildings a clear relation between the flood depth and the severity of the storm surge is observed.
+This suggests that to properly identify the risk and thus pave the way for deciding upon cost-effective adaptation measures, it is crucial to consider not only the median but the full range of hydrological projections.
+
 \begin{figure}[!hbpt]
 \begin{center}
-\includegraphics[width=\linewidth]{Floodmaps_Harbour.png}
-\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a 20 year return period (upper row) and a 100 year return period (lower row) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d).
+\includegraphics[width=\linewidth]{FloodMaps_Harbour_RP20_RP100_dif2.png}
+\caption{Flood extents and depths for the city of Esbjerg in year 2100 during storm surges with a return period of 20-years (RP20) and 100-years (RP100) with (a) no sea level rise and with sea level rise corresponding to RCP 85 (5th percentile (b), median (c) and 95th percentile (d)).
 }
@@
 In our analysis there are two more sources of uncertainty, namely the two regression models.
-Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen projections of sea level rise relative to 1999 under RCP 8.5.
-We see that the ensemble range is about 16 cm, whereas the overall uncertainty range is about 40 cm.
+Figure \ref{fig:unc} shows the single number (vertical black line), the ensemble spread (histogram), the uncertainty including only the global model (red) and the full uncertainty (blue) for Bergen and Esbjerg projections of sea level rise relative to 1999 under RCP 8.5.
+For Bergen, we see that the ensemble range is about 16 cm, whereas the overall uncertainty range is about 40 cm.
+While the values for Esbjerg are somewhat higher, there is slightly less uncertainty in the projections.
 
@@
 \begin{center}
-\includegraphics[width=0.5\linewidth]{unc.pdf}
-\caption{2050 Bergen sea level projections with uncertainty due to different sources for RCP 8.5.
+  \includegraphics[width=0.45\linewidth]{unc.pdf}
+  \includegraphics[width=0.45\linewidth]{ExbjergUncertainty.pdf}
+\caption{2050 Bergen (left) and Esbjerg (right) sea level projections with uncertainty due to different sources for RCP 8.5.
 The black vertical line is the median projection (with no uncertainty), while the gray histogram corresponds to the spread of the climate models, the red curve adds the uncertainty due to the relation between global temperature and global sea level, and the blue line that due to downscaling global sea level to Bergen.
@@
 \subsection{Damage costs}
-A simplistic analysis of projected damage costs at the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100.
-This yields a damage cost of NOK 338 million (the grey vertical line in Figure \ref{fig:costunc}).
-Similar results obtain when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}).
-However, allowing only the damage cost to vary yields a median cost of 3.85 billion (green dot on top of Figure \ref{fig:costunc}).
-The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their 2100 distributions.
-This corresponds to a median cost of 3.15 billion NOK, over 9 times higher than the simplistic value.
+A simplistic analysis of projected total damage costs for the year 2100, not taking into account the uncertainty, would use  the median  historical damage cost multiplied by  the median damage effect factor at 2100 at the median sea level rise projected for 2100.
+Performing this exercise for every year 2016-2100 yields a total (discounted) damage cost of 338 million NOK (the grey vertical line in Figure \ref{fig:costunc}).
+Similar results are obtained when allowing sea level or effect factor to vary, holding the other quantities at the median (yellow and purple dots on top of Figure \ref{fig:costunc}).
+However, allowing only the damage cost to vary yields a median total cost of 3.85 billion (green dot on top of Figure \ref{fig:costunc}).
+The appropriate uncertainty analysis for our model should draw each of sea level, effect factor and damage cost at random from their distributions for 2016-2100.
+This corresponds to a total median cost of 3.15 billion NOK, over 9 times higher than the simplistic value.
 Over 99\% of the costs in our simulation are higher than the simplistic median.
@@
 \begin{center}
-\includegraphics[width=0.8\linewidth]{UncertaintyLog.pdf}
-\caption{Simulated distribution of total damage cost for 2100 without adaptation on log-log scale (black histogram).
-We also show the distributions of costs varying only one aspect of the uncertainty (sea level rise in blue, effect multiplier in yellow, and damage cost in green), holding the other two at their median values.
+\includegraphics[width=\linewidth]{UncertaintyLog.pdf}
+\caption{Simulated distribution of total cumulated damage cost for 2016-2100 without adaptation on log-log scale (black histogram).
+We also show the distributions of costs varying only one aspect of the uncertainty (sea level rise in purple, effect multiplier in yellow, and damage cost in green), holding the other two at their median values.
 The grey vertical line is the result of holding all three factors at their median value.
@@
 
-%\subsection{...}
+\subsection{Risk assessment}
+The simple spatial analysis carried out for Esbjerg as a screening tool for identifying high-risk areas considers not only the median result but also higher and lower order percentiles as well as different storm surge intensities.
+Such a probabilistic approach is clearly needed in order to correctly identify areas susceptible to a high risk of flooding, which may subsequently be the subject of more detailed hydrological modelling related to the potential design and implementation of concrete adaptation measures.
 
-\section{Conclusions and discussion}
+Seen from the perspective of a decision-maker, however, adaption actions are likely to be related to an assessment of risk and not hazard.
+In this study we have not compounded our simple flood risk estimates with the categorical valuation scheme used by Esbjerg in the first phase of their adaptation plan to produce a risk mapping.
+Hence arguably the design of the original scheme does not discriminate sufficiently between assets to allow for an objective prioritization based on the underlying data, i.e.
+buildings are generally all assigned a high value, whereas all other assets are assigned significantly lower value.
+Using a more elaborate valuation model (such a model is currently being developed by Esbjerg municipality in collaboration with consultants) the uncertainties introduced by the valuation model should also be introduced into the probabilistic modelling chain, since the uncertainties introduced by the underlying economic assumptions and modelling \citep{Halsnaes&2015} may have significant implications on the final results of the risk mapping.
+One example of this, as already highlighted in the current scheme (Table \ref{tab:valuemap}), is the valuation of natural areas, which in the context of the Esbjerg case in particularly relates to the marshes close to the nearby city of Ribe that are protected under the code of Natura 2000\footnote{See \url{http://ec.europa.eu/environment/nature/natura2000/index_en.htm}.}.
+From interviews with the municipality it is evident that assigning an objective value to this area, whether by economic cost or in terms of a score, which aligns with the overall objectives of the adaptation plan is a difficult task, which must be taken into account when making the risk assessment including quantification of the uncertainties operational from the end-users' perspectives.
+
+\section{Conclusions}
 \label{disc}
-Our case studies demonstrate that it is possible to take uncertainty into account in deciding when and where to implement adaptation measures, even if one uses a light touch decision-making approach.
-If one fails to do so, bad scenarios, such as 95th percentiles, can be a order of magnitude worse than what the planners are expecting.
-It is likely worthwhile to be pessimistic in the planning and in the projections.
+Sea level rise has serious consequences for infrastructure in coastal areas.
+Our case studies demonstrate that it is possible to take uncertainty into account in deciding when and where to implement adaptation measures, even if one uses a light touch decision-making approaches.
+Failing to account for uncertainty can result in bad scenarios, such as 95th percentiles, to be an order of magnitude worse than what the planners are expecting.
+For climate change adaptation in general, it is likely worthwhile to be pessimistic in the planning and in the projections.
 
-We consider our case studies proofs of concept, which will be first steps in a sequence of interactions with local planners and other end-users.This has to be an iterative and interactive process, as the decision framework provides actionable information to decision-makers, who will then make their own decisions.
-These decision can then be incorporated into the current adaptation strategy.
+We consider our case studies proofs of concept and see these as first steps in a sequence of interactions with local planners and other end-users.The adaptation planning process has to be an iterative and interactive process, as the decision framework provides actionable information to decision-makers, who will then make their own decisions.
+These decisions can then be incorporated into the current adaptation strategy.
 Further simulation studies allow a continued loop to identify potential vulnerabilities of the approaches across a wide range of possible futures.
@@
 Down the line we plan to develop a flexible and easy-to-use tool kit for decision-making under uncertainty regarding sea level rise.
-An initial step in this direction is the software used in this analysis, which is publicly available and using free software.
+An initial step in this direction is the software used in this analysis, which is publicly available and uses free software.
 

### 2016-12-31 submission/paperMD2.tex
final edits on submission

--- 2016-12-30 submission/paper.tex
@@
 Sea level rise has serious consequences for harbor infrastructure, storm drains and sewer systems, and many other issues.
-Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing at what point in time the chosen strategy should be implemented.
-All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost of adaptation actions, and in the cost of no action.
-We develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark.
+Adapting to sea level rise requires comparing different possible adaptation strategies, comparing the cost of different actions (including no action), and assessing where and at what point in time the chosen strategy should be implemented.
+All these decisions must be made under considerable uncertainty--in the amount of sea level rise, in the cost and prioritization of adaptation actions, and in the implications of no action.
+Here we develop two illustrative examples: for Bergen on Norway's west coast and for Esbjerg on the west coast of Denmark, to highlight how technical efforts to understand and quantify uncertainties in hydrologic projections can be coupled with concrete decision-problems framed by the needs of the end-users using statistical formulations.
 Different components of uncertainty are visualized.
-We show that failing to take uncertainty into account can result in the median projected damage costs being an order of magnitude smaller.
+We demonstate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller.
 \end{abstract}
@@
 Esbjerg, on the southwest coast of Jutland, is the fifth-largest city in Denmark and the largest urban area in the region.
-The city hosts one of the largest harbours  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas.
+The city hosts one of the largest harbors  in Denmark, which serves as a focal point for offshore activities in the North Sea, including the continued development of offshore wind power and extensive activities related to the extraction of oil and gas.
 As a result critical infrastructures and commercial buildings figure prominently in the coastal zone.
-Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbour and the city.
-The highest since records began in 1874 was recorded in 1981, where the harbour was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.
+Esbjerg is frequently subject to substantial storms and storm surges, causing severe flooding of the harbor and the city.
+The highest since records began in 1874 was recorded in 1981, where the harbor was completely flooded and the water level reached  433 cm above the norm, causing massive economic losses.
 More generally, storm surges causing water levels in Esbjerg to rise to between 2 and 3 metres have quadrupled over the last four decades according to local records, whereas half of the most severe events have taken place since 1975.
@@
 
-\section{Danish decision framework}
+\section{Risk mapping}
 \label{decision_tools_PartII}
-In 2014 the municipality of Esbjerg adopted its climate adaptation plan {\color{blue}(ref)}, which will cover the period from 2014-2026, and which aims to reduce the risk of flooding caused by storm surges, heavy rainfall and rising ground water levels, respectively.
-In the following (and based on interviews with end-users from the municipality) we will focus on the harbour of Esbjerg and the nearby coastal areas, and we will only consider the hazard of coastal flooding, though evidently in some areas the flood risk is compounded.
+The initial scoping of climate adaptation in Esbjerg {\color{blue}(cite: Esbjerg2014)} has been informed by a limited set of floods maps representing different scenarios corresponding, respectively, to flooding caused by storm surges, heavy rainfall and rising ground water levels.
+The flood maps themselves are produced using simplified modelling approaches representing present day climate conditions, i.e.
+they do not consider the expected changes in e.g.
+sea level rise and rainfall characteristics caused by climate change.
+They also do include an explicit representation of the urban drainage system.
+Thus, in the context of making decisions on adaptation this preliminary mapping is primarily meant as a tool for identifying high-risk areas, e.g.
+in terms of avoiding future urban development into such areas, and as a precurser for much more detailed (and resource intensive) local hydrological and economical modelling efforts, e.g.
+to pave the way for deciding upon cost-effective adaptation measures in the second phase of the plan.
+As a result what is of relevance to the stakeholders at this stage is not the mapping of hydrological hazards by itself, but rather the mapping of risk, which compounds hazard with (economic) valuation of its consequences for any given map area or pixel.
+In general, this may be expressed as the probability, e.g.
+of a certain flood depth, derived from the urban flood model times a damage function associating the flood depth with the chosen measure of cost (e.g.
+{\color{blue}(cite:Halsnaes2015)}.
+The damage function - as in the case of Bergen - is often expressed in economic terms, however, in Esbjerg a categorical approach has been preferred, wherein a value between 1 and 10 is allocated to each map object, c.f.
+Table \ref{tab:valuemap}
 
-The initial scoping of the climate adaptation plan in Esbjerg includes a preliminary value and risk mapping, considering critical infrastructure and buildings of high cultural and societal value as identified by the municipality, while informed by spatial floods maps for different scenarios corresponding to each of the three different kinds of hydrological events (sea level rise/storm surges, pluvial flooding and rising ground water levels).
-In terms of coastal floods the mapping considers only one type of storm surge, corresponding to a 20-year return event (based on historical storm surge statistics), and increased sea level due to climate change was not accounted for.
-We will extend these flood maps to also deal with 100-year return events and quantiles of the projected sea level rise.
+\begin{table}
+\begin{center}
+\begin{tabular}[]{c | c | c | c }
 
-The risk for any given map area as the probability of, e.g., a certain flood depth, is derived from a hydrological flood model (not including urban drainage system) times a valuation of the consequences, which--similar to the Bergen case--is essentially a loss function associating the flood depth with a measure of cost.
+Map object&Description&Unit&Value\\
+\hline
+Important buildings&E.g. industry, hospitals and other public buildings&m2&10\\
+Other buildings&All other building types than the above&m2&8\\
+Critical infrastructure&Waste water treatment plant, railroads, etc.&m2&10\\
+Roads&Covers essentially all types of roads from main roads to minor paths&m2&6\\
+Cultural heritage&Including cemetaries, historical landscape&m2&0.1-4\\
+Natural systems&Included protected areas and sports facilities&m2&0.1\\
+\hline
+\end{tabular}
+\end{center}
+\caption{Examples of values allocated to different map objects in Esbjerg, prescribed by the end-users {\color{blue}(cite: Esbjerg2014)}.}
+\label{tab:valuemap}
+\end{table}
+
+Correspondingly, risk is mapped categorically on a scale from 1-10, where 10 indicates the highest risk level.
+What we observe is that buildings and critical infrastructure within this simplistic decision-framework are always associated with a high risk of economic losses, whereas all other objects are in practice given a lower priority.
+As a result distinguishing specific high risk areas in the built environment becomes crucially dependent on the results of the hazard modelling, which again means that considering the uncertainties can play a large role in what areas are identified.
+
+In the following we will expand on the existing light touch approach used in Esbjerg to identify areas susceptible to coastal floods, to test the importance of considering uncertainties related to projections of sea level rise and storm surges, as well as uncertainties related to the probability of occurence.
+The latter will be done by comparing the consequences of storm surge events of different severity: a moderately likely 20-years and an extreme 100-years return event (return periods calculated from historical storm surge statistics).
 
@@
 
-Storm surge data for Esbjerg ate obtained from the Danish Coastal Authority \citep{sealevel2012}.
+Storm surge data for Esbjerg are obtained from the Danish Coastal Authority \citep{sealevel2012}.
 
@@
 
-Figure \ref{fig:esbjerg} shows flooding maps corresponding to the entries in Table \ref{tab:stormsurge}. These maps can be used to further develop adaptation policies for Esbjerg.
+Figure \ref{fig:esbjerg} shows flood maps corresponding to the entries in Table \ref{tab:stormsurge}, based on data provided by the Danish Geodata Agency\footnote{See \url{http://download.kortforsyningen.dk/content/havstigning-410-500-cm}.} The figure indicates the expected flood depth, derived using a simple hydrologically adapted topographical model\footnote{For more information see \url{http://www.klimatilpasning.dk/kommuner/kortlaegning-til-brug-for-klimatilpasning/den-kommunale-risikokortlaegning/kyst.aspx} (in Danish).}, similar to the maps employed in the first phase of the Esbjerg climate adaptation plan.
+The blue and red color scales show the results for the 20-years and 100-years return events, whereas the frames left through right corresponds, respectively, to (a) no sea level rise/present day conditions (b) 5th percentile of the sea level rise distribution, (c) the median and (d) the 95th percentile.
+
+The most remarkable feature in Figure \ref{fig:esbjerg} is the apparent dike breach, which results in massive flooding to the north-west of Esbjerg, when considering the 95th percentile both for a 20-years and 100-years return event, and which does not appear when considering only the median.
+Secondly, it is also clear that the differences between a 20-years and a 100-years return storm surge event both in terms of the areas flooded and the flood depth overall are found to be relatively small, implying that already a moderate storm surge may potentially have high impacts.
+When further accouting for the projected sea level rise as shown in Figure  \ref{fig:ci} such events are likely to appear considerably more frequent in the future.
+
+Lastly, in terms of the harbour and the coastal areas which are dominated by high-value objects such as critical infrastructure, industry and buildings a clear relation between the flood depth and the severity of the storm surge is observed.
+This suggests that to properly identify the risk and thus pave the way for deciding upon cost-effective adaptation measures, it is crucial to consider not only the median but the full range of hydrological projections.
+
 \begin{figure}[!hbpt]
@@
 
+\subsection{Risk assessment}
+The simple spatial analysis carried out for Esbjerg as a screening tool for identifying high-risk areas considers not only the median result but also higher and lower order percentiles as well as different storm surge intensities.
+Such a probabilistic approach is clearly needed in order to correctly identify areas susceptible to a high risk of flooding, which may subsequently be the subject of more detailed hydrological modelling related to the potential design and implementation of concrete adaptation measures.
+
+Seen from the perspective of a decision-maker, however, adaption actions are likely to be related to an assessment of risk and not hazard.
+In this study we have not compounded our simple flood risk estimates with the categorical valuation scheme used by Esbjerg in the first phase of their adaptation plan to produce a risk mapping.
+Hence arguably the design of the original scheme does not discriminate sufficiently between assets to allow for an objective prioritization based on the underlying data, i.e.
+buildings are generally all assigned a high value, whereas all other assets are assigned significantly lower value.
+Using a more elaborate valuation model (such a model is currently being developed by Esbjerg municipality in collaboration with consultants) the uncertainties introduced by the valuation model should also be introduced into the probabilistic modelling chain, since the uncertainties introduced by the underlying economic assumptions and modelling ({\color{blue}(cite:Halsnaes2015)}) may have significant implications on the final results of the risk mapping.
+One example of this, as already highlighted in the current scheme (Table \ref{tab:valuemap}), is the valuation of natural areas, which in the context of the Esbjerg case in particularly relates to the marshes close to the nearby city of Ribe that are protected under the code of Natura 2000\footnote{See \url{http://ec.europa.eu/environment/nature/natura2000/index_en.htm}.}.
+From interviews with the municipality it is evident that assigning an objective value to this area, whether by economic cost or in terms of a score, which aligns with the overall objectives of the adaptation plan is a difficult task, which must be taken into account when making the risk assessment including quantification of the uncertainties operational from the end-users' perspectives.
+
 %\subsection{...}

### 2017-01-06 techreport/manual.tex
Technical Report

% $Id: manual.tex 239 2013-04-18 13:25:02Z wolfgang $
% Specify language of the report

\documentclass[note,screen,british,11pt]{nrdoc}

% Verbatim is already loaded by nrdoc. It is only
% included to make LaTeX2html work better.
\usepackage{verbatim}

% -------------------------------------------------
% BEGIN nrdoc special commands
\ifx\nrdocument\undefined\relax\else
\reportnumber{sand/08/05}
%\publicationyear{2003}
\keywords{\LaTeX, NR}
\projectnumber{}
\project{}
\availability{Internal}
\frontpagefigure{figs/texfriendly} % No file extension.
\target{All employees}
\researchfield{}
\fi
% END nrdoc special commands
% ------------------------------------------------

\makeindex

\begin{document}

\title{User manual for nrdoc: a \LaTeX\ class for notes and reports}
\author{Harald H. Soleng \email{harald.soleng@nr.no} \and 
         Anders Løland \email{anders.loland@nr.no}}
\aboutauthors{The authors are local \LaTeX\ gurus}
\date{\today}
\maketitle

% The abstract must always be present
\begin{abstract}
This document gives an example of the use of the \texttt{nrdoc} class. The
purpose of this document class is to provide a unified style for NR
notes and reports conforming with standard typographical practice in
Norwegian and English.  
The \texttt{nrdoc} class was originally
written by Harald H. Soleng and Anders Løland,  
redesigned by Lars Brusletto Sveen of Munch
Design, and again modified by Harald H. Soleng and Anders Løland.
\end{abstract}

\tableofcontents

\section{Introduction}
The \texttt{nrdoc} class was first written by \cite{NRDocSource03,NRDocManual03}
Harald H. Soleng and Anders Løland. When   Munch
Design redesigned the graphical profile of the Institute, the class was
modified
by Lars Brusletto Sveen. Later, 
the original authors have fine tuned the class 
and added more functionality.

This document gives an example of the use of the \texttt{nrdoc} class. 
In Appendix \ref{sec:code} the \LaTeX\ source code for this document is
given.

\section{Two ways to use the nrdoc class}

The \texttt{nrdoc} class can be used both by \LaTeX\ \index{latex@\LaTeX} 
\citep{latex}
(using the \texttt{latex} command) and 
pdf\LaTeX\ \index{pdflatex@pdf\LaTeX}
(using the \texttt{pdflatex} command). Please note that you need to use 
a reasonably recent tex distribution. Using the option
\textit{screen} \index{screen}\index{option!screen} together with
\texttt{pdflatex} or \texttt{latex} and
\texttt{dvipdfm}\index{dvipdfm} gives an \textbf{electronic report
  with hyperlinks}.\index{hyperlinks} An example of the hyperlinks is
given here: If this is an electronic report, Equations \eqref{eq:ey}
can be found by clicking the equation number. The option
\textit{screen} gives coloured hyperlinks, so
this option should not be used for a paper document.

By default the page layout is two-sided\footnote{On older \LaTeX\
installations, the default is one-sided. If you want to be sure to 
get two-sided output, specify the additional option
\textit{twoside}\index{option!twoside} in the \texttt{documentclass}
command. The option
\textit{oneside}\index{option!oneside} gives one-sided page layout.}.

\subsection{Note and report}
The default option \textit{note}\index{option!note} gives a NR
note. The option  \textit{report}\index{option!report} gives a NR
report. For a NR report, additional variables described in Section
\ref{sec:fronttitle} must be specified.

\subsection{Long reports}

The class option \textit{long}\index{option!long} 
gives access to part\index{part} and chapter\index{chapter} commands.
By default the highest level is section\index{section}.

\section{Front and title page\label{sec:fronttitle}}

For the title page to work, the abstract\index{abstract} (summary) must be
present. The title is set by \verb!\title!. \index{title}
\verb!\nrtitle! \index{nrtitle} can be used to specify the title in
the title page (this is convenient if the title on the front page
includes some line shifts). \verb!\shorttitle! \index{nrtitle}
specifies the text in the heading of each page. If \verb!\shorttitle!
\index{shorttitle} is not specified, the text in the heading will be
taken from \verb!\title!.

The date\index{date} is taken from \verb!\date!.

The note or publication number is given by
\verb+\reportnumber+\index{reportnumber@\verb+\reportnumber+}.
The keywords of the document are given by
\verb+\keywords+\index{keywords@\verb+\keywords+}.
The project title is given by 
\verb+\project+\index{project@\verb+\project+},
the project number by
\verb+\projectnumber+\index{projectnumber@\verb+\projectnumber+},
 and the target group by
\verb+\target+\index{target@\verb+\target+}.
The research field is specified using
\verb+\researchfield+\index{researchfield@\verb+\researchfield+}. Optionally, the name of a quality assurerer may be specified by
\verb+\qa+\index{quality assurance@\verb+\qa+}.

For reports, you also need to supply the isbn\index{isbn number} number. 
The isbn number consists
of a country code, a publisher code, 
a publication number and a control digit.
You can either use \texttt{isbn}\index{isbn@\verb+\isbn+} to supply 
the full number or 
\texttt{shortisbn}\index{shortisbn@\verb+\shortisbn+}
to supply just the last five digits:
\begin{verbatim}
\isbn{82-539-0481-9}
\shortisbn{0481-9}
\end{verbatim}

The availability level is given by
\verb+\availability+\index{availability@\verb+\availability+}. There
are four possible levels of availability:
\begin{itemize}
\item Open\index{availability!open}: Open for everyone. Available
from NR's web site.
\item Internal\index{availability!internal}: Open for everyone at NR. 
\item Confidential\index{availability!confidential}: Confidential. 
\item Strictlyconfidential\index{availability!strictly confidential}:
Strictly confidential. 
\item Statoil restricted\index{availability!Statoil restricted}: Statoil restricted. 
\end{itemize}

A typical preamble\index{preamble} then looks as follows:
\begin{verbatim}

\documentclass[note,screen,twoside,british,12pt]{nrdoc}
\reportnumber{sand/03/03}
\keywords{\LaTeX, NR}
\project{Fun with \LaTeX}
\projectnumber{314259}
\frontpagefigure{figs/texfriendly} % No file extension.
\target{All employees}
\researchfield{}
\availability{Internal}
\end{verbatim}
At the beginning of the document, we specify the title,
author(s) and abstract:
\begin{verbatim}
\makeindex
\begin{document}
\title{User manual for nrdoc: a \LaTeX\ class for notes and reports}
\author{Harald Soleng \email{harald.soleng@nr.no} \and 
        Anders Løland \email{anders.loland@nr.no}}

\aboutauthors{The authors are local \LaTeX\ Gurus}
\date{\today}

\maketitle
\end{verbatim}
Separate each author with the \verb+\and+\index{and@\verb+\and+}
command. The email\index{email@\verb+\email+} address and the
\verb+\aboutauthors+\index{aboutauthors@\verb+\aboutauthors+} command are optional.
\begin{verbatim}
\begin{abstract}
The abstract is obligatory!
\end{abstract}

\tableofcontents        % optional
\end{verbatim}

\section{Including a figure}

On the front page, a figure\index{cover figure} can be included with
\begin{verbatim}
\frontpagefigure{figs/my_figure}
\end{verbatim}
\index{frontpagefigure@\verb+\frontpagefigure+}
Just give the full path to the figure file,
without the file extension. The figure is automatically scaled to fit
into the front page. A more low-level commands called
\texttt{frontpagefigure\-command}%
\index{frontpagefigurecommand@\verb+\frontpagefigurecommand+} 
allows more control. With this commands one 
has to specify a full
figure inclusion command, e.g., 
\begin{verbatim}
\frontpagefigurecommand{%
\includegraphics[width=50mm,height=40mm]{figs/my_figure}}
\end{verbatim}
If you want more than one figure on the cover 
you either need to combine them
using a graphics manipulation program, or 
use \texttt{frontpage\-figure\-command}
with several 
\texttt{include\-graphics}%
\index{includegraphics@\verb+\includegraphics+} 
commands.

Other figures\index{graphics}\index{illustration} \index{figure}
are included by \texttt{includegraphics}. An example is
given here (and in Figure \ref{fig:logo}):
\begin{verbatim}
\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=0.3\linewidth,angle=45]{figs/fotball}
    \caption{\label{fig:logo} Rolling NR ball.}
  \end{center}
\end{figure}
\end{verbatim}
\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=0.3\linewidth,angle=45]{figs/fotball}
    \caption{\label{fig:logo} Rolling NR ball.}
  \end{center}
\end{figure}
Note that the extension of the figure file name should be dropped. The
reason is that \LaTeX\ will find the required file itself. 
The figure
files should have the following format:\index{format}
\begin{description}
\item [\LaTeX] postscript [.ps] or encapsulated postscript
  [.eps] \index{format!postscript}
\item [pdf\LaTeX] pdf  [.pdf], jpeg [.JPG/.JPEG/.jpeg/.jpg] or png [.png] formats
  \index{format!pdf} \index{format!jpeg}\index{format!png}
\end{description}
We recommend pdf\LaTeX, since this gives instant pdf files and enables
inclusion of jpeg and png images (in addition to pdf files).

\subsection{Format conversion}
postscript files can be converted to pdf using the \texttt{epstopdf}
command or Acrobat Distiller. 
pdf files can be converted to postscript using the \texttt{pdftops}
command or Adobe Acrobat.

\section{Fonts}

The default size is 11pt, but 10pt and 12pt sizes are also supported
through standard 
options.\index{option!10pt}\index{option!11pt}\index{option!12pt}

The standard font is type 1 versions of palatino. An important exception is
the typewriter font, which has been replaced with almost European fonts.

\section{Mathematics}\index{mathematics}

The \texttt{nrdoc} class loads the ams\index{amslatex@ams\LaTeX} 
packages so that the full power of these extensions are available.
\begin{align}\label{eq:ey}
\begin{split}
  EY(s)&=\sum_{i=1}^n \mathcal{K}_s(s_i)Ew(s_i),\\
  \gamma(s,s')&=\sum_{i=1}^n \mathcal{K}_s(s_i)
  \mathcal{K}_{s'}(s_i)Ew^2(s_i).
\end{split}
\end{align}

\section{The example environment\index{example}}

The example environment enables inclusion of numbered examples in your
notes and reports.
\begin{example}\caption{Example of example --  a quote!}
\begin{quote}
"Dirty-looking rascals, but I suppose every one has some little
immortal spark concealed about him. You would not think it, to look at
them. There is no a priori probability about it. A strange enigma is
man!"

"Winwood Reade is good upon the subject," said Holmes. "He
remarks that, while the individual man is an insoluble puzzle, in the
aggregate he becomes a mathematical certainty. You can, for example,
never foretell what any one man will do, but you can say with
precision what an average number will be up to. Individuals vary, but
percentages remain constant. So says the statistician."

\textit{--The Sign of Four, Arthur Conan Doyle}
\end{quote}
\end{example}
Note that each example must have a title, which is given in a caption;
\begin{verbatim}
\begin{example}\caption{Example of example --  a quote!}
\begin{quote}
"Dirty-looking rascals, but I suppose every one has some little
...
percentages remain constant. So says the statistician."

\textit{--The Sign of Four , Arthur Conan Doyle}
\end{quote}
\end{example}
\end{verbatim}
A list of examples can be generated using the command \verb+\listofexamples+\index{listofexamples@\verb+\listofexamples+}.

\section{Citation using {Bib}\TeX}

Citations\index{cite}
are handled by {Bib}\TeX\ 
\index{bibtex@BiB\TeX}\index{cite!BiB\TeX@bibtex} and 
\texttt{natbib} \citep{natbib:7.1}. By default 
an author--year citation 
scheme\footnote{With the options \texttt{citealphanumeric}\index{cite!numeric citation label}\index{option!citealphanumeric}
or \texttt{citenumeric}\index{option!citenumeric}
a scheme with numerical citation keys and 
citations appearing in alphabethic order or in the order of citations
is supported as well. This is intended for documents that need
numerical citations due to external demands.}  is used.
The required style files are 
automatically loaded by \texttt{nrdoc} and support both norsk
and english. Some examples are given in Table \ref{tab:cite}.    
\begin{table}[htb]
\centering
\begin{tabular}{lll}
\hline
usual citations &  
\protect\cite{mardia79} & 
\verb!\cite{mardia79}! \\
in parentheses  &  
\protect\citep{mardia79} & 
\verb!\citep{mardia79}!\\
with a see     &  
\protect\citep[see][]{mardia79} &
\verb!\citep[see][]{mardia79}!\\
with a section/page & 
\protect\citet[Section 2]{mardia79} & 
\verb!\citet{mardia79}!\\
will all authors & 
\protect\citet*{mardia79} & 
\verb!\citet*{mardia79}!\\
without parentheses & 
\protect\citealt{mardia79}&
\verb!\citealt{mardia79}!\\
only the year & 
\protect\citeyear{mardia79} & 
\verb!\citeyear{mardia79}!\\
only the author & 
\protect\citeauthor{mardia79} &
\verb!\citeauthor{mardia79}! \\
\hline
\end{tabular}
\caption{Examples of citations using {Bib}\TeX{}.}
\label{tab:cite} 
\end{table}

The citations are listed in a {Bib}\TeX{} file with the extension
\texttt{.bib}. An example (\texttt{ref.bib}) is given in Appendix
\ref{sec:ref.bib}. Use the \texttt{bibtex} command to run {Bib}\TeX{}.

\section{Languages}

The following options are available for 
different language\index{language} support:
\begin{description}
\item[\textit{american}] US english (default).\index{option!american} 
\item[\textit{british}] UK english. \index{option!british} 
\item[\textit{english}] US english. \index{option!english} 
\item[\textit{norsk}] Bokm\aa l.\index{option!norsk} 
\item[\textit{nynorsk}] Nynorsk.\index{option!nynorsk} 
\item[\textit{UKenglish}] UK english. \index{option!UKenglish}
\item[\textit{USenglish}] US english. \index{option!USenglish} 
\end{description}

\section{Additional options}
\subsection{The utf8 option}\index{option!utf8}

Allows use of utf8 font encoding. The current 
default is latin1.

\subsection{The draft option}\index{option!draft}

The draft option prevents figure inclusion, turns off the cover
generation and adds marks on oversized lines. This option also 
prints equation labels in the margin.

\subsection{The showlabels option}\index{option!showlabels}

This option prints equation labels in the margin.

\subsection{The indentpar option}\index{option!indentpar}

This option turns left intendation on for paragraphs.

\subsection{The nodesign option}\index{option!nodesign}

The nodesign option prevents inclusion of the design elements created
by Much Design. It is used to speed up processing in the writing stage.
This option must be removed in the final run.

\subsection{The dvips option}\index{option!dvips}

This package is designed to produce pdf output either via
\LaTeX\ followed by dvipdfm or directly by pdf\LaTeX.
By using the dvips option you can use \LaTeX\ followed by dvips to produce
postscript.

\subsection{The englishTitle option}\index{option!englishTitle}

When specifying both english and norsk as options for multi-lingual
documents the nrdoc style will choose norwegian as language for cover page
and title matter. The option \texttt{englishTitle} will choose english
as default language, including cover and title matter. Switching to
norwegian language is done by \texttt{\textbackslash
  selectlanguage\textbraceleft{norsk}\textbraceright}.

\section{Making an index}\index{index}

There is support for making indexes with makeindex.
Add the command \verb!\makeindex! in the preamble, and
run \texttt{makeindex} on the main tex file. This generates
an index file.

\section{Default packages}

The following packages are loaded by default (or by use of options):
\begin{description}
\item[\textit{a4wide}] For page geometry. \index{package!a4wide} 
\item[\textit{afterpage}] For float control. \index{package!afterpage}
\item[\textit{amsmath}] For advanced mathematical typesetting. \index{package!amsmath}\item[\textit{algoritm}] For algorithm (type of float). 
\index{package!algorithm}
\item[\textit{array}] For added tabular functionality \index{package!array}
\item[\textit{babel}] For language support. \index{package!babel} 
\item[\textit{calc}] For numerical programming constructs in \LaTeX. 
\index{package!calc}
\item[\textit{caption}] For formatting of captions.
\index{package!caption} 
\item[\textit{color}] For colour fonts. \index{package!color}
\item[\textit{fancyhdr}] For headers. \index{package!fancyhdr} 
\item[\textit{fontenc}] For font encoding (using T1). \index{package!fontenc}
\item[\textit{float}] For float control. \index{package!float} 
\item[\textit{graphicx}] for graphics inclusion. \index{package!graphicx} 
\item[\textit{hyperref}] For hyperlinks. \index{package!hyperref}
\item[\textit{ifpdf}] Provides tests for  pdf\LaTeX. 
\index{package!ifpdf} 
\item[\textit{ifthen}] Provides programming constructs in \LaTeX. \index{package!ifthen} 
\item[\textit{indentfirst}] Handles Norwegian type indentation.\index{package!firstindent} 
\item[\textit{inputenc}] Handles Norwegian
  letters. \index{package!inputenc} 
\item[\textit{makeidx}] For index. \index{package!makeidx}
\item[\textit{multicol}] For multiple columns. \index{package!multicol} 
\item[\textit{natbib}] For bibliography. \index{package!natbib} 
\item[\textit{paralist}] For list environments. \index{package!paralist}
\item[\textit{parskip}] For non-indenting paragraphs. \index{package!parskip} 
\item[\textit{rotating}] For text rotation. \index{package!rotating} 
\item[\textit{showlabels}] For printing of labels (in drafts). 
\index{package!showlabels} 
\item[\textit{textcase}] For text lower case and upper case functions.
\item[\textit{textcomp}] For special characters. 
\index{package!textcase}
\item[\textit{textpos}] For absolute positioning of text on front page.
\index{package!textpos} 
\item[\textit{thumbpdf}] For thumbnails in pdf.
\index{package!thumbpdf}
\item[\textit{titlesec}] For title formatting.
\index{package!titlesec} 
\item[\textit{geometry}] For page geometry. \index{package!geometry}

\item[\textit{verbatim}] For verbatim text. \index{package!verbatim} 
Also contains a useful \texttt{comment} environment.

\item[\textit{xspace}] For space control. \index{package!xspace}
\end{description} 
For more information about most of these packages,
consult the book by \cite{latexcompanion}.

\section{Shortcomings}

Pdf bookmarks does not work correctly for chapters and sections in appendices.
Some postprocessing of the bookmark file (file extension .out) 
may be needed to translate \LaTeX\ codes, since bookmarks must 
be written in PDFEncoding. To aid this process, the .out file 
is not rewritten by \LaTeX if it is edited to contain 
a line \let\WriteBookmarks\relax.

\bibliography{ref}

\clearpage
\appendix

\section{Setting Unix variables}
For the system to find the \texttt{nrdoc} class, the variables
\texttt{TEXINPUTS}\index{texinputs@TEXINPUTS} and 
\texttt{BSTINPUTS}\index{texinputs@BSTINPUTS} needs to be set. These
variable should have been set for you, but if you get the message 
\texttt{TEXINPUTS: Undefined variable.} after typing \texttt{echo
\$TEXINPUTS} in Unix/Linux, you need to set the variable:
If you use \texttt{tcsh} (that is
you have a \texttt{.cshrc} file), the  \texttt{.cshrc} file should
include this line:
\begin{verbatim}
setenv TEXINPUTS .:/nr/group/maler/nrdoc:
\end{verbatim}
If you have problems with {Bib}\TeX, it might help to set the variable
\texttt{BSTINPUTS}:
\begin{verbatim}
setenv BSTINPUTS .:/nr/group/maler/nrdoc:
\end{verbatim}

\section{The \LaTeX{} code for this document\label{sec:code}}
\verbatiminput{manual.tex}
\section{The {Bib}\TeX{} file (\texttt{ref.bib})}\label{sec:ref.bib}
\verbatiminput{ref.bib}

\end{document}

### 2017-01-06 techreport/paper.tex
Technical Report

--- 2016-12-31 submission/paper.tex
@@
-%% To submit your paper:
-\documentclass[draft,linenumbers]{agujournal}
-\draftfalse
+\documentclass[note,screen,british,11pt]{nrdoc}
 
-%% For final version.
-% \documentclass{agujournal}
+\usepackage{url, bbm, comment, verbatim}
 
-\journalname{Water Resources Research}
+% -------------------------------------------------
+% BEGIN nrdoc special commands
+\ifx\nrdocument\undefined\relax\else
+\reportnumber{samba/02/17}
+\keywords{Sea level rise; Adaptation; Decision making}
+\projectnumber{220730}
+\project{eSACP}
+\availability{Open}
+\frontpagefigure{sealevel-tim-marshall} 
+\target{Decision makers and scientists}
+\researchfield{Statistics, Climatology, Environmental Economics}
+\fi
+% END nrdoc special commands
+% ------------------------------------------------
 
-\usepackage{url, bbm, comment}
+\makeindex
 
@@
 
-\title{I don't know, are you sure we want to do this?\\
-Sea level adaptation decisions under uncertainty}
+\title{\huge I don't know, are you sure we want to do this? Sea level adaptation decisions under uncertainty}
 
-\authors{T. L. Thorarinsdottir\affil{1}, P. Guttorp\affil{1}, M. Drews\affil{2}, P. Skougaard Kaspersen\affil{2} and K. de Bruin\affil{3,4}}
+\nrtitle{I don't know, are you sure we want to do this? Sea level adaptation decisions under uncertainty}
 
-\affiliation{1}{Norwegian Computing Center, Oslo, Norway}
-\affiliation{2}{Technical University of Denmark, Copenhagen, Denmark}
-\affiliation{3}{Center for International Climate and Environmental Research, Oslo, Norway}
-\affiliation{4}{Wageningen Environmental Research, Wageningen, The Netherlands}
+\shorttitle{Sea level adaptation decisions under uncertainty}
 
-\correspondingauthor{T. L. Thorarinsdottir}{thordis@nr.no}
+\author{Thordis L.
+Thorarinsdottir \email{thordis@nr.no} \and Peter Guttorp \email{peter.guttorp@nr.no} \and Martin Drews \email{mard@dtu.dk} \and Per Skougaard Kaspersen \email{pskk@dtu.dk} \and Karianne de Bruin \email{karianne.debruin@cicero.oslo.no}}
 
-\begin{keypoints}
-\item Decisions on adaptation measures need to take careful account of all uncertainties
-\item Modeling local sea level rise is essential
-\item Decisions under uncertainty do not have to be be based on precise costs
-\end{keypoints}
+\aboutauthors{Thordis L.
+Thorarinsdottir is Chief Research Scientist and Peter Guttorp is Professor II at the Norwegian Computing Center in Oslo, Norway.
+Martin Drews is a Senior Scientist and Per Skougaard Kaspersen is Postdoc at DTU Management Engineering, Technical University of Denmark in Copenhagen, Denmark.
+Karianne de Bruin is a Senior Researcher at the Center for International Climate and Environmental Research in Oslo, Norway and Researcher at Wageningen Environmental Research in Wageningen, The Netherlands}
+
+\date{\today}
+
+\maketitle
 
@@
 We demonstrate the value of uncertainties and show for example that failing to take uncertainty into account can  result in the median projected damage costs being an order of magnitude smaller.
+
+  Front page photo is by Tim Marshall (\url{https://unsplash.com/@timmarshall}). 
 \end{abstract}
-
-%% ------------------------------------------------------------------------ %%
-%  BEGIN ARTICLE
-%% ------------------------------------------------------------------------ %%
 
@@
 For the i'th temperature projection $T_t^i$ we estimate the corresponding global mean sea level as
-\begin{linenomath*}
 \[H_t^{gl,i} = \int\limits_{{t_0}}^t {{\hat a} (T_u^i - {{\hat T}_0}} )du + {\varsigma _t},\]
-\end{linenomath*}
 where ${\hat a}$ and ${\hat T}_0$ are regression parameters of observed global sea level  rise on observed global temperature and $\varsigma_t$ the integrated time series regression error.
@@
 The local sea level projection based on the i'th climate model for years beyond 2000 is estimated as
-\begin{linenomath*}
 \[H_t^{loc,i} + \gamma (t -2000 ) = {\hat b} H_t^{gl,i}  + {\varepsilon _t},\]
-\end{linenomath*}
 where $\gamma$ is the annual land rise rate, $t$ denotes year,  $ {\hat b} $ is the regression coefficient relating global to local sea level and the $\varepsilon _t$ are Gaussian errors.
@@
 We model the distribution of annual damage, $F_{\textup{d}, t_0}$, for the year $t_0 = 2015$ by the three parameter Burr distribution \citep{Burr1942} with density
-\begin{linenomath*}
   \begin{equation}\label{eq:Burr}
@@
   \end{equation}
-\end{linenomath*}
 for $x > 0$, where $\alpha$ and $\gamma$ are shape parameters with $\alpha, \gamma > 0$, and $\theta >0$ is a scale parameter.
@@
 By repeating this process $J$ times, we obtain an empirical damage distribution for each future year $t_i$ given by
-\begin{linenomath*}
   \[
@@
   \]
-\end{linenomath*}
 for $i = 1, \ldots, 85$, where $r_{t_i}$ is the discount rate for year $t_i$. Alternatively, we obtain an empirical distribution of the total damage over the period $2016-2100$ by considering
-\begin{linenomath*}
   \[
@@
   \]
-\end{linenomath*}
 and similarly for the cumulative damage.
@@
 That is, for a sea level anomaly $s_{t_i}$ in year $t_i > 2015$ compared to the 2015 level, the annual damage cost becomes
-\begin{linenomath*}
   \[
@@
   \]
-\end{linenomath*}
 where $g( \cdot | \boldsymbol \beta )$ is a monotonic positive function with parameter vector  $\boldsymbol \beta$ such that $g(s|\boldsymbol \beta ) > 1$ for $s>0$ and $g(s|\boldsymbol \beta ) < 1$ for $s < 0$.
@@
 An empirical damage distribution for the future year $t_i$ that accounts for uncertainty in damage, sea level rise and its effect on the damage is then given by
-\begin{linenomath*}
   \begin{equation}\label{eq:FutureDamage}
@@
   \end{equation}
-\end{linenomath*}
 
@@
 This results in a damage distribution given by
-\begin{linenomath*}
   \[
@@
   \]
-\end{linenomath*}
 
@@
 %  ACKNOWLEDGMENTS
-\begin{acknowledgments}
+\section*{Acknowledgments}
 This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL).
 The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions/Code}.
-\end{acknowledgments}
 

### 2017-01-31 techreport/paper.tex
minor correction

@@
 Martin Drews is a Senior Scientist and Per Skougaard Kaspersen is Postdoc at DTU Management Engineering, Technical University of Denmark in Copenhagen, Denmark.
-Karianne de Bruin is a Senior Researcher at the Center for International Climate and Environmental Research in Oslo, Norway and Researcher at Wageningen Environmental Research in Wageningen, The Netherlands}
+Karianne de Bruin is Senior Researcher at the Center for International Climate and Environmental Research in Oslo, Norway and Researcher at Wageningen Environmental Research in Wageningen, The Netherlands}
 

### 2017-01-31 code/Bergen.R
cleaned up code on Bergen example

@@ -1,176 +1,0 @@
-setwd("~/GitHub/SeaLevelDecisions/code")
-require(maps)
-require(maptools)
-require(ncdf4)
-require(forecast)
-require(excursions)
-require(Matrix)
-require(chron)
-require(zoo)
-source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
-source('~/GitHub/SeaLevelDecisions/code/ci_bergen.R')
-norway=read.table("~/GitHub/SeaLevelProjections/Norway_sea_level/Norway.PSMSL.txt")#All Norwegian sea level stations with current data
-#from PSMSL.org
-norway[,1]=as.character(norway[,1])
-
-
-norwayurl=paste("http://www.psmsl.org/data/obtaining/rlr.monthly.data/",norway[,2],".rlrdata",sep="")
-
-N=length(norway[,2])
-norwaydata=list(N,NULL)
-
-for (i in 1:N){
-	tmp=read.table(norwayurl[i],na.strings="-99999",sep=";")
-	tmp=cbind(floor(tmp[,1]),round(12*(tmp[,1]-floor(tmp[,1]))+.5),tmp[,2:4])
-
-	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	NN=dim(tmp)[1]
-	if(tmp[NN,2]<12) tmp=tmp[-((NN+1-tmp[NN,2]):NN),]#delete final year if not full
-	tmp.rlr=tmp[,3]
-	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
-	tmp.mp=medpolish(tmp.rlr,na.rm=T)
-	tmp.mppred=(tmp.mp$overall+outer(tmp.mp$row,tmp.mp$col,"+"))
-	for(j in 1:12) tmp.rlr[which(is.na(tmp.rlr[,j])),j]=tmp.mppred[which(is.na(tmp.rlr[,j])),j]
-	tmp.ann=apply(tmp.rlr,1,"mean")/10 #cm
-	tmp.yrs=unique(floor(tmp[,1]))
-	tmp.ann=tmp.ann-mean(tmp.ann[which(is.element(tmp.yrs,1970:1999))],na.rm=T)
-	norwaydata[i]=list(data.frame(years=tmp.yrs,data=tmp.ann))
-	}
-	
-
[…]

### 2017-01-31 code/BergenDecisions/BergenDamageData (Autosaved).R

@@ -1,467 +1,0 @@
-### Decision analysis for Bergen. All the data files are located in the same
-### directory as this script. The R working directory should thus be set
-### to the current directory. 
-
-rm(list=ls())
-
-### Read in damage data obtained from Finance Norway's website:
-### www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/
-A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
-A <- A[-37,] ## Remove 2016 as data not complete
-V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
-V <- V[-37,] ## Remove 2016 as data not complete
-
-### Read in consumer price index (CPI).
-CPI <- read.csv("KPI.csv", header=TRUE)
-CPI <- CPI[-1,] ## Remove 2016 as data not yet available
-CPI$CPI <- CPI$CPI/CPI$CPI[1] ## Normalize to 2015 
-
-### Combine Hordaland and Rogaland data.
-### Standardize costs to 2015 values using CPI. 
-sf.nr <- c(A$Hordaland, A$Rogaland)
-sf.damage <- c(V$Hordaland/CPI$CPI[1:36], V$Rogaland/CPI$CPI[1:36])
-sf.damage <- sf.damage/1e3 ## Normalize to million NOK. 
-
-### Fit Burr distribution to damage data (in MNOK)  
-library(actuar)
-ind <- which(sf.damage==0)
-sf.damage[ind] <- 1e-04 ## Improves numerical stability of estimates. 
-log.lik <- function(theta)
-{
-  res <- sum(dburr(sf.damage, 
-                   shape1=theta[1], 
-                   shape2=theta[2], 
-                   rate=theta[3], 
-                   log=TRUE))
-  return(-res)
-}
-theta.start <- c(5, 0.5, 0.05)
-optim.out <- optim(theta.start, 
-                   fn=log.lik, 
-                   method="BFGS", 
-                   control=list(maxit=1000))
-
-### Get Hallegatte et al. (2013) data on relation between change in damage 
-### cost and change in sea level (csv files copied manually from
-### online supplementary information).   
-E <- read.csv("EuropeanLossData.csv", header=TRUE)
-E[,2:4] <- E[,2:4]/E[,2] 
-
[…]

### 2017-01-31 code/BergenDecisions/BergenDamageData.R

@@ -1,466 +1,75 @@
-### Decision analysis for Bergen. All the data files are located in the same
-### directory as this script. The R working directory should thus be set
-### to the current directory. 
+### Decision analysis for Bergen
+### Data files are located in the folder "data"
+### R working directory should be set to current directory 
 
 rm(list=ls())
 
-### Read in damage data obtained from Finance Norway's website:
-### www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/
-A <- read.csv("StormfloYearFylkeAntall.csv", header=TRUE)
-A <- A[-37,] ## Remove 2016 as data not complete
-V <- read.csv("StormfloYearFylkeVerdi.csv", header=TRUE)
-V <- V[-37,] ## Remove 2016 as data not complete
+library(actuar)
+library(RgoogleMaps)
+source("getDamageData.R")
+source("fitBurrDist.R")
+source("getDamageMult.R")
+source("getHallegatteData.R")
+source("getChangeSlopes.R")
+source("getDamageTrajectories.R")
+source("getDiscountRate.R")
+source("getAdaptationCost.R")
+source("getUncertaintyTrajectories.R")
+source("plotFigure1.R")
+source("plotFigure2.R")
+source("plotFigure3.R")
+source("plotFigure4.R")
+source("plotFigure5a.R")
+source("plotFigure5b.R")
 
-### Read in consumer price index (CPI).
-CPI <- read.csv("KPI.csv", header=TRUE)
-CPI <- CPI[-1,] ## Remove 2016 as data not yet available
-CPI$CPI <- CPI$CPI/CPI$CPI[1] ## Normalize to 2015 
-
-### Combine Hordaland and Rogaland data.
-### Standardize costs to 2015 values using CPI. 
-sf.nr <- c(A$Hordaland, A$Rogaland)
-sf.damage <- c(V$Hordaland/CPI$CPI[1:36], V$Rogaland/CPI$CPI[1:36])
-sf.damage <- sf.damage/1e3 ## Normalize to million NOK. 
+### Read in yearly damage data from Rogaland and Hordaland 1980 to 2015
+sf.damage <- getDamageData()
 
 ### Fit Burr distribution to damage data (in MNOK)  
-library(actuar)
-ind <- which(sf.damage==0)
-sf.damage[ind] <- 1e-04 ## Improves numerical stability of estimates. 
-log.lik <- function(theta)
-{
[…]

### 2017-01-31 code/BergenDecisions/fitBurrDist.R

@@ -1,0 +1,29 @@
+### Fit a Burr distribution to the damage data (in MNOK)
+### Uses functions from the package actuar
+
+fitBurrDist <- function(sf.damage, theta.start=c(5, 0.5, 0.05))
+{
+    ## Improve numerical stability of estimates by replacing zeros with a small value 
+    ind <- which(sf.damage==0)
+    sf.damage[ind] <- 1e-04
+
+    ## The negative log likelihood function to be optimized
+    log.lik <- function(theta)
+    {
+        res <- sum(dburr(sf.damage, 
+                         shape1=theta[1], 
+                         shape2=theta[2], 
+                         rate=theta[3], 
+                         log=TRUE))
+        return(-res)
+    }
+    
+    ## Optimize the loss function
+    optim.out <- optim(theta.start, 
+                       fn=log.lik, 
+                       method="BFGS", 
+                       control=list(maxit=1000))
+
+    ## Return parameter estimates
+    return(optim.out$par)
+}

### 2017-01-31 code/BergenDecisions/getAdaptationCost.R

@@ -1,0 +1,15 @@
+### Adaptation cost estimate was 1000 MNOK in 2009.
+### Normalized to 2015 values using CPI
+
+getAdaptationCost <- function(years=1980:2015, normYear=2015)
+{
+    CPI <- read.csv("data/KPI.csv", header=TRUE)
+    ind <- which(CPI[,1] %in% years) 
+    CPI <- CPI[ind,]
+    ind <- which(CPI[,1]==normYear)
+    CPI$CPI <- CPI$CPI/CPI$CPI[ind]
+
+    ind <- which(CPI[,1]==2009)
+
+    return(1000/CPI$CPI[ind])
+}

### 2017-01-31 code/BergenDecisions/getChangeSlopes.R

@@ -1,0 +1,24 @@
+### Hallegatte et al. (2013) provide data for 0, 20, 40 cm sea level rise.
+### Conservative extrapolation: Linear extrapolation of [20,40] beyond 40
+### and [0,20] below 0. 
+
+getChangeSlopes <- function(E)
+{
+    x <- c(0, 20, 40)
+    N <- dim(E)[1]
+    low.slope <- rep(NA, N)
+    high.slope <- rep(NA, N)
+    for(i in 1:N)
+    {
+        y <- as.double(E[i,2:4])
+        res <- lm(y[1:2]~x[1:2])
+        low.slope[i] <- res$coef[2]
+        res <- lm(y[2:3]~x[2:3])
+        high.slope[i] <- res$coef[2]
+    }
+
+    slopes <- list()
+    slopes$low <- low.slope
+    slopes$high <- high.slope
+    return(slopes)
+}

### 2017-01-31 code/BergenDecisions/getDamageData.R

@@ -1,0 +1,29 @@
+### Read in damage data obtained from Finance Norway's website:
+### www.finansnorge.no/statistikk/skadeforsikring/Naturskadestatistikk-NASK/
+### Standardize costs using CPI
+### Return standardized costs in MNOK
+
+getDamageData <- function(years=1980:2015, normYear=2015)
+{
+    ## Read in damage data 
+    V <- read.csv("data/StormfloYearFylkeVerdi.csv", header=TRUE) 
+    ## Select appropriate years
+    ind <- which(V[,1] %in% years) 
+    V <- V[ind,] 
+    
+    ## Read in consumer price index (CPI).
+    CPI <- read.csv("data/KPI.csv", header=TRUE)
+    ## Select same years are above
+    ind <- which(CPI[,1] %in% years) 
+    CPI <- CPI[ind,]
+    ind <- which(CPI[,1]==normYear)
+    CPI$CPI <- CPI$CPI/CPI$CPI[ind] ## Normalize to normYear 
+
+    ## Combine Hordaland and Rogaland data.
+    ## Standardize costs to normYear values using CPI. 
+    sf.damage <- c(V$Hordaland/CPI$CPI, V$Rogaland/CPI$CPI)
+    ## Normalize to MNOK 
+    sf.damage <- sf.damage/1e3 
+
+    return(sf.damage)
+}

### 2017-01-31 code/BergenDecisions/getDamageMult.R

@@ -1,0 +1,8 @@
+### Function that calculates the multiplicative change in damage for sea level x 
+### and Hallegatte et al. (2013) estimates for city k.  
+getDamageMult <- function(x, k, E=E, slopes=slopes)
+{
+    if(x < 0) return(1/(1 + abs(x) * slopes$low[k]))
+    else if(x < 20) return(1 + x * slopes$low[k])
+    else return(as.double(E[k,3]) + x * slopes$high[k])
+}

### 2017-01-31 code/BergenDecisions/getDamageTrajectories.R

@@ -1,0 +1,89 @@
+### Sample damage distributions for each year 2016-2100 (85 years)
+### under various settings. 
+
+### Calculates three different types of trajectories
+### -- No adaptation
+### -- No sea level change
+### -- Adaptation in year k for all k 
+
+getDamageTrajectories <- function(E, slopes, adaptation.cost)
+{
+    ## Read in sea level rise projections for Bergen. 
+    load("data/Simulation.Rdata")
+    I <- dim(sim)[1]
+    
+    ## Get discount rates 
+    discount.rate <- getDiscountRate()
+    accum.discount.rate <- cumprod(discount.rate)
+
+    ## Sample I iid damage costs for each year 2016-2100
+    ## This sample is used for all subsequent scenarios 
+    orig.damage <- array(rburr(I*85, 
+                               shape1=par.est[1], 
+                               shape2=par.est[2], 
+                               rate=par.est[3]), 
+                         dim=c(I, 85))
+    
+    ## Sample I damage change profiles
+    ## This sample is used for all subsequent scenarios 
+    damage.scenario <- sample(1:15, I, replace=TRUE)
+
+    ## Combine damage costs, damage change profiles and sea level
+    ## projections 
+    yearly.damage <- array(NA, dim=c(I, 85))
+    for(i in 1:I)
+    {
+        for(j in 1:85)
+        {
+            yearly.damage[i,j] <- orig.damage[i,j] * getDamageMult(sim[i, (j+16)],
+                                                                   damage.scenario[i])
+        }
+        yearly.damage[i,] <- yearly.damage[i,]/accum.discount.rate
+    }
+    
+    ## Combine damage costs, damage change profiles and sea level
+    ## projections assuming all subsequent years are like 2016. 
+    constant.damage <- array(NA, dim=c(I, 85))
+    for(i in 1:I)
+    {
+        for(j in 1:85)
+        {
+            constant.damage[i,j] <- orig.damage[i,j] * getDamageMult(sim[i, 16], 
+                                                                     damage.scenario[i])
+        }
+        constant.damage[i,] <- constant.damage[i,]/accum.discount.rate
+    }
+    
+    ## Combine damage costs, damage change profiles and sea level
+    ## projections under adaptation. Adaptation measure is assumed
+    ## to reduce 50% of the yearly damages to damages under 75 cm
[…]

### 2017-01-31 code/BergenDecisions/getDiscountRate.R

@@ -1,0 +1,6 @@
+### Discount rate based on Norwegian government recommendations.
+
+getDiscountRate <- function()
+{
+    return( c( rep(1.04, 40), rep(1.03, 35), rep(1.02, 10)) )
+}

### 2017-01-31 code/BergenDecisions/getHallegatteData.R

@@ -1,0 +1,11 @@
+### Get Hallegatte et al. (2013) data on relation between change in damage 
+### cost and change in sea level
+### 
+### Data in csv file is copied manually from online supplementary information  
+
+getHallegatteData <- function()
+{
+    E <- read.csv("data/EuropeanLossData.csv", header=TRUE)
+    E[,2:4] <- E[,2:4]/E[,2]
+    return(E)
+}

### 2017-01-31 code/BergenDecisions/getUncertaintyTrajectories.R

@@ -1,0 +1,89 @@
+### Investigation of the effects of uncertainty (no adaptation)
+
+getUncertaintyTrajectories <- function(E, slopes, orig.damage, damage.scenario)
+{
+    ## Cumulative damage with medians
+
+    ## Median damage scenario 
+    z <- c(0, 20, 40)
+    y <- apply(E[,2:4], 2, median)
+    res <- lm(y[1:2]~z[1:2])
+    low.med.slope <- res$coef[2]
+    res <- lm(y[2:3]~z[2:3])
+    high.med.slope <- res$coef[2]
+    
+    get.median.damage.mult <- function(x)
+    {
+        if(x < 0) return(1/(1 + abs(x) * low.med.slope))
+        else if(x < 20) return(1 + x * low.med.slope)
+        else return(y[2] + x * high.med.slope)
+    }
+
+    ## Read in sea level rise projections for Bergen. 
+    load("data/Simulation.Rdata")
+    I <- dim(sim)[1]
+    
+    ## Get discount rates 
+    discount.rate <- getDiscountRate()
+    accum.discount.rate <- cumprod(discount.rate)
+
+    ## Median sea level rise
+    median.sim <- apply(sim, 2, median)[17:101]
+
+    ## Median damage cost
+    median.damage <- apply(orig.damage, 2, median)
+
+    ## Putting the three together
+    yearly.median.damage <- rep(NA, 85)
+    for(j in 1:85)
+    {
+        yearly.median.damage[j] <- median.damage[j] * get.median.damage.mult(median.sim[j]) 
+    }
+    yearly.median.damage <- yearly.median.damage/accum.discount.rate
+    
+    ## Uncertainty in damage mult. fct. only 
+    yearly.mf.damage <- array(NA, dim=c(I, 85))
+    for(i in 1:I)
+    {
+        for(j in 1:85)
+        {
+            yearly.mf.damage[i,j] <- median.damage[j] * getDamageMult(median.sim[j],
+                                                                      damage.scenario[i],
+                                                                      E, slopes)
+        }
+        yearly.mf.damage[i,] <- yearly.mf.damage[i,]/accum.discount.rate
+    }
+    total.mf.damage <- apply(yearly.mf.damage,1,sum)
+    
+    ## Uncertainty in slr only 
+    yearly.slr.damage <- array(NA, dim=c(I, 85))
+    for(i in 1:I)
[…]

### 2017-01-31 code/BergenDecisions/plotFigure1.R

@@ -1,0 +1,14 @@
+## Plots showing the underlying data used in the adaptation decision analysis
+
+plotFigure1 <- function(sf.damage, par.est, ifPdf=TRUE, fileName="figures/DecisionAnalysisBergenPartI.pdf")
+{
+    if(ifPdf) pdf(file=fileName, width=5, height=5, points=12)
+    ## (a) Plot of damage data and fitted distribution
+    par(mex=0.75, mar=c(5,4,2,2)+0.1)
+    hist(sf.damage, main="", xlab="Amount (million NOK)", col="gray60", breaks=50,  xlim=c(0,65),
+         freq=FALSE)
+    x <- seq(0, 65, by=0.01)
+    y <- dburr(x, shape1=par.est[1], shape2=par.est[2], rate=par.est[3])
+    lines(x,y,col="red",lwd=2)
+    if(ifPdf) dev.off()
+}

### 2017-01-31 code/BergenDecisions/plotFigure2.R

@@ -1,0 +1,33 @@
+## Plots showing the underlying data used in the adaptation decision analysis
+
+plotFigure2 <- function(E, ifPdf=TRUE, fileName="figures/DecisionAnalysisBergenPartII.pdf")
+{
+    if(ifPdf) pdf(file="figures/DecisionAnalysisBergenPartII.pdf", width=5, height=5, points=12)
+    par(mex=0.75, mar=c(5,4,2,2)+0.1)
+    ## (b) Plot of extrapolated damage change profiles 
+    x <- c(0,20,40)
+    plot(x, E[1, 2:4], type="o", col="black", xlim=c(-100,130), ylim=c(-180,1000),
+         xlab="Sea level anomaly (cm)", ylab="Relative mean annual damage", main="", axes=FALSE)
+    box()
+    axis(1)
+    axis(2, at=c(-200, 1, 200, 400, 600, 800, 1000), labels=c(0.005, 1, 200, 400, 600, 800, 1000))
+    z <- c(40,130)
+    w <- c(-100,0)
+    for(i in 1:15)
+    {
+        lines(x, E[i, 2:4], type="o", col="black")
+        res <- lm(as.double(E[i,3:4])~x[2:3])
+        lines(z, res$coef[1] + z * res$coef[2], col="gray70")
+        res <- lm(as.double(E[i,2:3])~x[1:2])
+        lines(w, res$coef[1] + w * res$coef[2], col="gray70")
+    }
+    points(20, median(E[,3]), col="red", pch=15)
+    points(40, median(E[,4]), col="red", pch=15)
+    res <- lm(apply(E[,2:3],2,median)~x[1:2])
+    w <- c(-100, 20)
+    lines(w, res$coef[1] + w * res$coef[2], col="red", lwd=2)
+    res <- lm(apply(E[,3:4],2,median)~x[2:3])
+    z <- c(20,130)
+    lines(z, res$coef[1] + z * res$coef[2], col="red", lwd=2)
+    if(ifPdf) dev.off()
+}

### 2017-01-31 code/BergenDecisions/plotFigure3.R

@@ -1,0 +1,82 @@
+## Plot comparing accumulated damage costs under various scenarios
+
+plotFigure3 <- function(res, ifPdf=TRUE, fileName="figures/CumDamageCostsBergen.pdf")
+{
+    I <- 10000
+
+    ## Calculate cumulative damages over 2016-2100
+    ## We present the distribution through median and 90% interval
+
+    ## Cumulated additional damage due to sea level rise.
+    cumsum.yearly.damage <- array(NA, dim=dim(res$yearly))
+    cumsum.constant.damage <- array(NA, dim=dim(res$yearly))
+    for(i in 1:dim(res$yearly)[1]) 
+    {
+        cumsum.yearly.damage[i,] <- cumsum(res$yearly[i,])
+        cumsum.constant.damage[i,] <- cumsum(res$constant[i,])
+    }
+
+    upper.95.yearly.cumsum <- apply(cumsum.yearly.damage,2,quantile,0.95)
+    lower.5.yearly.cumsum <- apply(cumsum.yearly.damage,2,quantile,0.05)
+    median.yearly.cumsum <- apply(cumsum.yearly.damage,2,quantile,0.5)
+    
+    upper.95.constant.cumsum <- apply(cumsum.constant.damage,2,quantile,0.95)
+    lower.5.constant.cumsum <- apply(cumsum.constant.damage,2,quantile,0.05)
+    median.constant.cumsum <- apply(cumsum.constant.damage,2,quantile,0.5)
+
+    ## Cumulated damage for various adaptation options
+    cumsum.adapt.damage <- array(NA, dim=dim(res$adapted))
+    for(i in 1:85)
+    {
+        for(j in 1:I)
+        {
+            cumsum.adapt.damage[i,j,] <- cumsum(res$adapted[i,j,])
+        }
+    }
+    cumsum.yearly.damage <- array(NA, dim=dim(res$yearly))
+    for(i in 1:I)
+        cumsum.yearly.damage[i,] <- cumsum(res$yearly[i,])
+    
+    ## Calculate summary statistics for total damages for various
+    ## adaptation options
+    adapt.median <- apply(cumsum.adapt.damage[,,85],1,median)
+    adapt.upper.95 <- apply(cumsum.adapt.damage[,,85],1,quantile,0.95)
+    adapt.lower.05 <- apply(cumsum.adapt.damage[,,85],1,quantile,0.05)
+
+    ## Cumulated damage due to sea level rise with
[…]

### 2017-01-31 code/BergenDecisions/plotFigure4.R

@@ -1,0 +1,62 @@
+## Plot comparing total damage cost under different adaptation timings
+
+plotFigure4 <- function(res, ifPdf=TRUE, fileName="figures/TotalDamageCostsAdaptation.pdf")
+{
+    I <- 10000
+
+    ## Calculate cumulative damages over 2016-2100
+    ## We present the distribution through median and 90% interval
+
+    ## Cumulated additional damage due to sea level rise.
+    cumsum.yearly.damage <- array(NA, dim=dim(res$yearly))
+    for(i in 1:dim(res$yearly)[1]) 
+    {
+        cumsum.yearly.damage[i,] <- cumsum(res$yearly[i,])
+    }
+    
+    ## Cumulated damage for various adaptation options
+    cumsum.adapt.damage <- array(NA, dim=dim(res$adapted))
+    for(i in 1:85)
+    {
+        for(j in 1:I)
+        {
+            cumsum.adapt.damage[i,j,] <- cumsum(res$adapted[i,j,])
+        }
+    }
+    cumsum.yearly.damage <- array(NA, dim=dim(res$yearly))
+    for(i in 1:I)
+        cumsum.yearly.damage[i,] <- cumsum(res$yearly[i,])
+    
+    ## Calculate summary statistics for total damages for various
+    ## adaptation options
+    adapt.median <- apply(cumsum.adapt.damage[,,85],1,median)
+    adapt.upper.95 <- apply(cumsum.adapt.damage[,,85],1,quantile,0.95)
+    adapt.lower.05 <- apply(cumsum.adapt.damage[,,85],1,quantile,0.05)
+
+    ## Make the plot
+    
+    if(ifPdf) pdf(file=fileName, width=10, height=5, points=12)
+    par(mex=0.75)
+    a <- median(cumsum.yearly.damage[,85])/1e3
+    plot(c(2016,2100), c(a,a), type="l", ylim=c(0,14), xlab="Year of adaptation measure",
+         ylab="Total damage 2016-2100 (billion NOK)", main="")
+    year <- (2016:2100)
+    k <- 0.35
+    for(i in 1:85)
+    {
+        a <- year[i]
+        b <- adapt.upper.95[i]/1e3
+        c <- adapt.lower.05[i]/1e3
+        polygon(c(a-k, a+k, a+k, a-k), c(c, c, b, b), col="gray80", border="gray80")
[…]

### 2017-01-31 code/BergenDecisions/plotFigure5a.R

@@ -1,0 +1,29 @@
+### Plot comparing distributions of total damage for various uncertainty settings
+
+plotFigure5a <- function(res.unc, total.damage, ifPdf=TRUE, fileName="figures/Uncertainty.pdf")
+{
+    total.mf.damage <- res.unc$mf
+    total.slr.damage <- res.unc$slr
+    total.dam.damage <- res.unc$dam
+
+    if(ifPdf) pdf(file=fileName, width=10, height=5, points=12)
+    par(mex=0.75, mar=c(5,4,2,2)+0.1)
+    hist(total.damage/1e3, freq=FALSE, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
+         col="black", border="black", xlab="Total damage 2016-2100 (billion NOK)",
+         ylab="Density", xaxs="i", yaxs="i", main="") # black
+    hist(total.slr.damage/1e3, freq=FALSE, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
+         col="#7D26CD50", border="#7D26CD", add=TRUE) # purple
+    hist(total.mf.damage/1e3, freq=FALSE, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
+         col="#4876FF50", border="#4876FF", add=TRUE) # blue
+    hist(total.dam.damage/1e3, freq=FALSE, ylim=c(0,5), xlim=c(0,10), breaks=seq(0,170,by=0.05),
+         col="#008B4550", border="#008B45", add=TRUE) # green
+    abline(v=median(total.damage/1e3), col="black", lwd=3)
+    abline(v=median(total.slr.damage/1e3), col="#7D26CD", lwd=3)
+    abline(v=median(total.mf.damage/1e3), col="#4876FF", lwd=3)
+    abline(v=median(total.dam.damage/1e3), col="#008B45", lwd=3)
+    abline(v=sum(res.unc$yearly.median/1e3), col="yellow", lwd=3)
+    legend("topright", legend=c("Full uncertainty","SLR uncertainty","Effect uncertainty","Damage uncertainty", "No uncertainty"),
+           col=c("black", "#7D26CD", "#4876FF", "#008B45", "yellow"), lty=1, lwd=2)
+    box()
+    if(ifPdf) dev.off()
+}

### 2017-01-31 code/BergenDecisions/plotFigure5b.R

@@ -1,0 +1,39 @@
+### Plot comparing distributions of total damage for various uncertainty settings on a log-log scale
+
+plotFigure5b <- function(res.unc, total.damage, ifPdf=TRUE, fileName="figures/UncertaintyLog.pdf")
+{
+    total.mf.damage <- res.unc$mf
+    total.slr.damage <- res.unc$slr
+    total.dam.damage <- res.unc$dam
+
+    if(ifPdf) pdf(file="figures/UncertaintyLog.pdf", width=10, height=5, points=12)
+    par(mex=0.75, mar=c(5,4,2,2)+0.1)
+    buckets <- seq(log(50),log(165000), by=0.1)
+    buckets <- exp(buckets)
+    my.hist <- hist(total.damage, breaks=buckets, plot=FALSE)
+    my.slr.hist <- hist(total.slr.damage, breaks=buckets, plot=FALSE)
+    my.mf.hist <- hist(total.mf.damage, breaks=buckets, plot=FALSE)
+    my.dam.hist <- hist(total.dam.damage, breaks=buckets, plot=FALSE)
+    plot(log(my.slr.hist$breaks[-1]), log(my.slr.hist$counts), type="h", col="#7D26CD",
+         main="", ylab="Log frequency", xlab="Total damage 2016-2100 (million NOK)",
+         lwd=2, axes=FALSE)
+    ticks <- c(50, 150, 500, 1500, 5000, 15000, 50000, 150000)
+    axis(1, at = log(ticks), labels=ticks)
+    axis(2)
+    box()
+    lines(log(my.hist$breaks[-1])+0.02, log(my.hist$counts), col="black", type="h", lwd=2)
+    lines(log(my.mf.hist$breaks[-1])+0.04, log(my.mf.hist$counts), col="orange", type="h", lwd=2)
+    lines(log(my.dam.hist$breaks[-1])+0.06, log(my.dam.hist$counts), col="#008B45", type="h", lwd=2)
+    abline(v=log(sum(res.unc$yearly.median)),  col="gray50", lwd=2)
+    points(log(median(total.damage)), 7.8, col="black", pch=16)
+    points(log(median(total.slr.damage)),7.8,  col="#7D26CD", pch=16)
+    points(log(median(total.dam.damage)),7.8,  col="#008B45", pch=16)
+    points(log(sum(res.unc$yearly.median)),7.8,  col="gray50", pch=16)
+    points(log(median(total.mf.damage)),7.8,  col="orange", pch=16)
[…]

### 2017-01-31 code/Esbjerg.R

@@ -1,160 +1,0 @@
-setwd("~/GitHub/SeaLevelDecisions/code")
-require(maps)
-require(maptools)
-require(ncdf4)
-require(forecast)
-require(excursions)
-require(Matrix)
-require(chron)
-require(zoo)
-source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
-source('~/GitHub/SeaLevelDecisions/code/ci_bergen.R')
-denmark=read.table("~/GitHub/SeaLevelProjections/denmark_sea_level/denmark.PSMSL.txt")#All Danish sea level stations with current data]
-#from PSMSL.org
-denmark[,1]=as.character(denmark[,1])
-
-denmarkurl=paste("http://www.psmsl.org/data/obtaining/rlr.monthly.data/",denmark[,2],".rlrdata",sep="")
-
-N=length(denmark[,2])
-denmarkdata=list(N,NULL)
-for(i in 1:N){
-
-	tmp=read.table(denmarkurl[i],na.strings="-99999",sep=";")
-	tmp=cbind(floor(tmp[,1]),round(12*(tmp[,1]-floor(tmp[,1]))+.5),tmp[,2:4])
-
-	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	#if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
-	NN=dim(tmp)[1]
-	if(tmp[NN,2]<12) tmp=tmp[-((NN+1-tmp[NN,2]):N),]#delete final year if not full
-	tmp.rlr=tmp[,3]
-	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
-	tmp.mp=medpolish(tmp.rlr,na.rm=T)
-	tmp.mppred=(tmp.mp$overall+outer(tmp.mp$row,tmp.mp$col,"+"))
-	for(j in 1:12) tmp.rlr[which(is.na(tmp.rlr[,j])),j]=tmp.mppred[which(is.na(tmp.rlr[,j])),j]
-	tmp.ann=apply(tmp.rlr,1,"mean")/10 #cm
-	tmp.yrs=unique(floor(tmp[,1]))
-	tmp.ann=tmp.ann-mean(tmp.ann[which(is.element(tmp.yrs,1970:1999))],na.rm=T)
-	denmarkdata[i]=list(data.frame(years=unique(floor(tmp[,1])),data=tmp.ann))
-	}
-	
[…]

### 2017-01-31 code/SeaLevelProjections/Bergen.R

@@ -1,0 +1,176 @@
+setwd("~/GitHub/SeaLevelDecisions/code")
+require(maps)
+require(maptools)
+require(ncdf4)
+require(forecast)
+require(excursions)
+require(Matrix)
+require(chron)
+require(zoo)
+source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
+source('~/GitHub/SeaLevelDecisions/code/ci_bergen.R')
+norway=read.table("~/GitHub/SeaLevelProjections/Norway_sea_level/Norway.PSMSL.txt")#All Norwegian sea level stations with current data
+#from PSMSL.org
+norway[,1]=as.character(norway[,1])
+
+
+norwayurl=paste("http://www.psmsl.org/data/obtaining/rlr.monthly.data/",norway[,2],".rlrdata",sep="")
+
+N=length(norway[,2])
+norwaydata=list(N,NULL)
+
+for (i in 1:N){
+	tmp=read.table(norwayurl[i],na.strings="-99999",sep=";")
+	tmp=cbind(floor(tmp[,1]),round(12*(tmp[,1]-floor(tmp[,1]))+.5),tmp[,2:4])
+
+	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	NN=dim(tmp)[1]
+	if(tmp[NN,2]<12) tmp=tmp[-((NN+1-tmp[NN,2]):NN),]#delete final year if not full
+	tmp.rlr=tmp[,3]
+	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
+	tmp.mp=medpolish(tmp.rlr,na.rm=T)
+	tmp.mppred=(tmp.mp$overall+outer(tmp.mp$row,tmp.mp$col,"+"))
+	for(j in 1:12) tmp.rlr[which(is.na(tmp.rlr[,j])),j]=tmp.mppred[which(is.na(tmp.rlr[,j])),j]
+	tmp.ann=apply(tmp.rlr,1,"mean")/10 #cm
+	tmp.yrs=unique(floor(tmp[,1]))
+	tmp.ann=tmp.ann-mean(tmp.ann[which(is.element(tmp.yrs,1970:1999))],na.rm=T)
+	norwaydata[i]=list(data.frame(years=tmp.yrs,data=tmp.ann))
+	}
+	
+
[…]

### 2017-01-31 code/SeaLevelProjections/Esbjerg.R

@@ -1,0 +1,160 @@
+setwd("~/GitHub/SeaLevelDecisions/code")
+require(maps)
+require(maptools)
+require(ncdf4)
+require(forecast)
+require(excursions)
+require(Matrix)
+require(chron)
+require(zoo)
+source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
+source('~/GitHub/SeaLevelDecisions/code/ci_bergen.R')
+denmark=read.table("~/GitHub/SeaLevelProjections/denmark_sea_level/denmark.PSMSL.txt")#All Danish sea level stations with current data]
+#from PSMSL.org
+denmark[,1]=as.character(denmark[,1])
+
+denmarkurl=paste("http://www.psmsl.org/data/obtaining/rlr.monthly.data/",denmark[,2],".rlrdata",sep="")
+
+N=length(denmark[,2])
+denmarkdata=list(N,NULL)
+for(i in 1:N){
+
+	tmp=read.table(denmarkurl[i],na.strings="-99999",sep=";")
+	tmp=cbind(floor(tmp[,1]),round(12*(tmp[,1]-floor(tmp[,1]))+.5),tmp[,2:4])
+
+	if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	#if(tmp[1,2]>1) tmp=tmp[-(1:(13-tmp[1,2])),]#start at first full year
+	NN=dim(tmp)[1]
+	if(tmp[NN,2]<12) tmp=tmp[-((NN+1-tmp[NN,2]):N),]#delete final year if not full
+	tmp.rlr=tmp[,3]
+	tmp.rlr=matrix(tmp.rlr,ncol=12,byrow=T)
+	tmp.mp=medpolish(tmp.rlr,na.rm=T)
+	tmp.mppred=(tmp.mp$overall+outer(tmp.mp$row,tmp.mp$col,"+"))
+	for(j in 1:12) tmp.rlr[which(is.na(tmp.rlr[,j])),j]=tmp.mppred[which(is.na(tmp.rlr[,j])),j]
+	tmp.ann=apply(tmp.rlr,1,"mean")/10 #cm
+	tmp.yrs=unique(floor(tmp[,1]))
+	tmp.ann=tmp.ann-mean(tmp.ann[which(is.element(tmp.yrs,1970:1999))],na.rm=T)
+	denmarkdata[i]=list(data.frame(years=unique(floor(tmp[,1])),data=tmp.ann))
+	}
+	
[…]

### 2017-01-31 code/SeaLevelProjections/ci_bergen.R

@@ -1,0 +1,106 @@
+ci_bergen=function(models,sta,stafit,sta.common.years,alpha=.1)
+{
+load('~/GitHub/SeaLevelProjections/data&global.Rdata')#global temperature-sealevel relation, historical data,temperature projections etc.
+ source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
+s1999 = seal$sealevel[which(is.element(seal$years,1999))]
+K = length(models) 
+proj=which(is.element(cmip5$years,2000:2100))
+modelpath=matrix(0,K,101)
+count=0
+for (i in models) {
+  count=count+1
+  modelpath[count,]=cmip5$data[proj,i,4]
+}
+sd1999 =seal$se[which(is.element(seal$years,1999))]
+n.o =  length(sta.common.years)
+n.p=101
+n = n.o+n.p
+#Calculate covariance matrices for MA processes
+
+cov2 = ARMAcov(armodel$ar,n)
+cov1 = ARMAcov(stafit$ar,n)  
+Sigma2 = toeplitz(cov2) 
+Sigma1 = toeplitz(cov1)
+
+#covariance for non-differentiated global mean sea level:
+Sigma.i2 = apply(apply(Sigma2,1,cumsum),1,cumsum)
+
+#Total covariance for station
+Sigma = a4^2*Sigma.i2 + Sigma1
+#add the variance caused by the uncertainty in the starting value:
+
+Sigma = Sigma + diag(rep(sd1999^2,dim(Sigma)[1]))
+
+#add the variance due to gia
+
+times=(0:(n-1))-stafit$adj
+Sigma=Sigma+diag(sta$se.gia^2*times^2,dim(Sigma)[1])
+
+
+#Submatrices for observations and predictions:
+
+Sigma.o = Sigma[1:n.o,1:n.o]
+Sigma.p = Sigma[(n.o+1):n,(n.o+1):n]
+Sigma.op = Sigma[1:n.o,(n.o+1):n]
+
+#Calculate  precision matrix
+R.o = chol(Sigma.o)
+c = mean(seal$sealevel - cumsum(armodel$ar$coef[3] 
++armodel$ar$coef[4]*temperature$temp),na.rm=T)
[…]

### 2017-01-31 code/SeaLevelProjections/scratch.R

@@ -1,0 +1,1 @@
+## This is just an empty file for GitHub to include this folder...

### 2017-01-31 code/ci_bergen.R

@@ -1,106 +1,0 @@
-ci_bergen=function(models,sta,stafit,sta.common.years,alpha=.1)
-{
-load('~/GitHub/SeaLevelProjections/data&global.Rdata')#global temperature-sealevel relation, historical data,temperature projections etc.
- source('~/GitHub/SeaLevelProjections/Sealevel_functions.R')
-s1999 = seal$sealevel[which(is.element(seal$years,1999))]
-K = length(models) 
-proj=which(is.element(cmip5$years,2000:2100))
-modelpath=matrix(0,K,101)
-count=0
-for (i in models) {
-  count=count+1
-  modelpath[count,]=cmip5$data[proj,i,4]
-}
-sd1999 =seal$se[which(is.element(seal$years,1999))]
-n.o =  length(sta.common.years)
-n.p=101
-n = n.o+n.p
-#Calculate covariance matrices for MA processes
-
-cov2 = ARMAcov(armodel$ar,n)
-cov1 = ARMAcov(stafit$ar,n)  
-Sigma2 = toeplitz(cov2) 
-Sigma1 = toeplitz(cov1)
-
-#covariance for non-differentiated global mean sea level:
-Sigma.i2 = apply(apply(Sigma2,1,cumsum),1,cumsum)
-
-#Total covariance for station
-Sigma = a4^2*Sigma.i2 + Sigma1
-#add the variance caused by the uncertainty in the starting value:
-
-Sigma = Sigma + diag(rep(sd1999^2,dim(Sigma)[1]))
-
-#add the variance due to gia
-
-times=(0:(n-1))-stafit$adj
-Sigma=Sigma+diag(sta$se.gia^2*times^2,dim(Sigma)[1])
-
-
-#Submatrices for observations and predictions:
-
-Sigma.o = Sigma[1:n.o,1:n.o]
-Sigma.p = Sigma[(n.o+1):n,(n.o+1):n]
-Sigma.op = Sigma[1:n.o,(n.o+1):n]
-
-#Calculate  precision matrix
-R.o = chol(Sigma.o)
-c = mean(seal$sealevel - cumsum(armodel$ar$coef[3] 
-+armodel$ar$coef[4]*temperature$temp),na.rm=T)
[…]

### 2017-01-31 code/scratch.R

@@ -1,1 +1,0 @@
-## This is just an empty file for GitHub to include this folder...

### 2017-02-02 code/BergenDecisions/BergenDamageData.Rmd
html page for decision code

---
title: "Decision analysis for Bergen"
author: "Thordis L. Thorarinsdottir"
output:
    html_document:
    toc: TRUE
---

<hr></hr>
<br></br>

## Optimal timing of sea level rise adaptation measures in Bergen, Norway

Data files are located in the folder "data".
R working directory should be set to current directory.

```{r message=FALSE}
rm(list=ls())

library(actuar)
library(RgoogleMaps)
source("getDamageData.R")
source("fitBurrDist.R")
source("getDamageMult.R")
source("getHallegatteData.R")
source("getChangeSlopes.R")
source("getDamageTrajectories.R")
source("getDiscountRate.R")
source("getAdaptationCost.R")
source("getUncertaintyTrajectories.R")
source("plotFigure1.R")
source("plotFigure2.R")
source("plotFigure3.R")
source("plotFigure4.R")
source("plotFigure5a.R")
source("plotFigure5b.R")
```

Read in yearly damage data (in MNOK) from Rogaland and Hordaland 1980 to 2015.
```{r}
sf.damage <- getDamageData()
round(sf.damage, 3)
```

Fit Burr distribution to damage data.   
```{r warning=FALSE}
par.est <- fitBurrDist(sf.damage)
round(par.est, 3)
```

Plot histogram of damage data and estimated Burr distribution. 
```{r}
plotFigure1(sf.damage, par.est, ifPdf=FALSE)
```

Get normalized data from Hallegatte et al. (2013) on relation between change in damage cost and sea level.  
```{r}
E <- getHallegatteData()
E
```
Extrapolate to any sea level rise. 
```{r}
slopes <- getChangeSlopes(E)
slopes
```

Plot changes in damages as a function of sea level rise with data from Hallegatte et al. (2013) indicated by black circles and the median change indicated with a red line. 
```{r}
plotFigure2(E, ifPdf=FALSE)
```

Get adaptation cost for the construction of two inner barriers from Grief Foundation (2009) (in MNOK at 2015 level).   
```{r}
adaptation.cost <- getAdaptationCost() 
adaptation.cost
```

Create damage distributions for 2016-2100 under various scenarios. 
```{r}
## res <- getDamageTrajectories(E, slopes, adaptation.cost) 
## save(res, file="data/calculatedDamage.RData")
load("data/calculatedDamage.RData")
```

Create plot that compares accumulated damage costs under various scenarios: Median projected cumulative damage costs under constant sea level (gray line), under sea level rise according to RCP 8.5 with no adaptation (black line) and with adaptation finished in 2047 (red line). The shaded area denotes the 90% projection interval under constant sea level.  Dotted lines indicate 90% projection intervals with sea level rise according to RCP 8.5.   
```{r}
plotFigure3(res, ifPdf=FALSE) 
```

Plot comparing total damage cost 2016-2100 under different adaptation timings, showing median total costs (red lines) and 90% projection intervals. Median projected total damage under no adaptation is indicated with a black line and the associated 90% projection interval with black dotted lines. 
```{r}
plotFigure4(res, ifPdf=FALSE) 
```

Investigation of the effects of uncertainty (no adaptation)
```{r}
res.unc <- getUncertaintyTrajectories(E=E, slopes=slopes, orig.damage=res$orig, damage.scenario=res$scenario)
total.damage <- apply(res$yearly, 1, sum)
summary(total.damage)
```

Plot comparing distributions of total damage for various uncertainty settings
```{r}
plotFigure5b(res.unc, total.damage, ifPdf=FALSE)
```

### 2017-02-02 code/BergenDecisions/BergenDamageData.Rmd
Minor changes to Rmd file

@@
----
-title: "Decision analysis for Bergen"
-author: "Thordis L. Thorarinsdottir"
-output:
-    html_document:
-    toc: TRUE
----
-
-<hr></hr>
-<br></br>
-
 ## Optimal timing of sea level rise adaptation measures in Bergen, Norway
+### Author: Thordis L. Thorarinsdottir
 
@@
 R working directory should be set to current directory.
+
+To run the analysis, run the main script BergenDamageData.R as shown below.
 
@@
 
-Investigation of the effects of uncertainty (no adaptation)
+Investigation of the effects of uncertainty (no adaptation).
 ```{r}
@@
 
-Plot comparing distributions of total damage for various uncertainty settings
+Plot comparing distributions of total damage for various uncertainty settings.
 ```{r}

### 2017-02-02 techreport/paper.tex
Updated readme files

@@
 This work was funded by NordForsk through project number 74456 ``Statistical Analysis of Climate Projections'' (eSACP) and The Research Council of Norway through project number 243953 ``Physical and Statistical Analysis of Climate Extremes in Large Datasets'' (ClimateXL).
-The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions/Code}.
+The source code for the analysis is implemented in the statistical programming language {\tt R} (\citet{R}) and is available on GitHub at \url{http://github.com/eSACP/SeaLevelDecisions}.
 

### 2017-05-15 code/BergenDecisions/BergenDamageData.Rmd
Updated code for Bergen case

@@
 source("plotFigure4.R")
-source("plotFigure5a.R")
-source("plotFigure5b.R")
+source("plotFigure5.R")
 ```
@@
 ```
-
-Fit Burr distribution to damage data.   
-```{r warning=FALSE}
-par.est <- fitBurrDist(sf.damage)
-round(par.est, 3)
-```
-
-Plot histogram of damage data and estimated Burr distribution. 
-```{r}
-plotFigure1(sf.damage, par.est, ifPdf=FALSE)
-```
-
 Get normalized data from Hallegatte et al. (2013) on relation between change in damage cost and sea level.  
@@
 ```
-
 Plot changes in damages as a function of sea level rise with data from Hallegatte et al. (2013) indicated by black circles and the median change indicated with a red line. 
@@
 plotFigure2(E, ifPdf=FALSE)
+```
+
+Fit Burr distribution to damage data.   
+```{r warning=FALSE}
+par.est.res <- fitBurrDist(sf.damage, E)   
+par.est <- par.est.res$par ## parameter estimates
+round(par.est, 3)
+mult <- par.est.res$mult ## multiplicative damage data corrections
+round(mult, 3)
+```
+
+Plot histogram of damage data and estimated Burr distribution. 
+```{r}
+plotFigure1(sf.damage/mult, par.est, ifPdf=FALSE)
 ```
@@
 ```{r}
-## res <- getDamageTrajectories(E, slopes, adaptation.cost) 
-## save(res, file="data/calculatedDamage.RData")
-load("data/calculatedDamage.RData")
+## res26 <- getDamageTrajectories(E, par.est, slopes, adaptation.cost, inputfile="data/Simulation_2_6.Rdata") 
+## save(res26, file="data/calculatedDamage_nonstat26.RData")
+## res45 <- getDamageTrajectories(E, par.est, slopes, adaptation.cost, inputfile="data/Simulation_4_5.Rdata") 
+## save(res45, file="data/calculatedDamage_nonstat45.RData")
+## res <- getDamageTrajectories(E, par.est, slopes, adaptation.cost, inputfile="data/Simulation_8_5.Rdata") 
+## save(res, file="data/calculatedDamage_nonstat.RData")
+load("data/calculatedDamage_nonstat.RData")
+load("data/calculatedDamage_nonstat26.RData")
+load("data/calculatedDamage_nonstat45.RData")
 ```
 
-Create plot that compares accumulated damage costs under various scenarios: Median projected cumulative damage costs under constant sea level (gray line), under sea level rise according to RCP 8.5 with no adaptation (black line) and with adaptation finished in 2047 (red line).
-The shaded area denotes the 90% projection interval under constant sea level.
+Create plot that compares accumulated damage costs under various scenarios: Median projected cumulative damage costs under constant sea level (blue line), under sea level rise according to RCP 8.5 with no adaptation (black line) and with adaptation finished in 2047 (red line).
+The blue shaded area denotes the 90% projection interval under constant sea level.
 Dotted lines indicate 90% projection intervals with sea level rise according to RCP 8.5.
@@
 
-Plot comparing total damage cost 2016-2100 under different adaptation timings, showing median total costs (red lines) and 90% projection intervals.
+Plot comparing total damage cost 2016-2100 under different adaptation timings, showing median total costs (blue lines) and 90% projection intervals.
 Median projected total damage under no adaptation is indicated with a black line and the associated 90% projection interval with black dotted lines.
@@
 ```{r}
-res.unc <- getUncertaintyTrajectories(E=E, slopes=slopes, orig.damage=res$orig, damage.scenario=res$scenario)
+res.unc <- getUncertaintyTrajectories(E=E,
+                                      slopes=slopes,
+                                      orig.damage=res$orig,
+                                      damage.scenario=res$scenario,
+                                      inputfile="data/Simulation_8_5.Rdata")
 total.damage <- apply(res$yearly, 1, sum)
 summary(total.damage)
+res.unc26 <- getUncertaintyTrajectories(E=E,
+                                        slopes=slopes,
+                                        orig.damage=res26$orig,
+                                        damage.scenario=res26$scenario,
+                                        inputfile="data/Simulation_2_6.Rdata")
+total.damage26 <- apply(res26$yearly, 1, sum)
+summary(total.damage26)
+res.unc45 <- getUncertaintyTrajectories(E=E,
+                                        slopes=slopes,
+                                        orig.damage=res45$orig,
+                                        damage.scenario=res45$scenario,
+                                        inputfile="data/Simulation_4_5.Rdata")
+total.damage45 <- apply(res45$yearly, 1, sum)
+summary(total.damage45)
 ```
@@
 ```{r}
-plotFigure5b(res.unc, total.damage, ifPdf=FALSE)
+plotFigure5(res.unc, total.damage, res.unc45, total.damage45, res.unc26, total.damage26, ifPdf=FALSE, fileName="figures/UncertaintyLog.pdf")
 ```

### 2017-05-15 code/BergenDecisions/.#fitBurrDist.R

@@ -1,0 +1,1 @@
+disa@disa-VirtualBox.6190:1494832380

### 2017-05-15 code/BergenDecisions/BergenDamageData.R

@@ -19,17 +19,10 @@ source("plotFigure1.R")
 source("plotFigure2.R")
 source("plotFigure3.R")
 source("plotFigure4.R")
-source("plotFigure5a.R")
-source("plotFigure5b.R")
+source("plotFigure5.R")
 
 ### Read in yearly damage data from Rogaland and Hordaland 1980 to 2015
 sf.damage <- getDamageData()
-
-### Fit Burr distribution to damage data (in MNOK)  
-par.est <- fitBurrDist(sf.damage)
-
-### Plot showing the damage data 
-plotFigure1(sf.damage, par.est)
 
 ### Get data on relation between change in damage cost and sea level 
 ### Get data from Hallegatte et al. (2013)
@@ -37,16 +30,27 @@ E <- getHallegatteData()
 ### Extrapolate to any sea level rise
 slopes <- getChangeSlopes(E)
 
+### Get adaptation cost d
+adaptation.cost <- getAdaptationCost() 
+
+### Fit Burr distribution to damage data (in MNOK)  
+par.est.res <- fitBurrDist(sf.damage, E)   
+par.est <- par.est.res$par ## parameter estimates
+mult <- par.est.res$mult ## multiplicative damage data corrections
+
+### Plot showing the damage data 
+plotFigure1(sf.damage/mult, par.est)
+
 ### Plot showing the changes in damages with sea level rise
 plotFigure2(E)
 
-### Get adaptation cost 
-adaptation.cost <- getAdaptationCost() 
-
 ### Damage distributions for 2016-2100 under various scenarios
-## res <- getDamageTrajectories(E, slopes, adaptation.cost) 
-## save(res, file="data/calculatedDamage.RData")
-load("data/calculatedDamage.RData")
+## res26 <- getDamageTrajectories(E, par.est, slopes, adaptation.cost, inputfile="data/Simulation_2_6.Rdata") 
+## save(res26, file="data/calculatedDamage_nonstat26.RData")
+
+load("data/calculatedDamage_nonstat.RData")
+load("data/calculatedDamage_nonstat26.RData")
+load("data/calculatedDamage_nonstat45.RData")
 
 ### Plot comparing accumulated damage costs under various scenarios
 plotFigure3(res) 
@@ -55,13 +59,27 @@ plotFigure3(res)
 plotFigure4(res) 
 
 ### Investigation of the effects of uncertainty (no adaptation)
[…]

### 2017-05-15 code/BergenDecisions/fitBurrDist.R

@@ -1,8 +1,29 @@
 ### Fit a Burr distribution to the damage data (in MNOK)
 ### Uses functions from the package actuar
 
-fitBurrDist <- function(sf.damage, theta.start=c(5, 0.5, 0.05))
+fitBurrDist <- function(sf.damage, E, theta.start=c(5, 0.5, 0.05))
 {
+    
+    load("data/ObsSeaLevel.RData")
+    mult <- rep(NA, 72)
+    level <- rep(BergenSeaLevel, 2)
+    ## Median damage scenario 
+    z <- c(0, 20, 40)
+    y <- apply(E[,2:4], 2, median)
+    res <- lm(y[1:2]~z[1:2])
+    low.med.slope <- res$coef[2]
+    res <- lm(y[2:3]~z[2:3])
+    high.med.slope <- res$coef[2]
+    get.median.damage.mult <- function(x)
+    {
+        if(x < 0) return(1/(1 + abs(x) * low.med.slope))
+        else if(x < 20) return(1 + x * low.med.slope)
+        else return(y[2] + x * high.med.slope)
+    }
+    for(i in 1:72) mult[i] <- get.median.damage.mult(level[i])
+
+    sf.damage <- sf.damage/mult
+    
     ## Improve numerical stability of estimates by replacing zeros with a small value 
     ind <- which(sf.damage==0)
     sf.damage[ind] <- 1e-04
@@ -25,5 +46,5 @@ fitBurrDist <- function(sf.damage, theta.start=c(5, 0.5, 0.05))
                        control=list(maxit=1000))
 
     ## Return parameter estimates
-    return(optim.out$par)
+    return(list(par=optim.out$par, mult=mult))
 }

### 2017-05-15 code/BergenDecisions/getDamageMult.R

@@ -2,7 +2,7 @@
 ### and Hallegatte et al. (2013) estimates for city k.  
 getDamageMult <- function(x, k, E=E, slopes=slopes)
 {
-    if(x < 0) return(1/(1 + abs(x) * slopes$low[k]))
-    else if(x < 20) return(1 + x * slopes$low[k])
-    else return(as.double(E[k,3]) + x * slopes$high[k])
+    if(x < 0){ return(1/(1 + abs(x) * slopes$low[k]) )
+    } else if(x < 20){ return(1 + x * slopes$low[k])
+    } else{ return(as.double(E[k,3]) + x * slopes$high[k])}
 }

### 2017-05-15 code/BergenDecisions/getDamageTrajectories.R

@@ -6,10 +6,10 @@
 ### -- No sea level change
 ### -- Adaptation in year k for all k 
 
-getDamageTrajectories <- function(E, slopes, adaptation.cost)
+getDamageTrajectories <- function(E, par.est, slopes, adaptation.cost, inputfile)
 {
     ## Read in sea level rise projections for Bergen. 
-    load("data/Simulation.Rdata")
+    load(inputfile)
     I <- dim(sim)[1]
     
     ## Get discount rates 
@@ -36,7 +36,8 @@ getDamageTrajectories <- function(E, slopes, adaptation.cost)
         for(j in 1:85)
         {
             yearly.damage[i,j] <- orig.damage[i,j] * getDamageMult(sim[i, (j+16)],
-                                                                   damage.scenario[i])
+                                                                   damage.scenario[i],
+                                                                   E=E, slopes=slopes)
         }
         yearly.damage[i,] <- yearly.damage[i,]/accum.discount.rate
     }
@@ -49,7 +50,8 @@ getDamageTrajectories <- function(E, slopes, adaptation.cost)
         for(j in 1:85)
         {
             constant.damage[i,j] <- orig.damage[i,j] * getDamageMult(sim[i, 16], 
-                                                                     damage.scenario[i])
+                                                                     damage.scenario[i],
+                                                                     E=E, slopes=slopes)
         }
         constant.damage[i,] <- constant.damage[i,]/accum.discount.rate
     }
@@ -69,9 +71,10 @@ getDamageTrajectories <- function(E, slopes, adaptation.cost)
             for(j in 1:85)
             {
                 adapted.damage[k,i,j] <- orig.damage[i,j]/2 *
-                    getDamageMult(sim[i, (j+16)], damage.scenario[i]) +
+                    getDamageMult(sim[i, (j+16)], damage.scenario[i], E=E, slopes=slopes) +
                     orig.damage[i,j]/2 *
-                    getDamageMult(sim[i, (j+16)]-adapted.slr[j], damage.scenario[i])
+                    getDamageMult(sim[i, (j+16)]-adapted.slr[j], damage.scenario[i],
+                                  E=E, slopes=slopes)
             }
             m <- which(adapted.slr==75)[1]
             adapted.damage[k,i,m] <- adapted.damage[k,i,m] + adaptation.cost

### 2017-05-15 code/BergenDecisions/getUncertaintyTrajectories.R

@@ -1,6 +1,6 @@
 ### Investigation of the effects of uncertainty (no adaptation)
 
-getUncertaintyTrajectories <- function(E, slopes, orig.damage, damage.scenario)
+getUncertaintyTrajectories <- function(E, slopes, orig.damage, damage.scenario, inputfile)
 {
     ## Cumulative damage with medians
 
@@ -20,7 +20,7 @@ getUncertaintyTrajectories <- function(E, slopes, orig.damage, damage.scenario)
     }
 
     ## Read in sea level rise projections for Bergen. 
-    load("data/Simulation.Rdata")
+    load(inputfile)
     I <- dim(sim)[1]
     
     ## Get discount rates 

### 2017-05-15 code/BergenDecisions/getUncertaintyTrajectoriesAllRCP.R

@@ -1,0 +1,89 @@
+### Investigation of the effects of uncertainty (no adaptation)
+
+getUncertaintyTrajectoriesAllRCP <- function(E, slopes, orig.damage, damage.scenario)
+{
+    ## Cumulative damage with medians
+
+    ## Median damage scenario 
+    z <- c(0, 20, 40)
+    y <- apply(E[,2:4], 2, median)
+    res <- lm(y[1:2]~z[1:2])
+    low.med.slope <- res$coef[2]
+    res <- lm(y[2:3]~z[2:3])
+    high.med.slope <- res$coef[2]
+    
+    get.median.damage.mult <- function(x)
+    {
+        if(x < 0) return(1/(1 + abs(x) * low.med.slope))
+        else if(x < 20) return(1 + x * low.med.slope)
+        else return(y[2] + x * high.med.slope)
+    }
+
+    ## Read in sea level rise projections for Bergen. 
+    load("data/Simulation_8_5.Rdata")
+    I <- dim(sim)[1]
+    
+    ## Get discount rates 
+    discount.rate <- getDiscountRate()
+    accum.discount.rate <- cumprod(discount.rate)
+
+    ## Median sea level rise
+    median.sim <- apply(sim, 2, median)[17:101]
+
+    ## Median damage cost
+    median.damage <- apply(orig.damage, 2, median)
+
+    ## Putting the three together
+    yearly.median.damage <- rep(NA, 85)
+    for(j in 1:85)
+    {
+        yearly.median.damage[j] <- median.damage[j] * get.median.damage.mult(median.sim[j]) 
+    }
+    yearly.median.damage <- yearly.median.damage/accum.discount.rate
+    
+    ## Uncertainty in damage mult. fct. only 
+    yearly.mf.damage <- array(NA, dim=c(I, 85))
+    for(i in 1:I)
+    {
+        for(j in 1:85)
+        {
+            yearly.mf.damage[i,j] <- median.damage[j] * getDamageMult(median.sim[j],
+                                                                      damage.scenario[i],
+                                                                      E, slopes)
+        }
+        yearly.mf.damage[i,] <- yearly.mf.damage[i,]/accum.discount.rate
+    }
+    total.mf.damage <- apply(yearly.mf.damage,1,sum)
+    
+    ## Uncertainty in slr only 
+    yearly.slr.damage <- array(NA, dim=c(I, 85))
[…]

### 2017-05-15 code/BergenDecisions/plotFigure1.R

@@ -2,13 +2,24 @@
 
 plotFigure1 <- function(sf.damage, par.est, ifPdf=TRUE, fileName="figures/DecisionAnalysisBergenPartI.pdf")
 {
-    if(ifPdf) pdf(file=fileName, width=5, height=5, points=12)
+    if(ifPdf) pdf(file=fileName, width=10, height=5, points=12)
     ## (a) Plot of damage data and fitted distribution
-    par(mex=0.75, mar=c(5,4,2,2)+0.1)
-    hist(sf.damage, main="", xlab="Amount (million NOK)", col="gray60", breaks=50,  xlim=c(0,65),
+    par(mex=0.75, mar=c(5,4,2,2)+0.1, mfrow=c(1,2))
+    hist(sf.damage, main="", xlab="Cost (million NOK)", col="gray60", breaks=100,  xlim=c(0,25),
          freq=FALSE)
-    x <- seq(0, 65, by=0.01)
+    x <- seq(0, 25, by=0.01)
     y <- dburr(x, shape1=par.est[1], shape2=par.est[2], rate=par.est[3])
     lines(x,y,col="red",lwd=2)
+
+    ## (b) QQ-plot of obs vs. estimated quantiles
+    quants <- seq(0.02,0.99, by=0.01)
+    q.obs <- as.numeric(quantile(sf.damage, quants))
+    q.mod <- qburr(quants, shape1=par.est[1], shape2=par.est[2], rate=par.est[3])
+    q.max <- 25
+    plot(q.mod, q.obs, type="b", xlim=c(0,q.max), ylim=c(0,q.max), main="",
+         xlab="Estimated quantiles", ylab="Observed quantiles", axes=FALSE)
+    axis(1)
+    axis(2)
+    lines(c(0,q.max), c(0,q.max), col="blue")
     if(ifPdf) dev.off()
 }

### 2017-05-15 code/BergenDecisions/plotFigure2.R

@@ -7,9 +7,9 @@ plotFigure2 <- function(E, ifPdf=TRUE, fileName="figures/DecisionAnalysisBergenP
     ## (b) Plot of extrapolated damage change profiles 
     x <- c(0,20,40)
     plot(x, E[1, 2:4], type="o", col="black", xlim=c(-100,130), ylim=c(-180,1000),
-         xlab="Sea level anomaly (cm)", ylab="Relative mean annual damage", main="", axes=FALSE)
+         xlab="Sea level anomaly (m)", ylab="Relative mean annual cost", main="", axes=FALSE)
     box()
-    axis(1)
+    axis(1, at=c(-100, -50, 0, 50, 100), labels=c(-1, -0.5, 0, 0.5, 1))
     axis(2, at=c(-200, 1, 200, 400, 600, 800, 1000), labels=c(0.005, 1, 200, 400, 600, 800, 1000))
     z <- c(40,130)
     w <- c(-100,0)
@@ -31,3 +31,5 @@ plotFigure2 <- function(E, ifPdf=TRUE, fileName="figures/DecisionAnalysisBergenP
     lines(z, res$coef[1] + z * res$coef[2], col="red", lwd=2)
     if(ifPdf) dev.off()
 }
+
+ 

### 2017-05-15 code/BergenDecisions/plotFigure3.R

@@ -44,18 +44,18 @@ plotFigure3 <- function(res, ifPdf=TRUE, fileName="figures/CumDamageCostsBergen.
     adapt.lower.05 <- apply(cumsum.adapt.damage[,,85],1,quantile,0.05)
 
     ## Cumulated damage due to sea level rise with
-    ## adaptation in 2047 (most cost effective adaptation). 
-    cumsum.adapt.damage.2047 <- array(NA, dim=dim(res$adapted[32,,]))
+    ## adaptation in 2059 (most cost effective adaptation). 
+    cumsum.adapt.damage.2059 <- array(NA, dim=dim(res$adapted[44,,]))
     
-    for(i in 1:dim(res$adapted[32,,])[1])
+    for(i in 1:dim(res$adapted[44,,])[1])
     {
-        cumsum.adapt.damage.2047[i,] <- cumsum(res$adapted[32,i,])
+        cumsum.adapt.damage.2059[i,] <- cumsum(res$adapted[44,i,])
     }
-    upper.95.adapt.cumsum.2047 <- apply(cumsum.adapt.damage.2047,
+    upper.95.adapt.cumsum.2059 <- apply(cumsum.adapt.damage.2059,
                                         2,quantile,0.95)
-    lower.5.adapt.cumsum.2047 <- apply(cumsum.adapt.damage.2047,
+    lower.5.adapt.cumsum.2059 <- apply(cumsum.adapt.damage.2059,
                                        2,quantile,0.05)
-    median.adapt.cumsum.2047 <- apply(cumsum.adapt.damage.2047,
+    median.adapt.cumsum.2059 <- apply(cumsum.adapt.damage.2059,
                                       2,quantile,0.5)
 
     ## Making the plot 
@@ -63,20 +63,21 @@ plotFigure3 <- function(res, ifPdf=TRUE, fileName="figures/CumDamageCostsBergen.
     par(mex=0.75)
     ## Cumulative damage costs without adaptation 
     plot(2016:2100, upper.95.yearly.cumsum/1e3,type="l",
-         ylim=c(0,14), xlab="Year",
-         ylab="Accumulated damage (billion NOK)", main="", lty=3)
+         ## ylim=c(0,14), xlab="Year",
+         ylim=c(0,4.5), xlab="Year",
+         ylab="Accumulated cost (billion NOK)", main="", lty=3)
     ## Cumulative damage costs for no sea level rise
[…]

### 2017-05-15 code/BergenDecisions/plotFigure4.R

@@ -7,13 +7,6 @@ plotFigure4 <- function(res, ifPdf=TRUE, fileName="figures/TotalDamageCostsAdapt
     ## Calculate cumulative damages over 2016-2100
     ## We present the distribution through median and 90% interval
 
-    ## Cumulated additional damage due to sea level rise.
-    cumsum.yearly.damage <- array(NA, dim=dim(res$yearly))
-    for(i in 1:dim(res$yearly)[1]) 
-    {
-        cumsum.yearly.damage[i,] <- cumsum(res$yearly[i,])
-    }
-    
     ## Cumulated damage for various adaptation options
     cumsum.adapt.damage <- array(NA, dim=dim(res$adapted))
     for(i in 1:85)
@@ -38,8 +31,8 @@ plotFigure4 <- function(res, ifPdf=TRUE, fileName="figures/TotalDamageCostsAdapt
     if(ifPdf) pdf(file=fileName, width=10, height=5, points=12)
     par(mex=0.75)
     a <- median(cumsum.yearly.damage[,85])/1e3
-    plot(c(2016,2100), c(a,a), type="l", ylim=c(0,14), xlab="Year of adaptation measure",
-         ylab="Total damage 2016-2100 (billion NOK)", main="")
+    plot(c(2016,2100), c(a,a), type="l", ylim=c(0,4.5), xlab="Year of adaptation measure",
+         ylab="Total cost 2016-2100 (billion NOK)", main="")
     year <- (2016:2100)
     k <- 0.35
     for(i in 1:85)
@@ -47,16 +40,16 @@ plotFigure4 <- function(res, ifPdf=TRUE, fileName="figures/TotalDamageCostsAdapt
         a <- year[i]
         b <- adapt.upper.95[i]/1e3
         c <- adapt.lower.05[i]/1e3
-        polygon(c(a-k, a+k, a+k, a-k), c(c, c, b, b), col="gray80", border="gray80")
+        polygon(c(a-k, a+k, a+k, a-k), c(c, c, b, b), col="#6594ED60", border="#6594ED60")
         d <- adapt.median[i]/1e3
-        lines(c(a-k, a+k), c(d,d), lwd=2, col="red")
+        lines(c(a-k, a+k), c(d,d), lwd=3, col="#6594ED")
     }
     a <- median(cumsum.yearly.damage[,85])/1e3
[…]

### 2017-05-15 code/BergenDecisions/plotFigure5.R

@@ -1,0 +1,71 @@
+### Plot comparing 90% intervals of total damage for various uncertainty settings on a log scale
+
+plotFigure5 <- function(res.unc, total.damage, res.unc45, total.damage45, res.unc26, total.damage26, ifPdf=TRUE, fileName="figures/UncertaintyLog.pdf")
+{
+    quant.mf.damage <- as.numeric(quantile(res.unc$mf, c(0.05, 0.95)))
+    quant.slr.damage <- as.numeric(quantile(res.unc$slr, c(0.05, 0.95)))
+    quant.dam.damage <- as.numeric(quantile(res.unc$dam, c(0.05, 0.95)))
+    quant.damage <- as.numeric(quantile(total.damage, c(0.05, 0.95)))
+
+    quant.mf.damage26 <- as.numeric(quantile(res.unc26$mf, c(0.05, 0.95)))
+    quant.slr.damage26 <- as.numeric(quantile(res.unc26$slr, c(0.05, 0.95)))
+    quant.dam.damage26 <- as.numeric(quantile(res.unc26$dam, c(0.05, 0.95)))
+    quant.damage26 <- as.numeric(quantile(total.damage26, c(0.05, 0.95)))
+
+    quant.mf.damage45 <- as.numeric(quantile(res.unc45$mf, c(0.05, 0.95)))
+    quant.slr.damage45 <- as.numeric(quantile(res.unc45$slr, c(0.05, 0.95)))
+    quant.dam.damage45 <- as.numeric(quantile(res.unc45$dam, c(0.05, 0.95)))
+    quant.damage45 <- as.numeric(quantile(total.damage45, c(0.05, 0.95)))
+
+    
+    if(ifPdf) pdf(file="figures/UncertaintyLog.pdf", width=10, height=5, points=12)
+    par(mex=0.75, mar=c(5,2,2,2)+0.1)
+    n <- 8
+    plot(1, type="n", main="", ylab="", xlab="Total cost 2016-2100 (million NOK)",
+         xlim=c(log(n),log(5000)), ylim=c(0,18), axes=FALSE)
+    ticks <- c(10, 20, 50, 150, 500, 1500, 5000)
+    axis(1, at = log(ticks), labels=ticks)
+    k <- 0.4
+    polygon(rep(c(log(n), log(5000)), each=2), c(0,6,6,0), col="gray75", border=FALSE)
[…]

### 2017-05-15 code/BergenDecisions/testStationarity.R

@@ -1,0 +1,58 @@
+### Testing an inclusion of observed sea level when calculating
+### parameter estimates of Burr distribution
+
+testStationarity <- function(sf.damage, theta.start=c(5, 0.5, 0.05, 0))
+{
+    ## Improve numerical stability of estimates by replacing zeros with a small value 
+    ind <- which(sf.damage==0)
+    sf.damage[ind] <- 1e-04
+
+    load("data/ObsSeaLevel.RData")
+    obs <- rep(BergenSeaLevel, 2)
+    ind.obs.neg <- which(obs < 0)
+     
+    ## The negative log likelihood function to be optimized
+    log.lik <- function(theta)
+    {
+        mult.factor <- 1 + abs(obs) * theta[4]
+        mult.factor[ind.obs.neg] <- 1/mult.factor[ind.obs.neg]
+        data <- sf.damage/mult.factor
+        res <- sum(dburr(data, 
+                         shape1=theta[1], 
+                         shape2=theta[2], 
+                         rate=theta[3], 
+                         log=TRUE))
+        return(-res)
+    }
+    
+    ## Optimize the loss function
+    optim.out <- optim(theta.start, 
+                       fn=log.lik, 
+                       method="BFGS", 
+                       control=list(maxit=1000))
+
+    ## Return parameter estimates
+    return(optim.out$par)
+}
+
+### Running some tests 
+
+quants <- c(0.02, seq(0.05,0.95, by=0.05), 0.98)
+mult <- rep(NA, 36)
+x11()
+par(mfrow=c(3,5), mar=c(4,4,1,1))
+for(t in 1:15)
+{
+    for(i in 1:36) mult[i] <- getDamageMult(BergenSeaLevel[i], t, E, slopes)
+    par.est.t <- fitBurrDist(sf.damage/c(mult,mult))   
+    q.mod <- qburr(quants, shape1=par.est[1],
+                   shape2=par.est[2], rate=par.est[3])
+    q.mod.t <- qburr(quants, shape1=par.est.t[1],
+                   shape2=par.est.t[2], rate=par.est.t[3])
+    q.max <- 75
+    plot(q.mod, q.mod.t, type="b", xlim=c(0,q.max), ylim=c(0,q.max), main="",
+     xlab="Original estimate", ylab="New estimate", axes=FALSE)
+    axis(1)
+    axis(2)
+    lines(c(0,q.max), c(0,q.max), col="blue")
+}

### 2017-05-15 code/BergenDecisions/.#fitBurrDist.R
Minor changes

@@ -1,1 +1,0 @@
-disa@disa-VirtualBox.6190:1494832380

### 2017-05-15 code/BergenDecisions/fitBurrDist.R

@@ -3,7 +3,7 @@
 
 fitBurrDist <- function(sf.damage, E, theta.start=c(5, 0.5, 0.05))
 {
-    
+    ## Account for the effect of observed sea level 
     load("data/ObsSeaLevel.RData")
     mult <- rep(NA, 72)
     level <- rep(BergenSeaLevel, 2)
@@ -21,7 +21,6 @@ fitBurrDist <- function(sf.damage, E, theta.start=c(5, 0.5, 0.05))
         else return(y[2] + x * high.med.slope)
     }
     for(i in 1:72) mult[i] <- get.median.damage.mult(level[i])
-
     sf.damage <- sf.damage/mult
     
     ## Improve numerical stability of estimates by replacing zeros with a small value 
